### 2023-10-19 paper.qmd
First commit with quarto skeleton

---
title: "paper"
---

## Quarto

Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.

```{r}
1 + 1
```

### 2023-10-19 paper.qmd
Add frontmatter and outline

@@
 ---
-title: "paper"
+title: "XRONOS: an open data infrastructure for archaeological chronometry"
+#date: today
+author:
+  - name: Joe Roe
+    id: jr
+    orcid: 0000-0002-1011-1244
+    email: joseph.roe@unibe.ch
+    affiliation: 
+    - ref: iaw
+  - name: Martin Hinz
+    id: mh
+    orcid:  0000-0002-9904-6548 
+    email: martin.hinz@unibe.ch
+    affiliation: 
+    - ref: iaw
+affiliations:
+  - id: iaw
+    name: University of Bern
+    department: Institute of Archaeological Sciences
+    country: Switzerland
+    url: https://www.iaw.unibe.ch/
+abstract: |
+  Chronometric data has been one of the principle domains in which open data and open science approaches have been applied over the last ten years.
+Comprehensive compilations of radiocarbon dates have become available for many parts of the world in the last decade and, as natural next step, there are now several initiatives to collate this data globally, including the retrieval tool c14bazAAR [@SchmidEtAl2019], the IntChron exchange format [@BronkRamseyEtAl2019], and the synthetic database p3kc14 [@BirdEtAl2022].
+This move from archiving individual datasets to producing a living open data infrastructure for archaeological chronologies raises a number of challenges.
+Radiocarbon datasets are still sorely lacking for many parts of the world and, even in those regions with good coverage, the quality of data is highly uneven and largely undocumented.
+There exists no central repository ensuring the long-term sustainability and completeness of these datasets, and the potential of placing other sources of chronometric information (e.g.
+dendrochronology, typological dating) in an open data framework has hardly been realised at all.
+
+  Here, building on and complementing these initiatives, we present XRONOS: a new web-based platform for chronometric data from archaeological contexts worldwide, combining an open data repository with tools for importing, curating and analysing chronometric information from diverse sources.
+The principal design goals of the software are to combine all available sources of radiocarbon and other chronometric data in single database; develop robust tools for the continuous ingestion and refinement of this data; and disseminate this data within an open and FAIR framework, embedding it in the wider world of Linked Open Data in archaeology and beyond.
+
+  This paper outlines the conceptual and technical infrastructure developed to realise these goals in XRONOS’ initial phases of development (2019 and 2021–2023), including a generalised data model for site and radiocarbon information, extendable to other chronometric data; an R- and Ruby-based pipeline for continuous ingestion of data from a variety of sources; continuous, semi-automated data cleaning protocols; a Ruby-on-Rails application providing a web-based frontend to the data and a REST API for programmatic access; and an R package for interfacing with the API.
+We believe the XRONOS framework provides more open, more reliable, and more comprehensive access to chronometric data than previously available, as well as a foundation for its continuous expansion and refinement.
+copyright: 
+  holder: the authors
+  year: 2023
+license: "CC BY"
+# citation: 
+#  container-title: Archaeometry
+#  volume: 1
+#  issue: 1
+#  doi: 10.5555/12345678
+funding: "This work was supported by the Swiss National Science Foundation ([SNSF Project #198152](https://data.snf.ch/grants/grant/198153))"
+bibliography: references.bib
+format:
+  html:
+    page-layout: article
+    toc: true
 ---
 
-## Quarto
+## Abstract
 
-Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.
+    Briefly summarize the software's purpose, key features, and potential impact.
+    Highlight the paper's objectives, methodology, and key findings.
 
-```{r}
-1 + 1
-```
+## Introduction
+
+    Introduce the context and significance of the scientific problem your software addresses.
+    Explain the need for new software tools in this domain.
+    Present an overview of the paper's structure.
+
+## Background
+
+    Review relevant literature, discussing existing software tools and their limitations.
+    Explain the scientific principles or theories that underlie the problem your software aims to solve.
+    Highlight the gap in existing tools and the need for your software.
+
+## Software architecture
+
+    Describe the software's architecture, algorithms, and data handling mechanisms.
+    Explain the programming languages, frameworks, and libraries used in its development.
+    Detail the key components and functionalities.
+
+## Design and user interfaces
+
+    Explain the software's user interface design and features.
+    Describe the user experience and how the interface enhances usability.
+    Include screenshots or diagrams to illustrate the interface.
+
+## Features and capabilities
+
+    Provide an in-depth overview of the software's features and capabilities.
+    Highlight its unique functionalities that set it apart from existing tools.
+    Discuss how these features address the scientific problem.
+
+## Performance evaluation
+
+    Present results from performance tests, such as speed, accuracy, and resource usage.
+    Compare your software's performance with existing alternatives if applicable.
+    Discuss any benchmark datasets or experiments used for evaluation.
+
+## Case studies and use cases
+
+    Provide real-world examples of how your software can be applied.
+    Describe case studies or experiments that showcase its effectiveness.
+    Include user testimonials or feedback.
+
+## Documentation and support
+
+    Discuss the availability of documentation, tutorials, and user manuals.
+    Explain how users can seek support or assistance when using the software.
+    Highlight any plans for future updates or enhancements.
+
+## Availability and licensing
+
+    Describe how and where users can access and download the software.
+    Specify the software's licensing model (e.g., open source, proprietary).
+    Explain any restrictions or usage terms.
+
+## Conclusion
+
+    Summarize the key findings and contributions of your software.
+    Discuss the significance of your software in advancing scientific knowledge.
+    Mention any limitations or challenges encountered during development.
+
+## Future work
+
+    Share your vision for the future development of the software.
+    Describe potential enhancements, new features, or research directions.
+    Address any planned updates or community contributions.
+
+## Acknowledgements {.appendix}
+
+* Setareh Ebrahimiabareghi
+* Tomasz J. Chmielewski

### 2023-10-19 paper.qmd
Expand outline

@@
 ---
-title: "XRONOS: an open data infrastructure for archaeological chronometry"
+title: "XRONOS: an open infrastructure for chronometric data in archaeology"
 #date: today
@@
 abstract: |
+  XRONOS (<https://xronos.ch>) is an open data infrastructure for the backbone of the archaeological record – chronology.
+It provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world.
+
   Chronometric data has been one of the principle domains in which open data and open science approaches have been applied over the last ten years.
@@
 
-## Abstract
-
-    Briefly summarize the software's purpose, key features, and potential impact.
-    Highlight the paper's objectives, methodology, and key findings.
-
 ## Introduction
@@
 
+* Radiocarbon dates were open to begin with (*Radiocarbon*'s "date lists"), but this didn't scale and fell out of use
+* Even though the generation of this data is relatively centralised (in XX radiocarbon labs), those seeking a wider overview are now caught in a labour-intensive process of re-gathering it from literature into databases
+  * The coverage of these databases is uneven, systematically biases, and duplicative of each other
+  * Databases are put together *ad hoc* and rarely continuously maintained over long periods
+  * Even though the reporting of radiocarbon determinations is relatively standardised [@Millard2014], integration of these datasets is hampered by minor variations in encoding and wide variation in data quality/completeness
+* Other forms of chronometric data (dendro, typology, sea level, etc.) remain relatively untouched
+
+In this paper we:
+
+* Review the state of the art in radiocarbon data compilation (and other chronometric data, such as they are), highlighting the need for XRONOS
+* Describe XRONOS' software architecture, design goals, and currently-implemented features
+  * With a focus on the choices we made in representing/modelling chronological data
+* Assess the current state of the software and database, and give case studies of how it can be used
+* Conclude with our future plans
+
 ## Background
@@
     Highlight the gap in existing tools and the need for your software.
+
+* There are a very large number of regional/specialist radiocarbon databases
+  * Summarise coverage in table
+* Also several efforts to bring this together: c14bazAAR, IntChron, p3kc14
 

### 2024-01-30 paper.qmd
Redo outline

--- 2023-10-19 paper.qmd
@@
 ---
-title: "paper"
+title: "XRONOS: an open data infrastructure for archaeological chronometry"
+author:
+  - name: Joe Roe
+    orcid: 0000-0002-1011-1244
+    url: "https://joeroe.io"
+    email: joseph.roe@unibe.ch
+    affiliation:
+    - ref: iaw
+    roles: [writing, editing, software, analysis]
+  - name: Martin Hinz
+    orcid: 0000-0002-9904-6548
+    url: "https://martinhinz.info"
+    email: martin.hinz@unibe.ch
+    corresponding: true
+    affiliation:
+    - ref: iaw
+    roles: [conceptualization, funding, supervision, editing, software]
+affiliations:
+  - id: iaw
+    name: University of Bern
+    department: Institute of Archaeological Sciences
+    country: Switzerland
+    url: https://www.iaw.unibe.ch/
+abstract: |
+  XRONOS (<https://xronos.ch>) is an open data infrastructure for the backbone of the archaeological record – chronology.
+It provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world.
+
+  Chronometric data has been one of the principle domains in which open data and open science approaches have been applied over the last ten years.
+Comprehensive compilations of radiocarbon dates have become available for many parts of the world in the last decade and, as natural next step, there are now several initiatives to collate this data globally, including the retrieval tool c14bazAAR [@SchmidEtAl2019], the IntChron exchange format [@BronkRamseyEtAl2019], and the synthetic database p3kc14 [@BirdEtAl2022].
+This move from archiving individual datasets to producing a living open data infrastructure for archaeological chronologies raises a number of challenges.
+Radiocarbon datasets are still sorely lacking for many parts of the world and, even in those regions with good coverage, the quality of data is highly uneven and largely undocumented.
+There exists no central repository ensuring the long-term sustainability and completeness of these datasets, and the potential of placing other sources of chronometric information (e.g.
+dendrochronology, typological dating) in an open data framework has hardly been realised at all.
+
+  Here, building on and complementing these initiatives, we present XRONOS: a new web-based platform for chronometric data from archaeological contexts worldwide, combining an open data repository with tools for importing, curating and analysing chronometric information from diverse sources.
+The principal design goals of the software are to combine all available sources of radiocarbon and other chronometric data in single database; develop robust tools for the continuous ingestion and refinement of this data; and disseminate this data within an open and FAIR framework, embedding it in the wider world of Linked Open Data in archaeology and beyond.
+
+  This paper outlines the conceptual and technical infrastructure developed to realise these goals in XRONOS’ initial phases of development (2019 and 2021–2023), including a generalised data model for site and radiocarbon information, extendable to other chronometric data; an R- and Ruby-based pipeline for continuous ingestion of data from a variety of sources; continuous, semi-automated data cleaning protocols; a Ruby-on-Rails application providing a web-based frontend to the data and a REST API for programmatic access; and an R package for interfacing with the API.
+We believe the XRONOS framework provides more open, more reliable, and more comprehensive access to chronometric data than previously available, as well as a foundation for its continuous expansion and refinement.
+copyright: 
+  holder: the authors
+  year: 2023
+license: "CC BY"
+# citation: 
+#  container-title: Archaeometry
+#  volume: 1
+#  issue: 1
+#  doi: 10.5555/12345678
+funding: "This work was funded by the Swiss National Science Foundation ([SNSF Project #198152](https://data.snf.ch/grants/grant/198153))"
+bibliography: references.bib
+format:
+  html:
+    page-layout: article
+    toc: true
 ---
 
-## Quarto
+## Introduction
 
-Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.
+* Chronometric modelling depends on ready access to chronometric data!
+* Luckily, we do have a lot of it.
+* Moving from 'open data' (uploading a CSV) to 'open infrastructure' [@CITE]
 
-```{r}
-1 + 1
-```
+In this paper we:
+
+* Review the state of the art in radiocarbon data compilation (and other chronometric data, such as they are), highlighting the need for XRONOS
+* Describe XRONOS' software architecture, design goals, and currently-implemented features
+  * With a focus on the choices we made in representing/modelling chronological data
+* Assess the current state of the software and database, and give case studies of how it can be used
+* Conclude with our future plans
+
+## Compilations of radiocarbon dates
+
+* Radiocarbon dates used to be regularly compiled and published in *Radiocarbon* (so it was 'open data' before it was cool), but this is no longer practical.
+* Plus the bare measurements are of limited use without contextual information, which labs don't usually have access to.
+* So even though radiocarbon data comes from a relatively limited number of sources [@IntCalLabList] and has relatively standardised reporting conventions [@Millard2014], in practice data has to be manually compiled from literature – this is time-consuming!
+* Since at least 19XX, there have been efforts to systematically collect and compile this information
+* Graph of number of dates/databases through time
+* But this has been fragmentary: we counted at least XX databases
+* And patchy: highly uneven spatial and geographic coverage, different data collected, different conventions for recording data
+
+## Time for a global radiocarbon database?
+
+* Regional databases are put together *ad hoc* and rarely continuously maintained over long periods
+* The coverage of these databases is uneven, systematically biased, and duplicative of each other
+* Other forms of chronometric data (dendro, typology, sea level, etc.) remain relatively untouched
+
+* Recent moves towards a global synthesis: c14bazAAR [@SchmidEtAl2019], IntChron [@BronkRamseyEtAl2019], p3k14c [@BirdEtAl2022]
+* Each taking a different approach, pros and cons
+* We see a need for something that is comprehensive, unified, living
+* And we don't see why it should be limited to radiocarbon: what about other radiometric dates? Dendro? Typologies?
+
+## XRONOS
+
+### Design goals and principles
+
+* Inspired by GBIF, which manages to integrate large amounts of data from diverse sources into a signal, consistent database
+
+### Data model
+
+* Basic data structure inherited from RADON.
+* Chrons: coordinates linking a time, a place, and human activity
+  * Radiocarbon dates, typos, dendro etc.
+  * Departs from the usual convention of treating e.g. a "period" as a different sort of information to a scientific date
+  * In fact they can all be seen as the same kind of data
+  * Time element can be represented as a probability distribution
+  * Facilitates downstream modelling in e.g. OxCal, aoristic
+* Contextual information: sample, context, site
+* Bibliographic references
+* Paper trail
+
+### Software architecture
+
+* CRUD Rails app
+  * Not so exciting, but try to keep up with modern standards, remain accessible, etc.
+  * Data browser
+* API
+  * R package
+* Linked data
+
+## Implementation
+
+The first prototype of XRONOS was developed in 2019.
+It has been under sustained development 2021–2024 with the first public release in July 2021.
+
+* Assessment of XRONOS' coverage
+* Tale from the trenches: duplicate processing?
+* Tale from the trenches: cleaning species and countries?
+* Case study of use?
+
+## Conclusion
+
+* Chronology is the backbone of the archaeological record
+* XRONOS could be a backbone for a universe of linked open data resources...
+* ...but sustainability is a challenge.
+  * We're coming to the end of our project funds
+  * We'll try to get more, but you never know (grant-based science and long-term investments in digital infrastructure don't work so well together!)
+  * Future sustainability
+
+## Acknowledgements
+
+Direct contributors to XRONOS (or invite as coauthors?):
+
+* Clemens Schmid
+* Caroline Heitz
+* Albert Hafner
+* Setareh Ebrahimiabareghi
+* Chiara Huwiler
+* Rivana Moser

### 2024-01-30 paper.qmd
Render self-contained HTML

@@
     toc: true
+    embed-resources: true
 ---

### 2024-02-21 paper.qmd
Add coauthors and expand draft

@@
     url: "https://joeroe.io"
-    email: joseph.roe@unibe.ch
     affiliation:
     - ref: iaw
-    roles: [writing, editing, software, analysis]
+    roles: [writing, editing, curation, software, analysis]
+  - name: Clemens Schmid
+    orcid: 0000-0003-3448-5715
+    affiliation:
+    - ref: mpg
+    roles: [curation, software, editing]
+  - name: Setareh Ebrahimiabareghi
+    orcid: 0000-0003-3749-3147
+    affiliation:
+    - ref: iaw
+    roles: [curation, investigation, editing]
+  - name: Caroline Heitz
+    orcid: 0000-0001-7188-6775
+    affiliation:
+    - ref: iaw
+    roles: [conceptualization, funding, editing]
   - name: Martin Hinz
@@
     url: https://www.iaw.unibe.ch/
+  - id: mpg
+    name: Max Planck Institute for Geoanthropology
+    country: Germany
+    url: https://www.shh.mpg.de/
 abstract: |
@@
 
-* Basic data structure inherited from RADON.
-* Chrons: coordinates linking a time, a place, and human activity
-  * Radiocarbon dates, typos, dendro etc.
-  * Departs from the usual convention of treating e.g. a "period" as a different sort of information to a scientific date
-  * In fact they can all be seen as the same kind of data
+TODO: Figure: conceptual scheme, centred on chrons (i.e. an idealised form of the database).
+
+XRONOS data model is based on the concept of a set of coordinates in time or *chron*.
+We define a chron as an assertion that links human activity with a particular point in space and time.
+More concretely, our data model currently encompasses three types of chron: radiocarbon dates, typological dates (e.g. 'Early Neolithic') and dendrochronological dates.
+However we anticipate that the concept will accommodate all sorts of absolute and relative dating techniques, as the scope of the database expands.
+
+Chrons are useful conceptually because they emphasise that dates from different sources still consist of essentially the same kind of information: a statement locating an event in space and time.
+We thereby avoid privileging certain sources of chronological data over (as might be the case if, for example, we treated 'period' as a fixed attribute of a site) and can accommodate contradictory (e.g.
+differences of opinion on typological classification).
+This is important given that XRONOS aspires to be an authoratative 'backbone' with a global scope, so we cannot realistically impose a single chronological scheme or resolve conflicting information provided by specialists.
+They are useful practically because they expose a common interface for attributes that all types of chronological information share, such as a *terminus post quem*, *terminus ante quem*, and midpoint estimate.
+This allows applications that use XRONOS' data model (including XRONOS itself) to collate chronological data from multiple sources, without necessarily having to be aware of the pecularities of each type of dating.
+
+Key to 'unifying' chrons in this way is a common model of time.
 * Time element can be represented as a probability distribution
-  * Facilitates downstream modelling in e.g. OxCal, aoristic
+  * Inspired by OxCal, aoristic analysis
+  * How you get to the PD representation can be very different (e.g. calibration vs. aoristic), but the important thing is the endpoint is the same.
+
+Chrons are located in space through association to a *sample* – the physical object from which a chronological determination was made.
+The location of samples is represented with geographical coordinates and an associated coordinate reference system (CRS), though since in practice the precise location of single samples is rarely available, this property is usually inherited from the site.
+We also record relevant metadata on the nature of the sample.
+For radiocarbon dates, for example, we follow established conventions [@Millard2014] in recording the type (e.g.
+charcoal, charred seed) and, where applicable, taxonomic designation (e.g.
+TODO, TODO) of the organic material used for dating.
+For typological dates, an ideal scenario would be for the sample to represent the particular object from which an inference was made (e.g. 'Natufian' might be inferred from 'lunate-type microlith').
+In practice, the best we can glean from most published datasets is the type of material used (e.g. 'pottery', 'lithics').
+The same sample can be associated with multiple chrons, including different types of chron.
+This is useful, for example, for representing replicate radiocarbon dates on the same sample, or radiocarbon dates and dendrochronological made on the same section of wood for wiggle-matching.
+
 * Contextual information: sample, context, site
+  * Different types of xrons can be attached to the same sample, context, or site, as appropriate
 * Bibliographic references
@@
 
-* Clemens Schmid
-* Caroline Heitz
 * Albert Hafner
-* Setareh Ebrahimiabareghi
 * Chiara Huwiler
 * Rivana Moser
+* Tomasz
+* Stephanie Döppler

### 2024-02-21 paper.qmd
Add @MartinHinz's text from Antiquity draft

@@
     - ref: iaw
-    roles: [conceptualization, funding, supervision, editing, software]
+    roles: [conceptualization, funding, supervision, writing, editing, software]
 affiliations:
@@
 #  doi: 10.5555/12345678
-funding: "This work was funded by the Swiss National Science Foundation ([SNSF Project #198152](https://data.snf.ch/grants/grant/198153))"
+funding: "This work was funded by the Swiss National Science Foundation ([SNSF Project #198152](https://data.snf.ch/grants/grant/198153)) and the University of Bern (Initiator Grant, Caroline Heitz, Project 'Time and Temporality')"
 bibliography: references.bib
@@
 
+We live in interesting times: Today, archaeological research can no longer complain that too few absolute dates exist for the phenomena of the past.
+Large-scale dating projects such as \[example\] and \[example\] are producing extensive data series in a short time, which make it possible to situate individual sites in time with unprecedented precision.
+But also due to the dating methods that have become affordable even for smaller budgets in the meantime and the integration of natural scientific dating that is almost mandatory, such large initiatives are accompanied by a great amount of individual and scattered chronological information that represents a rich source of data for supra-regional investigations.
+Based on that, data modelling can be used as a method to draw conclusions from a variety of individually information-weak data.
+
+The problem, however, is that these data are very scattered.
+Traditionally, radiocarbon data were disseminated by the laboratories, with special attention paid to publication in the form of "date lists" in the journal "Radiocarbon".
+Today, unfortunately, this has become very rare, on the one hand because the quantity and complexity of the data limit the possibilities of publication on paper.
+On the other hand, data in printed form are hardly suitable for reuse today.
+As a result, access to chronological information, especially a comprehensive overview, is made much more difficult [@bronkramsey2019].
+Furthermore, data is presented in a variety of different formats and granularities, and critical information is often missing, although standards exist [@millard2014].
+
+Without data repositories, most comprehensive analyses are no longer feasible.
+So far, there has been no overarching database that constantly collects both 14C and dendrochronological data and makes them accessible in the long term.
+In the spirit of Open Science, it is the goal of XRONOS to establish such an Open Access database.
+Our primary ambition is to collect 14C and dendrochronological data for Switzerland and Europe as extensively and comprehensively as possible.
+However, the project is designed in such a way that an expansion beyond the originally defined spatial focus is already a reality.
+
 ## Compilations of radiocarbon dates
@@
 
+Currently, all scientific disciplines face a revolution in the way data is disseminated and used (Bronk Ramsey et al., 2019).
+The access to these data, and methods integrate individual data into larger scientific questions of social relevance are both are decisive, but not yet sufficiently implemented prerequisites to reach an understanding of the temporal phenomena and dynamics in past societies (Kristiansen, 2014).
+This novel, cross-disciplinary circulation and utilisation makes it possible to deal with completely new scientific questions and makes archaeological data relevant in thoroughly new contexts.
+For example, analysis of past climatic changes and their long-term effects rely on information from archaeologists to estimate the impact on human and biotic activities (Kintigh et al., 2015, 7--8).
+An important source of such information is absolute chronological data.However, the basis of any empirically based modelling is the availability of data.
+
+In recent years, there have been various initiatives, especially to collect 14C data and make it available to the research public.
+This has resulted in some publicly accessible databases.
+Unfortunately, however, those databases that currently exist have so far been insufficient for large-scale and reliable analyses for various reasons.
+For radiocarbon dating, there are only databases that are limited by their specific focus
+
+-   Some databases, such as KIK-IRPA, ORAU and BANADORA, which are maintained by 14C laboratories, only contain their own measurement data and are therefore of very limited use for overarching research projects.
+
+-   Others, such as CONTEXT, RADON, RADON-B, 14SEA and CARD, have a limited regional or temporal focus, and the rich data available e.g.
+in Switzerland and surrounding countries is not sufficiently represented in these databases.
+In addition, some of them are project-specific and their continued existence can be described as uncertain due to funding constraints.
+
+-   Recently, [@bronkramsey2019] introduced IntChron, a data structure designed to ensure easy availability of radiocarbon data from published articles.
+We think this approach is very commendable and it inspires us in terms of reproducibility of dating.
+But we want to go beyond that and put more emphasis on the evaluation and dissemination of relevant information.
+"IntChron is not intended to be a stand-alone database" [@bronkramsey2019, 5], but XRONOS is.
+An overarching database is critical for any research application in terms of data retrievability.
+Nevertheless, interoperability with other data services, including IntChron in particular, is very important to us.
+
+For European dendrochronological data, the situation is even less favourable for archaeological purposes:
+
+-   NOAA stores some global datasets, but has its main focus in the United States and, with its only 34 European prehistoric datasets, is not useful for the study of the European past.
+
+-   The ADS database is compiled and maintained by the UK-based Vernacular Architecture Group and is limited to data published annually in the journal Vernacular Architecture, containing only medieval or later dates for the UK.
+
+-   DendroDB has a European focus but only provides data for historical periods.
+
+-   The "Dendrochronological Database" of the Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) has a worldwide focus on collected natural wood samples and has different goals than a database for archaeological wood material (furthermore, although announced a long time ago, this database is still not functional).
+
+-   Although "Digital Collaboratory for Cultural Dendrochronology (DCCD).
+An international digital data library for dendrochronology" is well connected in the context of archaeological data services (e.g.
+ARIADNE), it is focused on the Netherlands (more than 2/3 of the data, while 0.08% of the Quercus data refer to Switzerland) and contains very few prehistoric data (about 2.5%).
+Furthermore, the overall activity in this database has decreased from 3846 new project records in 2010-2014 to only 83 since then (until the end of 2019).
+
+What is completely missing worldwide so far, is a database that encompasses both, 14C and dendro data.
+However, to efficiently conduct data driven modelling of human-environmental dependencies and dynamics on the past, such a data base seems indispensable for future research.
+
 ## XRONOS
 
+XRONOS was originally built on the basic structure of the RADON database. 
+At the time, this was one of the largest 14C data collections for absolute chronological information in Europe. 
+RADON and RADON-B can be considered a complete success, as they have been the basis for several of our own publications (Hinz, 2015; Hinz, Feeser, et al., 2012; Krossa et al., 2017; Weinelt et al., 2015) and as a data repository for high-level publications by other research groups (Bevan et al, 2017, 2019; Dolukhanov and Shrukov, 2004; Henkner et al., 2018; Lemmen and Wirtz, 2014; Manning et al, 2016; Pinhasi et al, 2005; Puchol and García, 2017; Sanders, 2012; Shennan et al, 2013; Timpson et al, 2014), and inspired other projects (Schmid, 2018; IsoMemo; Schmid et al, 2018).
+However, the focus of XRONOS goes far beyond the structures established there, which made extensive new development necessary. 
+Thanks to funding from an Initiator Grant from the University of Bern (Caroline Heitz, project "Time and Temporality"), a first operational version of XRONOS and further activities could already be implemented in 2020 by Clemens Schmid (Max Planck Institute for the Study of Human History, Jena).
+The current database is modular and, in its current stage of development, contains all the structures necessary for storing 14C data. 
+In addition, the integration of dendrochronological information into the structure is already prepared, and a security and data protection architecture has been prepared for the final stage of development.
+The software is available as open source on Github (<https://github.com/xronos-ch),> and an R package for easy access, data retrieval and evaluation is also already available. 
+XRONOS currently holds 92,129 14C data from 22,225 sites. 
+The primary focus is on Switzerland and Europe, but the database also contains information from Africa, South America, Asia and Australia. 
+The collection is subject to continuous addition.
+
 ### Design goals and principles
+
+Our concrete goals in the implementation were and are the following:
+
+1. Provision of access to 14C and dendrochronological data with meta information on archaeological contexts via web-based Open Access frontend
+2. Web frontend with processing functions (filters and queries, export to various data formats, 14C calibration & chronology formation) for in place analytical functions
+3. machine readable interface --- API --- for direct access e.g. by open source statistical environments (eg. R) for analytical purposes or dendroanalytical software
+4. licensing and versioning of individual data using eg. DOI
+5. Storage in a repository with user and role-specific access rights as well as encryption for data privacy
+6. Ensuring sustainable data security and usage through compliance with standards, appropriate backup solutions and high-availability technologies
+
+Points 1-3 are now largely implemented and ready to be used.
+The results of such a search is easily exportable in formats that are simple and generally usable, and represent common, open standards (CSV, JSON).
+Such formats can be conveniently imported into common evaluation environments (R, Matlab, Python, OxCal, Chronomodel), or even better, directly accessed via the API.
+This can serve to establish a single gateway to the different sources of information and eliminate the necessity for data publishers to maintain information in multiple formats.
+With regard to the 14C data, this analytical functions include the live calibration of the individual dates, allowing the most recent calibration curve to be considered.
+In addition there are extended calibration possibilities like sum, as known from the standard software solutions (e.g.
+OxCal).
+Furthermore, the Bayesian calibration in relation to phase models (sequential calibration under consideration of stratigraphy) may be implemented as far as possible, whereby this will take place in close connection with the endeavours of the further development of corresponding R packages.
+For the dendrochronological data, especially the implementation of visualizations as well as the possibility to create mean curves and to identify impact phases will represent extended possibilities of immediate analysis.
+This portfolio is to be expanded in consultation with the user group, especially the state (cantonal) dendrochronological laboratories of Switzerland.
+More elaborated functionality should be made available primarily via an R Package, since this would be too complex for the user to be channelled via an web interface.
+
+An increasing number of research funders legitimately demand that the data that is used in the course of the research they fund become available to other researchers.
+This disclosure enables new knowledge to be extracted through comparative studies, data mining, etc.; it also enables better transferability and reproducibility of the research process, which increases the quality of research.
+All data in XRONOS should be uniquely referenceable.
+Common publication practices, which so far led to ambiguities regarding the data, must be overcome.
+Therefore, each individual data set should be permanently archived in its current version, and changes resulting from updated information or changed boundary conditions (updated calibration curves) should be provided in a new version with versioned unique identification.
+In this way, it can be ensured that the exact state of a reference to a date can be unambiguously stored.
+Only in this way can the research based on these data truly become traceable.
+One means for this is a Digital object identifier (DOI), which combines a unique and machine-readable as well as persistent referencing with easy retrievability.
+This functionality is still to be implemented.
+In order to make such a sustainable data referencing possible, XRONOS strives for a cooperation with longterm data repositories, so that a permanent storage of the relevant information can be guaranteed.
+This ensures that the collected information will be available to the research and public community beyond the lifetime of XRONOS and thus beyond a possible end of the project.
+
+The ultimate goal of XRONOS is to create an open data structure for the exchange and evaluation of absolute chronological data by the entire research community.
+However, the data management with regard to dendrochronological data in particular is associated with different interests of the institutions involved, so that compromises have to be made here.
+For all dendrochronological laboratories control over data disclosure is indispensable.
+These are mainly publication-related reasons and data is used as "exchange capital" between laboratories.
+However, in order to provide an incentive for central data management, it must be ensured that users retain full sovereignty over their data.
+To this end, this project plans to keep the data fully encrypted in such a way that even the database administrators have no insight into the data.
+For this purpose, methods of asymmetric cryptography (public/private key encryption) based on passwords will be integrated into the database, which make it possible to make data available to individual other users or to the general public.
+
+We hope that the combination of easy and open data access, the provision of analytical functions within the web interface and by means of R packages, the sustainable, transparent and unambiguous data management, the long-term archiving already considered in the initial phase, together with the possibility for individual users to have full sovereignty over their data, will distinguish XRONOS as an outstanding data source and repository, and that we will thereby contribute to the sustainable provision and use of big data in archaeology and advance the use and improvement of standards in this field.
 

### 2024-05-08 paper.qmd
Update text

@@
 ---
-title: "XRONOS: an open data infrastructure for archaeological chronometry"
+title: "XRONOS: an open infrastructure for archaeological chronometric data"
+short-title: XRONOS
 author:
@@
     orcid: 0000-0002-1011-1244
-    url: "https://joeroe.io"
     affiliation:
@@
     orcid: 0000-0002-9904-6548
-    url: "https://martinhinz.info"
     email: martin.hinz@unibe.ch
@@
 It provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world.
+keywords: 
+  - open data
+  - chronometry
+  - radiocarbon dating
+  - dendrochronology
+  - typological dating
+  - chronological modelling
+bibliography: references.bib
+format:
+  hikmah-pdf:
+    papersize: a4
+    mainfont: "Linux Libertine"
+    sansfont: "URW Gothic"
+    biblatex-chicago: true
+    biblio-style: authordate
+    biblatexoptions:
+      - backend=biber
+      - autolang=hyphen
+      - isbn=false
+      - uniquename=false
+#  html:
+#    page-layout: article
+#    toc: true
+#    embed-resources: true
+---
+
+<!-- 
+For submission to Archaeometry. Guidelines:
+
+- Max 8000 words including all figures etc.
+- Max 6 figures/tables
+- Abstract max 100 words
+- Max 10 keywords
+- Word 2002 or earlier or RTF (???)
+- Tables seperately in XLS
+- Figures seperately in TIFF or EPS
+- double spaced, wide
+  -->
+
+## Introduction
+
+* Chronometric modelling depends on ready access to chronometric data!
+* Luckily, we do have a lot of it.
+* Moving from 'open data' (uploading a CSV) to 'open infrastructure' [@Kintigh2006]
 
@@
 Here, building on and complementing these initiatives, we present XRONOS: a new web-based platform for chronometric data from archaeological contexts worldwide, combining an open data repository with tools for importing, curating and analysing chronometric information from diverse sources.
-The principal design goals of the software are to combine all available sources of radiocarbon and other chronometric data in single database; develop robust tools for the continuous ingestion and refinement of this data; and disseminate this data within an open and FAIR framework, embedding it in the wider world of Linked Open Data in archaeology and beyond.
+The principal design goals of the software are to combine all available sources of radiocarbon and other chronometric data in single database; 
+develop robust tools for the continuous ingestion and refinement of this data; 
+and disseminate this data within an open and FAIR framework, embedding it in the wider world of Linked Open Data in archaeology and beyond.
 
-  This paper outlines the conceptual and technical infrastructure developed to realise these goals in XRONOS’ initial phases of development (2019 and 2021–2023), including a generalised data model for site and radiocarbon information, extendable to other chronometric data; an R- and Ruby-based pipeline for continuous ingestion of data from a variety of sources; continuous, semi-automated data cleaning protocols; a Ruby-on-Rails application providing a web-based frontend to the data and a REST API for programmatic access; and an R package for interfacing with the API.
+This paper outlines the conceptual and technical infrastructure developed to realise these goals in XRONOS’ initial phases of development (2019 and 2021–2024), including a generalised data model for site and radiocarbon information, extendable to other chronometric data;
+an R- and Ruby-based pipeline for continuous ingestion of data from a variety of sources; continuous, semi-automated data cleaning protocols; 
+a Ruby-on-Rails application providing a web-based frontend to the data and a REST API for programmatic access; 
+and an R package for interfacing with the API. 
 We believe the XRONOS framework provides more open, more reliable, and more comprehensive access to chronometric data than previously available, as well as a foundation for its continuous expansion and refinement.
-copyright: 
-  holder: the authors
-  year: 2023
-license: "CC BY"
-# citation: 
-#  container-title: Archaeometry
-#  volume: 1
-#  issue: 1
-#  doi: 10.5555/12345678
-funding: "This work was funded by the Swiss National Science Foundation ([SNSF Project #198152](https://data.snf.ch/grants/grant/198153)) and the University of Bern (Initiator Grant, Caroline Heitz, Project 'Time and Temporality')"
-bibliography: references.bib
-format:
-  html:
-    page-layout: article
-    toc: true
-    embed-resources: true
----
-
-## Introduction
-
-* Chronometric modelling depends on ready access to chronometric data!
-* Luckily, we do have a lot of it.
-* Moving from 'open data' (uploading a CSV) to 'open infrastructure' [@CITE]
 
@@
 * Review the state of the art in radiocarbon data compilation (and other chronometric data, such as they are), highlighting the need for XRONOS
-* Describe XRONOS' software architecture, design goals, and currently-implemented features
+* Describe XRONOS' design goals and software architecture.
+* Describe the implementation of some representative features
   * With a focus on the choices we made in representing/modelling chronological data
@@
 * Conclude with our future plans
+
+## State of the art
 
@@
 The problem, however, is that these data are very scattered.
-Traditionally, radiocarbon data were disseminated by the laboratories, with special attention paid to publication in the form of "date lists" in the journal "Radiocarbon".
+Traditionally, radiocarbon data were disseminated by the laboratories, with special attention paid to publication in the form of "date lists" in the journal *Radiocarbon*.
 Today, unfortunately, this has become very rare, on the one hand because the quantity and complexity of the data limit the possibilities of publication on paper.
 On the other hand, data in printed form are hardly suitable for reuse today.
-As a result, access to chronological information, especially a comprehensive overview, is made much more difficult [@bronkramsey2019].
-Furthermore, data is presented in a variety of different formats and granularities, and critical information is often missing, although standards exist [@millard2014].
+As a result, access to chronological information, especially a comprehensive overview, is made much more difficult [@BronkRamseyEtAl2019].
+Furthermore, data is presented in a variety of different formats and granularities, and critical information is often missing, although standards exist [@Millard2014].
 
@@
 
-## Compilations of radiocarbon dates
+Currently, all scientific disciplines face a revolution in the way data is disseminated and used [@BronkRamseyEtAl2019].
+The access to these data, and methods integrate individual data into larger scientific questions of social relevance are both are decisive, but not yet sufficiently implemented prerequisites to reach an understanding of the temporal phenomena and dynamics in past societies [@Kristiansen2014].
+This novel, cross-disciplinary circulation and utilisation makes it possible to deal with completely new scientific questions and makes archaeological data relevant in thoroughly new contexts. 
+For example, analysis of past climatic changes and their long-term effects rely on information from archaeologists to estimate the impact on human and biotic activities [@KintighEtAl2015, 7-8]<!-- From Antiquity paper, not sure which paper exactly this is -->.
+An important source of such information is absolute chronological data.
+However, the basis of any empirically based modelling is the availability of data.
 
-* Radiocarbon dates used to be regularly compiled and published in *Radiocarbon* (so it was 'open data' before it was cool), but this is no longer practical.
-* Plus the bare measurements are of limited use without contextual information, which labs don't usually have access to.
-* So even though radiocarbon data comes from a relatively limited number of sources [@IntCalLabList] and has relatively standardised reporting conventions [@Millard2014], in practice data has to be manually compiled from literature – this is time-consuming!
-* Since at least 19XX, there have been efforts to systematically collect and compile this information
+### Compilations of radiocarbon dates
+
+* Where do radiocarbon dates come from?
+* Where are they published?
+
+Although 'open data' is often thought to be a new phenomenon in archaeology [@CITE], it has a long history in radiometric dating.
+Starting in the 19XXs, radiocarbon laboratories shared and compiled as 'date lists' published on a regular basis in the journal *Radiocarbon* [@CITE].
+However, as the number of labs and volume of radiocarbon dates being produced grew, this paper-based format became impractical: the last 'date list' appeared in 19XX, and unfortunately was not replaced by another form of systematic data-sharing or dissemination.
+Additionally, because the date lists were sourced from radiocarbon laboratories directly—not those who collected the sample—they typically included only very limited contextual information.
+Thus, even though radiocarbon data comes from a relatively limited number of sources [@IntCalLabList] and has relatively standardised reporting conventions [@Millard2014], in practice the only way to produce aggregated datasets recently has been to manually search through relevant literature for dates reported by the submitter of the sample.
+This already laborious process is further hampered by a significant inconsistency in how much authors adhere to measurement reporting conventions [@Millard2014], the lack of any convention on the reporting of contextual information, and weak or nonexistent disciplinary norms regarding the responsibility to publish results openly in a timely fashion.
+
 * Graph of number of dates/databases through time
-* But this has been fragmentary: we counted at least XX databases
+* Table of databases
+
+Despite these difficulties, there have been many efforts to systematically compile radiocarbon dates in this way over the last few decades [@FIG].
+
+E.g. RADON:
+At the time, this was one of the largest 14C data collections for absolute chronological information in Europe. 
+RADON and RADON-B can be considered a complete success, as they have been the basis for several of our own publications (Hinz, 2015; Hinz, Feeser, et al., 2012; Krossa et al., 2017; Weinelt et al., 2015) and as a data repository for high-level publications by other research groups (Bevan et al, 2017, 2019; Dolukhanov and Shrukov, 2004; Henkner et al., 2018; Lemmen and Wirtz, 2014; Manning et al, 2016; Pinhasi et al, 2005; Puchol and García, 2017; Sanders, 2012; Shennan et al, 2013; Timpson et al, 2014), and inspired other projects (Schmid, 2018; IsoMemo; Schmid et al, 2018).
+
+We are aware of XX existing radiocarbon databases or published datasets, comprising an estimated XXXX individual records [@TABLE] – though this is likely an underestimate.
+The majority of these resources have a specific regional (XX%) or temporal (XX%) scope.
+Others are associated with individual radiocarbon labs (e.g. KIK-IRPA, ORAU, BANADORA) and a handful aim at global coverage (XX%, discussed further below).
+
+* How are they published? As databases, SI, etc.?
+* How many are open?
+
+While these compilations have greatly advanced access to chronometric data and enabled new modes of analysis (e.g.
+'dates as data', @CITE), there are some obvious limitations to the fragmentary way in which they have been compiled.
+
 * And patchy: highly uneven spatial and geographic coverage, different data collected, different conventions for recording data
-
-## Time for a global radiocarbon database?
-
 * Regional databases are put together *ad hoc* and rarely continuously maintained over long periods
 * The coverage of these databases is uneven, systematically biased, and duplicative of each other
-* Other forms of chronometric data (dendro, typology, sea level, etc.) remain relatively untouched
+* Not all are open
+* Not all are machine readable
 
@@
 
-Currently, all scientific disciplines face a revolution in the way data is disseminated and used (Bronk Ramsey et al., 2019).
-The access to these data, and methods integrate individual data into larger scientific questions of social relevance are both are decisive, but not yet sufficiently implemented prerequisites to reach an understanding of the temporal phenomena and dynamics in past societies (Kristiansen, 2014).
-This novel, cross-disciplinary circulation and utilisation makes it possible to deal with completely new scientific questions and makes archaeological data relevant in thoroughly new contexts.
-For example, analysis of past climatic changes and their long-term effects rely on information from archaeologists to estimate the impact on human and biotic activities (Kintigh et al., 2015, 7--8).
-An important source of such information is absolute chronological data.However, the basis of any empirically based modelling is the availability of data.
-
-In recent years, there have been various initiatives, especially to collect 14C data and make it available to the research public.
-This has resulted in some publicly accessible databases.
-Unfortunately, however, those databases that currently exist have so far been insufficient for large-scale and reliable analyses for various reasons.
-For radiocarbon dating, there are only databases that are limited by their specific focus
-
--   Some databases, such as KIK-IRPA, ORAU and BANADORA, which are maintained by 14C laboratories, only contain their own measurement data and are therefore of very limited use for overarching research projects.
-
--   Others, such as CONTEXT, RADON, RADON-B, 14SEA and CARD, have a limited regional or temporal focus, and the rich data available e.g.
+* Some databases, such as KIK-IRPA, ORAU and BANADORA, which are maintained by 14C laboratories, only contain their own measurement data and are therefore of very limited use for overarching research projects.
+* Others, such as CONTEXT, RADON, RADON-B, 14SEA and CARD, have a limited regional or temporal focus, and the rich data available e.g.
 in Switzerland and surrounding countries is not sufficiently represented in these databases.
 In addition, some of them are project-specific and their continued existence can be described as uncertain due to funding constraints.
-
--   Recently, [@bronkramsey2019] introduced IntChron, a data structure designed to ensure easy availability of radiocarbon data from published articles.
+* Recently, [@BronkRamseyEtAl2019] introduced IntChron, a data structure designed to ensure easy availability of radiocarbon data from published articles.
 We think this approach is very commendable and it inspires us in terms of reproducibility of dating.
 But we want to go beyond that and put more emphasis on the evaluation and dissemination of relevant information.
-"IntChron is not intended to be a stand-alone database" [@bronkramsey2019, 5], but XRONOS is.
+"IntChron is not intended to be a stand-alone database" [@BronkRamseyEtAl2019, 5], but XRONOS is.
 An overarching database is critical for any research application in terms of data retrievability.
@@
 
+### Beyond radiocarbon
+
+* There are lots of other types of chronometric data!
+  * Dendro - in certain places (e.g. Switzerland!), way more important than radiocarbon
+  * Other radiometric/absolute dating methods, OSL, etc.
+  * Typology - still the backbone of many chronologies
+  * Application-specific: sea level dating, rock patina, etc.
+* Compilation of other types of dates has been much more limited.
+  * Exceptions: ...?
+
+<!-- MH to do: -->
 For European dendrochronological data, the situation is even less favourable for archaeological purposes:
 
--   NOAA stores some global datasets, but has its main focus in the United States and, with its only 34 European prehistoric datasets, is not useful for the study of the European past.
-
--   The ADS database is compiled and maintained by the UK-based Vernacular Architecture Group and is limited to data published annually in the journal Vernacular Architecture, containing only medieval or later dates for the UK.
-
--   DendroDB has a European focus but only provides data for historical periods.
-
--   The "Dendrochronological Database" of the Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) has a worldwide focus on collected natural wood samples and has different goals than a database for archaeological wood material (furthermore, although announced a long time ago, this database is still not functional).
-
--   Although "Digital Collaboratory for Cultural Dendrochronology (DCCD).
+* NOAA stores some global datasets, but has its main focus in the United States and, with its only 34 European prehistoric datasets, is not useful for the study of the European past.
+* The ADS database is compiled and maintained by the UK-based Vernacular Architecture Group and is limited to data published annually in the journal Vernacular Architecture, containing only medieval or later dates for the UK.
+* DendroDB has a European focus but only provides data for historical periods.
+* The "Dendrochronological Database" of the Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) has a worldwide focus on collected natural wood samples and has different goals than a database for archaeological wood material (furthermore, although announced a long time ago, this database is still not functional).
+* Although "Digital Collaboratory for Cultural Dendrochronology (DCCD).
 An international digital data library for dendrochronology" is well connected in the context of archaeological data services (e.g.
@@
 
+* Typology
+  * perio.do
+  * Often embedded in other types of database (e.g. gazetteers)
+  * But nobody has treated it as a form of chronometric data in its own right?
+
 What is completely missing worldwide so far, is a database that encompasses both, 14C and dendro data.
@@
 
-## XRONOS
+## Concept
 
-XRONOS was originally built on the basic structure of the RADON database. 
-At the time, this was one of the largest 14C data collections for absolute chronological information in Europe. 
-RADON and RADON-B can be considered a complete success, as they have been the basis for several of our own publications (Hinz, 2015; Hinz, Feeser, et al., 2012; Krossa et al., 2017; Weinelt et al., 2015) and as a data repository for high-level publications by other research groups (Bevan et al, 2017, 2019; Dolukhanov and Shrukov, 2004; Henkner et al., 2018; Lemmen and Wirtz, 2014; Manning et al, 2016; Pinhasi et al, 2005; Puchol and García, 2017; Sanders, 2012; Shennan et al, 2013; Timpson et al, 2014), and inspired other projects (Schmid, 2018; IsoMemo; Schmid et al, 2018).
-However, the focus of XRONOS goes far beyond the structures established there, which made extensive new development necessary. 
-Thanks to funding from an Initiator Grant from the University of Bern (Caroline Heitz, project "Time and Temporality"), a first operational version of XRONOS and further activities could already be implemented in 2020 by Clemens Schmid (Max Planck Institute for the Study of Human History, Jena).
-The current database is modular and, in its current stage of development, contains all the structures necessary for storing 14C data. 
-In addition, the integration of dendrochronological information into the structure is already prepared, and a security and data protection architecture has been prepared for the final stage of development.
-The software is available as open source on Github (<https://github.com/xronos-ch),> and an R package for easy access, data retrieval and evaluation is also already available. 
-XRONOS currently holds 92,129 14C data from 22,225 sites. 
-The primary focus is on Switzerland and Europe, but the database also contains information from Africa, South America, Asia and Australia. 
-The collection is subject to continuous addition.
+XRONOS builds on the structure of RADON (@RADON), one of the earliest comprehensive databases of radiocarbon dates.
 
-### Design goals and principles
+### Design goals
 
-Our concrete goals in the implementation were and are the following:
+Our concrete goals in the design and implementation of XRONOS are the following:
 
@@
 Key to 'unifying' chrons in this way is a common model of time.
+
 * Time element can be represented as a probability distribution
@@
 
-* CRUD Rails app
-  * Not so exciting, but try to keep up with modern standards, remain accessible, etc.
-  * Data browser
+The XRONOS data model is implemented as a standard relational database using the free and open source database management system PostgreSQL.
+However, apart from backups and other routine maintenance procedures, all interaction with the database is via a web application, which thus forms the core of XRONOS' architecture.
+The XRONOS web application is written in Ruby and uses well-known CRUD (Create, Read, Update, Delete) and MVC (Model, View, Controller) patterns as implemented in the Ruby on Rails framework.
+The choice of this somewhat traditional or 'boring' architecture—as opposed to the non-relational/semantic technologies that are more *à la mode* in the digital humanities [@Fan2018; @Hyvonen2020; @SchloenProsser2023]—was purposeful, motivated by a sense that it offers a better chance of longevity and maintanability.
+Our aim is for the software architecture to be as boring as possible so that the scientific contents can be as interesting as possible.
+
+The XRONOS web application exposes two distinct user interfaces: a graphical user interface accessed through a web browser; and an application programming interface (API). 
+Both interfaces follow a REST (Representational State Transfer) pattern [@VerborghEtAl2015], where each resource (e.g.
+a single radiocarbon date, a single bibliographic reference, or a single user) is statelessly mapped to a single address.
+Users can then interact with resources at these addresses using a preditable and uniform interface based on HTTP verbs.
+For example, the radiocarbon date RTD-8904 is represented by the address <https://xronos.ch/c14s/156205>.
+Users can view information on this resource by sending a GET request to that address, regardless of which interface they are using, and authorised users can modify it using POST, PATCH, or DELETE requests.
+The bibliographic reference associated with this date [@RichterEtAl2017] is similarly represented at the address <https://xronos.ch/references/17778>, and can be accessed at that address using the same interface as the radiocarbon date.
+These uniform RESTful interfaces are another example of a boring architectural choice that make it easier for us to enrich the scientific contents of XRONOS, by  adding new types of modular resource that represent new scientific entities.
+
+The graphical user interface to XRONOS, accessed through a web browser (e.g.
+at https://xronos.ch), uses RESTful resources as the building blocks for various interfaces through which users can browse, search, retrieve, and analyse chronometric data.
+It also provides 'backstage' interfaces (currently not public) for creating, editing, and deleting data, and monitoring data quality.
+
+* RESTful
+* One record one page
+* Data
+* No sharp public/private boundary
+  * Eventually have multiple access levels, semi-public access to editing features, ability to have private data
+
+XRONOS' API ...
+  * Data browser -> implementation
 * API
+  * JSON
+  - RESTFUL
   * R package
 * Linked data
+
+* Do stuff on the server side
+  * Because JS is fragile
+  * And then we can use it in multiple interfaces
 
@@
 
-* Assessment of XRONOS' coverage
+<!-- TODO: update/rewrite status of development -->
+The current database is modular and, in its current stage of development, contains all the structures necessary for storing 14C data. 
+In addition, the integration of dendrochronological information into the structure is already prepared, and a security and data protection architecture has been prepared for the final stage of development.
+The software is available as open source on Github (<https://github.com/xronos-ch),> and an R package for easy access, data retrieval and evaluation is also already available. 
+XRONOS currently holds 92,129 14C data from 22,225 sites. 
+The primary focus is on Switzerland and Europe, but the database also contains information from Africa, South America, Asia and Australia. 
+The collection is subject to continuous addition.
+
+Due to space limitations, and because we hope and expect XRONOS to continuously evolve, we don't attempt a comprehensive description of its implementation here.
+Instead we offer a selection of representative features which illustrate our design goals and key aspects of our approach.
+
+* Data import / assessment of XRONOS' coverage
+* The data browser
+* XRONOS R package
+* Tale from the trenches: cleaning species and countries?
 * Tale from the trenches: duplicate processing?
-* Tale from the trenches: cleaning species and countries?
 * Case study of use?
@@
 
+## Data availability statement
+
+TODO
+
+## Funding statement
+
+This work was funded by the Swiss National Science Foundation ([SNSF Project #198152](https://data.snf.ch/grants/grant/198153)) and the University of Bern (Initiator Grant, Caroline Heitz, Project 'Time and Temporality').
+
 ## Acknowledgements
 
-Direct contributors to XRONOS (or invite as coauthors?):
+Albert Hafner. Plus contributors to XRONOS:
 
-* Albert Hafner
 * Chiara Huwiler

### 2024-05-09 paper.qmd
Expand text

@@
 
-XRONOS data model is based on the concept of a set of coordinates in time or *chron*.
-We define a chron as an assertion that links human activity with a particular point in space and time.
-More concretely, our data model currently encompasses three types of chron: radiocarbon dates, typological dates (e.g. 'Early Neolithic') and dendrochronological dates.
-However we anticipate that the concept will accommodate all sorts of absolute and relative dating techniques, as the scope of the database expands.
+At the base of the XRONOS data model are sets of spatiotemporal coordinates or, as we call them, *chrons*.
+In an archaeological context, we conceptualise a chron as an assertion linking human activity with a particular point in space and time.
+Our data model currently encompasses three types of chron: radiocarbon dates, typological dates (e.g. 'Early Neolithic') and dendrochronological dates.
+However we anticipate that the concept will accommodate other types of absolute and relative dating techniques, as the scope of the database expands.
 
-Chrons are useful conceptually because they emphasise that dates from different sources still consist of essentially the same kind of information: a statement locating an event in space and time.
+Chrons are conceptually useful because they emphasise that different types of archaeological 'dates', drawn from different sources, have essentially the same information content: the location of an event in space and time.
 We thereby avoid privileging certain sources of chronological data over (as might be the case if, for example, we treated 'period' as a fixed attribute of a site) and can accommodate contradictory (e.g.
@@
 This is important given that XRONOS aspires to be an authoratative 'backbone' with a global scope, so we cannot realistically impose a single chronological scheme or resolve conflicting information provided by specialists.
-They are useful practically because they expose a common interface for attributes that all types of chronological information share, such as a *terminus post quem*, *terminus ante quem*, and midpoint estimate.
+They are useful practically because they expose a common interface for attributes that all types of chronological information share, such as a *terminus post quem* (TPQ), *terminus ante quem* (TAQ), and midpoint estimate.
 This allows applications that use XRONOS' data model (including XRONOS itself) to collate chronological data from multiple sources, without necessarily having to be aware of the pecularities of each type of dating.
 
-Key to 'unifying' chrons in this way is a common model of time.
-
-* Time element can be represented as a probability distribution
-  * Inspired by OxCal, aoristic analysis
-  * How you get to the PD representation can be very different (e.g. calibration vs. aoristic), but the important thing is the endpoint is the same.
+In order to unify chronological information in the form of a chron, we need a common chronological 'coordinate system'.
+The natural choice is a *calendar probability distribution*, which expresses the probability that an event occurred as a function of time on a calendric scale.
+Most archaeologists are familiar with working with this kind of representation in the form of calibrated radiocarbon dates, but it can be extended and generalised to essentially any kind of chronological information [@CITE].
+For example, in aoristic analysis [@CITE], a periodic time estimate (e.g.
+the event occurred in the Neolithic) is conceptualised as a uniform probability distribution over the timespan between the known start and end dates of that period.
+A similar model is used in OxCal [@OxCal, a direct inspiration for our approach] to integrate prior chronological information from diverse sources.
+In practical terms, this model means that the canonical representation the time component of any chron in XRONOS', regardless of source, is a probably distribution over the set of calendar years (arbitrarily measured in years Before Present) in which it could have plausibly occurred.
+Further statistics, e.g. a midpoint estimate or TPQ/TAQ range, can be derived from this distribution using well-known methods.
+In this way, we can support many different types of date and much of the implementation of XRONOS can be agnostic to the source of chronological information.
 
@@
 
-* Contextual information: sample, context, site
-  * Different types of xrons can be attached to the same sample, context, or site, as appropriate
+Further contextual information is associated with *contexts* and *sites*.
+The site is the primary geographic container for chronological information.
+As already mentioned, we typically record the spatial location of chrons using this entity, though it is possible to modify this by providing specific coordinates at the sample level.
+Sites also have attributes describing their conventional name or names in different languages and are associated with a flexible 'site type' typology that combines information on their form and function.
+
+A context represents the specific find-context of a sample, e.g. an architectural feature, stratigraphic unit, or phase.
+Since the units and conventions for recording such information vary greatly between different regions and archaeological traditions---and XRONOS is designed with global data in mind---we leave the question of what a context precisely represents open, and only record an unstandardised, free text label for it.
+Crucially, however, contexts can have a self-referential association to other contexts belonging to the same site.
+This allows it to encode arbitrary relational structures between contexts, whether they be hierarchical (e.g. phases and sub-phases) or graphical (e.g. stratigraphic).
+In this way, it can serve as a foundation for chronological modelling. <!-- TODO: reword, maybe a bit vague? -->
+
+The series of relations `[chron] > sample > context > site` links the chronological and contextual sides of the XRONOS data model.
+Each step is a many-to-many association, meaning for example that it is possible to attach multiple chrons to the same sample (e.g.
+replicated radiocarbon dates on the same material), multiple *types* of chrons to the same sample (e.g.
+radiocarbon dates on tree-rings for wiggle-matching).
+Since this kind of information is rarely systematically recorded in our source databases, there are currently few actual records that make use of this feature of the data model.
+However, we hope it will provide a foundation for ... ? <!-- TODO -->
+
 * Bibliographic references
@@
 
+* LOD??
+
 ### Software architecture
 
-The XRONOS data model is implemented as a standard relational database using the free and open source database management system PostgreSQL.
+The XRONOS data model is implemented as a relational database using the free and open source database management system PostgreSQL.
 However, apart from backups and other routine maintenance procedures, all interaction with the database is via a web application, which thus forms the core of XRONOS' architecture.
-The XRONOS web application is written in Ruby and uses well-known CRUD (Create, Read, Update, Delete) and MVC (Model, View, Controller) patterns as implemented in the Ruby on Rails framework.
+The XRONOS web application is written in Ruby and uses CRUD (Create, Read, Update, Delete) and MVC (Model, View, Controller) patterns as implemented in the Ruby on Rails framework.
 The choice of this somewhat traditional or 'boring' architecture—as opposed to the non-relational/semantic technologies that are more *à la mode* in the digital humanities [@Fan2018; @Hyvonen2020; @SchloenProsser2023]—was purposeful, motivated by a sense that it offers a better chance of longevity and maintanability.
@@
 The bibliographic reference associated with this date [@RichterEtAl2017] is similarly represented at the address <https://xronos.ch/references/17778>, and can be accessed at that address using the same interface as the radiocarbon date.
-These uniform RESTful interfaces are another example of a boring architectural choice that make it easier for us to enrich the scientific contents of XRONOS, by  adding new types of modular resource that represent new scientific entities.
+These uniform REST interfaces are another example of a boring architectural choice that make it easier for us to enrich the scientific contents of XRONOS, by  adding new types of modular resource that represent new scientific entities.
 
-The graphical user interface to XRONOS, accessed through a web browser (e.g.
-at https://xronos.ch), uses RESTful resources as the building blocks for various interfaces through which users can browse, search, retrieve, and analyse chronometric data.
-It also provides 'backstage' interfaces (currently not public) for creating, editing, and deleting data, and monitoring data quality.
+This basic REST pattern is augmented by seven 'actions' (following the standard pattern in Rails application) that express different ways of interacting with a resource: index, show, destroy, new, create, edit, and update.
+The 'show' action represents interaction with a single resource, as described above. 
+The 'index' action, which lists resources of a given type (e.g.
+<https://xronos.ch/c14s> for radiocarbon dates), is worth special mention because it is through this that the filtering logic at the core of XRONOS' two interfaces is implemented.
+By passing a query as HTTP GET parameters to the index action of a resource, the list returned the user is modified to only include records that match that query.
+For example, <https://xronos.ch/sites?site[country_code]=CH> (the part of the URL after the `?` character encodes the SQL WHERE clause `country = 'CH'` as a GET parameter) lists sites in Switzerland <!-- TODO: check this actually works -->.
+More complex queries can be executed using nested parameters.
+For example, <https://xronos.ch/c14s?c14[sample][material][name]=charcoal> (encoding that the `c14` table should be joined to the `material` table via `sample`, followed by the WHERE clause `material.name = 'charcoal'`) lists radiocarbon dates obtained from charcoal samples <!-- TODO: check that this actually works...
+-->.
+Uniquely, index actions can also respond with the result in a tabular data format (i.e. `.csv`).
 
-* RESTful
-* One record one page
-* Data
-* No sharp public/private boundary
-  * Eventually have multiple access levels, semi-public access to editing features, ability to have private data
+The graphical user interface (GUI) to XRONOS, accessed through a web browser (e.g.
+at https://xronos.ch), uses REST resources and actions as the building blocks for various interfaces through which users can browse, search, retrieve, and analyse chronometric data.
+Each action on each resource is represented by a page, though not all of these are publicly accessible.
+Pages representing REST resources directly are supplemented by a number of synthetic interfaces, for example the 'data browser' (<https://xronos.ch/data>), which facilitates more complex filtering, or the search interface (<https://xronos.ch/search>).
+The GUI also includes several resources which are not part of XRONOS' scientific data model, for example documentation pages, user profiles and news articles; these are currently not exposed in the API.
 
-XRONOS' API ...
-  * Data browser -> implementation
-* API
-  * JSON
-  - RESTFUL
-  * R package
-* Linked data
+Access to various 'backstage' interfaces for creating, editing, and deleting data, and monitoring data quality is managed using a user permissions system.
+Currently only authorised users affiliated with the XRONOS project can access these, but in the future we intend to support open registration and expose editing interfaces to all authenticated users.
+For this reason, there is no sharp division between a 'public' and 'private' areas – viewing/querying data and editing/curating data share the same architecture and interface patterns.
+<!-- TODO: say something about private data, even though it's not implemented yet? -->
 
-* Do stuff on the server side
-  * Because JS is fragile
-  * And then we can use it in multiple interfaces
+The XRONOS API uses the same addresses as the web-based GUI (with the exception of some of the synthetic interfaces mentioned above) but responds with machine-readable data in JSON format, rather than a HTML page.
+This response can be triggered by appending `.json` to the address or by including a HTTP `content-format` header in the request.
+Though users can make such requests manually and parse the data with one of several off-the-shelf tools, the primary intended uses of this interface is to provide access for 1) programmatic clients to XRONOS and 2) other web services.
+The XRONOS R package (discussed in more detail below) is an example of the former use, which could be extended to other programming environments used for scientific computing, e.g. Python or Haskell.
+<!-- TODO: example of a web service that might use the API? -->
+
+An overarching principle of this software architecture is that all interaction with XRONOS' data store, and as much of the data processing and 'business logic' of responding to REST requests as possible, is directed through the same server-side routines.
+First and foremost, this allows us to provide multiple interfaces (i.e. the GUI and API, perhaps more in the future) without duplicating these elements of our codebase.
+It also improves accessibility for users accessing XRONOS through devices with limited processing capability or through text-only browsers.
+More broadly, avoiding reliance on client-side processing, e.g.
+with Javascript or Web Assembly (WASM)---which would be the other option---allows us keep our client interfaces simple (in most cases plain, semantic HTML pages and self-contained stylesheets) and therefore, we hope, sustainable in the face of constantly-evolving client-side technologies and standards.
+It does have the weakness that, in practical terms, XRONOS is difficult to run and relies on the continued existence of a maintained external server.
+We have however tried to mitigate this by providing clearly-documented source code and regular data dumps so that, if our instance of XRONOS disappears, or if one simply does not want to use it, it is possible for others to host a XRONOS server of their own.
+As the sustainability of scientific software and data infrastructures is a pressing problem (a point we will return to in the conclusion), in the future it may be desirable to support further decentralisation through, for example, a federated server-server model [@CITE].
 

### 2024-07-29 paper.qmd
Rewrite for JCAA

@@
 - double spaced, wide
+
+OR as a JCAA 'case study' - 5,000 words
   -->

### 2024-07-29 paper_jcaa.qmd
Rewrite for JCAA

--- 2024-05-09 paper.qmd
@@
 ---
-title: "XRONOS: an open infrastructure for archaeological chronometric data"
+title: "XRONOS: an open data infrastructure for archaeological chronology"
 short-title: XRONOS
@@
     orcid: 0000-0002-1011-1244
+    email: joseph.a.roe@unibe.ch
+    corresponding: true
     affiliation:
@@
     orcid: 0000-0002-9904-6548
-    email: martin.hinz@unibe.ch
-    corresponding: true
     affiliation:
@@
   XRONOS (<https://xronos.ch>) is an open data infrastructure for the backbone of the archaeological record – chronology.
-It provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world.
+It aims provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world.
+Collating a number of recent regional and global compilations of dates, XRONOS offers the most comprehensive radiocarbon database yet published, with over 350,000 date and 75,000 site records.
+It also provides a foundation for expanding the systematic collection of chronometric information beyond radiocarbon, with support for typological and dendrochronological dates and a generalisable data model that can be adapted to other methods of absolute dating.
+Automated and semi-automated quality control processes ensure that data from diverse sources is continuously integrated and standardised, making it significantly easier to filter data of interest and reducing the need for manual data cleaning by end users.
+In this paper we describe the concept and implementation of XRONOS in relation to the state of the art in chronometric data-sharing, and evaluate its potential as a general-purpose open repository and curation platform for archaeological chronology.
 keywords: 
   - open data
+  - chronology
   - chronometry
@@
   - typological dating
-  - chronological modelling
+published: "Manuscript for submission to the *Journal of Computer Applications in Archaeology*"
 bibliography: references.bib
@@
 <!--
-For submission to Archaeometry. Guidelines:
-
-- Max 8000 words including all figures etc.
-- Max 6 figures/tables
-- Abstract max 100 words
-- Max 10 keywords
-- Word 2002 or earlier or RTF (???)
-- Tables seperately in XLS
-- Figures seperately in TIFF or EPS
-- double spaced, wide
+* [ ] First draft:
+  * [ ] Introduction [500]
+  * [ ] State of the art [2000]
+    * [ ] Compilations of radiocarbon dates [1000]
+    * [x] Open data infrastructures [500]
+    * [ ] Beyond radiocarbon [500]
+  * [ ] Concept [1000]
+    * [x] Inherits a common structure from RADON
+    * [ ] Design goals [250]
+    * [x] Data model [1150/500]
+    * [x] LOD [250/250]
+  * [x] Implementation [1000]
+    * [x] Status of development [100/100]
+    * [x] Software architecture [550/300]
+    * [x] Data ingestion and integration [500/300]
+    * [x] User interfaces [650/300]
+  * [ ] Evaluation [500]
+* [ ] Add references:
+  * [ ] Introduction
+  * [ ] State of the art
+  * [ ] Concept
+  * [ ] Implementation
+  * [ ] Evaluation
+* [ ] Figures and tables
+* [ ] Copyedit:
+  * [ ] Introduction
+  * [ ] State of the art
+  * [ ] Concept
+  * [ ] Implementation
+  * [ ] Evaluation
   -->
@@
 
-* Chronometric modelling depends on ready access to chronometric data!
-* Luckily, we do have a lot of it.
-* Moving from 'open data' (uploading a CSV) to 'open infrastructure' [@Kintigh2006]
+* Chronology is the backbone of the archaeological record – we need it to be open!
 
-Chronometric data has been one of the principle domains in which open data and open science approaches have been applied over the last ten years. 
-Comprehensive compilations of radiocarbon dates have become available for many parts of the world in the last decade and, as natural next step, there are now several initiatives to collate this data globally, including the retrieval tool c14bazAAR [@SchmidEtAl2019], the IntChron exchange format [@BronkRamseyEtAl2019], and the synthetic database p3kc14 [@BirdEtAl2022].
-This move from archiving individual datasets to producing a living open data infrastructure for archaeological chronologies raises a number of challenges. 
-Radiocarbon datasets are still sorely lacking for many parts of the world and, even in those regions with good coverage, the quality of data is highly uneven and largely undocumented. 
-There exists no central repository ensuring the long-term sustainability and completeness of these datasets, and the potential of placing other sources of chronometric information (e.g.
-dendrochronology, typological dating) in an open data framework has hardly been realised at all.
-
-Here, building on and complementing these initiatives, we present XRONOS: a new web-based platform for chronometric data from archaeological contexts worldwide, combining an open data repository with tools for importing, curating and analysing chronometric information from diverse sources.
-The principal design goals of the software are to combine all available sources of radiocarbon and other chronometric data in single database; 
-develop robust tools for the continuous ingestion and refinement of this data; 
-and disseminate this data within an open and FAIR framework, embedding it in the wider world of Linked Open Data in archaeology and beyond.
-
-This paper outlines the conceptual and technical infrastructure developed to realise these goals in XRONOS’ initial phases of development (2019 and 2021–2024), including a generalised data model for site and radiocarbon information, extendable to other chronometric data;
-an R- and Ruby-based pipeline for continuous ingestion of data from a variety of sources; continuous, semi-automated data cleaning protocols; 
-a Ruby-on-Rails application providing a web-based frontend to the data and a REST API for programmatic access; 
-and an R package for interfacing with the API. 
-We believe the XRONOS framework provides more open, more reliable, and more comprehensive access to chronometric data than previously available, as well as a foundation for its continuous expansion and refinement.
-
-In this paper we:
-
-* Review the state of the art in radiocarbon data compilation (and other chronometric data, such as they are), highlighting the need for XRONOS
-* Describe XRONOS' design goals and software architecture.
-* Describe the implementation of some representative features
-  * With a focus on the choices we made in representing/modelling chronological data
-* Assess the current state of the software and database, and give case studies of how it can be used
-* Conclude with our future plans
-
-## State of the art
-
-We live in interesting times: Today, archaeological research can no longer complain that too few absolute dates exist for the phenomena of the past.
-Large-scale dating projects such as \[example\] and \[example\] are producing extensive data series in a short time, which make it possible to situate individual sites in time with unprecedented precision.
-But also due to the dating methods that have become affordable even for smaller budgets in the meantime and the integration of natural scientific dating that is almost mandatory, such large initiatives are accompanied by a great amount of individual and scattered chronological information that represents a rich source of data for supra-regional investigations.
-Based on that, data modelling can be used as a method to draw conclusions from a variety of individually information-weak data.
-
-The problem, however, is that these data are very scattered.
-Traditionally, radiocarbon data were disseminated by the laboratories, with special attention paid to publication in the form of "date lists" in the journal *Radiocarbon*.
-Today, unfortunately, this has become very rare, on the one hand because the quantity and complexity of the data limit the possibilities of publication on paper.
-On the other hand, data in printed form are hardly suitable for reuse today.
-As a result, access to chronological information, especially a comprehensive overview, is made much more difficult [@BronkRamseyEtAl2019].
-Furthermore, data is presented in a variety of different formats and granularities, and critical information is often missing, although standards exist [@Millard2014].
-
-Without data repositories, most comprehensive analyses are no longer feasible.
-So far, there has been no overarching database that constantly collects both 14C and dendrochronological data and makes them accessible in the long term.
-In the spirit of Open Science, it is the goal of XRONOS to establish such an Open Access database.
-Our primary ambition is to collect 14C and dendrochronological data for Switzerland and Europe as extensively and comprehensively as possible.
-However, the project is designed in such a way that an expansion beyond the originally defined spatial focus is already a reality.
-
-Currently, all scientific disciplines face a revolution in the way data is disseminated and used [@BronkRamseyEtAl2019].
-The access to these data, and methods integrate individual data into larger scientific questions of social relevance are both are decisive, but not yet sufficiently implemented prerequisites to reach an understanding of the temporal phenomena and dynamics in past societies [@Kristiansen2014].
-This novel, cross-disciplinary circulation and utilisation makes it possible to deal with completely new scientific questions and makes archaeological data relevant in thoroughly new contexts. 
-For example, analysis of past climatic changes and their long-term effects rely on information from archaeologists to estimate the impact on human and biotic activities [@KintighEtAl2015, 7-8]<!-- From Antiquity paper, not sure which paper exactly this is -->.
-An important source of such information is absolute chronological data.
-However, the basis of any empirically based modelling is the availability of data.
-
-### Compilations of radiocarbon dates
-
-* Where do radiocarbon dates come from?
-* Where are they published?
-
+<!-- Prehistory of open radiocarbon 
+TODO: shorten -->
 Although 'open data' is often thought to be a new phenomenon in archaeology [@CITE], it has a long history in radiometric dating.
@@
 
-* Graph of number of dates/databases through time
-* Table of databases
+* Move to open compilations in the last ~10 years
+* Range of other chronological information used by archaeologists is largely untouched
+* From uploading CSVs to (social) infrastructure
 
-Despite these difficulties, there have been many efforts to systematically compile radiocarbon dates in this way over the last few decades [@FIG].
+## State of the art
 
-E.g. RADON:
-At the time, this was one of the largest 14C data collections for absolute chronological information in Europe. 
-RADON and RADON-B can be considered a complete success, as they have been the basis for several of our own publications (Hinz, 2015; Hinz, Feeser, et al., 2012; Krossa et al., 2017; Weinelt et al., 2015) and as a data repository for high-level publications by other research groups (Bevan et al, 2017, 2019; Dolukhanov and Shrukov, 2004; Henkner et al., 2018; Lemmen and Wirtz, 2014; Manning et al, 2016; Pinhasi et al, 2005; Puchol and García, 2017; Sanders, 2012; Shennan et al, 2013; Timpson et al, 2014), and inspired other projects (Schmid, 2018; IsoMemo; Schmid et al, 2018).
+### Compilations of radiocarbon dates
 
-We are aware of XX existing radiocarbon databases or published datasets, comprising an estimated XXXX individual records [@TABLE] – though this is likely an underestimate.
-The majority of these resources have a specific regional (XX%) or temporal (XX%) scope.
-Others are associated with individual radiocarbon labs (e.g. KIK-IRPA, ORAU, BANADORA) and a handful aim at global coverage (XX%, discussed further below).
+* Where do radiocarbon dates come from and how are they published?
+* History of systematic compilations
+* Summary of existing databases
+  * FIGURE: graph of number of dates/databases through time
+  * FIGURE: map of radiocarbon database coverage
+  * TABLE: radiocarbon databases
+* Problems with current state of the art
 
-* How are they published? As databases, SI, etc.?
-* How many are open?
+### Open data infrastructures
 
-While these compilations have greatly advanced access to chronometric data and enabled new modes of analysis (e.g.
-'dates as data', @CITE), there are some obvious limitations to the fragmentary way in which they have been compiled.
+The profusion of radiocarbon compilations over the last decade has naturally prompted many to think more globally:
 
-* And patchy: highly uneven spatial and geographic coverage, different data collected, different conventions for recording data
-* Regional databases are put together *ad hoc* and rarely continuously maintained over long periods
-* The coverage of these databases is uneven, systematically biased, and duplicative of each other
-* Not all are open
-* Not all are machine readable
+- The first available radiocarbon 'metadatabase' was **c14bazAAR** [@CITE], an R package that provides an index of openly published radiocarbon databases and a common interface for retrieving them and performing basic data cleaning.
+Because c14bazAAR downloads data from its original source repositories, rather than mirroring it, it only includes resources that have been published in a fully open and machine-accessible format.
+Despite this limitation, it has global coverage and is still the largest collection of radiocarbon dates available (see @FIG).
 
-* Recent moves towards a global synthesis: c14bazAAR [@SchmidEtAl2019], IntChron [@BronkRamseyEtAl2019], p3k14c [@BirdEtAl2022]
-  * Each taking a different approach, pros and cons
-  * We see a need for something that is comprehensive, unified, living
-* And we don't see why it should be limited to radiocarbon: what about other radiometric dates? Dendro? Typologies?
+- Another indexical approach is the **IntChron** project [@CITE], which exposes data from multiple sources and exposes them with a common JSON-based web interface.
+The IntChron specification is open, meaning that radiocarbon labs or compilation projects can implement it independently and thereby allow end users to access their data through a common itnerface (though to our knowledge it has so far only been adopted by databases associated with the Oxford Radiocarbon Lab).
+The JSON format also lends itself to the implementation of wrapper libraries, for example the 'rIntChron' package gives direct access to IntChron-indexed databases in R [@CITE].
 
-* Some databases, such as KIK-IRPA, ORAU and BANADORA, which are maintained by 14C laboratories, only contain their own measurement data and are therefore of very limited use for overarching research projects.
-* Others, such as CONTEXT, RADON, RADON-B, 14SEA and CARD, have a limited regional or temporal focus, and the rich data available e.g.
-in Switzerland and surrounding countries is not sufficiently represented in these databases.
-In addition, some of them are project-specific and their continued existence can be described as uncertain due to funding constraints.
-* Recently, [@BronkRamseyEtAl2019] introduced IntChron, a data structure designed to ensure easy availability of radiocarbon data from published articles.
-We think this approach is very commendable and it inspires us in terms of reproducibility of dating.
-But we want to go beyond that and put more emphasis on the evaluation and dissemination of relevant information.
-"IntChron is not intended to be a stand-alone database" [@BronkRamseyEtAl2019, 5], but XRONOS is.
-An overarching database is critical for any research application in terms of data retrievability.
-Nevertheless, interoperability with other data services, including IntChron in particular, is very important to us.
+* **p3k14c** [@CITE] instead compiles multiple source databases into a single flat file dataset, with a similar level of coverage to c14bazAAR.
+The major advantage of this approach is that the data is made internally consistent and has been manually cleaned to an extent, which makes it particularly well-suited to global analyses.
+The downside is that without the continuous link to the source databases present in the c14bazAAR and IntChron, it can only be kept up to date manually with periodic re-releases.
+An accompanying package (@CITE) provides direct access to the p3k14c dataset in R.
+
+XRONOS blends aspects of all three approaches to achieve the same aim of providing access to the global radiocarbon data through a common interface.
+Like c14bazAAR and IntChron, it is a 'metadatabase' that draws from existing data resources and maintains an explicit link to them.
+But like p3k14c, it integrates these into a single database and applies data curation processes to harmonise them and improve the quality of the information.
+It has a wider scope than c14bazAAR or IntChron, as it mirrors rather than directly retrieves the source data (allowing us to use resources that aren't openly published), and does not rely on the authors of these sources to implement a common specification.
+It also goes beyond the functionality of p3k14c by providing systems for the contunious ingestion and curation of new data as it is published.
+However in general we see the approaches as complementary.
+A c14bazAAR parser for XRONOS is in development (https://github.com/ropensci/c14bazAAR/pull/150), and we also aim to provide an IntChron interface to XRONOS' data in the near future.
+New data and corrections from p3k14c are incorporated into XRONOS as they are released.
+
+XRONOS is our answer to Kintigh's call [@Kintigh2006] for digital infrastructures that don't just provide access to chronological data but enables researchers to "archive, access, integrate, and mine disparate data sets".
+It parallels and draws inspiration from several similar initiatives in archaeology, such as <!-- TODO examples -->
+And outside of it: <!-- TODO examples GBIF -->
 
@@
 
-* There are lots of other types of chronometric data!
-  * Dendro - in certain places (e.g. Switzerland!), way more important than radiocarbon
-  * Other radiometric/absolute dating methods, OSL, etc.
-  * Typology - still the backbone of many chronologies
-  * Application-specific: sea level dating, rock patina, etc.
-* Compilation of other types of dates has been much more limited.
-  * Exceptions: ...?
-
-<!-- MH to do: -->
-For European dendrochronological data, the situation is even less favourable for archaeological purposes:
-
-* NOAA stores some global datasets, but has its main focus in the United States and, with its only 34 European prehistoric datasets, is not useful for the study of the European past.
-* The ADS database is compiled and maintained by the UK-based Vernacular Architecture Group and is limited to data published annually in the journal Vernacular Architecture, containing only medieval or later dates for the UK.
-* DendroDB has a European focus but only provides data for historical periods.
-* The "Dendrochronological Database" of the Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) has a worldwide focus on collected natural wood samples and has different goals than a database for archaeological wood material (furthermore, although announced a long time ago, this database is still not functional).
-* Although "Digital Collaboratory for Cultural Dendrochronology (DCCD).
-An international digital data library for dendrochronology" is well connected in the context of archaeological data services (e.g.
-ARIADNE), it is focused on the Netherlands (more than 2/3 of the data, while 0.08% of the Quercus data refer to Switzerland) and contains very few prehistoric data (about 2.5%).
-Furthermore, the overall activity in this database has decreased from 3846 new project records in 2010-2014 to only 83 since then (until the end of 2019).
-
+* Dendro
 * Typology
-  * perio.do
-  * Often embedded in other types of database (e.g. gazetteers)
-  * But nobody has treated it as a form of chronometric data in its own right?
-
-What is completely missing worldwide so far, is a database that encompasses both, 14C and dendro data.
-However, to efficiently conduct data driven modelling of human-environmental dependencies and dynamics on the past, such a data base seems indispensable for future research.
+* Other radiometric methods
+* Application-specific 
+* XRONOS aims to bring all these together, starting with radiocarbon, dendro-, and typology
 
@@
 
-XRONOS builds on the structure of RADON (@RADON), one of the earliest comprehensive databases of radiocarbon dates.
+XRONOS inherits its basic structure from RADON (@CITE), with a database-backed web application and a data model that separates radiocarbon dates, contextual information, and sites.
+Our overall aims in developing XRONOS is to bring this model, which RADON has operated on for XX years, up to date, to generalise it to other types of chronometric information, and to transform it from an online database to a data infrastructure that supports the continuous ingestion, curation, and open dissemination of archaeological chronologies from diverse sources.
 
@@
 
-Our concrete goals in the design and implementation of XRONOS are the following:
-
-1. Provision of access to 14C and dendrochronological data with meta information on archaeological contexts via web-based Open Access frontend
-2. Web frontend with processing functions (filters and queries, export to various data formats, 14C calibration & chronology formation) for in place analytical functions
-3. machine readable interface --- API --- for direct access e.g. by open source statistical environments (eg. R) for analytical purposes or dendroanalytical software
-4. licensing and versioning of individual data using eg. DOI
-5. Storage in a repository with user and role-specific access rights as well as encryption for data privacy
-6. Ensuring sustainable data security and usage through compliance with standards, appropriate backup solutions and high-availability technologies
-
-Points 1-3 are now largely implemented and ready to be used.
-The results of such a search is easily exportable in formats that are simple and generally usable, and represent common, open standards (CSV, JSON).
-Such formats can be conveniently imported into common evaluation environments (R, Matlab, Python, OxCal, Chronomodel), or even better, directly accessed via the API.
-This can serve to establish a single gateway to the different sources of information and eliminate the necessity for data publishers to maintain information in multiple formats.
-With regard to the 14C data, this analytical functions include the live calibration of the individual dates, allowing the most recent calibration curve to be considered.
-In addition there are extended calibration possibilities like sum, as known from the standard software solutions (e.g.
-OxCal).
-Furthermore, the Bayesian calibration in relation to phase models (sequential calibration under consideration of stratigraphy) may be implemented as far as possible, whereby this will take place in close connection with the endeavours of the further development of corresponding R packages.
-For the dendrochronological data, especially the implementation of visualizations as well as the possibility to create mean curves and to identify impact phases will represent extended possibilities of immediate analysis.
-This portfolio is to be expanded in consultation with the user group, especially the state (cantonal) dendrochronological laboratories of Switzerland.
-More elaborated functionality should be made available primarily via an R Package, since this would be too complex for the user to be channelled via an web interface.
-
-An increasing number of research funders legitimately demand that the data that is used in the course of the research they fund become available to other researchers.
-This disclosure enables new knowledge to be extracted through comparative studies, data mining, etc.; it also enables better transferability and reproducibility of the research process, which increases the quality of research.
-All data in XRONOS should be uniquely referenceable.
-Common publication practices, which so far led to ambiguities regarding the data, must be overcome.
-Therefore, each individual data set should be permanently archived in its current version, and changes resulting from updated information or changed boundary conditions (updated calibration curves) should be provided in a new version with versioned unique identification.
-In this way, it can be ensured that the exact state of a reference to a date can be unambiguously stored.
-Only in this way can the research based on these data truly become traceable.
-One means for this is a Digital object identifier (DOI), which combines a unique and machine-readable as well as persistent referencing with easy retrievability.
-This functionality is still to be implemented.
-In order to make such a sustainable data referencing possible, XRONOS strives for a cooperation with longterm data repositories, so that a permanent storage of the relevant information can be guaranteed.
-This ensures that the collected information will be available to the research and public community beyond the lifetime of XRONOS and thus beyond a possible end of the project.
-
-The ultimate goal of XRONOS is to create an open data structure for the exchange and evaluation of absolute chronological data by the entire research community.
-However, the data management with regard to dendrochronological data in particular is associated with different interests of the institutions involved, so that compromises have to be made here.
-For all dendrochronological laboratories control over data disclosure is indispensable.
-These are mainly publication-related reasons and data is used as "exchange capital" between laboratories.
-However, in order to provide an incentive for central data management, it must be ensured that users retain full sovereignty over their data.
-To this end, this project plans to keep the data fully encrypted in such a way that even the database administrators have no insight into the data.
-For this purpose, methods of asymmetric cryptography (public/private key encryption) based on passwords will be integrated into the database, which make it possible to make data available to individual other users or to the general public.
-
-We hope that the combination of easy and open data access, the provision of analytical functions within the web interface and by means of R packages, the sustainable, transparent and unambiguous data management, the long-term archiving already considered in the initial phase, together with the possibility for individual users to have full sovereignty over their data, will distinguish XRONOS as an outstanding data source and repository, and that we will thereby contribute to the sustainable provision and use of big data in archaeology and advance the use and improvement of standards in this field.
-
-* Inspired by GBIF, which manages to integrate large amounts of data from diverse sources into a signal, consistent database
+<!-- TODO: Look at Martin's grant application and drafts -->
 
@@
 
-TODO: Figure: conceptual scheme, centred on chrons (i.e. an idealised form of the database).
+* FIGURE: conceptual scheme
+
+<!-- TODO: shorten -->
 
@@
 
-* Bibliographic references
-* Paper trail
+Metadata is incorporated into XRONOS' data model at the level of the individual records (e.g.
+all records store their data of creation and last modification) and through two additional nodes: bibliographic references and versions.
+Bibliographic references store information on the source of a record in the standard BibTeX (@CITE) format and can be linked through many-to-many associations to
+sites or chrons.
+Versions are a special type of record that are associated with all other records (including bibliographic references) and store the previous versions of those records as a series of changesets.
+In this way all changes to data are recorded and can be reconstructed (or reversed) precisely.
+This 'paper trail' also stores contextual metadata, e.g. who made the change and why.
+It also means that records that are deleted can be reviewed or restored from their stored version history, which is never discarded.
+Together these two systems provide a transparent record where the data in XRONOS comes from and how it has been altered, which we view as essential in a scientific data infrastructure.
 
-* LOD??
+### Linked data
+
+The XRONOS data model presents several opportunities to link to other resources as linked open data (LOD) [@CITE].
+We use controlled vocabularies in the form of the GBIF Backbone Taxonomy (for taxonomic descriptions of samples) via its API, and in future plan to extend this to for example site types (using the Getty and/or DAI<!-- TODO: check & cite -->).
+Standardising field using these vocabularies to automatically then link the two entities.
+Beyond this, there is also a concordance between XRONOS entities and other specialised data infrastructures in archaeology, such as Perio.do (@CITE, mapping to typological chrons) and gazetters like Pleiades or Vici (@CITE, mapping to site names).
+Moving beyond archaeology, Wikidata (@CITE) already includes many of the concepts represented in the XRONOS model (e.g. archaeological sites, Q<!-- TODO -->).
+Linking to Wikidata is especially useful because it dissimenates—and thereby preserves—the data compiled in XRONOS beyond the narrow field of archaeology.
+It also allows us to enrich the database with contextual information that otherwise be beyond our scope and resources, for example embedding multilingual descriptions of sites from Wikipedia articles on site records where the site has been linked to a Wikidata item.
+
+Conversely, we encourage others to use XRONOS as a linked open data resource by providing stable URLs and a machine-readable interface for every record.
+Such usages could look like, for example, using a XRONOS URL as a canonical representation of a single radiocarbon date (e.g. <!-- TODO -->).
+We also plan to implement an IntChron ([@CITE]) interface to XRONOS, to allow it to be indexed through that existing standard.<!-- TODO: also mention RDF? -->
+
+## Implementation
+
+Following a short pilot project in 2019, the first phase of development of XRONOS was completed in 2021–2024, supported by a project grant from the Swiss National Science Foundation.
+The web interface (<https://xronos.ch>) has been publicly accessible since July 2021.
+Though we envisage XRONOS as a continuously-developing open source project and 'living database', the following offers a snapshot of progress at the end of our first grant-funded implementation phase.
+We do not aim to be comprehensive, but rather to describe some key elements of XRONOS' current implementation that illustrate how our concept has been realised in practice.
 
 ### Software architecture
+<!-- TODO: shorten -->
 
+<!-- Overview -->
 The XRONOS data model is implemented as a relational database using the free and open source database management system PostgreSQL.
@@
 
+<!-- REST -->
 The XRONOS web application exposes two distinct user interfaces: a graphical user interface accessed through a web browser; and an application programming interface (API). 
@@
 
+<!-- Actions and queries -->
 This basic REST pattern is augmented by seven 'actions' (following the standard pattern in Rails application) that express different ways of interacting with a resource: index, show, destroy, new, create, edit, and update.
@@
 
+### Data ingestion and curation
+
+The chronological data in XRONOS comes from a variety of sources, including published structured datasets in repositories and journal supplements, other online databases, literature review, and direct input from collaborators.
+Our aim is not just to 'mirror' these sources as they are, but integrate them into a single curated and continuously updated database.
+<!-- TODO: too detailed?
+-->For the purposes of ingestion, we classify data resources into three categories: static resources, such as supplementary data in published papers, which are imported once; versioned resources, updated on a periodic basis, which we import after each new version; and live resources, which are continuously updated and therefore continuously imported.
+Records of each type are imported into XRONOS in as close to their original state as possible, i.e. without any corrections or standardisation applied.
+This ensures that any subsequent changes (even immediate, automatic ones) are entered into the record's version history, so that the source of any deviations or potential errors can always be reconstructed.
+The version history also records the direct source of the data for attribution purposes.
+In addition, a bibliographic reference to the original resource is attached to ensure that the source is clearly attributed even if the record is merged with another one.
+
+Once ingested, we apply a number of automated and semi-automated quality control processes to integrate new data into the existing database.
+Controlled vocabularies are used in a number of places in the data model (@FIG?), and we use thesauruses to automatically standardise these fields as much as possible.
+For example, the taxonomic description of samples is controlled using GBIF's backbone taxonomy (@CITE), and we also use a thesaurus service provided by GBIF to automatically change variant or obsolete taxonomic names to the canonical version.
+If the system is not able to standardise a field using the available thesaurus, it is flagged for manual correction.
+A wide variety of other potential data quality issues (e.g.
+missing data on what country a site is in) are also flagged for human review by this system, which canoften be semi-automated (e.g.
+suggesting close matches in the thesaurus or the country indicated by the record's coordinates).
+
+A final critical component of XRONOS' data curation system is duplicate handling.
+We import data from many overlapping resources (many of which incorporate each other either in whole or in part), so duplicate records are common.
+The end result of standardising and correcting a record is also often to create a duplicate: e.g.
+the same sample imported from one source as 'oak' but another as '*Quercus* sp.' will become a duplicate pair as '*Quercus*', and thus be recognised as a single sample.
+Such exact duplicates can be merged automatically, with the oldest record becoming the authoratitive version, but detecting fuzzier duplicated information (e.g.
+differences in the spelling of site names) has proved a more difficult problem.
+As of writing there are therefore still many duplicate records in XRONOS that need to be manually resolved, but we hope to automate much more of this work in the future.
+
+### User interfaces
+
+<!-- TODO: figures? -->
+<!-- TODO: add in more concrete examples -->
+
+<!-- Public GUI -->
 The graphical user interface (GUI) to XRONOS, accessed through a web browser (e.g.
@@
 
+<!-- Non-public GUI -->
 Access to various 'backstage' interfaces for creating, editing, and deleting data, and monitoring data quality is managed using a user permissions system.
@@
 
+<!-- API / R package -->
 The XRONOS API uses the same addresses as the web-based GUI (with the exception of some of the synthetic interfaces mentioned above) but responds with machine-readable data in JSON format, rather than a HTML page.
@@
 Though users can make such requests manually and parse the data with one of several off-the-shelf tools, the primary intended uses of this interface is to provide access for 1) programmatic clients to XRONOS and 2) other web services.
-The XRONOS R package (discussed in more detail below) is an example of the former use, which could be extended to other programming environments used for scientific computing, e.g. Python or Haskell.
-<!-- TODO: example of a web service that might use the API? -->
+The XRONOS R package [@CITE] is an example of a programmatic client; it uses the API to facilitate direct querying and retrieval of data from XRONOS in the R statistical programming language [@CITE], which is widely used for computational applications in archaeology [@CITE].
+Similar libraries could be developed in other programming environments used for scientific computing, such as Python or Haskell.
+The API also provides the foundation for other web services to access XRONOS directly, to embed chronological information in other contexts or otherwise make use of its data resources.
 
+<!-- Conclusion. TODO: do we need this? -->
 An overarching principle of this software architecture is that all interaction with XRONOS' data store, and as much of the data processing and 'business logic' of responding to REST requests as possible, is directed through the same server-side routines.
@@
 
-## Implementation
+## Evaluation
 
-The first prototype of XRONOS was developed in 2019.
-It has been under sustained development 2021–2024 with the first public release in July 2021.
-
-<!-- TODO: update/rewrite status of development -->
-The current database is modular and, in its current stage of development, contains all the structures necessary for storing 14C data. 
-In addition, the integration of dendrochronological information into the structure is already prepared, and a security and data protection architecture has been prepared for the final stage of development.
-The software is available as open source on Github (<https://github.com/xronos-ch),> and an R package for easy access, data retrieval and evaluation is also already available. 
-XRONOS currently holds 92,129 14C data from 22,225 sites. 
-The primary focus is on Switzerland and Europe, but the database also contains information from Africa, South America, Asia and Australia. 
-The collection is subject to continuous addition.
-
-Due to space limitations, and because we hope and expect XRONOS to continuously evolve, we don't attempt a comprehensive description of its implementation here.
-Instead we offer a selection of representative features which illustrate our design goals and key aspects of our approach.
-
-* Data import / assessment of XRONOS' coverage
-* The data browser
-* XRONOS R package
-* Tale from the trenches: cleaning species and countries?
-* Tale from the trenches: duplicate processing?
-* Case study of use?
-
-## Conclusion
-
-* Chronology is the backbone of the archaeological record
-* XRONOS could be a backbone for a universe of linked open data resources...
-* ...but sustainability is a challenge.
-  * We're coming to the end of our project funds
-  * We'll try to get more, but you never know (grant-based science and long-term investments in digital infrastructure don't work so well together!)
-  * Future sustainability
-
-## Data availability statement
-
-TODO
-
-## Funding statement
-
-This work was funded by the Swiss National Science Foundation ([SNSF Project #198152](https://data.snf.ch/grants/grant/198153)) and the University of Bern (Initiator Grant, Caroline Heitz, Project 'Time and Temporality').
-
-## Acknowledgements
-
-Albert Hafner. Plus contributors to XRONOS:
-
-* Chiara Huwiler
-* Rivana Moser
-* Tomasz
-* Stephanie Döppler
+* XRONOS is useable, use XRONOS!
+* Immediate development goals
+* The sustainability challenge

### 2024-08-03 paper_jcaa.qmd
Add lit review and data analysis

@@
 published: "Manuscript for submission to the *Journal of Computer Applications in Archaeology*"
+execute:
+  echo: false
+number-sections: true
 bibliography: references.bib
@@
 ---
+
+```{r setup}
+library("countrycode")
+library("cowplot")
+library("dplyr", warn.conflicts = FALSE)
+library("giscoR")
+library("ggplot2")
+library("glue")
+library("khroma")
+library("patchwork", warn.conflicts = FALSE)
+library("readr")
+library("sf")
+library("tidyr")
+library("tinytable")
+
+#' Count number of rows in a filtered data frame
+n_filter <- function(.data, ...) {
+  nrow(dplyr::filter(.data, ...))
+}
+```
 
@@
 * [ ] Figures and tables
+  * [x] FIGURE: graph of number of datasets through time
+  * [x] FIGURE: map of number of dates in datasets
+  * [ ] FIGURE: distribution of radiocarbon dates in c14bazAAR / IntChron / p3k14c / XRONOS
+  * [ ] FIGURE: data model
+  * [ ] FIGURE: GUI screenshots (data browser, c14 show view, version history, data curation)
+  * [ ] TABLE: list of 'issues'
+  * [ ] APPENDIX: table of datasets
 * [ ] Copyedit:
@@
 
-## Introduction
+# Introduction
 
-* Chronology is the backbone of the archaeological record – we need it to be open!
+Chronology is the backbone of the archaeological record.
+It is needed for X, Y, and Z.
+Open access to chronological data is therefore a critical...
 
-<!-- Prehistory of open radiocarbon 
-TODO: shorten -->
-Although 'open data' is often thought to be a new phenomenon in archaeology [@CITE], it has a long history in radiometric dating.
-Starting in the 19XXs, radiocarbon laboratories shared and compiled as 'date lists' published on a regular basis in the journal *Radiocarbon* [@CITE].
-However, as the number of labs and volume of radiocarbon dates being produced grew, this paper-based format became impractical: the last 'date list' appeared in 19XX, and unfortunately was not replaced by another form of systematic data-sharing or dissemination.
-Additionally, because the date lists were sourced from radiocarbon laboratories directly—not those who collected the sample—they typically included only very limited contextual information.
-Thus, even though radiocarbon data comes from a relatively limited number of sources [@IntCalLabList] and has relatively standardised reporting conventions [@Millard2014], in practice the only way to produce aggregated datasets recently has been to manually search through relevant literature for dates reported by the submitter of the sample.
-This already laborious process is further hampered by a significant inconsistency in how much authors adhere to measurement reporting conventions [@Millard2014], the lack of any convention on the reporting of contextual information, and weak or nonexistent disciplinary norms regarding the responsibility to publish results openly in a timely fashion.
-
-* Move to open compilations in the last ~10 years
+* Profusion of open C14 data in the last 20–30 years calls for global solutions
 * Range of other chronological information used by archaeologists is largely untouched
@@
 
-## State of the art
+# State of the art
 
-### Compilations of radiocarbon dates
+## Compilations of radiocarbon dates
 
-* Where do radiocarbon dates come from and how are they published?
-* History of systematic compilations
-* Summary of existing databases
-  * FIGURE: graph of number of dates/databases through time
-  * FIGURE: map of radiocarbon database coverage
-  * TABLE: radiocarbon databases
-* Problems with current state of the art
+Though 'open data' in archaeology is a relatively recent phenomenon [@CITE], the open publication of compiled radiocarbon dates has a substantial prehistory.
+Starting in the 19XXs, radiocarbon laboratories shared and compiled as 'date lists' published on a regular basis in the journal *Radiocarbon* [@CITE].
+However, as the number of labs and volume of radiocarbon dates being produced grew, this paper-based format became impractical: the last 'date list' appeared in 19XX, and was not replaced by another form of systematic data-sharing or dissemination.
+<!-- TODO: actually continued in *Archaeometry* -->
+Additionally, because the date lists were sourced from radiocarbon laboratories directly—not those who collected the sample—they typically included only very limited contextual information.
+On the eve of the AMS revolution there was an effort to create a computerised 'International Radiocarbon Database' [@Kra1988]—already by 1989 described as a "much needed, long overdue enterprise" [@Kra1989]—but it never came to fruition.
 
-### Open data infrastructures
+Thus, even though radiocarbon data comes from a relatively limited number of sources [@IntCalLabList] and has relatively standardised reporting conventions [@Millard2014; @Bayliss2015], in practice the only way to produce aggregated datasets in recent decades has been to manually search through relevant literature for dates reported by the submitter of the sample.
+This already laborious process is further hampered by a significant inconsistency in how much authors adhere to measurement reporting conventions, a lack of conventions on the reporting of *contextual* information, and weak or nonexistent disciplinary norms regarding the responsibility to publish results openly in a timely fashion.
 
-The profusion of radiocarbon compilations over the last decade has naturally prompted many to think more globally:
+```{r data-c14-datasets}
+c14_datasets <- read_tsv("data/c14_datasets.tsv", show_col_types = FALSE)
 
-- The first available radiocarbon 'metadatabase' was **c14bazAAR** [@CITE], an R package that provides an index of openly published radiocarbon databases and a common interface for retrieving them and performing basic data cleaning.
+n_c14_datasets <- nrow(c14_datasets)
+n_c14_datasets_sans_publication_year <- n_filter(c14_datasets, is.na(publication_year))
+```
+
+Despite these inefficiencies, there have been a profusion of published radiocarbon compilations since the decline of the date list.
+Our review of the literature identified `r n_c14_datasets` published since 1994.
+This is almost certainly an undercount, because our firsthand knowledge of regional literature was limited to Europe and West Asia and many resources only ever existed in 'grey' formats (e.g.
+websites that were not indexed and no longer exist).
+We also restricted ourselves to structured datasets disseminated primarily in a digital format; 
+'date lists' in printed periodicals and gazetteers were excluded.
+A full list of the datasets we identified is presented in appendix XXXX<!-- TODO -->.
+
+```{r fig-c14-datasets-time}
+#| fig-cap: Cumulative number of radiocarbon compilations published since 1995
+c14_datasets |>
+  drop_na(publication_year) |>
+  count(publication_year, sort = TRUE) |>
+  arrange(publication_year) |>
+  mutate(cumn = cumsum(n)) |>
+  ggplot(aes(publication_year, cumn)) +
+  geom_line() +
+  scale_x_continuous(breaks = scales::breaks_width(5)) +
+  labs(x = NULL, y = "Datasets*",
+       caption = glue("* Excludes {n_c14_datasets_sans_publication_year} datasets with an unknown year of publication")) +
+  theme_half_open()
+```
+
+The number of available compilations has steadily increased since around 1995 [@fig-c14-datasets-time].
+The first generation came around the turn of the century and consisted mostly of online databases with a web frontend.
+These included some databases operated by radiocarbon labs, for example the Oxford Radiocarbon Lab (ORAU) and the Belgian Royal Institute for Cultural Heritage (KIK-IRPA), and essentially represented a continuation of their date lists in a digital format.
+The majority, however, were compiled from the literature and individual researchers interested in a particular region.
+Notable early examples include ANDES 14C in 1994 [Central Andes, @MichczynskiEtAl1995], CARD [Canada, @GajewskiEtAl2011] and RADON [Europe, @Raetzel-Fabian1999] in 1999, and CANEW in 2001 [Near East, @ReingruberThissen2005].
+From 2010, coinciding with broader shifts in scientific publishing [@TenopirEtAl2011], it became more common to publish standalone 'open data' products in the form of journal supplements, archives in repositories and/or data papers;
+the *[Journal of Open Archaeology Data](<https://openarchaeologydata.metajnl.com>)*, launched in 2012, has been a prominent venue for this latter category.
+Most recently there has been a trend towards providing version-controlled plain text data via platforms such as [GitHub](https://github.com), reflecting the broader adoption of these tools amongst computational archaeologists over the last decade [@BatistRoe2024].
+The shift from online databases towards more static but more preservable open data products is welcome, given how many databases from the first generation have subsequently ceased to be accessible.
+Version-controlled repositories are particular well-suited to data compilation projects because they allow for continued updates whilst still providing snapshot 'releases' that are citeable and can be archived in long-term repositories.
+
+```{r data-basemap}
+coast <- gisco_get_coastallines()
+m49_regions <- gisco_get_countries() |>
+  mutate(
+    m49_macroregion = countrycode(ISO3_CODE, "iso3c", "un.region.name"),
+    m49_subregion = countrycode(ISO3_CODE, "iso3c", "un.regionsub.name"),
+    m49_intregion = countrycode(ISO3_CODE, "iso3c", "un.regionintermediate.name"),
+    m49_region = coalesce(m49_intregion, m49_subregion)
+  ) |>
+  summarise(geometry = st_union(geometry), .by = m49_region)
+
+```
+
+```{r fig-c14-datasets-map}
+#| fig-cap: Geographic coverage of published regional radiocarbon compilations
+c14_datasets |>
+  separate_longer_delim(m49_region, "; ") |>
+  drop_na(m49_region) |>
+  summarise(n_dates = sum(n_dates, na.rm = TRUE), .by = m49_region) |>
+  right_join(m49_regions) |>
+  mutate(n_dates = replace_na(n_dates, 0)) |>
+  st_as_sf() |>
+  ggplot() +
+  geom_sf(aes(fill = n_dates), colour = NA) +
+  geom_sf(data = coast, fill = NA) +
+  scale_fill_turku(reverse = TRUE, labels = scales::comma) +
+  labs(fill = "Dates") +
+  coord_sf(crs = "+proj=moll") +
+  theme_map()
+```
+
+Although this body of work has greatly improved the accessibility of radiocarbon dates and supported significant methodological advances [@CremaBevan2021], some limitations are apparent.
+
+The geographic coverage of regional radiocarbon compilations is markedly uneven (@fig-c14-datasets-map).
+Europe and, especially, North America are very over-represented [as has already been observed by @ChaputGajewski2016 and others].
+South America, West Asia, and East Asia are reasonably well-covered, but there practically no systematically compiled dates from East or West Africa, Central or South Asia, or Mainland Southeast Asia.
+This is probably explained in part by a lower volume of archaeological research and access to radiocarbon dating in these regions, but a lack of attention in compilation work must also be a factor.
+For example, radiocarbon dating has been an established part of Indian archaeology since at least 1961 [@KusumgarEtAl1963], but we have not able to locate a single systematic compilation of dates from the Indian subcontinent.
+
+```{r data-c14-datasets-maintainance}
+n_c14_datasets_maintained <- n_filter(c14_datasets, maintained)
+n_c14_datasets_recently_updated <- n_filter(c14_datasets, maintained, last_updated > 2022)
+mean_c14_datasets_lifespan <- mean(c14_datasets$last_updated - c14_datasets$publication_year, na.rm = TRUE)
+```
+
+Compilations also become out of date almost immediately upon publication, due the the constant stream of new dates.
+Unfortunately this applies to many databases that are in theory continuously updated, as it is common to see them become unmaintained and or unexpectedly become unavailable.
+Of the `r n_c14_datasets` published datasets we identified, `r n_c14_datasets_maintained` were intended to be continuously updated, but only `r n_c14_datasets_recently_updated` had received updates in the last two years.
+The average 'lifespan' of a dataset from its publication to its last update is `r mean_c14_datasets_lifespan` years.
+Most radiocarbon datasets we reviewed were compiled with a specific goal in mind (e.g.
+a particular analysis) and, even where there is the intention to keep them updated afterwards, the exigencies of scientific production combined with the labour-intensive nature of the process make that difficult to achieve in practice.
+
+The fragmentation of the radiocarbon record into regional datasets also hinders analysis at larger scales.
+Although the core elements of a radiocarbon date—laboratory identifier, radiocarbon age, measurement error—are more or less standardised, there is no such consistency in contextual information on the sample or site.
+Such contextual information is important not just for the interpretation of dates, but for 'chronometric hygiene' [filtering out unreliable dates based on sample information, see e.g.
+@PettittEtAl2003] and for correcting for known systematic errors such as the marine reservoir effect [@AlvesEtAl2018].
+Most published datasets incorporate all or part of earlier compilations, duplicate records are also very common, but deduplicating them is not a trivial problem due to format variations (see @sec-implementation-data).
+These issues are by no means impossible to overcome, but adds a significant amount of data-cleaning effort to a process that is otherwise very amenable to standardisation.
+
+## Open data infrastructures
+
+The profusion of radiocarbon compilations over the last decade has naturally prompted many to think globally:
+
+- The first available synthetic radiocarbon database was **c14bazAAR** [@CITE], an R package that provides an index of openly published radiocarbon databases and a common interface for retrieving them and performing basic data cleaning.
 Because c14bazAAR downloads data from its original source repositories, rather than mirroring it, it only includes resources that have been published in a fully open and machine-accessible format.
@@
 
+<!-- TODO: brief evaluation of the coverage of each -->
+
+<!-- TODO: probably should be in design goals or evaluation -->
 XRONOS blends aspects of all three approaches to achieve the same aim of providing access to the global radiocarbon data through a common interface.
@@
 However in general we see the approaches as complementary.
-A c14bazAAR parser for XRONOS is in development (https://github.com/ropensci/c14bazAAR/pull/150), and we also aim to provide an IntChron interface to XRONOS' data in the near future.
+A c14bazAAR parser for XRONOS is in development (<https://github.com/ropensci/c14bazAAR/pull/150>), and we also aim to provide an IntChron interface to XRONOS' data in the near future.
 New data and corrections from p3k14c are incorporated into XRONOS as they are released.
 
-XRONOS is our answer to Kintigh's call [@Kintigh2006] for digital infrastructures that don't just provide access to chronological data but enables researchers to "archive, access, integrate, and mine disparate data sets".
-It parallels and draws inspiration from several similar initiatives in archaeology, such as <!-- TODO examples -->
-And outside of it: <!-- TODO examples GBIF -->
-
-### Beyond radiocarbon
+## Beyond radiocarbon
 
@@
 
-## Concept
+# Concept
 
@@
 
-### Design goals
+## Design goals
+
+XRONOS is our answer to Kintigh's call [@Kintigh2006] for digital infrastructures that don't just provide access to chronological data but enables researchers to "archive, access, integrate, and mine disparate data sets".
+It parallels and draws inspiration from several similar initiatives in archaeology, such as <!-- TODO examples -->
+And outside of it: <!-- TODO examples GBIF -->
 
@@
 
-### Data model
+* Link to limitations identified in the state of the art section
+
+## Data model
 
@@
 
-### Linked data
+## Linked data
 
@@
 
-## Implementation
+# Implementation
 
@@
 
-### Software architecture
+## Software architecture
 <!-- TODO: shorten -->
@@
 
-### Data ingestion and curation
+## Data ingestion and curation {#sec-implementation-data}
 
@@
 
-### User interfaces
+## User interfaces
 
@@
 
-## Evaluation
+# Evaluation
 

### 2024-10-13 paper_jcaa.qmd
Expand draft

@@
 library("glue")
+library("gt")
 library("khroma")
@@
 library("tidyr")
-library("tinytable")
 
@@
   * [ ] State of the art [2000]
-    * [ ] Compilations of radiocarbon dates [1000]
+    * [x] Compilations of radiocarbon dates [1000]
     * [x] Open data infrastructures [500]
@@
   * [ ] Evaluation
+  * [ ] Appendix table
 * [ ] Figures and tables
@@
   coord_sf(crs = "+proj=moll") +
-  theme_map()
+  theme_minimal_grid()
 ```
 
-Although this body of work has greatly improved the accessibility of radiocarbon dates and supported significant methodological advances [@CremaBevan2021], some limitations are apparent.
+Although this body of work has greatly improved the accessibility of radiocarbon dates and supported significant methodological advances [@Crema2022; @CremaEtAl2024], some limitations are apparent.
 
@@
 
-Compilations also become out of date almost immediately upon publication, due the the constant stream of new dates.
+Datasets based on literature review also become out of date almost immediately upon publication, due the the constant stream of new dates.
 Unfortunately this applies to many databases that are in theory continuously updated, as it is common to see them become unmaintained and or unexpectedly become unavailable.
@@
 a particular analysis) and, even where there is the intention to keep them updated afterwards, the exigencies of scientific production combined with the labour-intensive nature of the process make that difficult to achieve in practice.
+
+<!-- TODO: lab databases don't go out of date, but have arbitary coverage? -->
+
+* The coverage of these databases is uneven, systematically biased, and duplicative of each other [@CITE Africa paper]?
 
@@
 
+<!-- TODO: -->
+* Not all are open
+* Not all are machine readable
+
 ## Open data infrastructures
@@
 
-- The first available synthetic radiocarbon database was **c14bazAAR** [@CITE], an R package that provides an index of openly published radiocarbon databases and a common interface for retrieving them and performing basic data cleaning.
+- The first available synthetic radiocarbon database was **c14bazAAR** [@SchmidEtAl2019], an R package that provides an index of openly published radiocarbon databases and a common interface for retrieving them and performing basic data cleaning.
 Because c14bazAAR downloads data from its original source repositories, rather than mirroring it, it only includes resources that have been published in a fully open and machine-accessible format.
-Despite this limitation, it has global coverage and is still the largest collection of radiocarbon dates available (see @FIG).
+Despite this limitation, it has global coverage and is still the largest collection of radiocarbon dates available (@FIG).
 
-- Another indexical approach is the **IntChron** project [@CITE], which exposes data from multiple sources and exposes them with a common JSON-based web interface.
+- Another indexical approach is the **IntChron** project [@BronkRamseyEtAl2019], which exposes data from multiple sources and exposes them with a common JSON-based web interface.
 The IntChron specification is open, meaning that radiocarbon labs or compilation projects can implement it independently and thereby allow end users to access their data through a common itnerface (though to our knowledge it has so far only been adopted by databases associated with the Oxford Radiocarbon Lab).
@@
 
-* **p3k14c** [@CITE] instead compiles multiple source databases into a single flat file dataset, with a similar level of coverage to c14bazAAR.
+* **p3k14c** [@BirdEtAl2022] instead compiles multiple source databases into a single flat file dataset, with a similar level of coverage to c14bazAAR.
 The major advantage of this approach is that the data is made internally consistent and has been manually cleaned to an extent, which makes it particularly well-suited to global analyses.
@@
 <!-- TODO: brief evaluation of the coverage of each -->
-
-<!-- TODO: probably should be in design goals or evaluation -->
-XRONOS blends aspects of all three approaches to achieve the same aim of providing access to the global radiocarbon data through a common interface.
-Like c14bazAAR and IntChron, it is a 'metadatabase' that draws from existing data resources and maintains an explicit link to them.
-But like p3k14c, it integrates these into a single database and applies data curation processes to harmonise them and improve the quality of the information.
-It has a wider scope than c14bazAAR or IntChron, as it mirrors rather than directly retrieves the source data (allowing us to use resources that aren't openly published), and does not rely on the authors of these sources to implement a common specification.
-It also goes beyond the functionality of p3k14c by providing systems for the contunious ingestion and curation of new data as it is published.
-However in general we see the approaches as complementary.
-A c14bazAAR parser for XRONOS is in development (<https://github.com/ropensci/c14bazAAR/pull/150>), and we also aim to provide an IntChron interface to XRONOS' data in the near future.
-New data and corrections from p3k14c are incorporated into XRONOS as they are released.
 
@@
 
+* There are lots of other types of chronometric data!
+  * Dendro - in certain places (e.g. Switzerland!), way more important than radiocarbon
+  * Other radiometric/absolute dating methods, OSL, etc.
+  * Typology - still the backbone of many chronologies
+  * Application-specific: sea level dating, rock patina, etc.
+* Compilation of other types of dates has been much more limited.
+  * Exceptions: ...?
+
+<!-- MH to do: -->
+For European dendrochronological data, the situation is even less favourable for archaeological purposes:
+
+* NOAA stores some global datasets, but has its main focus in the United States and, with its only 34 European prehistoric datasets, is not useful for the study of the European past.
+* The ADS database is compiled and maintained by the UK-based Vernacular Architecture Group and is limited to data published annually in the journal Vernacular Architecture, containing only medieval or later dates for the UK.
+* DendroDB has a European focus but only provides data for historical periods.
+* The "Dendrochronological Database" of the Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) has a worldwide focus on collected natural wood samples and has different goals than a database for archaeological wood material (furthermore, although announced a long time ago, this database is still not functional).
+* Although "Digital Collaboratory for Cultural Dendrochronology (DCCD).
+An international digital data library for dendrochronology" is well connected in the context of archaeological data services (e.g.
+ARIADNE), it is focused on the Netherlands (more than 2/3 of the data, while 0.08% of the Quercus data refer to Switzerland) and contains very few prehistoric data (about 2.5%).
+Furthermore, the overall activity in this database has decreased from 3846 new project records in 2010-2014 to only 83 since then (until the end of 2019).
+
+* Typology
+  * perio.do
+  * Often embedded in other types of database (e.g. gazetteers)
+  * But nobody has treated it as a form of chronometric data in its own right?
+
+What is completely missing worldwide so far, is a database that encompasses both, 14C and dendro data.
+However, to efficiently conduct data driven modelling of human-environmental dependencies and dynamics on the past, such a data base seems indispensable for future research.
+
 # Concept
 
-XRONOS inherits its basic structure from RADON (@CITE), with a database-backed web application and a data model that separates radiocarbon dates, contextual information, and sites.
+XRONOS inherits its basic structure from RADON [@Raetzel-Fabian1999; @RADON; @RADONB; @RADONB2024], with a database-backed web application and a data model that separates radiocarbon dates, contextual information, and sites.
 Our overall aims in developing XRONOS is to bring this model, which RADON has operated on for XX years, up to date, to generalise it to other types of chronometric information, and to transform it from an online database to a data infrastructure that supports the continuous ingestion, curation, and open dissemination of archaeological chronologies from diverse sources.
@@
 
-<!-- TODO: Look at Martin's grant application and drafts -->
+The principal design goals of the software are to combine all available sources of radiocarbon and other chronometric data in single database; 
+develop robust tools for the continuous ingestion and refinement of this data; 
+and disseminate this data within an open and FAIR framework, embedding it in the wider world of Linked Open Data in archaeology and beyond.
+
+Our concrete goals in the design and implementation of XRONOS are the following:
+
+1. Provision of access to 14C and dendrochronological data with meta information on archaeological contexts via web-based Open Access frontend
+2. Web frontend with processing functions (filters and queries, export to various data formats, 14C calibration & chronology formation) for in place analytical functions
+3. machine readable interface --- API --- for direct access e.g. by open source statistical environments (eg. R) for analytical purposes or dendroanalytical software
+4. licensing and versioning of individual data using eg. DOI
+5. Storage in a repository with user and role-specific access rights as well as encryption for data privacy
+6. Ensuring sustainable data security and usage through compliance with standards, appropriate backup solutions and high-availability technologies
 
@@
 
+```{r tbl-issues}
+#| tbl-cap: Automatically-recognised data quality issues currently implemented in XRONOS
+tribble(
+  ~entity,                    ~issue_code,            ~n,     ~issue_description,
+  "Sites",                    "MISSING_COORDINATES",  4452,   "Missing geographic coordinates",
+  "Sites",                    "INVALID_COORDINATES",  0,      "Geographic coordinates fall outside the earth's ellipsoid",
+  "Sites",                    "MISSING_COUNTRY_CODE", 1221,   "Missing data on what country the site is in",
+  "Samples",                  "MISSING_MATERIAL",     138177, "Missing data on the sample material",
+  "Samples",                  "MISSING_TAXON",        138182, "Missing data on the sample taxon",
+  "Samples",                  "MISSING_CRS",          0,      "Sample has coordinates but the coordinate reference system is not given",
+  "Taxons",                   "UNKNOWN_TAXON",        9260,   "Sample taxon has not been matched to the GBIF Backbone Taxonomy",
+  "Taxons",                   "LONG_TAXON",           509,    "Description of the sample taxon is implausibly long",
+  "Radiocarbon dates",        "MISSING_C14_AGE",      447,    "Missing radiocarbon age",
+  "Radiocarbon dates",        "VERY_OLD_C14",         76,     "Radiocarbon age older than the effective range of the method (50 ka)",
+  "Radiocarbon dates",        "MISSING_C14_ERROR",    585,    "Missing measurement error",
+  "Radiocarbon dates",        "MISSING_D14C",         238875, "Missing δ13C measurement",
+  "Radiocarbon dates",        "MISSING_D14C_ERROR",   238875, "Missing δ13C measurement error",
+  "Radiocarbon dates",        "MISSING_C14_METHOD",   242233, "Missing data on radiocarbon dating method (conventional, AMS, etc.)",
+  "Radiocarbon dates",        "MISSING_C14_LAB_ID",   1233,   "Missing laboratory identifier",
+  "Radiocarbon dates",        "INVALID_LAB_ID",       16053,  "Laboratory identifier does not match the standard format (e.g. 'Abc-1234')",
+  "Radiocarbon dates",        "MISSING_C14_LAB",      350190, "Missing data on the radiocarbon laboratory",
+  "Bibliographic references", "MIXED_REFERENCE",      21933,  "Bibliographic reference appears to combine multiple publications",
+  "Bibliographic references", "MISSING_BIBTEX",       42109,  "Bibliographic reference without structured data in BibTeX format",
+) |>
+  group_by(entity) |>
+  gt() |>
+  cols_label(issue_code = "Issue", n = "N", issue_description = "Description") |>
+  cols_width(everything() ~ pct(33)) |> # TODO: fix...
+  tab_options(
+    table.font.size = 13
+  )
+```
+
 Once ingested, we apply a number of automated and semi-automated quality control processes to integrate new data into the existing database.
@@
 A wide variety of other potential data quality issues (e.g.
-missing data on what country a site is in) are also flagged for human review by this system, which canoften be semi-automated (e.g.
+missing data on what country a site is in) are also flagged for human review by this system [@tbl-issues], which can often be semi-automated (e.g.
 suggesting close matches in the thesaurus or the country indicated by the record's coordinates).
@@
 
+This paper outlines the conceptual and technical infrastructure developed to realise these goals in XRONOS’ initial phases of development (2019 and 2021–2024), including a generalised data model for site and radiocarbon information, extendable to other chronometric data;
+an R- and Ruby-based pipeline for continuous ingestion of data from a variety of sources; continuous, semi-automated data cleaning protocols; 
+a Ruby-on-Rails application providing a web-based frontend to the data and a REST API for programmatic access; 
+and an R package for interfacing with the API. 
+We believe the XRONOS framework provides more open, more reliable, and more comprehensive access to chronometric data than previously available, as well as a foundation for its continuous expansion and refinement.
+
 * XRONOS is useable, use XRONOS!
@@
 * The sustainability challenge
+
+XRONOS blends aspects of all three approaches to achieve the same aim of providing access to the global radiocarbon data through a common interface.
+Like c14bazAAR and IntChron, it is a 'metadatabase' that draws from existing data resources and maintains an explicit link to them.
+But like p3k14c, it integrates these into a single database and applies data curation processes to harmonise them and improve the quality of the information.
+It has a wider scope than c14bazAAR or IntChron, as it mirrors rather than directly retrieves the source data (allowing us to use resources that aren't openly published), and does not rely on the authors of these sources to implement a common specification.
+It also goes beyond the functionality of p3k14c by providing systems for the contunious ingestion and curation of new data as it is published.
+However in general we see the approaches as complementary.
+A c14bazAAR parser for XRONOS is in development (<https://github.com/ropensci/c14bazAAR/pull/150>), and we also aim to provide an IntChron interface to XRONOS' data in the near future.
+New data and corrections from p3k14c are incorporated into XRONOS as they are released.
+
+# Acknowledgements
+
+The XRONOS project was carried under the direction of Albert Hafner. 
+Individual contributors to the XRONOS database to date include Chiara Huwiler, Rivana Moser, Tomasz Chmielewski, and Stephanie Döppler. 
+A complete and up-to-date list of acknowledgements for the project can be found at <https://xronos.ch/about/acknowledgements>.
+
+# Funding statement
+
+This work was funded by the Swiss National Science Foundation ([SNSF Project #198152](https://data.snf.ch/grants/grant/198153)) and the University of Bern (Initiator Grant, Caroline Heitz, Project 'Time and Temporality').

### 2024-10-13 paper_jcaa.qmd
Add introduction

@@
 It aims provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world.
-Collating a number of recent regional and global compilations of dates, XRONOS offers the most comprehensive radiocarbon database yet published, with over 350,000 date and 75,000 site records.
+By collating a large number of existing regional and global compilations of dates, XRONOS offers the most comprehensive radiocarbon database yet published, with over 350,000 radiocarbon and 75,000 site records.
 It also provides a foundation for expanding the systematic collection of chronometric information beyond radiocarbon, with support for typological and dendrochronological dates and a generalisable data model that can be adapted to other methods of absolute dating.
@@
 * [ ] First draft:
-  * [ ] Introduction [500]
+  * [x] Introduction [530/500]
   * [ ] State of the art [2000]
@@
   * [ ] Evaluation [500]
-* [ ] Add references:
-  * [ ] Introduction
+* [ ] Complete references:
+  * [x] Introduction
   * [ ] State of the art
@@
 Chronology is the backbone of the archaeological record.
-It is needed for X, Y, and Z.
-Open access to chronological data is therefore a critical...
+As a necessary prerequisite to understanding the context of any past event or process [@Lucas2004], it is has unsurprisingly been at the forefront of methodological development in archaeology for as long as the discipline has existed: from putting finds and events in sequence [@Rowley-Conwy2007] to using scientific methods to place them on a real timescale [@Libby1955] and increasingly sophisticated statistical methods to refine them [@BronkRamsey2009].
+If archaeology is to be an open science [@Lake2012], it is therefore critical that effective open access to chronological information be placed front and centre.
 
-* Profusion of open C14 data in the last 20–30 years calls for global solutions
-* Range of other chronological information used by archaeologists is largely untouched
-* From uploading CSVs to (social) infrastructure
+Over the last two decades, archaeologists have answered this call by publishing an increasing number of compilations of dates from archaeological contexts as open data.
+These efforts have facilitated major reevaluations of previously-established chronologies [e.g.
+@HighamEtAl2014; @LoftusEtAl2019; @PratesEtAl2020; @KatsianisEtAl2020], important new insights into past processes [@RirisEtAl2024], and the development of novel ways of using chronological data [@Crema2022].
+<!-- TODO: Would welcome more citations here -->
+The focus has been overwhelmingly on radiocarbon dating and most compilations focus on a single region and/or period.
+The profusion of open radiocarbon data in particular has prompted several initiatives towards a global synthesis [e.g. @SchmidEtAl2019; @BronkRamseyEtAl2019; @BirdEtAl2022].
+
+At the same time, the broad range of other types of chronological information used in archaeology—from other radiometric methods to dendrochronology to typological seriation and numismatics—remains relatively difficult to access as open data.
+Even when it comes to radiocarbon data, the coverage of available compilations is patchy both geographically and in time and of variable quality (see @sec-c14-compilation).
+The publication of many overlapping, non-standardised and mostly static open data resources means that it is still difficult to obtain reliable and up-to-date chronological datasets, especially for applications that crosscut convential geographic and temporal domains of research.
+Initiatives towards synthesis this have improved this situation, but the goal of a global dataset that is both comprehensive and up-to-date remains elusive.
+
+XRONOS <https://xronos.ch> is a new open data infrastructure that aims to provide access to published radiocarbon dates and other chronometric data from any period, anywhere in the world.
+It is our attempt to move the state of the art in open archaeological chronology beyond the publication of static, one-off resources ['uploading CSVs', @Batist2023, pp.
+188-189], and towards a living digital infrastructure [@Kintigh2006] embedded in a transparent and sustainable collabrative network.
+The core of XRONOS is a server application that ingests chronological data from diverse sources, facilitates semi-automated and manual curation of this data, and makes it available via both a web-based graphical user interface (GUI) and machine-readable application programming interface (API).
+The web frontend can be accessed via <https://xronos.ch> and all components of the software are developed as free and open source software with source code available at <https://github.com/xronos-ch>.
+
+In the remainder of this paper, we describe the concept and implementation of XRONOS in relation to the state of the art in open chronometric data in archaeology, and evaluate our progress in achieving these goals as of writing.
+Since we envisage both XRONOS as a dataset and XRONOS as software to be continually developing resources, the description here should be read as a 'snapshot' rather than the final state of the project.
 
@@
 
-## Compilations of radiocarbon dates
+## Compilations of radiocarbon dates {#sec-c14-compilation}
 
-Though 'open data' in archaeology is a relatively recent phenomenon [@CITE], the open publication of compiled radiocarbon dates has a substantial prehistory.
+Though an explicit emphasis on 'open data' is a relatively recent phenomenon in archaeology [@Lake2012], the open publication of compiled radiocarbon dates has a substantial prehistory.
 Starting in the 19XXs, radiocarbon laboratories shared and compiled as 'date lists' published on a regular basis in the journal *Radiocarbon* [@CITE].
@@
 XRONOS inherits its basic structure from RADON [@Raetzel-Fabian1999; @RADON; @RADONB; @RADONB2024], with a database-backed web application and a data model that separates radiocarbon dates, contextual information, and sites.
-Our overall aims in developing XRONOS is to bring this model, which RADON has operated on for XX years, up to date, to generalise it to other types of chronometric information, and to transform it from an online database to a data infrastructure that supports the continuous ingestion, curation, and open dissemination of archaeological chronologies from diverse sources.
+Our overall aims in developing XRONOS is to bring this model, which RADON has operated on for more than twenty years, up to date, to generalise it to other types of chronometric information, and to transform it from an online database to a data infrastructure that supports the continuous ingestion, curation, and open dissemination of archaeological chronologies from diverse sources.
 

### 2024-10-13 paper_jcaa.qmd
Add design goals

@@
     * [x] Inherits a common structure from RADON
-    * [ ] Design goals [250]
+    * [x] Design goals [250/250]
     * [x] Data model [1150/500]
@@
 
-## Open data infrastructures
+## Open data infrastructures {#sec-global-compilations}
 
@@
 XRONOS is our answer to Kintigh's call [@Kintigh2006] for digital infrastructures that don't just provide access to chronological data but enables researchers to "archive, access, integrate, and mine disparate data sets".
-It parallels and draws inspiration from several similar initiatives in archaeology, such as <!-- TODO examples -->
-And outside of it: <!-- TODO examples GBIF -->
+It parallels and draws inspiration from several similar initiatives within and outwith archaeology, such as <!-- TODO examples -->.
+To improve upon existing global syntheses of radiocarbon dates (see @sec-global-compilations), we aimed to develop a living infrastructure that both continually collected data from diverse sources and presented a seamless single database to the user.
+The Global Biodiversity Information Facility [GBIF, <https://gbif.org>, @CanhosEtAl2004]—which provides a single, consistent interface to many sources of global biodiversity data—has served as an exemplar for us in this regard.
 
-The principal design goals of the software are to combine all available sources of radiocarbon and other chronometric data in single database; 
-develop robust tools for the continuous ingestion and refinement of this data; 
-and disseminate this data within an open and FAIR framework, embedding it in the wider world of Linked Open Data in archaeology and beyond.
+Our principal goals for the software were therefore to:
 
-Our concrete goals in the design and implementation of XRONOS are the following:
+1. Combine all available sources of radiocarbon and other chronometric data in single database; 
+2. Develop robust tools for the continuous ingestion, collation and curation of this data;
+3. Disseminate the collated and curated data as linked open data within a FAIR framework.
 
-1. Provision of access to 14C and dendrochronological data with meta information on archaeological contexts via web-based Open Access frontend
-2. Web frontend with processing functions (filters and queries, export to various data formats, 14C calibration & chronology formation) for in place analytical functions
-3. machine readable interface --- API --- for direct access e.g. by open source statistical environments (eg. R) for analytical purposes or dendroanalytical software
-4. licensing and versioning of individual data using eg. DOI
-5. Storage in a repository with user and role-specific access rights as well as encryption for data privacy
-6. Ensuring sustainable data security and usage through compliance with standards, appropriate backup solutions and high-availability technologies
-
-* Link to limitations identified in the state of the art section
+Meeting these goals required principally the development of a) a conceptual data model, including links to other open data resources, that is flexible enough for all forms of chronometric data; and b) an software implementation that supports the main functions of ingesting, curating, and disseminating this data.
+The individual components of this work are described in more depth below but, briefly, consist of a relational data model implemented in a PostgreSQL database; a Ruby application providing server-based tools for ingestion, curation and dissemination of data; and multiple graphical and programmatic interfaces to the resulting dataset.
 

### 2024-10-13 paper_jcaa.qmd
Add summary of other dating methods

@@
 
-At the same time, the broad range of other types of chronological information used in archaeology—from other radiometric methods to dendrochronology to typological seriation and numismatics—remains relatively difficult to access as open data.
+At the same time, the broad range of other types of chronological information used in archaeology—from other radiometric methods to dendrochronology to typological dating and epigraphy—remains relatively difficult to access as open data.
 Even when it comes to radiocarbon data, the coverage of available compilations is patchy both geographically and in time and of variable quality (see @sec-c14-compilation).
@@
 
-* Dendro
-* Typology
-* Other radiometric methods
-* Application-specific 
-* XRONOS aims to bring all these together, starting with radiocarbon, dendro-, and typology
+Radiocarbon has been by far the most active area of open data compilation, but archaeological chronology incorporates a much more diverse range of sources of information [@Harding1999].
+In periods beyond the practical limit of radiocarbon dating (c. 50,000 BP), other types of radiometric (K–Ar, U–Pb, etc.), chemical or luminescence dating offer an alternative [@Aitken1999].
+Conversely, in historic periods, radiocarbon is often relatively underused compared to conventional typological dating (based on artefact characteristics), which in these periods can offer comparable or better temporal resolution, or direct dating based on epigraphy [@HermankovaEtAl2021], numismatics [@KemmersMyrberg2011] or historical sources.
+In places where it is widely available, dendrochronology [@Baillie2014] also produces significantly better resolved chronologies and therefore tends to be the main source of chronometric data.
+Other more application-specific chronological methods include shoreline dating [@Brogger1905; @Roalkvam2023], lichenometry [@Benedict2009] and rock weathering dating [@Whitley2012; @Bednarik2020].
+There have been comparatively few examples of systematic, open compilations of any of these types of data.
 
-* There are lots of other types of chronometric data!
-  * Dendro - in certain places (e.g. Switzerland!), way more important than radiocarbon
-  * Other radiometric/absolute dating methods, OSL, etc.
-  * Typology - still the backbone of many chronologies
-  * Application-specific: sea level dating, rock patina, etc.
-* Compilation of other types of dates has been much more limited.
-  * Exceptions: ...?
+<!-- TODO: brief para on other radiometric/OSL -->
 
-<!-- MH to do: -->
+<!-- TODO: MH -->
 For European dendrochronological data, the situation is even less favourable for archaeological purposes:
@@
   * But nobody has treated it as a form of chronometric data in its own right?
+    * Malta database?
+
+* XRONOS aims to bring all these together, starting with radiocarbon, dendro-, and typology
 
@@
 We have however tried to mitigate this by providing clearly-documented source code and regular data dumps so that, if our instance of XRONOS disappears, or if one simply does not want to use it, it is possible for others to host a XRONOS server of their own.
-As the sustainability of scientific software and data infrastructures is a pressing problem (a point we will return to in the conclusion), in the future it may be desirable to support further decentralisation through, for example, a federated server-server model [@CITE].
+As the sustainability of scientific software and data infrastructures is a pressing problem, in the future it may be desirable to support further decentralisation through, for example, a federated server-server model [@CITE].
 

### 2024-10-13 paper_jcaa.qmd
Add paragraph on other scientific dates

@@
 Other more application-specific chronological methods include shoreline dating [@Brogger1905; @Roalkvam2023], lichenometry [@Benedict2009] and rock weathering dating [@Whitley2012; @Bednarik2020].
-There have been comparatively few examples of systematic, open compilations of any of these types of data.
 
-<!-- TODO: brief para on other radiometric/OSL -->
+Compared to radiocarbon, there are few examples of systematic, open compilations of any of these other types of data.
+
+This is most striking when it comes to other radiometric/scientific dating methods, as the data structures and publication modes are very similar to radiocarbon.
+The 'Radiocarbon Palaeolithic Europe Database' [@Vermeersch2020], despite the name, includes a significant number of TL (1177 items), OSL (1086 items), ESR, Th/U and AAR.
+Similarly, the AustArch database [@WilliamsEtAl2014] includes luminescence dates alongside radiocarbon, but is limited to Australia and was last updated in 2013.
+Other than this, we are not aware of any other significant published open datasets.
 

### 2024-10-14 paper_jcaa.qmd
Add typology open data

@@
     * [x] Open data infrastructures [500]
-    * [ ] Beyond radiocarbon [500]
+    * [x] Beyond radiocarbon [850/500]
   * [ ] Concept [1000]
@@
 
-XRONOS <https://xronos.ch> is a new open data infrastructure that aims to provide access to published radiocarbon dates and other chronometric data from any period, anywhere in the world.
+XRONOS is a new open data infrastructure that aims to provide access to published radiocarbon dates and other chronometric data from any period, anywhere in the world.
 It is our attempt to move the state of the art in open archaeological chronology beyond the publication of static, one-off resources ['uploading CSVs', @Batist2023, pp.
@@
 This is most striking when it comes to other radiometric/scientific dating methods, as the data structures and publication modes are very similar to radiocarbon.
-The 'Radiocarbon Palaeolithic Europe Database' [@Vermeersch2020], despite the name, includes a significant number of TL (1177 items), OSL (1086 items), ESR, Th/U and AAR.
+The 'Radiocarbon Palaeolithic Europe Database' [@Vermeersch2020], despite the name, includes a significant number of thermo- and optically stimulated luminescence, electon spin resonance, uranium–thorium and amino acid dates.
 Similarly, the AustArch database [@WilliamsEtAl2014] includes luminescence dates alongside radiocarbon, but is limited to Australia and was last updated in 2013.
-Other than this, we are not aware of any other significant published open datasets.
+Apart from these and a few other exceptions where other scientific dates are collected alongside radiocarbon, we are not aware of any open compilations of them.
 
-<!-- TODO: MH -->
+<!-- TODO: MH, update and turn into prose -->
 For European dendrochronological data, the situation is even less favourable for archaeological purposes:
@@
 
-* Typology
-  * perio.do
-  * Often embedded in other types of database (e.g. gazetteers)
-  * But nobody has treated it as a form of chronometric data in its own right?
-    * Malta database?
+Typological dates—i.e.
+relative, expertise- or seriation-based dating based on artefact characteristics—is an interesting case because they are ubiquitous but rarely treated as a form of chronometric data in their own right.
+For example, the majority of the radiocarbon datasets we reviewed (@sec-c14-compilation) included some form of typological chronological information in the form of a 'period' or 'culture' column.
+This is also typically present in many other forms of systematic compilation work in archaeology, for example site gazetteers [@CITE].
+Aggregated typological information from such sources are often used in aoristic analysis and related methods [@Mischka2004; @Crema2024].
+What is lacking in this presentation of typological dating is metadata on how the determination was made and how exactly it is to be understood.
+Like any archaeological date, a typological date is derived from a physical sample – the object or set of object from which a chronological estimate was derived.
+Typological dates on one class of object may well clash with other classes of object, or for that matter with scientific dates, but without this kind of metadata such inconsistencies are difficult to resolve.
+Similarly the real calendar range corresponding to a typological date (e.g.
+"Late Neolithic") can be interpreted in multiple ways depending on the region and intentions of the expert making the determination.
+PeriodO [@RabinowitzEtAl2016] is a linked open data infrastructure that includes a shared vocabulary of typological periods and corresponding calendar age estimates, and an important step towards addressing the latter problem.
+However, it remains to be systematically linked to actual compilations of typological dates.
 
-* XRONOS aims to bring all these together, starting with radiocarbon, dendro-, and typology
-
-What is completely missing worldwide so far, is a database that encompasses both, 14C and dendro data.
-However, to efficiently conduct data driven modelling of human-environmental dependencies and dynamics on the past, such a data base seems indispensable for future research.
+What is completely missing to date is a general-purpose infrastructure for combining all of these types of chronometric information on a global scale.
+This is the gap that XRONOS aims to fill, starting with three methods: radiocarbon, dendrochronology, and typological dating.
+These were chosen because they are widely used and relatively advanced in terms of open data, but an important aim of the project is to develop a generalisable data model that can easily scale to any and all types of archaeological chronology (see @sec-data-model).
 
@@
 
-## Data model
+## Data model {#sec-data-model}
 

### 2024-10-23 paper_jcaa.qmd
Add reference to @ReiterEtAl2024

@@
 @PettittEtAl2003] and for correcting for known systematic errors such as the marine reservoir effect [@AlvesEtAl2018].
-Most published datasets incorporate all or part of earlier compilations, duplicate records are also very common, but deduplicating them is not a trivial problem due to format variations (see @sec-implementation-data).
+Most published datasets incorporate all or part of earlier compilations, meaning duplicate records are also very common, but deduplicating them is not a trivial problem due to format variations (see @sec-implementation-data).
 These issues are by no means impossible to overcome, but adds a significant amount of data-cleaning effort to a process that is otherwise very amenable to standardisation.
@@
 A final critical component of XRONOS' data curation system is duplicate handling.
-We import data from many overlapping resources (many of which incorporate each other either in whole or in part), so duplicate records are common.
+We import data from many overlapping resources (many of which incorporate each other either in whole or in part), so duplicate records are common [as recently discussed by @ReiterEtAl2024].
 The end result of standardising and correcting a record is also often to create a duplicate: e.g.

### 2024-11-19 paper_jcaa.qmd
Incorporate suggestions from @martinHinz

@@
 Chronology is the backbone of the archaeological record.
-As a necessary prerequisite to understanding the context of any past event or process [@Lucas2004], it is has unsurprisingly been at the forefront of methodological development in archaeology for as long as the discipline has existed: from putting finds and events in sequence [@Rowley-Conwy2007] to using scientific methods to place them on a real timescale [@Libby1955] and increasingly sophisticated statistical methods to refine them [@BronkRamsey2009].
+As a necessary prerequisite to understanding the context of any past event or process [@Lucas2004], it is has unsurprisingly been at the forefront of methodological development in archaeology for as long as the discipline has existed: from putting finds and events in sequence [@Rowley-Conwy2007] to using scientific methods to place them on an absolute timescale [@Libby1955] and increasingly sophisticated statistical methods to refine them [@BronkRamsey2009].
 If archaeology is to be an open science [@Lake2012], it is therefore critical that effective open access to chronological information be placed front and centre.
@@
 
-The number of available compilations has steadily increased since around 1995 [@fig-c14-datasets-time].
+The number of available compilations has steadily increased since around 1995 (@fig-c14-datasets-time).
 The first generation came around the turn of the century and consisted mostly of online databases with a web frontend.
@@
 The shift from online databases towards more static but more preservable open data products is welcome, given how many databases from the first generation have subsequently ceased to be accessible.
+<!-- TODO: can this be quantified or visualised?
+-MH -->
 Version-controlled repositories are particular well-suited to data compilation projects because they allow for continued updates whilst still providing snapshot 'releases' that are citeable and can be archived in long-term repositories.
@@
 - Another indexical approach is the **IntChron** project [@BronkRamseyEtAl2019], which exposes data from multiple sources and exposes them with a common JSON-based web interface.
-The IntChron specification is open, meaning that radiocarbon labs or compilation projects can implement it independently and thereby allow end users to access their data through a common itnerface (though to our knowledge it has so far only been adopted by databases associated with the Oxford Radiocarbon Lab).
+The IntChron specification is open, meaning that radiocarbon labs or compilation projects can implement it independently and thereby allow end users to access their data through a common interface (though to our knowledge it has so far only been adopted by databases associated with the Oxford Radiocarbon Lab).
 The JSON format also lends itself to the implementation of wrapper libraries, for example the 'rIntChron' package gives direct access to IntChron-indexed databases in R [@CITE].
@@
 Typological dates on one class of object may well clash with other classes of object, or for that matter with scientific dates, but without this kind of metadata such inconsistencies are difficult to resolve.
-Similarly the real calendar range corresponding to a typological date (e.g.
+<!-- TODO: clarify - MH -->
+Similarly the absolute date range corresponding to a typological determination (e.g.
 "Late Neolithic") can be interpreted in multiple ways depending on the region and intentions of the expert making the determination.
@@
 The series of relations `[chron] > sample > context > site` links the chronological and contextual sides of the XRONOS data model.
-Each step is a many-to-many association, meaning for example that it is possible to attach multiple chrons to the same sample (e.g.
+Each step is a many-to-onone association, meaning for example that it is possible to attach multiple chrons to the same sample (e.g.
 replicated radiocarbon dates on the same material), multiple *types* of chrons to the same sample (e.g.
@@
 It has a wider scope than c14bazAAR or IntChron, as it mirrors rather than directly retrieves the source data (allowing us to use resources that aren't openly published), and does not rely on the authors of these sources to implement a common specification.
-It also goes beyond the functionality of p3k14c by providing systems for the contunious ingestion and curation of new data as it is published.
+It also goes beyond the functionality of p3k14c by providing systems for the continuous ingestion and curation of new data as it is published.
 However in general we see the approaches as complementary.

### 2024-11-24 paper_jcaa.qmd
updated Dendro Part and turned it into prose

@@
 
-<!-- TODO: MH, update and turn into prose -->
+<!-- TODO: MH, update and turn into prose. From MH: Should be done now, OK like this? -->
 For European dendrochronological data, the situation is even less favourable for archaeological purposes:
 
-* NOAA stores some global datasets, but has its main focus in the United States and, with its only 34 European prehistoric datasets, is not useful for the study of the European past.
-* The ADS database is compiled and maintained by the UK-based Vernacular Architecture Group and is limited to data published annually in the journal Vernacular Architecture, containing only medieval or later dates for the UK.
-* DendroDB has a European focus but only provides data for historical periods.
-* The "Dendrochronological Database" of the Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) has a worldwide focus on collected natural wood samples and has different goals than a database for archaeological wood material (furthermore, although announced a long time ago, this database is still not functional).
-* Although "Digital Collaboratory for Cultural Dendrochronology (DCCD).
-An international digital data library for dendrochronology" is well connected in the context of archaeological data services (e.g.
-ARIADNE), it is focused on the Netherlands (more than 2/3 of the data, while 0.08% of the Quercus data refer to Switzerland) and contains very few prehistoric data (about 2.5%).
-Furthermore, the overall activity in this database has decreased from 3846 new project records in 2010-2014 to only 83 since then (until the end of 2019).
+Some databases provide valuable resources for dendrochronological studies in general but are not primarily designed to collect tree-ring data from archaeological contexts.
+For instance, Dendro4Art specializes in dendrochronological data related to wooden art objects, such as sculptures and panel paintings.
+While this focus serves art historians and conservationists well, its utility for studying prehistoric datasets is minimal.
+Similarly, the Dendrochronological Picture Database, maintained by the Swiss Federal Institute for Snow and Avalanche Research (SLF), offers a visual archive of approximately 1,400 images documenting dendrochronological phenomena.
+Although valuable as an educational resource, it does not provide raw data necessary for chronological or archaeological analyses.
+Additionally, the OLDLIST and Eastern OLDLIST databases focus on documenting the maximum ages of trees worldwide.
+Their emphasis on biological longevity, while significant for ecological research, limits their applicability to archaeological or prehistoric investigations.
+
+Among databases that do provide dendrochronological data, the degree to which they support prehistoric research varies substantially.
+The NOAA International Tree-Ring Data Bank (ITRDB) serves as a global repository of tree-ring measurements.
+However, its focus remains predominantly on North America, with only 34 datasets representing European prehistoric contexts.
+This restricts its relevance for studying the European past.
+Similarly, the ADS database, maintained by the UK-based Vernacular Architecture Group, compiles dendrochronological data from the UK but is limited to medieval and later periods, making it unsuitable for prehistoric studies.
+
+DendroDB, hosted by the Swiss Federal Institute for Forest, Snow and Landscape Research (WSL), emphasizes ecological and climate studies over archaeological wood material.
+While it claims a broad scope, the database remains non-functional, rendering it ineffective for research needs.
+The CFS-TRenD database, managed by the Canadian Forest Service, compiles over 4,600 datasets from Canadian forests, primarily focusing on boreal ecosystems.
+Despite its extensive coverage for North America, its geographical specificity and lack of open access restrict its utility for European prehistoric contexts.
+Similarly, the QUB Dendrochronology Database, managed by Queen’s University Belfast, offers valuable datasets for Ireland and the UK but lacks significant representation of prehistoric material, limiting its application in broader archaeological investigations.
+The Building Archaeology Research Database (BARD) contains over 24,000 records, including dendrochronological data from more than 2,700 buildings.
+However, its focus on medieval and post-medieval timber-frame construction further narrows its utility for studies involving prehistoric wood samples.
+
+The Digital Collaboratory for Cultural Dendrochronology (DCCD) presents itself as a potentially valuable international platform for dendrochronology, particularly through its integration with archaeological data services such as ARIADNE.
+However, it remains heavily biased toward datasets from the Netherlands, which account for more than two-thirds of its entries, while only 0.08% of its Quercus data pertain to Switzerland and just 2.5% represent prehistoric datasets.
+Notably, these estimates date back to 2021, and an updated assessment is currently unattainable following the platform’s migration to DataverseNL, which now charges an annual fee of nearly €6,000 for access.
+Furthermore, database activity has declined significantly, from 3,846 new project records between 2010 and 2014 to only 83 by the end of 2019.
+Although 519 additional records have been reported since 2021, it remains unclear whether this figure includes revisions to pre-existing entries, potentially inflating the count.
+The database’s narrow focus and restrictive access model significantly limit its broader utility for prehistoric research.
+
+Finally, the Strategic Environmental Archaeology Database (SEAD), hosted by Umeå University, integrates multiple environmental proxies, including dendrochronological data.
+However, the dendrochronological component is largely confined to Swedish data, with limited relevance to prehistoric contexts.
+While SEAD aims for broader applications, its dendro component has limited utility for studies outside Sweden.
+
+In conclusion, the utility of dendrochronological databases for prehistoric research varies widely.
+Global resources such as NOAA’s ITRDB and DCCD offer substantial datasets but face significant limitations in practical geographical and temporal scope.
+Similarly, platforms like DendroDB and BARD primarily cater to historical studies, leaving critical gaps in prehistoric coverage.
+Specialized resources like OLDLIST, Dendro4Art, and the Dendrochronological Picture Database provide valuable contributions but lack the direct relevance necessary for archaeological tree-ring analysis.
+Consequently, researchers focused on prehistoric dendrochronology must navigate a fragmented landscape of databases, each offering distinct strengths and limitations.
+Addressing these gaps remains crucial for advancing the field.
 

### 2024-11-25 paper_jcaa.qmd
Add remaining missing prose

@@
 ```{r setup}
+#| include: false
 library("countrycode")
@@
 library("glue")
-library("gt")
+library("gt", warn.conflicts = FALSE)
 library("khroma")
@@
 <!--
-* [ ] First draft:
-  * [x] Introduction [530/500]
-  * [ ] State of the art [2000]
-    * [x] Compilations of radiocarbon dates [1000]
-    * [x] Open data infrastructures [500]
-    * [x] Beyond radiocarbon [850/500]
-  * [ ] Concept [1000]
-    * [x] Inherits a common structure from RADON
-    * [x] Design goals [250/250]
-    * [x] Data model [1150/500]
-    * [x] LOD [250/250]
-  * [x] Implementation [1000]
-    * [x] Status of development [100/100]
-    * [x] Software architecture [550/300]
-    * [x] Data ingestion and integration [500/300]
-    * [x] User interfaces [650/300]
-  * [ ] Evaluation [500]
+* [x] First draft
 * [ ] Complete references:
@@
   * [ ] APPENDIX: table of datasets
+* [ ] Other TODOs
 * [ ] Copyedit:
@@
 Though an explicit emphasis on 'open data' is a relatively recent phenomenon in archaeology [@Lake2012], the open publication of compiled radiocarbon dates has a substantial prehistory.
+<!-- TODO: check and update this about date lists... -->
 Starting in the 19XXs, radiocarbon laboratories shared and compiled as 'date lists' published on a regular basis in the journal *Radiocarbon* [@CITE].
@@
 ```{r data-basemap}
+#| warning: false
 coast <- gisco_get_coastallines()
@@
   summarise(geometry = st_union(geometry), .by = m49_region)
-
 ```
@@
 
-<!-- TODO: lab databases don't go out of date, but have arbitary coverage? -->
+Laboratory databases solve the problem of currency, but tend to have more arbitary coverage, since the inclusion of data is determined by who submits dates to that lab, not any form of principled curation.
+There are also comparatively few of them – most active labs no longer directly publish dates that they produce (if they ever did).
 
-* The coverage of these databases is uneven, systematically biased, and duplicative of each other [@CITE Africa paper]?
+In addition, the temporal and geographic coverage of these resources is uneven, systematically biased, and duplicative of each other [@CITE Africa paper].
+<!-- TODO: example, X databases for Europe, Y for the rest of the world? -->
+By our count, <!-- TODO: X --> of the databases are not 'open' according to the Open Knowledge Foundation's definition of data openness ["Open data and content can be freely used, modified, and shared by anyone for any purpose", @CITE], which both limits the access to and reuse potential of these datasets.
+<!-- TODO: X --> are not currently available in readily machine-readable formats (e.g. plain text or database files rather than PDFs or hypertext).
 
@@
 
-<!-- TODO: -->
-* Not all are open
-* Not all are machine readable
+## Global radiocarbon compilations {#sec-global-compilations}
 
-## Open data infrastructures {#sec-global-compilations}
+<!-- TODO: Include CalPal?? -->
 
-The profusion of radiocarbon compilations over the last decade has naturally prompted many to think globally:
+The profusion of radiocarbon compilations over the last decade has naturally prompted many to think globally.
+Three existing initiatives are particularly similar to XRONOS in their aims: c14bazAAR, IntChron, and p3k14c.
 
-- The first available synthetic radiocarbon database was **c14bazAAR** [@SchmidEtAl2019], an R package that provides an index of openly published radiocarbon databases and a common interface for retrieving them and performing basic data cleaning.
+The first available synthetic radiocarbon database was c14bazAAR [@SchmidEtAl2019], an R package that provides an index of openly published radiocarbon databases and a common interface for retrieving them and performing basic data cleaning.
 Because c14bazAAR downloads data from its original source repositories, rather than mirroring it, it only includes resources that have been published in a fully open and machine-accessible format.
@@
 
-- Another indexical approach is the **IntChron** project [@BronkRamseyEtAl2019], which exposes data from multiple sources and exposes them with a common JSON-based web interface.
+Another indexical approach is the IntChron project [@BronkRamseyEtAl2019], which exposes data from multiple sources and exposes them with a common JSON-based web interface.
 The IntChron specification is open, meaning that radiocarbon labs or compilation projects can implement it independently and thereby allow end users to access their data through a common interface (though to our knowledge it has so far only been adopted by databases associated with the Oxford Radiocarbon Lab).
@@
 
-* **p3k14c** [@BirdEtAl2022] instead compiles multiple source databases into a single flat file dataset, with a similar level of coverage to c14bazAAR.
+p3k14c [@BirdEtAl2022] instead compiles multiple source databases into a single flat file dataset, with a similar level of coverage to c14bazAAR. 
 The major advantage of this approach is that the data is made internally consistent and has been manually cleaned to an extent, which makes it particularly well-suited to global analyses. 
@@
 
-<!-- TODO: brief evaluation of the coverage of each -->
+As of writing, c14bazAAR includes <!-- TODO: N --> radiocarbon dates with unique laboratory identifiers, IntChron <!-- TODO: N --> (excluding those from non-archaeological contexts), and p3k14c <!-- TODO: N -->.
+The geographic distribution of dates from each is similar (@FIG), reflecting the large degree in overlap between the sources of c14bazAAR and p3k14c in particular.
+IntChron, which in practice is currently only used to publish dates associated with the Oxford Radiocarbon Lab, has dates from more diverse contexts, but is much smaller.
 
@@
 
-<!-- TODO: MH, update and turn into prose. From MH: Should be done now, OK like this? -->
-For European dendrochronological data, the situation is even less favourable for archaeological purposes:
+### Dendrochronology
 
-Some databases provide valuable resources for dendrochronological studies in general but are not primarily designed to collect tree-ring data from archaeological contexts.
+With regard to tree-ring data, some databases provide valuable resources for dendrochronological studies in general but are not primarily designed to collect tree-ring data from archaeological contexts.
 For instance, Dendro4Art specializes in dendrochronological data related to wooden art objects, such as sculptures and panel paintings.
@@
 
-In conclusion, the utility of dendrochronological databases for prehistoric research varies widely.
+The utility of dendrochronological databases for prehistoric research varies widely.
 Global resources such as NOAA’s ITRDB and DCCD offer substantial datasets but face significant limitations in practical geographical and temporal scope.
@@
 Addressing these gaps remains crucial for advancing the field.
+
+### Typological dating
 
@@
 
-* FIGURE: conceptual scheme
+<!-- TODO: data model figure -->
 
@@
 The graphical user interface (GUI) to XRONOS, accessed through a web browser (e.g.
-at https://xronos.ch), uses REST resources and actions as the building blocks for various interfaces through which users can browse, search, retrieve, and analyse chronometric data.
+at <https://xronos.ch>), uses REST resources and actions as the building blocks for various interfaces through which users can browse, search, retrieve, and analyse chronometric data.
 Each action on each resource is represented by a page, though not all of these are publicly accessible.
@@
 
-This paper outlines the conceptual and technical infrastructure developed to realise these goals in XRONOS’ initial phases of development (2019 and 2021–2024), including a generalised data model for site and radiocarbon information, extendable to other chronometric data;
-an R- and Ruby-based pipeline for continuous ingestion of data from a variety of sources; continuous, semi-automated data cleaning protocols; 
+In this paper we have outlined the conceptual and technical infrastructure developed during the initial phases of development on XRONOS in 2019 and 2021–2024.
+These include a generalised data model for radiocarbon and typological dates, extendable to other chronometric information, and associated contextual information;
+an R- and Ruby-based pipeline for continuous ingestion of data from a variety of sources; 
+continuous, semi-automated data cleaning protocols; 
 a Ruby-on-Rails application providing a web-based frontend to the data and a REST API for programmatic access; 
 and an R package for interfacing with the API. 
-We believe the XRONOS framework provides more open, more reliable, and more comprehensive access to chronometric data than previously available, as well as a foundation for its continuous expansion and refinement.
+These systems are in production and publicly accessible at <https://xronos.ch>.
 
-* XRONOS is useable, use XRONOS!
-* Immediate development goals
-* The sustainability challenge
-
-XRONOS blends aspects of all three approaches to achieve the same aim of providing access to the global radiocarbon data through a common interface.
+At the current stage of implementation, we argue XRONOS provides a framework for access to chronometric data that is more open, more reliable, and more comprehensive than the previously available global radiocarbon compilations.
+XRONOS blends aspects of the three existing approaches (c14bazAAR, IntChron, and p3k14c) to achieve the same aim of providing access to the global radiocarbon data through a common interface.
 Like c14bazAAR and IntChron, it is a 'metadatabase' that draws from existing data resources and maintains an explicit link to them.
@@
 It also goes beyond the functionality of p3k14c by providing systems for the continuous ingestion and curation of new data as it is published.
-However in general we see the approaches as complementary.
+With that said, we see the approaches followed by as complementary rather than competing.
 A c14bazAAR parser for XRONOS is in development (<https://github.com/ropensci/c14bazAAR/pull/150>), and we also aim to provide an IntChron interface to XRONOS' data in the near future.
 New data and corrections from p3k14c are incorporated into XRONOS as they are released.
+
+We intend to continue development of XRONOS into the foreseeable future.
+Our immediate development goals include the incorporation of dendrochronology into the database, further refinement of our data curation pipelines, and the public release of the editing interfaces.
+Looking beyond the near term, we have endeavoured to create a sustainable infrastructure that can be maintained by a wider scholarly commons – though we must acknowledge that this is a difficult problem, and one that is more organisational than technical.
+The source code for all the software components of the system are available online (at <https://github.com/xronos-ch>) under open licenses.
+The databases of the instance at <https://xronos.ch> are also archived with Zenodo (<!-- TODO: link -->).
+With these two resources we reduce XRONOS' 'bus factor'; if we are not able to continue operating XRONOS, somebody else can fully recreate it.
+Equally importantly, we enable the 'right to fork' should others wish to take the software and/or database in another direction.
+But aside from these extreme scenarios, the long-term sustainability of XRONOS is contingent on the existence of a community of scholars that use and contribute back to it.
+In this we can only lead the horse to water; we cannot make it drink.<!-- TODO: too glib? -->
 

### 2024-11-25 paper_jcaa.qmd
Suppress sf message

@@
 ```{r data-basemap}
-#| warning: false
+#| include: false
 coast <- gisco_get_coastallines()

### 2024-11-25 paper.qmd
Remove old draft

--- 2024-05-09 paper.qmd
@@
   XRONOS (<https://xronos.ch>) is an open data infrastructure for the backbone of the archaeological record – chronology.
-It provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world.
+It aims provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world.
+By collating a large number of existing regional and global compilations of dates, XRONOS offers the most comprehensive radiocarbon database yet published, with over 350,000 radiocarbon and 75,000 site records.
+It also provides a foundation for expanding the systematic collection of chronometric information beyond radiocarbon, with support for typological and dendrochronological dates and a generalisable data model that can be adapted to other methods of absolute dating.
+Automated and semi-automated quality control processes ensure that data from diverse sources is continuously integrated and standardised, making it significantly easier to filter data of interest and reducing the need for manual data cleaning by end users.
+In this paper we describe the concept and implementation of XRONOS in relation to the state of the art in chronometric data-sharing, and evaluate its potential as a general-purpose open repository and curation platform for archaeological chronology.
 keywords: 
   - open data
+  - chronology
   - chronometry
@@
   - typological dating
-  - chronological modelling
+published: "Manuscript for submission to the *Journal of Computer Applications in Archaeology*"
+execute:
+  echo: false
+number-sections: true
 bibliography: references.bib
@@
 
+```{r setup}
+#| include: false
+library("countrycode")
+library("cowplot")
+library("dplyr", warn.conflicts = FALSE)
+library("giscoR")
+library("ggplot2")
+library("glue")
+library("gt", warn.conflicts = FALSE)
+library("khroma")
+library("patchwork", warn.conflicts = FALSE)
+library("readr")
+library("sf")
+library("tidyr")
+
+#' Count number of rows in a filtered data frame
+n_filter <- function(.data, ...) {
+  nrow(dplyr::filter(.data, ...))
+}
+```
+
 <!--
-For submission to Archaeometry. Guidelines:
-
-- Max 8000 words including all figures etc.
-- Max 6 figures/tables
-- Abstract max 100 words
-- Max 10 keywords
-- Word 2002 or earlier or RTF (???)
-- Tables seperately in XLS
-- Figures seperately in TIFF or EPS
-- double spaced, wide
+* [x] First draft
+* [ ] Complete references:
+  * [x] Introduction
+  * [ ] State of the art
+  * [ ] Concept
+  * [ ] Implementation
+  * [ ] Evaluation
+  * [ ] Appendix table
+* [ ] Figures and tables
+  * [x] FIGURE: graph of number of datasets through time
+  * [x] FIGURE: map of number of dates in datasets
+  * [ ] FIGURE: distribution of radiocarbon dates in c14bazAAR / IntChron / p3k14c / XRONOS
+  * [ ] FIGURE: data model
+  * [ ] FIGURE: GUI screenshots (data browser, c14 show view, version history, data curation)
+  * [ ] TABLE: list of 'issues'
+  * [ ] APPENDIX: table of datasets
+* [ ] Other TODOs
+* [ ] Copyedit:
+  * [ ] Introduction
+  * [ ] State of the art
+  * [ ] Concept
+  * [ ] Implementation
+  * [ ] Evaluation
   -->
 
-## Introduction
+# Introduction
 
-* Chronometric modelling depends on ready access to chronometric data!
-* Luckily, we do have a lot of it.
-* Moving from 'open data' (uploading a CSV) to 'open infrastructure' [@Kintigh2006]
+Chronology is the backbone of the archaeological record.
+As a necessary prerequisite to understanding the context of any past event or process [@Lucas2004], it is has unsurprisingly been at the forefront of methodological development in archaeology for as long as the discipline has existed: from putting finds and events in sequence [@Rowley-Conwy2007] to using scientific methods to place them on an absolute timescale [@Libby1955] and increasingly sophisticated statistical methods to refine them [@BronkRamsey2009].
+If archaeology is to be an open science [@Lake2012], it is therefore critical that effective open access to chronological information be placed front and centre.
 
-Chronometric data has been one of the principle domains in which open data and open science approaches have been applied over the last ten years. 
-Comprehensive compilations of radiocarbon dates have become available for many parts of the world in the last decade and, as natural next step, there are now several initiatives to collate this data globally, including the retrieval tool c14bazAAR [@SchmidEtAl2019], the IntChron exchange format [@BronkRamseyEtAl2019], and the synthetic database p3kc14 [@BirdEtAl2022].
-This move from archiving individual datasets to producing a living open data infrastructure for archaeological chronologies raises a number of challenges. 
-Radiocarbon datasets are still sorely lacking for many parts of the world and, even in those regions with good coverage, the quality of data is highly uneven and largely undocumented. 
-There exists no central repository ensuring the long-term sustainability and completeness of these datasets, and the potential of placing other sources of chronometric information (e.g.
-dendrochronology, typological dating) in an open data framework has hardly been realised at all.
+Over the last two decades, archaeologists have answered this call by publishing an increasing number of compilations of dates from archaeological contexts as open data.
+These efforts have facilitated major reevaluations of previously-established chronologies [e.g.
+@HighamEtAl2014; @LoftusEtAl2019; @PratesEtAl2020; @KatsianisEtAl2020], important new insights into past processes [@RirisEtAl2024], and the development of novel ways of using chronological data [@Crema2022].
+<!-- TODO: Would welcome more citations here -->
+The focus has been overwhelmingly on radiocarbon dating and most compilations focus on a single region and/or period.
+The profusion of open radiocarbon data in particular has prompted several initiatives towards a global synthesis [e.g. @SchmidEtAl2019; @BronkRamseyEtAl2019; @BirdEtAl2022].
 
-Here, building on and complementing these initiatives, we present XRONOS: a new web-based platform for chronometric data from archaeological contexts worldwide, combining an open data repository with tools for importing, curating and analysing chronometric information from diverse sources.
-The principal design goals of the software are to combine all available sources of radiocarbon and other chronometric data in single database; 
-develop robust tools for the continuous ingestion and refinement of this data; 
-and disseminate this data within an open and FAIR framework, embedding it in the wider world of Linked Open Data in archaeology and beyond.
+At the same time, the broad range of other types of chronological information used in archaeology—from other radiometric methods to dendrochronology to typological dating and epigraphy—remains relatively difficult to access as open data.
+Even when it comes to radiocarbon data, the coverage of available compilations is patchy both geographically and in time and of variable quality (see @sec-c14-compilation).
+The publication of many overlapping, non-standardised and mostly static open data resources means that it is still difficult to obtain reliable and up-to-date chronological datasets, especially for applications that crosscut convential geographic and temporal domains of research.
+Initiatives towards synthesis this have improved this situation, but the goal of a global dataset that is both comprehensive and up-to-date remains elusive.
 
-This paper outlines the conceptual and technical infrastructure developed to realise these goals in XRONOS’ initial phases of development (2019 and 2021–2024), including a generalised data model for site and radiocarbon information, extendable to other chronometric data;
-an R- and Ruby-based pipeline for continuous ingestion of data from a variety of sources; continuous, semi-automated data cleaning protocols; 
-a Ruby-on-Rails application providing a web-based frontend to the data and a REST API for programmatic access; 
-and an R package for interfacing with the API. 
-We believe the XRONOS framework provides more open, more reliable, and more comprehensive access to chronometric data than previously available, as well as a foundation for its continuous expansion and refinement.
+XRONOS is a new open data infrastructure that aims to provide access to published radiocarbon dates and other chronometric data from any period, anywhere in the world.
+It is our attempt to move the state of the art in open archaeological chronology beyond the publication of static, one-off resources ['uploading CSVs', @Batist2023, pp.
+188-189], and towards a living digital infrastructure [@Kintigh2006] embedded in a transparent and sustainable collabrative network.
+The core of XRONOS is a server application that ingests chronological data from diverse sources, facilitates semi-automated and manual curation of this data, and makes it available via both a web-based graphical user interface (GUI) and machine-readable application programming interface (API).
+The web frontend can be accessed via <https://xronos.ch> and all components of the software are developed as free and open source software with source code available at <https://github.com/xronos-ch>.
 
-In this paper we:
+In the remainder of this paper, we describe the concept and implementation of XRONOS in relation to the state of the art in open chronometric data in archaeology, and evaluate our progress in achieving these goals as of writing.
+Since we envisage both XRONOS as a dataset and XRONOS as software to be continually developing resources, the description here should be read as a 'snapshot' rather than the final state of the project.
 
-* Review the state of the art in radiocarbon data compilation (and other chronometric data, such as they are), highlighting the need for XRONOS
-* Describe XRONOS' design goals and software architecture.
-* Describe the implementation of some representative features
-  * With a focus on the choices we made in representing/modelling chronological data
-* Assess the current state of the software and database, and give case studies of how it can be used
-* Conclude with our future plans
+# State of the art
 
-## State of the art
+## Compilations of radiocarbon dates {#sec-c14-compilation}
 
-We live in interesting times: Today, archaeological research can no longer complain that too few absolute dates exist for the phenomena of the past.
-Large-scale dating projects such as \[example\] and \[example\] are producing extensive data series in a short time, which make it possible to situate individual sites in time with unprecedented precision.
-But also due to the dating methods that have become affordable even for smaller budgets in the meantime and the integration of natural scientific dating that is almost mandatory, such large initiatives are accompanied by a great amount of individual and scattered chronological information that represents a rich source of data for supra-regional investigations.
-Based on that, data modelling can be used as a method to draw conclusions from a variety of individually information-weak data.
+Though an explicit emphasis on 'open data' is a relatively recent phenomenon in archaeology [@Lake2012], the open publication of compiled radiocarbon dates has a substantial prehistory.
+<!-- TODO: check and update this about date lists... -->
+Arnold and Libby [-@ArnoldLibby1951] initiated the tradition of regularly publishing all the dates they had obtained.
+This practice was subsequently continued as radiocarbon laboratories periodically shared and compiled their own 'date lists', published mainly in the journals *Radiocarbon* and *Archaeometry*.
+However, as the number of labs and volume of radiocarbon dates being produced grew, this paper-based format became impractical and mostly disappeared (with exceptions, e.g.
+@NdeyeEtAl2022) without being replaced by another form of systematic data-sharing or dissemination [@BronkRamseyEtAl2019].
+Additionally, because the date lists were sourced from radiocarbon laboratories directly—not those who collected the sample—they typically included only very limited contextual information.
+On the eve of the AMS revolution there was an effort to create a computerised 'International Radiocarbon Database' [@Kra1988]—already by 1989 described as a "much needed, long overdue enterprise" [@Kra1989]—but it never came to fruition.
 
-The problem, however, is that these data are very scattered.
-Traditionally, radiocarbon data were disseminated by the laboratories, with special attention paid to publication in the form of "date lists" in the journal *Radiocarbon*.
-Today, unfortunately, this has become very rare, on the one hand because the quantity and complexity of the data limit the possibilities of publication on paper.
-On the other hand, data in printed form are hardly suitable for reuse today.
-As a result, access to chronological information, especially a comprehensive overview, is made much more difficult [@BronkRamseyEtAl2019].
-Furthermore, data is presented in a variety of different formats and granularities, and critical information is often missing, although standards exist [@Millard2014].
+Thus, even though radiocarbon data comes from a relatively limited number of sources (one of the 172 active labs, @RadiocarbonLabList) and has relatively standardised reporting conventions [@Millard2014; @Bayliss2015], in practice the only way to produce aggregated datasets in recent decades has been to manually search through relevant literature for dates reported by the submitter of the sample.
+This already laborious process is further hampered by a significant inconsistency in how much authors adhere to measurement reporting conventions, a lack of conventions on the reporting of *contextual* information, and weak or nonexistent disciplinary norms regarding the responsibility to publish results openly in a timely fashion.
 
-Without data repositories, most comprehensive analyses are no longer feasible.
-So far, there has been no overarching database that constantly collects both 14C and dendrochronological data and makes them accessible in the long term.
-In the spirit of Open Science, it is the goal of XRONOS to establish such an Open Access database.
-Our primary ambition is to collect 14C and dendrochronological data for Switzerland and Europe as extensively and comprehensively as possible.
-However, the project is designed in such a way that an expansion beyond the originally defined spatial focus is already a reality.
+```{r data-c14-datasets}
+c14_datasets <- read_tsv("data/c14_datasets.tsv", show_col_types = FALSE)
 
-Currently, all scientific disciplines face a revolution in the way data is disseminated and used [@BronkRamseyEtAl2019].
-The access to these data, and methods integrate individual data into larger scientific questions of social relevance are both are decisive, but not yet sufficiently implemented prerequisites to reach an understanding of the temporal phenomena and dynamics in past societies [@Kristiansen2014].
-This novel, cross-disciplinary circulation and utilisation makes it possible to deal with completely new scientific questions and makes archaeological data relevant in thoroughly new contexts. 
-For example, analysis of past climatic changes and their long-term effects rely on information from archaeologists to estimate the impact on human and biotic activities [@KintighEtAl2015, 7-8]<!-- From Antiquity paper, not sure which paper exactly this is -->.
-An important source of such information is absolute chronological data.
-However, the basis of any empirically based modelling is the availability of data.
+n_c14_datasets <- nrow(c14_datasets)
+n_c14_datasets_sans_publication_year <- n_filter(c14_datasets, is.na(publication_year))
+```
 
-### Compilations of radiocarbon dates
+Despite these inefficiencies, there have been a profusion of published radiocarbon compilations since the decline of the date list.
+Our review of the literature identified `r n_c14_datasets` published since 1994.
+This is almost certainly an undercount, because our firsthand knowledge of regional literature was limited to Europe and West Asia and many resources only ever existed in 'grey' formats (e.g.
+websites that were not indexed and no longer exist).
+We also restricted ourselves to structured datasets disseminated primarily in a digital format; 
+'date lists' in printed periodicals and gazetteers were excluded.
+A full list of the datasets we identified is presented in appendix XXXX<!-- TODO -->.
 
-* Where do radiocarbon dates come from?
-* Where are they published?
+```{r fig-c14-datasets-time}
+#| fig-cap: Cumulative number of radiocarbon compilations published since 1995
+c14_datasets |>
+  drop_na(publication_year) |>
+  count(publication_year, sort = TRUE) |>
+  arrange(publication_year) |>
+  mutate(cumn = cumsum(n)) |>
+  ggplot(aes(publication_year, cumn)) +
+  geom_line() +
+  scale_x_continuous(breaks = scales::breaks_width(5)) +
+  labs(x = NULL, y = "Datasets*",
+       caption = glue("* Excludes {n_c14_datasets_sans_publication_year} datasets with an unknown year of publication")) +
+  theme_half_open()
+```
 
-Although 'open data' is often thought to be a new phenomenon in archaeology [@CITE], it has a long history in radiometric dating.
-Starting in the 19XXs, radiocarbon laboratories shared and compiled as 'date lists' published on a regular basis in the journal *Radiocarbon* [@CITE].
-However, as the number of labs and volume of radiocarbon dates being produced grew, this paper-based format became impractical: the last 'date list' appeared in 19XX, and unfortunately was not replaced by another form of systematic data-sharing or dissemination.
-Additionally, because the date lists were sourced from radiocarbon laboratories directly—not those who collected the sample—they typically included only very limited contextual information.
-Thus, even though radiocarbon data comes from a relatively limited number of sources [@IntCalLabList] and has relatively standardised reporting conventions [@Millard2014], in practice the only way to produce aggregated datasets recently has been to manually search through relevant literature for dates reported by the submitter of the sample.
-This already laborious process is further hampered by a significant inconsistency in how much authors adhere to measurement reporting conventions [@Millard2014], the lack of any convention on the reporting of contextual information, and weak or nonexistent disciplinary norms regarding the responsibility to publish results openly in a timely fashion.
+The number of available compilations has steadily increased since around 1995 (@fig-c14-datasets-time).
+The first generation came around the turn of the century and consisted mostly of online databases with a web frontend.
+These included some databases operated by radiocarbon labs, for example the Oxford Radiocarbon Lab (ORAU) and the Belgian Royal Institute for Cultural Heritage (KIK-IRPA), and essentially represented a continuation of their date lists in a digital format.
+The majority, however, were compiled from the literature and individual researchers interested in a particular region.
+Notable early examples include ANDES 14C in 1994 [Central Andes, @MichczynskiEtAl1995], CARD [Canada, @GajewskiEtAl2011] and RADON [Europe, @Raetzel-Fabian1999] in 1999, and CANEW in 2001 [Near East, @ReingruberThissen2005].
+From 2010, coinciding with broader shifts in scientific publishing [@TenopirEtAl2011], it became more common to publish standalone 'open data' products in the form of journal supplements, archives in repositories and/or data papers;
+the *[Journal of Open Archaeology Data](<https://openarchaeologydata.metajnl.com>)*, launched in 2012, has been a prominent venue for this latter category.
+Most recently there has been a trend towards providing version-controlled plain text data via platforms such as [GitHub](https://github.com), reflecting the broader adoption of these tools amongst computational archaeologists over the last decade [@BatistRoe2024].
+The shift from online databases towards more static but more preservable open data products is welcome, given how many databases from the first generation have subsequently ceased to be accessible.
+<!-- TODO: can this be quantified or visualised?
+-MH -->
+Version-controlled repositories are particular well-suited to data compilation projects because they allow for continued updates whilst still providing snapshot 'releases' that are citeable and can be archived in long-term repositories.
 
-* Graph of number of dates/databases through time
-* Table of databases
+```{r data-basemap}
+#| include: false
+coast <- gisco_get_coastallines()
+m49_regions <- gisco_get_countries() |>
+  mutate(
+    m49_macroregion = countrycode(ISO3_CODE, "iso3c", "un.region.name"),
+    m49_subregion = countrycode(ISO3_CODE, "iso3c", "un.regionsub.name"),
+    m49_intregion = countrycode(ISO3_CODE, "iso3c", "un.regionintermediate.name"),
+    m49_region = coalesce(m49_intregion, m49_subregion)
+  ) |>
+  summarise(geometry = st_union(geometry), .by = m49_region)
+```
 
-Despite these difficulties, there have been many efforts to systematically compile radiocarbon dates in this way over the last few decades [@FIG].
+```{r fig-c14-datasets-map}
+#| fig-cap: Geographic coverage of published regional radiocarbon compilations
+c14_datasets |>
+  separate_longer_delim(m49_region, "; ") |>
+  drop_na(m49_region) |>
+  summarise(n_dates = sum(n_dates, na.rm = TRUE), .by = m49_region) |>
+  right_join(m49_regions) |>
+  mutate(n_dates = replace_na(n_dates, 0)) |>
+  st_as_sf() |>
+  ggplot() +
+  geom_sf(aes(fill = n_dates), colour = NA) +
+  geom_sf(data = coast, fill = NA) +
+  scale_fill_turku(reverse = TRUE, labels = scales::comma) +
+  labs(fill = "Dates") +
+  coord_sf(crs = "+proj=moll") +
+  theme_minimal_grid()
+```
 
-E.g. RADON:
-At the time, this was one of the largest 14C data collections for absolute chronological information in Europe. 
-RADON and RADON-B can be considered a complete success, as they have been the basis for several of our own publications (Hinz, 2015; Hinz, Feeser, et al., 2012; Krossa et al., 2017; Weinelt et al., 2015) and as a data repository for high-level publications by other research groups (Bevan et al, 2017, 2019; Dolukhanov and Shrukov, 2004; Henkner et al., 2018; Lemmen and Wirtz, 2014; Manning et al, 2016; Pinhasi et al, 2005; Puchol and García, 2017; Sanders, 2012; Shennan et al, 2013; Timpson et al, 2014), and inspired other projects (Schmid, 2018; IsoMemo; Schmid et al, 2018).
+Although this body of work has greatly improved the accessibility of radiocarbon dates and supported significant methodological advances [@Crema2022; @CremaEtAl2024], some limitations are apparent.
 
-We are aware of XX existing radiocarbon databases or published datasets, comprising an estimated XXXX individual records [@TABLE] – though this is likely an underestimate.
-The majority of these resources have a specific regional (XX%) or temporal (XX%) scope.
-Others are associated with individual radiocarbon labs (e.g. KIK-IRPA, ORAU, BANADORA) and a handful aim at global coverage (XX%, discussed further below).
+The geographic coverage of regional radiocarbon compilations is markedly uneven (@fig-c14-datasets-map).
+Europe and, especially, North America are very over-represented [as has already been observed by @ChaputGajewski2016 and others].
+South America, West Asia, and East Asia are reasonably well-covered, but there practically no systematically compiled dates from East or West Africa, Central or South Asia, or Mainland Southeast Asia.
+This is probably explained in part by a lower volume of archaeological research and access to radiocarbon dating in these regions, but a lack of attention in compilation work must also be a factor.
+For example, radiocarbon dating has been an established part of Indian archaeology since at least 1961 [@KusumgarEtAl1963], but we have not able to locate a single systematic compilation of dates from the Indian subcontinent.
 
-* How are they published? As databases, SI, etc.?
-* How many are open?
+```{r data-c14-datasets-maintainance}
+n_c14_datasets_maintained <- n_filter(c14_datasets, maintained)
+n_c14_datasets_recently_updated <- n_filter(c14_datasets, maintained, last_updated > 2022)
+mean_c14_datasets_lifespan <- mean(c14_datasets$last_updated - c14_datasets$publication_year, na.rm = TRUE)
+```
 
-While these compilations have greatly advanced access to chronometric data and enabled new modes of analysis (e.g.
-'dates as data', @CITE), there are some obvious limitations to the fragmentary way in which they have been compiled.
+Datasets based on literature review also become out of date almost immediately upon publication, due the the constant stream of new dates.
+Unfortunately this applies to many databases that are in theory continuously updated, as it is common to see them become unmaintained and or unexpectedly become unavailable.
+Of the `r n_c14_datasets` published datasets we identified, `r n_c14_datasets_maintained` were intended to be continuously updated, but only `r n_c14_datasets_recently_updated` had received updates in the last two years.
+The average 'lifespan' of a dataset from its publication to its last update is `r mean_c14_datasets_lifespan` years.
+Most radiocarbon datasets we reviewed were compiled with a specific goal in mind (e.g.
+a particular analysis) and, even where there is the intention to keep them updated afterwards, the exigencies of scientific production combined with the labour-intensive nature of the process make that difficult to achieve in practice.
 
-* And patchy: highly uneven spatial and geographic coverage, different data collected, different conventions for recording data
-* Regional databases are put together *ad hoc* and rarely continuously maintained over long periods
-* The coverage of these databases is uneven, systematically biased, and duplicative of each other
-* Not all are open
-* Not all are machine readable
+Laboratory databases solve the problem of currency, but tend to have more arbitary coverage, since the inclusion of data is determined by who submits dates to that lab, not any form of principled curation.
+There are also comparatively few of them – most active labs no longer directly publish dates that they produce (if they ever did).
 
-* Recent moves towards a global synthesis: c14bazAAR [@SchmidEtAl2019], IntChron [@BronkRamseyEtAl2019], p3k14c [@BirdEtAl2022]
-  * Each taking a different approach, pros and cons
-  * We see a need for something that is comprehensive, unified, living
-* And we don't see why it should be limited to radiocarbon: what about other radiometric dates? Dendro? Typologies?
+In addition, the temporal and geographic coverage of these resources is uneven [@AlcantaraPedrozainpress], systematically biased [@ClistEtAl2023], and duplicative of each other.
+<!-- TODO: example, X databases for Europe, Y for the rest of the world? -->
+By our count, <!-- TODO: X --> of the databases are not 'open' according to the Open Knowledge Foundation's definition of data openness ["Open data and content can be freely used, modified, and shared by anyone for any purpose", @OpenKnowledgeFoundation], which both limits the access to and reuse potential of these datasets.
+<!-- TODO: X --> are not currently available in readily machine-readable formats (e.g. plain text or database files rather than PDFs or hypertext).
 
-* Some databases, such as KIK-IRPA, ORAU and BANADORA, which are maintained by 14C laboratories, only contain their own measurement data and are therefore of very limited use for overarching research projects.
-* Others, such as CONTEXT, RADON, RADON-B, 14SEA and CARD, have a limited regional or temporal focus, and the rich data available e.g.
-in Switzerland and surrounding countries is not sufficiently represented in these databases.
-In addition, some of them are project-specific and their continued existence can be described as uncertain due to funding constraints.
-* Recently, [@BronkRamseyEtAl2019] introduced IntChron, a data structure designed to ensure easy availability of radiocarbon data from published articles.
-We think this approach is very commendable and it inspires us in terms of reproducibility of dating.
-But we want to go beyond that and put more emphasis on the evaluation and dissemination of relevant information.
-"IntChron is not intended to be a stand-alone database" [@BronkRamseyEtAl2019, 5], but XRONOS is.
-An overarching database is critical for any research application in terms of data retrievability.
-Nevertheless, interoperability with other data services, including IntChron in particular, is very important to us.
+The fragmentation of the radiocarbon record into regional datasets also hinders analysis at larger scales.
+Although the core elements of a radiocarbon date—laboratory identifier, radiocarbon age, measurement error—are more or less standardised, there is no such consistency in contextual information on the sample or site.
+Such contextual information is important not just for the interpretation of dates, but for 'chronometric hygiene' [filtering out unreliable dates based on sample information, see e.g.
+@PettittEtAl2003] and for correcting for known systematic errors such as the marine reservoir effect [@AlvesEtAl2018].
+Most published datasets incorporate all or part of earlier compilations, meaning duplicate records are also very common, but deduplicating them is not a trivial problem due to format variations (see @sec-implementation-data).
+These issues are by no means impossible to overcome, but adds a significant amount of data-cleaning effort to a process that is otherwise very amenable to standardisation.
 
-### Beyond radiocarbon
+## Global radiocarbon compilations {#sec-global-compilations}
 
-* There are lots of other types of chronometric data!
-  * Dendro - in certain places (e.g. Switzerland!), way more important than radiocarbon
-  * Other radiometric/absolute dating methods, OSL, etc.
-  * Typology - still the backbone of many chronologies
-  * Application-specific: sea level dating, rock patina, etc.
-* Compilation of other types of dates has been much more limited.
-  * Exceptions: ...?
+<!-- TODO: Include CalPal?? -->
 
-<!-- MH to do: -->
-For European dendrochronological data, the situation is even less favourable for archaeological purposes:
+The profusion of radiocarbon compilations over the last decade has naturally prompted many to think globally.
+Three existing initiatives are particularly similar to XRONOS in their aims: c14bazAAR, IntChron, and p3k14c.
 
-* NOAA stores some global datasets, but has its main focus in the United States and, with its only 34 European prehistoric datasets, is not useful for the study of the European past.
-* The ADS database is compiled and maintained by the UK-based Vernacular Architecture Group and is limited to data published annually in the journal Vernacular Architecture, containing only medieval or later dates for the UK.
-* DendroDB has a European focus but only provides data for historical periods.
-* The "Dendrochronological Database" of the Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) has a worldwide focus on collected natural wood samples and has different goals than a database for archaeological wood material (furthermore, although announced a long time ago, this database is still not functional).
-* Although "Digital Collaboratory for Cultural Dendrochronology (DCCD).
-An international digital data library for dendrochronology" is well connected in the context of archaeological data services (e.g.
-ARIADNE), it is focused on the Netherlands (more than 2/3 of the data, while 0.08% of the Quercus data refer to Switzerland) and contains very few prehistoric data (about 2.5%).
-Furthermore, the overall activity in this database has decreased from 3846 new project records in 2010-2014 to only 83 since then (until the end of 2019).
+The first available synthetic radiocarbon database was c14bazAAR [@SchmidEtAl2019], an R package that provides an index of openly published radiocarbon databases and a common interface for retrieving them and performing basic data cleaning.
+Because c14bazAAR downloads data from its original source repositories, rather than mirroring it, it only includes resources that have been published in a fully open and machine-accessible format.
+Despite this limitation, it has global coverage and is still the largest collection of radiocarbon dates available (@FIG).
 
-* Typology
-  * perio.do
-  * Often embedded in other types of database (e.g. gazetteers)
-  * But nobody has treated it as a form of chronometric data in its own right?
+Another indexical approach is the IntChron project [@BronkRamseyEtAl2019], which exposes data from multiple sources and exposes them with a common JSON-based web interface.
+The IntChron specification is open, meaning that radiocarbon labs or compilation projects can implement it independently and thereby allow end users to access their data through a common interface (though to our knowledge it has so far only been adopted by databases associated with the Oxford Radiocarbon Lab).
+The JSON format also lends itself to the implementation of wrapper libraries, for example the rIntChron package gives direct access to IntChron-indexed databases in R [@Roe2024].
 
-What is completely missing worldwide so far, is a database that encompasses both, 14C and dendro data.
-However, to efficiently conduct data driven modelling of human-environmental dependencies and dynamics on the past, such a data base seems indispensable for future research.
+p3k14c [@BirdEtAl2022] instead compiles multiple source databases into a single flat file dataset, with a similar level of coverage to c14bazAAR. 
+The major advantage of this approach is that the data is made internally consistent and has been manually cleaned to an extent, which makes it particularly well-suited to global analyses. 
+The downside is that without the continuous link to the source databases present in the c14bazAAR and IntChron, it can only be kept up to date manually with periodic re-releases. 
+An accompanying package [@BirdEtAl2024] provides direct access to the p3k14c dataset in R.
 
-## Concept
+As of writing, c14bazAAR includes <!-- TODO: N --> radiocarbon dates with unique laboratory identifiers, IntChron <!-- TODO: N --> (excluding those from non-archaeological contexts), and p3k14c <!-- TODO: N -->.
+The geographic distribution of dates from each is similar (@FIG), reflecting the large degree in overlap between the sources of c14bazAAR and p3k14c in particular.
+IntChron, which in practice is currently only used to publish dates associated with the Oxford Radiocarbon Lab, has dates from more diverse contexts, but is much smaller.
 
-XRONOS builds on the structure of RADON (@RADON), one of the earliest comprehensive databases of radiocarbon dates.
+## Beyond radiocarbon
 
-### Design goals
+Radiocarbon has been by far the most active area of open data compilation, but archaeological chronology incorporates a much more diverse range of sources of information [@Harding1999].
+In periods beyond the practical limit of radiocarbon dating (c. 50,000 BP), other types of radiometric (K–Ar, U–Pb, etc.), chemical or luminescence dating offer an alternative [@Aitken1999].
+Conversely, in historic periods, radiocarbon is often relatively underused compared to conventional typological dating (based on artefact characteristics), which in these periods can offer comparable or better temporal resolution, or direct dating based on epigraphy [@HermankovaEtAl2021], numismatics [@KemmersMyrberg2011] or historical sources.
+In places where it is widely available, dendrochronology [@Baillie2014] also produces significantly better resolved chronologies and therefore tends to be the main source of chronometric data.
+Other more application-specific chronological methods include shoreline dating [@Brogger1905; @Roalkvam2023], lichenometry [@Benedict2009] and rock weathering dating [@Whitley2012; @Bednarik2020].
 
-Our concrete goals in the design and implementation of XRONOS are the following:
+Compared to radiocarbon, there are few examples of systematic, open compilations of any of these other types of data.
 
-1. Provision of access to 14C and dendrochronological data with meta information on archaeological contexts via web-based Open Access frontend
-2. Web frontend with processing functions (filters and queries, export to various data formats, 14C calibration & chronology formation) for in place analytical functions
-3. machine readable interface --- API --- for direct access e.g. by open source statistical environments (eg. R) for analytical purposes or dendroanalytical software
-4. licensing and versioning of individual data using eg. DOI
-5. Storage in a repository with user and role-specific access rights as well as encryption for data privacy
-6. Ensuring sustainable data security and usage through compliance with standards, appropriate backup solutions and high-availability technologies
+This is most striking when it comes to other radiometric/scientific dating methods, as the data structures and publication modes are very similar to radiocarbon.
+The 'Radiocarbon Palaeolithic Europe Database' [@Vermeersch2020], despite the name, includes a significant number of thermo- and optically stimulated luminescence, electon spin resonance, uranium–thorium and amino acid dates.
+Similarly, the AustArch database [@WilliamsEtAl2014] includes luminescence dates alongside radiocarbon, but is limited to Australia and was last updated in 2013.
+Apart from these and a few other exceptions where other scientific dates are collected alongside radiocarbon, we are not aware of any open compilations of them.
 
-Points 1-3 are now largely implemented and ready to be used.
-The results of such a search is easily exportable in formats that are simple and generally usable, and represent common, open standards (CSV, JSON).
-Such formats can be conveniently imported into common evaluation environments (R, Matlab, Python, OxCal, Chronomodel), or even better, directly accessed via the API.
-This can serve to establish a single gateway to the different sources of information and eliminate the necessity for data publishers to maintain information in multiple formats.
-With regard to the 14C data, this analytical functions include the live calibration of the individual dates, allowing the most recent calibration curve to be considered.
-In addition there are extended calibration possibilities like sum, as known from the standard software solutions (e.g.
-OxCal).
-Furthermore, the Bayesian calibration in relation to phase models (sequential calibration under consideration of stratigraphy) may be implemented as far as possible, whereby this will take place in close connection with the endeavours of the further development of corresponding R packages.
-For the dendrochronological data, especially the implementation of visualizations as well as the possibility to create mean curves and to identify impact phases will represent extended possibilities of immediate analysis.
-This portfolio is to be expanded in consultation with the user group, especially the state (cantonal) dendrochronological laboratories of Switzerland.
-More elaborated functionality should be made available primarily via an R Package, since this would be too complex for the user to be channelled via an web interface.
+### Dendrochronology
 
-An increasing number of research funders legitimately demand that the data that is used in the course of the research they fund become available to other researchers.
-This disclosure enables new knowledge to be extracted through comparative studies, data mining, etc.; it also enables better transferability and reproducibility of the research process, which increases the quality of research.
-All data in XRONOS should be uniquely referenceable.
-Common publication practices, which so far led to ambiguities regarding the data, must be overcome.
-Therefore, each individual data set should be permanently archived in its current version, and changes resulting from updated information or changed boundary conditions (updated calibration curves) should be provided in a new version with versioned unique identification.
-In this way, it can be ensured that the exact state of a reference to a date can be unambiguously stored.
-Only in this way can the research based on these data truly become traceable.
-One means for this is a Digital object identifier (DOI), which combines a unique and machine-readable as well as persistent referencing with easy retrievability.
-This functionality is still to be implemented.
-In order to make such a sustainable data referencing possible, XRONOS strives for a cooperation with longterm data repositories, so that a permanent storage of the relevant information can be guaranteed.
-This ensures that the collected information will be available to the research and public community beyond the lifetime of XRONOS and thus beyond a possible end of the project.
+With regard to tree-ring data, some databases provide valuable resources for dendrochronological studies in general but are not primarily designed to collect tree-ring data from archaeological contexts.
+For instance, Dendro4Art specializes in dendrochronological data related to wooden art objects, such as sculptures and panel paintings.
+While this focus serves art historians and conservationists well, its utility for studying prehistoric datasets is minimal.
+Similarly, the Dendrochronological Picture Database, maintained by the Swiss Federal Institute for Snow and Avalanche Research (SLF), offers a visual archive of approximately 1,400 images documenting dendrochronological phenomena.
+Although valuable as an educational resource, it does not provide raw data necessary for chronological or archaeological analyses.
+Additionally, the OLDLIST and Eastern OLDLIST databases focus on documenting the maximum ages of trees worldwide.
+Their emphasis on biological longevity, while significant for ecological research, limits their applicability to archaeological or prehistoric investigations.
 
-The ultimate goal of XRONOS is to create an open data structure for the exchange and evaluation of absolute chronological data by the entire research community.
-However, the data management with regard to dendrochronological data in particular is associated with different interests of the institutions involved, so that compromises have to be made here.
-For all dendrochronological laboratories control over data disclosure is indispensable.
-These are mainly publication-related reasons and data is used as "exchange capital" between laboratories.
-However, in order to provide an incentive for central data management, it must be ensured that users retain full sovereignty over their data.
-To this end, this project plans to keep the data fully encrypted in such a way that even the database administrators have no insight into the data.
-For this purpose, methods of asymmetric cryptography (public/private key encryption) based on passwords will be integrated into the database, which make it possible to make data available to individual other users or to the general public.
+Among databases that do provide dendrochronological data, the degree to which they support prehistoric research varies substantially.
+The NOAA International Tree-Ring Data Bank (ITRDB) serves as a global repository of tree-ring measurements.
+However, its focus remains predominantly on North America, with only 34 datasets representing European prehistoric contexts.
+This restricts its relevance for studying the European past.
+Similarly, the ADS database, maintained by the UK-based Vernacular Architecture Group, compiles dendrochronological data from the UK but is limited to medieval and later periods, making it unsuitable for prehistoric studies.
 
-We hope that the combination of easy and open data access, the provision of analytical functions within the web interface and by means of R packages, the sustainable, transparent and unambiguous data management, the long-term archiving already considered in the initial phase, together with the possibility for individual users to have full sovereignty over their data, will distinguish XRONOS as an outstanding data source and repository, and that we will thereby contribute to the sustainable provision and use of big data in archaeology and advance the use and improvement of standards in this field.
+DendroDB, hosted by the Swiss Federal Institute for Forest, Snow and Landscape Research (WSL), emphasizes ecological and climate studies over archaeological wood material.
+While it claims a broad scope, the database remains non-functional, rendering it ineffective for research needs.
+The CFS-TRenD database, managed by the Canadian Forest Service, compiles over 4,600 datasets from Canadian forests, primarily focusing on boreal ecosystems.
+Despite its extensive coverage for North America, its geographical specificity and lack of open access restrict its utility for European prehistoric contexts.
+Similarly, the QUB Dendrochronology Database, managed by Queen’s University Belfast, offers valuable datasets for Ireland and the UK but lacks significant representation of prehistoric material, limiting its application in broader archaeological investigations.
+The Building Archaeology Research Database (BARD) contains over 24,000 records, including dendrochronological data from more than 2,700 buildings.
+However, its focus on medieval and post-medieval timber-frame construction further narrows its utility for studies involving prehistoric wood samples.
 
-* Inspired by GBIF, which manages to integrate large amounts of data from diverse sources into a signal, consistent database
+The Digital Collaboratory for Cultural Dendrochronology (DCCD) presents itself as a potentially valuable international platform for dendrochronology, particularly through its integration with archaeological data services such as ARIADNE.
+However, it remains heavily biased toward datasets from the Netherlands, which account for more than two-thirds of its entries, while only 0.08% of its Quercus data pertain to Switzerland and just 2.5% represent prehistoric datasets.
+Notably, these estimates date back to 2021, and an updated assessment is currently unattainable following the platform’s migration to DataverseNL, which now charges an annual fee of nearly €6,000 for access.
+Furthermore, database activity has declined significantly, from 3,846 new project records between 2010 and 2014 to only 83 by the end of 2019.
+Although 519 additional records have been reported since 2021, it remains unclear whether this figure includes revisions to pre-existing entries, potentially inflating the count.
+The database’s narrow focus and restrictive access model significantly limit its broader utility for prehistoric research.
 
-### Data model
+Finally, the Strategic Environmental Archaeology Database (SEAD), hosted by Umeå University, integrates multiple environmental proxies, including dendrochronological data.
+However, the dendrochronological component is largely confined to Swedish data, with limited relevance to prehistoric contexts.
+While SEAD aims for broader applications, its dendro component has limited utility for studies outside Sweden.
 
-TODO: Figure: conceptual scheme, centred on chrons (i.e. an idealised form of the database).
+The utility of dendrochronological databases for prehistoric research varies widely.
+Global resources such as NOAA’s ITRDB and DCCD offer substantial datasets but face significant limitations in practical geographical and temporal scope.
+Similarly, platforms like DendroDB and BARD primarily cater to historical studies, leaving critical gaps in prehistoric coverage.
+Specialized resources like OLDLIST, Dendro4Art, and the Dendrochronological Picture Database provide valuable contributions but lack the direct relevance necessary for archaeological tree-ring analysis.
+Consequently, researchers focused on prehistoric dendrochronology must navigate a fragmented landscape of databases, each offering distinct strengths and limitations.
+Addressing these gaps remains crucial for advancing the field.
+
+### Typological dating
+
+Typological dates—i.e.
+relative, expertise- or seriation-based dating based on artefact characteristics—is an interesting case because they are ubiquitous but rarely treated as a form of chronometric data in their own right.
+For example, the majority of the radiocarbon datasets we reviewed (@sec-c14-compilation) included some form of typological chronological information in the form of a 'period' or 'culture' column.
+This is also typically present in many other forms of systematic compilation work in archaeology, for example site gazetteers.
+Aggregated typological information from such sources are often used in aoristic analysis and related methods [@Mischka2004; @Crema2024].
+What is lacking in this presentation of typological dating is metadata on how the determination was made and how exactly it is to be understood.
+Like any archaeological date, a typological date is derived from a physical sample – the object or set of object from which a chronological estimate was derived.
+Typological dates on one class of object may well clash with other classes of object, or for that matter with scientific dates, but without this kind of metadata such inconsistencies are difficult to resolve.
+<!-- TODO: clarify - MH -->
+Similarly the absolute date range corresponding to a typological determination (e.g.
+"Late Neolithic") can be interpreted in multiple ways depending on the region and intentions of the expert making the determination.
+PeriodO [@RabinowitzEtAl2016] is a linked open data infrastructure that includes a shared vocabulary of typological periods and corresponding calendar age estimates, and an important step towards addressing the latter problem.
+However, it remains to be systematically linked to actual compilations of typological dates.
+
+What is completely missing to date is a general-purpose infrastructure for combining all of these types of chronometric information on a global scale.
+This is the gap that XRONOS aims to fill, starting with three methods: radiocarbon, dendrochronology, and typological dating.
+These were chosen because they are widely used and relatively advanced in terms of open data, but an important aim of the project is to develop a generalisable data model that can easily scale to any and all types of archaeological chronology (see @sec-data-model).
+
+# Concept
+
+XRONOS inherits its basic structure from RADON [@Raetzel-Fabian1999; @RADON; @RADONB; @RADONB2024], with a database-backed web application and a data model that separates radiocarbon dates, contextual information, and sites.
+Our overall aims in developing XRONOS is to bring this model, which RADON has operated on for more than twenty years, up to date, to generalise it to other types of chronometric information, and to transform it from an online database to a data infrastructure that supports the continuous ingestion, curation, and open dissemination of archaeological chronologies from diverse sources.
+
+## Design goals
+
+XRONOS is our answer to Kintigh's call [@Kintigh2006] for digital infrastructures that don't just provide access to chronological data but enables researchers to "archive, access, integrate, and mine disparate data sets".
+It parallels and draws inspiration from several similar initiatives within and outwith archaeology, such as <!-- TODO examples -->.
+To improve upon existing global syntheses of radiocarbon dates (see @sec-global-compilations), we aimed to develop a living infrastructure that both continually collected data from diverse sources and presented a seamless single database to the user.
+The Global Biodiversity Information Facility [GBIF, <https://gbif.org>, @CanhosEtAl2004]—which provides a single, consistent interface to many sources of global biodiversity data—has served as an exemplar for us in this regard.
+
+Our principal goals for the software were therefore to:
+
+1. Combine all available sources of radiocarbon and other chronometric data in single database; 
+2. Develop robust tools for the continuous ingestion, collation and curation of this data;
+3. Disseminate the collated and curated data as linked open data within a FAIR framework.
+
+Meeting these goals required principally the development of a) a conceptual data model, including links to other open data resources, that is flexible enough for all forms of chronometric data; and b) an software implementation that supports the main functions of ingesting, curating, and disseminating this data.
+The individual components of this work are described in more depth below but, briefly, consist of a relational data model implemented in a PostgreSQL database; a Ruby application providing server-based tools for ingestion, curation and dissemination of data; and multiple graphical and programmatic interfaces to the resulting dataset.
+
+## Data model {#sec-data-model}
+
+<!-- TODO: data model figure -->
+
+<!-- TODO: shorten -->
 
@@
 The natural choice is a *calendar probability distribution*, which expresses the probability that an event occurred as a function of time on a calendric scale.
-Most archaeologists are familiar with working with this kind of representation in the form of calibrated radiocarbon dates, but it can be extended and generalised to essentially any kind of chronological information [@CITE].
-For example, in aoristic analysis [@CITE], a periodic time estimate (e.g.
+Most archaeologists are familiar with working with this kind of representation in the form of calibrated radiocarbon dates, but it can be extended and generalised to essentially any kind of chronological information.
+For example, in aoristic analysis [@Mischka2004], a periodic time estimate (e.g.
 the event occurred in the Neolithic) is conceptualised as a uniform probability distribution over the timespan between the known start and end dates of that period.
-A similar model is used in OxCal [@OxCal, a direct inspiration for our approach] to integrate prior chronological information from diverse sources.
+A similar model is used in OxCal [@BronkRamsey2024, a direct inspiration for our approach] to integrate prior chronological information from diverse sources.
 In practical terms, this model means that the canonical representation the time component of any chron in XRONOS', regardless of source, is a probably distribution over the set of calendar years (arbitrarily measured in years Before Present) in which it could have plausibly occurred.
@@
 
-* Bibliographic references
-* Paper trail
+Metadata is incorporated into XRONOS' data model at the level of the individual records (e.g.
+all records store their data of creation and last modification) and through two additional nodes: bibliographic references and versions.
+Bibliographic references store information on the source of a record in BibTeX format and can be linked through many-to-many associations to
+sites or chrons.
+Versions are a special type of record that are associated with all other records (including bibliographic references) and store the previous versions of those records as a series of changesets.
+In this way all changes to data are recorded and can be reconstructed (or reversed) precisely.
+This 'paper trail' also stores contextual metadata, e.g. who made the change and why.
+It also means that records that are deleted can be reviewed or restored from their stored version history, which is never discarded.
+Together these two systems provide a transparent record where the data in XRONOS comes from and how it has been altered, which we view as essential in a scientific data infrastructure.
 
-* LOD??
+## Linked data
 
-### Software architecture
+The XRONOS data model presents several opportunities to link to other resources as linked open data (LOD).
+We use controlled vocabularies in the form of the GBIF Backbone Taxonomy (for taxonomic descriptions of samples) via its API, and in future plan to extend this to for example site types (using the Getty and/or DAI<!-- TODO: check & cite -->).
+Standardising field using these vocabularies to automatically then link the two entities.
+Beyond this, there is also a concordance between XRONOS entities and other specialised data infrastructures in archaeology, such as Perio.do (@RabinowitzEtAl2016, mapping to typological chrons) and gazetters like Pleiades [@Pleiades] or Vici (@Voorburg2012, mapping to site names).
+Moving beyond archaeology, Wikidata (<https://wikidata.org>) already includes many of the concepts represented in the XRONOS model (e.g. archaeological sites, Q<!-- TODO -->).
+Linking to Wikidata is especially useful because it dissimenates—and thereby preserves—the data compiled in XRONOS beyond the narrow field of archaeology.
+It also allows us to enrich the database with contextual information that otherwise be beyond our scope and resources, for example embedding multilingual descriptions of sites from Wikipedia articles on site records where the site has been linked to a Wikidata item.
 
+Conversely, we encourage others to use XRONOS as a linked open data resource by providing stable URLs and a machine-readable interface for every record.
+Such usages could look like, for example, using a XRONOS URL as a canonical representation of a single radiocarbon date (e.g. <!-- TODO -->).
+We also plan to implement an IntChron interface to XRONOS, to allow it to be indexed through that existing standard.<!-- TODO: also mention RDF? -->
+
+# Implementation
+
+Following a short pilot project in 2019, the first phase of development of XRONOS was completed in 2021–2024, supported by a project grant from the Swiss National Science Foundation.
+The web interface (<https://xronos.ch>) has been publicly accessible since July 2021.
+Though we envisage XRONOS as a continuously-developing open source project and 'living database', the following offers a snapshot of progress at the end of our first grant-funded implementation phase.
+We do not aim to be comprehensive, but rather to describe some key elements of XRONOS' current implementation that illustrate how our concept has been realised in practice.
+
+## Software architecture
+<!-- TODO: shorten -->
+
+<!-- Overview -->
 The XRONOS data model is implemented as a relational database using the free and open source database management system PostgreSQL.
@@
 
+## Data ingestion and curation {#sec-implementation-data}
+
+The chronological data in XRONOS comes from a variety of sources, including published structured datasets in repositories and journal supplements, other online databases, literature review, and direct input from collaborators.
+Our aim is not just to 'mirror' these sources as they are, but integrate them into a single curated and continuously updated database.
+<!-- TODO: too detailed?
+-->For the purposes of ingestion, we classify data resources into three categories: static resources, such as supplementary data in published papers, which are imported once; versioned resources, updated on a periodic basis, which we import after each new version; and live resources, which are continuously updated and therefore continuously imported.
+Records of each type are imported into XRONOS in as close to their original state as possible, i.e. without any corrections or standardisation applied.
+This ensures that any subsequent changes (even immediate, automatic ones) are entered into the record's version history, so that the source of any deviations or potential errors can always be reconstructed.
+The version history also records the direct source of the data for attribution purposes.
+In addition, a bibliographic reference to the original resource is attached to ensure that the source is clearly attributed even if the record is merged with another one.
+
+```{r tbl-issues}
+#| tbl-cap: Automatically-recognised data quality issues currently implemented in XRONOS
+tribble(
+  ~entity,                    ~issue_code,            ~n,     ~issue_description,
+  "Sites",                    "MISSING_COORDINATES",  4452,   "Missing geographic coordinates",
+  "Sites",                    "INVALID_COORDINATES",  0,      "Geographic coordinates fall outside the earth's ellipsoid",
+  "Sites",                    "MISSING_COUNTRY_CODE", 1221,   "Missing data on what country the site is in",
+  "Samples",                  "MISSING_MATERIAL",     138177, "Missing data on the sample material",
+  "Samples",                  "MISSING_TAXON",        138182, "Missing data on the sample taxon",
+  "Samples",                  "MISSING_CRS",          0,      "Sample has coordinates but the coordinate reference system is not given",
+  "Taxons",                   "UNKNOWN_TAXON",        9260,   "Sample taxon has not been matched to the GBIF Backbone Taxonomy",
+  "Taxons",                   "LONG_TAXON",           509,    "Description of the sample taxon is implausibly long",
+  "Radiocarbon dates",        "MISSING_C14_AGE",      447,    "Missing radiocarbon age",
+  "Radiocarbon dates",        "VERY_OLD_C14",         76,     "Radiocarbon age older than the effective range of the method (50 ka)",
+  "Radiocarbon dates",        "MISSING_C14_ERROR",    585,    "Missing measurement error",
+  "Radiocarbon dates",        "MISSING_D14C",         238875, "Missing δ13C measurement",
+  "Radiocarbon dates",        "MISSING_D14C_ERROR",   238875, "Missing δ13C measurement error",
+  "Radiocarbon dates",        "MISSING_C14_METHOD",   242233, "Missing data on radiocarbon dating method (conventional, AMS, etc.)",
+  "Radiocarbon dates",        "MISSING_C14_LAB_ID",   1233,   "Missing laboratory identifier",
+  "Radiocarbon dates",        "INVALID_LAB_ID",       16053,  "Laboratory identifier does not match the standard format (e.g. 'Abc-1234')",
+  "Radiocarbon dates",        "MISSING_C14_LAB",      350190, "Missing data on the radiocarbon laboratory",
+  "Bibliographic references", "MIXED_REFERENCE",      21933,  "Bibliographic reference appears to combine multiple publications",
+  "Bibliographic references", "MISSING_BIBTEX",       42109,  "Bibliographic reference without structured data in BibTeX format",
+) |>
+  group_by(entity) |>
+  gt() |>
+  cols_label(issue_code = "Issue", n = "N", issue_description = "Description") |>
+  cols_width(everything() ~ pct(33)) |> # TODO: fix...
+  tab_options(
+    table.font.size = 13
+  )
+```
+
+Once ingested, we apply a number of automated and semi-automated quality control processes to integrate new data into the existing database.
+Controlled vocabularies are used in a number of places in the data model (@FIG), and we use thesauruses to automatically standardise these fields as much as possible.
+For example, the taxonomic description of samples is controlled using GBIF's backbone taxonomy [@GBIF2023], and we also use a thesaurus service provided by GBIF to automatically change variant or obsolete taxonomic names to the canonical version.
+If the system is not able to standardise a field using the available thesaurus, it is flagged for manual correction.
+A wide variety of other potential data quality issues (e.g.
+missing data on what country a site is in) are also flagged for human review by this system [@tbl-issues], which can often be semi-automated (e.g.
+suggesting close matches in the thesaurus or the country indicated by the record's coordinates).
+
+A final critical component of XRONOS' data curation system is duplicate handling.
+We import data from many overlapping resources (many of which incorporate each other either in whole or in part), so duplicate records are common [as recently discussed by @ReiterEtAl2024].
+The end result of standardising and correcting a record is also often to create a duplicate: e.g.
+the same sample imported from one source as 'oak' but another as '*Quercus* sp.' will become a duplicate pair as '*Quercus*', and thus be recognised as a single sample.
+Such exact duplicates can be merged automatically, with the oldest record becoming the authoratitive version, but detecting fuzzier duplicated information (e.g.
+differences in the spelling of site names) has proved a more difficult problem.
+As of writing there are therefore still many duplicate records in XRONOS that need to be manually resolved, but we hope to automate much more of this work in the future.
+
+## User interfaces
+
+<!-- TODO: figures? -->
+<!-- TODO: add in more concrete examples -->
+
+<!-- Public GUI -->
 The graphical user interface (GUI) to XRONOS, accessed through a web browser (e.g.
-at https://xronos.ch), uses REST resources and actions as the building blocks for various interfaces through which users can browse, search, retrieve, and analyse chronometric data.
+at <https://xronos.ch>), uses REST resources and actions as the building blocks for various interfaces through which users can browse, search, retrieve, and analyse chronometric data.
 Each action on each resource is represented by a page, though not all of these are publicly accessible.
@@
 Though users can make such requests manually and parse the data with one of several off-the-shelf tools, the primary intended uses of this interface is to provide access for 1) programmatic clients to XRONOS and 2) other web services.
-The XRONOS R package (discussed in more detail below) is an example of the former use, which could be extended to other programming environments used for scientific computing, e.g. Python or Haskell.
-<!-- TODO: example of a web service that might use the API? -->
+The XRONOS R package [@HinzRoe2024] is an example of a programmatic client; it uses the API to facilitate direct querying and retrieval of data from XRONOS in the R statistical programming language [@R2024], which is widely used for computational applications in archaeology [@BatistRoe2024].
+Similar libraries could be developed in other programming environments used for scientific computing, such as Python or Haskell.
+The API also provides the foundation for other web services to access XRONOS directly, to embed chronological information in other contexts or otherwise make use of its data resources.
 
+<!-- Conclusion. TODO: do we need this? -->
 An overarching principle of this software architecture is that all interaction with XRONOS' data store, and as much of the data processing and 'business logic' of responding to REST requests as possible, is directed through the same server-side routines.
@@
 We have however tried to mitigate this by providing clearly-documented source code and regular data dumps so that, if our instance of XRONOS disappears, or if one simply does not want to use it, it is possible for others to host a XRONOS server of their own.
-As the sustainability of scientific software and data infrastructures is a pressing problem (a point we will return to in the conclusion), in the future it may be desirable to support further decentralisation through, for example, a federated server-server model [@CITE].
+As the sustainability of scientific software and data infrastructures is a pressing problem, in the future it may be desirable to support further decentralisation through, for example, a federated server-server model.
 
-## Implementation
+# Evaluation
 
-The first prototype of XRONOS was developed in 2019.
-It has been under sustained development 2021–2024 with the first public release in July 2021.
+In this paper we have outlined the conceptual and technical infrastructure developed during the initial phases of development on XRONOS in 2019 and 2021–2024.
+These include a generalised data model for radiocarbon and typological dates, extendable to other chronometric information, and associated contextual information;
+an R- and Ruby-based pipeline for continuous ingestion of data from a variety of sources; 
+continuous, semi-automated data cleaning protocols; 
+a Ruby-on-Rails application providing a web-based frontend to the data and a REST API for programmatic access; 
+and an R package for interfacing with the API. 
+These systems are in production and publicly accessible at <https://xronos.ch>.
 
-<!-- TODO: update/rewrite status of development -->
-The current database is modular and, in its current stage of development, contains all the structures necessary for storing 14C data. 
-In addition, the integration of dendrochronological information into the structure is already prepared, and a security and data protection architecture has been prepared for the final stage of development.
-The software is available as open source on Github (<https://github.com/xronos-ch),> and an R package for easy access, data retrieval and evaluation is also already available. 
-XRONOS currently holds 92,129 14C data from 22,225 sites. 
-The primary focus is on Switzerland and Europe, but the database also contains information from Africa, South America, Asia and Australia. 
-The collection is subject to continuous addition.
+At the current stage of implementation, we argue XRONOS provides a framework for access to chronometric data that is more open, more reliable, and more comprehensive than the previously available global radiocarbon compilations.
+XRONOS blends aspects of the three existing approaches (c14bazAAR, IntChron, and p3k14c) to achieve the same aim of providing access to the global radiocarbon data through a common interface.
+Like c14bazAAR and IntChron, it is a 'metadatabase' that draws from existing data resources and maintains an explicit link to them.
+But like p3k14c, it integrates these into a single database and applies data curation processes to harmonise them and improve the quality of the information.
+It has a wider scope than c14bazAAR or IntChron, as it mirrors rather than directly retrieves the source data (allowing us to use resources that aren't openly published), and does not rely on the authors of these sources to implement a common specification.
+It also goes beyond the functionality of p3k14c by providing systems for the continuous ingestion and curation of new data as it is published.
+With that said, we see the approaches followed by as complementary rather than competing.
+A c14bazAAR parser for XRONOS is in development (<https://github.com/ropensci/c14bazAAR/pull/150>), and we also aim to provide an IntChron interface to XRONOS' data in the near future.
+New data and corrections from p3k14c are incorporated into XRONOS as they are released.
 
-Due to space limitations, and because we hope and expect XRONOS to continuously evolve, we don't attempt a comprehensive description of its implementation here.
-Instead we offer a selection of representative features which illustrate our design goals and key aspects of our approach.
+We intend to continue development of XRONOS into the foreseeable future.
+Our immediate development goals include the incorporation of dendrochronology into the database, further refinement of our data curation pipelines, and the public release of the editing interfaces.
+Looking beyond the near term, we have endeavoured to create a sustainable infrastructure that can be maintained by a wider scholarly commons – though we must acknowledge that this is a difficult problem, and one that is more organisational than technical.
+The source code for all the software components of the system are available online (at <https://github.com/xronos-ch>) under open licenses.
+The databases of the instance at <https://xronos.ch> are also archived with Zenodo (<!-- TODO: link -->).
+With these two resources we reduce XRONOS' 'bus factor'; if we are not able to continue operating XRONOS, somebody else can fully recreate it.
+Equally importantly, we enable the 'right to fork' should others wish to take the software and/or database in another direction.
+But aside from these extreme scenarios, the long-term sustainability of XRONOS is contingent on the existence of a community of scholars that use and contribute back to it.
+In this we can only lead the horse to water; we cannot make it drink.<!-- TODO: too glib? -->
 
-* Data import / assessment of XRONOS' coverage
-* The data browser
-* XRONOS R package
-* Tale from the trenches: cleaning species and countries?
-* Tale from the trenches: duplicate processing?
-* Case study of use?
+# Acknowledgements
 
-## Conclusion
+The XRONOS project was carried under the direction of Albert Hafner. 
+Individual contributors to the XRONOS database to date include Chiara Huwiler, Rivana Moser, Tomasz Chmielewski, and Stephanie Döppler. 
+A complete and up-to-date list of acknowledgements for the project can be found at <https://xronos.ch/about/acknowledgements>.
 
-* Chronology is the backbone of the archaeological record
-* XRONOS could be a backbone for a universe of linked open data resources...
-* ...but sustainability is a challenge.
-  * We're coming to the end of our project funds
-  * We'll try to get more, but you never know (grant-based science and long-term investments in digital infrastructure don't work so well together!)
-  * Future sustainability
-
-## Data availability statement
-
-TODO
-
-## Funding statement
+# Funding statement
 
 This work was funded by the Swiss National Science Foundation ([SNSF Project #198152](https://data.snf.ch/grants/grant/198153)) and the University of Bern (Initiator Grant, Caroline Heitz, Project 'Time and Temporality').
-
-## Acknowledgements
-
-Albert Hafner. Plus contributors to XRONOS:
-
-* Chiara Huwiler
-* Rivana Moser
-* Tomasz
-* Stephanie Döppler

### 2024-11-29 paper.qmd
changed font to make it render locally, silenced fig-c14-datasets-map

@@
     papersize: a4
-    mainfont: "Linux Libertine"
+    mainfont: "Linux Libertine O" # had to adjust, otherwise was not rendering
     sansfont: "URW Gothic"
@@
 
-```{r fig-c14-datasets-map}
+```{r fig-c14-datasets-map, message=FALSE}
 #| fig-cap: Geographic coverage of published regional radiocarbon compilations

### 2024-11-29 paper.qmd
added reference to suppMat to captions figures

@@
 ```{r fig-c14-datasets-time}
-#| fig-cap: Cumulative number of radiocarbon compilations published since 1995
+#| fig-cap: Cumulative number of radiocarbon compilations published since 1995 according to our survey (see Supplementary Material).
 c14_datasets |>
@@
 ```{r fig-c14-datasets-map, message=FALSE}
-#| fig-cap: Geographic coverage of published regional radiocarbon compilations
+#| fig-cap: Geographic coverage of published regional radiocarbon compilations according to our survey (see Supplementary Material).
 c14_datasets |>

### 2024-11-29 paper.qmd
added some examples for kintighs call

@@
 XRONOS is our answer to Kintigh's call [@Kintigh2006] for digital infrastructures that don't just provide access to chronological data but enables researchers to "archive, access, integrate, and mine disparate data sets".
-It parallels and draws inspiration from several similar initiatives within and outwith archaeology, such as <!-- TODO examples -->.
+It parallels and draws inspiration from several similar initiatives within and outwith archaeology, such as <!-- TODO examples
+Examples:
+https://link.springer.com/article/10.1007/s10816-018-9399-6
+https://link.springer.com/article/10.1007/s10816-010-9098-4
+https://www.cambridge.org/core/journals/latin-american-antiquity/article/public-database-of-archaeological-resources-on-easter-island-rapa-nui-using-google-earth/F30677F6FC99B4C2BC762BCC98FD0966
+https://www.cambridge.org/core/journals/advances-in-archaeological-practice/article/data-integration-in-the-service-of-synthetic-research/E297C57441239AB35053F987AA758EB9
+https://www.cambridge.org/core/journals/advances-in-archaeological-practice/article/tdar/FCE1056A210A452E5FE65560763D4CC7
+https://www.cambridge.org/core/journals/advances-in-archaeological-practice/article/googlebased-freeware-solution-for-archaeological-field-survey-and-onsite-artifact-analysis/7914F673D1E1CAD96E5270E794CDFFD9
+https://link.springer.com/article/10.1007/s10816-015-9272-9
+https://link.springer.com/article/10.1007/s10816-015-9251-1
+https://link.springer.com/article/10.1007/s10816-015-9240-4
+https://anatomypubs.onlinelibrary.wiley.com/doi/10.1002/ar.23130
+https://academic.oup.com/dsh/article/29/3/349/2938128
+https://www.sciencedirect.com/science/article/pii/S0167739X13000678?via%3Dihub
+
+-->.
 To improve upon existing global syntheses of radiocarbon dates (see @sec-global-compilations), we aimed to develop a living infrastructure that both continually collected data from diverse sources and presented a seamless single database to the user.

### 2024-11-29 paper.qmd
added some comments to comments

@@
 charcoal, charred seed) and, where applicable, taxonomic designation (e.g.
-TODO, TODO) of the organic material used for dating.
+TODO, TODO<!--sorry, not fully clear what goes here.
+MH.-->) of the organic material used for dating.
 For typological dates, an ideal scenario would be for the sample to represent the particular object from which an inference was made (e.g. 'Natufian' might be inferred from 'lunate-type microlith').
@@
 The series of relations `[chron] > sample > context > site` links the chronological and contextual sides of the XRONOS data model.
-Each step is a many-to-onone association, meaning for example that it is possible to attach multiple chrons to the same sample (e.g.
+Each step is a many-to-one association, meaning for example that it is possible to attach multiple chrons to the same sample (e.g.
 replicated radiocarbon dates on the same material), multiple *types* of chrons to the same sample (e.g.
@@
 Since this kind of information is rarely systematically recorded in our source databases, there are currently few actual records that make use of this feature of the data model.
-However, we hope it will provide a foundation for ... ? <!-- TODO -->
+However, we hope it will provide a foundation for ... ? <!-- TODO Not fully clear what need to go here. MH. -->
 

### 2024-11-29 paper.qmd
removed some double (( and added stonehenge

@@
 Standardising field using these vocabularies to automatically then link the two entities.
-Beyond this, there is also a concordance between XRONOS entities and other specialised data infrastructures in archaeology, such as Perio.do (@RabinowitzEtAl2016, mapping to typological chrons) and gazetters like Pleiades [@Pleiades] or Vici (@Voorburg2012, mapping to site names).
-Moving beyond archaeology, Wikidata (<https://wikidata.org>) already includes many of the concepts represented in the XRONOS model (e.g. archaeological sites, Q<!-- TODO -->).
+Beyond this, there is also a concordance between XRONOS entities and other specialised data infrastructures in archaeology, such as Perio.do [@RabinowitzEtAl2016, mapping to typological chrons] and gazetters like Pleiades [@Pleiades] or Vici [@Voorburg2012, mapping to site names].
+Moving beyond archaeology, Wikidata (<https://wikidata.org>) already includes many of the concepts represented in the XRONOS model (e.g. archaeological sites, Q<!-- TODO Eg Q39671? MH.-->).
 Linking to Wikidata is especially useful because it dissimenates—and thereby preserves—the data compiled in XRONOS beyond the narrow field of archaeology.

### 2024-11-29 paper.qmd
added comments to comments and nerdy latin plural

@@
 Conversely, we encourage others to use XRONOS as a linked open data resource by providing stable URLs and a machine-readable interface for every record.
-Such usages could look like, for example, using a XRONOS URL as a canonical representation of a single radiocarbon date (e.g. <!-- TODO -->).
+Such usages could look like, for example, using a XRONOS URL as a canonical representation of a single radiocarbon date (e.g. <!-- TODO Not fully clear what need to go here. MH. -->).
 We also plan to implement an IntChron interface to XRONOS, to allow it to be indexed through that existing standard.<!-- TODO: also mention RDF? -->
@@
 Once ingested, we apply a number of automated and semi-automated quality control processes to integrate new data into the existing database.
-Controlled vocabularies are used in a number of places in the data model (@FIG), and we use thesauruses to automatically standardise these fields as much as possible.
+Controlled vocabularies are used in a number of places in the data model (@FIG), and we use thesauri to automatically standardise these fields as much as possible.
 For example, the taxonomic description of samples is controlled using GBIF's backbone taxonomy [@GBIF2023], and we also use a thesaurus service provided by GBIF to automatically change variant or obsolete taxonomic names to the canonical version.

### 2024-11-29 paper.qmd
suggesting an ESTER Mentioning

@@
 Our immediate development goals include the incorporation of dendrochronology into the database, further refinement of our data curation pipelines, and the public release of the editing interfaces.
+<!--In parallel, the planned ERC Consolidator Grant Project ESTER, focusing on the estimation of population development in prehistory, will over the next five years play a crucial role in supporting the continuation and further development of the XRONOS database, ensuring its sustainability and fostering interdisciplinary collaboration.
+-->
 Looking beyond the near term, we have endeavoured to create a sustainable infrastructure that can be maintained by a wider scholarly commons – though we must acknowledge that this is a difficult problem, and one that is more organisational than technical.

### 2024-11-29 paper.qmd
corrected further citations in brackets

@@
 This practice was subsequently continued as radiocarbon laboratories periodically shared and compiled their own 'date lists', published mainly in the journals *Radiocarbon* and *Archaeometry*.
-However, as the number of labs and volume of radiocarbon dates being produced grew, this paper-based format became impractical and mostly disappeared (with exceptions, e.g.
-@NdeyeEtAl2022) without being replaced by another form of systematic data-sharing or dissemination [@BronkRamseyEtAl2019].
+However, as the number of labs and volume of radiocarbon dates being produced grew, this paper-based format became impractical and mostly disappeared [with exceptions, e.g.
+@NdeyeEtAl2022] without being replaced by another form of systematic data-sharing or dissemination [@BronkRamseyEtAl2019].
 Additionally, because the date lists were sourced from radiocarbon laboratories directly—not those who collected the sample—they typically included only very limited contextual information.
@@
 
-Thus, even though radiocarbon data comes from a relatively limited number of sources (one of the 172 active labs, @RadiocarbonLabList) and has relatively standardised reporting conventions [@Millard2014; @Bayliss2015], in practice the only way to produce aggregated datasets in recent decades has been to manually search through relevant literature for dates reported by the submitter of the sample.
+Thus, even though radiocarbon data comes from a relatively limited number of sources [one of the 172 active labs, @RadiocarbonLabList] and has relatively standardised reporting conventions [@Millard2014; @Bayliss2015], in practice the only way to produce aggregated datasets in recent decades has been to manually search through relevant literature for dates reported by the submitter of the sample.
 This already laborious process is further hampered by a significant inconsistency in how much authors adhere to measurement reporting conventions, a lack of conventions on the reporting of *contextual* information, and weak or nonexistent disciplinary norms regarding the responsibility to publish results openly in a timely fashion.

### 2024-11-29 paper.qmd
Add fig-data-model

@@
 library("dplyr", warn.conflicts = FALSE)
+library("dm")
 library("giscoR")
@@
 library("khroma")
+library("magick")
 library("patchwork", warn.conflicts = FALSE)
 library("readr")
+library("RPostgres")
 library("sf")
 library("tidyr")
+library("webshot2")
 
@@
   * [ ] FIGURE: distribution of radiocarbon dates in c14bazAAR / IntChron / p3k14c / XRONOS
-  * [ ] FIGURE: data model
+  * [x] FIGURE: data model
   * [ ] FIGURE: GUI screenshots (data browser, c14 show view, version history, data curation)
@@
 
-<!-- TODO: data model figure -->
+```{r fig-data-model}
+#| fig-cap: Simplified entity relationship diagram showing the XRONOS data model
+# Requires a running XRONOS dev database
+# See https://github.com/xronos-ch/xronos.rails for setup instructions
+xronos_dev_db <- DBI::dbConnect(
+  RPostgres::Postgres(),
+  user= "xronos",
+  password = "xronos",
+  dbname = "xronos_dev",
+  host = "localhost"
+)
+
+xronos_dm <- dm_from_con(xronos_dev_db, learn_keys = TRUE) # doesn't learn much
+
+xronos_dm_svg <- xronos_dm |>
+  dm_select_tbl(c("c14_labs", "c14s", "cals", "citations", "contexts",
+                  "materials", "references", "samples", "site_names",
+                  "site_types", "site_types_sites", "sites", "taxons", "typos", 
+                  "users", "versions", "wikidata_links")) |>
+  # Relations
+  dm_add_fk("c14s", "c14_lab_id", "c14_labs") |>
+  dm_add_fk("c14s", "sample_id", "samples") |>
+  dm_add_fk("cals", c("c14_age", "c14_error"), "c14s", c("bp", "std")) |>
+  dm_add_fk("citations", "reference_id", "references") |>
+  dm_add_fk("citations", "citing_id", "sites") |>
+  dm_add_fk("citations", "citing_id", "c14s") |>
+  dm_add_fk("citations", "citing_id", "typos") |>
+  dm_add_fk("contexts", "site_id", "sites") |>
+  dm_add_fk("samples", "context_id", "contexts") |>
+  dm_add_fk("samples", "material_id", "materials") |>
+  dm_add_fk("samples", "taxon_id", "taxons") |>
+  # dm_add_fk("site_names", "site_id", "sites") |> # autodetected
+  dm_add_fk("site_types_sites", "site_id", "sites") |>
+  dm_add_fk("site_types_sites", "site_type_id", "site_types") |>
+  dm_add_fk("typos", "sample_id", "samples") |>
+  dm_add_fk("wikidata_links", "wikidata_linkable_id", "sites") |>
+  dm_add_fk("wikidata_links", "wikidata_linkable_id", "c14s") |>
+  dm_add_fk("wikidata_links", "wikidata_linkable_id", "c14_labs") |>
+  dm_add_fk("wikidata_links", "wikidata_linkable_id", "materials") |>
+  dm_add_fk("wikidata_links", "wikidata_linkable_id", "taxons") |>
+  dm_add_fk("versions", "whodunnit", "users", "id") |>
+  dm_add_fk("versions", "item_id", "c14_labs") |>
+  dm_add_fk("versions", "item_id", "c14s") |>
+  dm_add_fk("versions", "item_id", "contexts") |>
+  dm_add_fk("versions", "item_id", "materials") |>
+  dm_add_fk("versions", "item_id", "references") |>
+  dm_add_fk("versions", "item_id", "samples") |>
+  dm_add_fk("versions", "item_id", "site_names") |>
+  dm_add_fk("versions", "item_id", "site_types") |>
+  dm_add_fk("versions", "item_id", "sites") |>
+  dm_add_fk("versions", "item_id", "taxons") |>
+  dm_add_fk("versions", "item_id", "typos") |>
+  # TODO: Self-references (primarily persuadable models), or too messy?
+  # dm_add_fk("contexts", "superseded_by", "contexts") |>
+  dm_draw("BT", view_type = "title_only", column_types = TRUE) |>
+  DiagrammeRsvg::export_svg()
+
+# Dear god why
+xronos_dm_svg_file <- tempfile("xronos_dm_", fileext = ".svg")
+write(xronos_dm_svg, xronos_dm_svg_file)
+ggdraw() + draw_image(xronos_dm_svg_file)
+```
 
@@
 
-At the base of the XRONOS data model are sets of spatiotemporal coordinates or, as we call them, *chrons*.
+At the base of the XRONOS data model (@fig-data-model) are sets of spatiotemporal coordinates or, as we call them, *chrons*.
 In an archaeological context, we conceptualise a chron as an assertion linking human activity with a particular point in space and time.

### 2024-11-29 paper.qmd
Add fig-ui

@@
   * [x] FIGURE: data model
-  * [ ] FIGURE: GUI screenshots (data browser, c14 show view, version history, data curation)
+  * [x] FIGURE: GUI screenshots (data browser, c14 show view, version history, data curation)
   * [ ] TABLE: list of 'issues'
@@
 Once ingested, we apply a number of automated and semi-automated quality control processes to integrate new data into the existing database.
-Controlled vocabularies are used in a number of places in the data model (@FIG), and we use thesauri to automatically standardise these fields as much as possible.
+Controlled vocabularies are used in a number of places in the data model (@fig-data-model), and we use thesauri to automatically standardise these fields as much as possible.
 For example, the taxonomic description of samples is controlled using GBIF's backbone taxonomy [@GBIF2023], and we also use a thesaurus service provided by GBIF to automatically change variant or obsolete taxonomic names to the canonical version.
@@
 
-<!-- TODO: figures? -->
-<!-- TODO: add in more concrete examples -->
+::: {#fig-ui layout-nrow=2}
+![Data browser](figures/xronos-ui-data-browser.png){#fig-ui-data-browser}
+
+![Site record](figures/xronos-ui-show.png){#fig-ui-show}
+
+![Record change log](figures/xronos-ui-papertrail.png){#fig-ui-papertrail}
+
+![Curation interface](figures/xronos-ui-curation.png){#fig-ui-curation}
+
+Views in the XRONOS web graphical user interface
+:::
 
@@
 The graphical user interface (GUI) to XRONOS, accessed through a web browser (e.g.
-at <https://xronos.ch>), uses REST resources and actions as the building blocks for various interfaces through which users can browse, search, retrieve, and analyse chronometric data.
+at <https://xronos.ch>), uses REST resources and actions as the building blocks for various interfaces through which users can browse, search, retrieve, and analyse chronometric data (@fig-ui).
 Each action on each resource is represented by a page, though not all of these are publicly accessible.
-Pages representing REST resources directly are supplemented by a number of synthetic interfaces, for example the 'data browser' (<https://xronos.ch/data>), which facilitates more complex filtering, or the search interface (<https://xronos.ch/search>).
+The most import of these are the 'index' views, which list and summaries all instances of a resource and support filtering and sorting;
+and 'show' views (e.g.
+@fig-ui-show for a site), which give a more comprehensive information on an individual record along with visualisations, links to related records, external linked open data resources, and a log of changes made to the data since it was imported into XRONOS (@fig-ui-papertrail).
+Pages representing REST resources directly are supplemented by a number of synthetic interfaces, for example the 'data browser' (<https://xronos.ch/data>, @fig-ui-data-browser), which facilitates more complex filtering, or the search interface (<https://xronos.ch/search>).
 The GUI also includes several resources which are not part of XRONOS' scientific data model, for example documentation pages, user profiles and news articles; these are currently not exposed in the API.

### 2024-11-29 paper.qmd
Resolve @MartinHinz's suggestions and other TODOs

@@
 charcoal, charred seed) and, where applicable, taxonomic designation (e.g.
-TODO, TODO<!--sorry, not fully clear what goes here.
-MH.-->) of the organic material used for dating.
+*Quercus*, *Triticum dicoccum*) of the organic material used for dating.
 For typological dates, an ideal scenario would be for the sample to represent the particular object from which an inference was made (e.g. 'Natufian' might be inferred from 'lunate-type microlith').
@@
 Since this kind of information is rarely systematically recorded in our source databases, there are currently few actual records that make use of this feature of the data model.
-However, we hope it will provide a foundation for ... ? <!-- TODO Not fully clear what need to go here. MH. -->
+However, we hope it will provide a foundation for more nuanced chronological modelling in the future.
 
@@
 Beyond this, there is also a concordance between XRONOS entities and other specialised data infrastructures in archaeology, such as Perio.do [@RabinowitzEtAl2016, mapping to typological chrons] and gazetters like Pleiades [@Pleiades] or Vici [@Voorburg2012, mapping to site names].
-Moving beyond archaeology, Wikidata (<https://wikidata.org>) already includes many of the concepts represented in the XRONOS model (e.g. archaeological sites, Q<!-- TODO Eg Q39671? MH.-->).
+Moving beyond archaeology, Wikidata (<https://wikidata.org>) already includes many of the concepts represented in the XRONOS model (e.g.
+an 'archaeological site', <https://www.wikidata.org/wiki/Q839954>).
 Linking to Wikidata is especially useful because it dissimenates—and thereby preserves—the data compiled in XRONOS beyond the narrow field of archaeology.
@@
 Conversely, we encourage others to use XRONOS as a linked open data resource by providing stable URLs and a machine-readable interface for every record.
-Such usages could look like, for example, using a XRONOS URL as a canonical representation of a single radiocarbon date (e.g. <!-- TODO Not fully clear what need to go here. MH. -->).
-We also plan to implement an IntChron interface to XRONOS, to allow it to be indexed through that existing standard.<!-- TODO: also mention RDF? -->
+Such usages could look like, for example, using a XRONOS URL as a canonical representation of a single radiocarbon date (e.g. <https://xronos.ch/c14s/SMU-257> -> <https://xronos.ch/c14s/23410>).
+We also plan to implement additional view formats to facilitate this, such as IntChron-format JSON, to allow it to be indexed through the IntChron service, or RDF, which is widely used by many linked open data resources in the digital humanities.
 
@@
 
-In this paper we have outlined the conceptual and technical infrastructure developed during the initial phases of development on XRONOS in 2019 and 2021–2024.
-These include a generalised data model for radiocarbon and typological dates, extendable to other chronometric information, and associated contextual information;
+In this paper we have outlined the conceptual and technical infrastructure developed during the initial, SNF-funded phases of development on XRONOS in 2019 and 2021–2024.
+These include a generalised data model for radiocarbon and typological dates, extendible to other chronometric information, and associated contextual information;
 an R- and Ruby-based pipeline for continuous ingestion of data from a variety of sources; 
@@
 
-We intend to continue development of XRONOS into the foreseeable future.
+Development of XRONOS will continue within the framework of 'ESTER', an ERC-funded research project on estimation of prehistoric population development from large, multiproxy datasets.
 Our immediate development goals include the incorporation of dendrochronology into the database, further refinement of our data curation pipelines, and the public release of the editing interfaces.
-<!--In parallel, the planned ERC Consolidator Grant Project ESTER, focusing on the estimation of population development in prehistory, will over the next five years play a crucial role in supporting the continuation and further development of the XRONOS database, ensuring its sustainability and fostering interdisciplinary collaboration.
--->
+
 Looking beyond the near term, we have endeavoured to create a sustainable infrastructure that can be maintained by a wider scholarly commons – though we must acknowledge that this is a difficult problem, and one that is more organisational than technical.
@@
 But aside from these extreme scenarios, the long-term sustainability of XRONOS is contingent on the existence of a community of scholars that use and contribute back to it.
-In this we can only lead the horse to water; we cannot make it drink.<!-- TODO: too glib? -->
 

### 2024-12-02 paper.qmd
Add fig-c14-global

@@
 published: "Manuscript for submission to the *Journal of Computer Applications in Archaeology*"
-execute:
-  echo: false
 number-sections: true
@@
 #    embed-resources: true
+execute:
+  echo: false
+  cache: true
 ---
@@
 library("gt", warn.conflicts = FALSE)
+library("here")
 library("khroma")
@@
 library("patchwork", warn.conflicts = FALSE)
+library("purrr")
 library("readr")
@@
 library("sf")
+library("spatstat")
+library("stars")
 library("tidyr")
@@
 
-In addition, the temporal and geographic coverage of these resources is uneven [@AlcantaraPedrozainpress], systematically biased [@ClistEtAl2023], and duplicative of each other.
+In addition, the temporal and geographic coverage of these resources is uneven [@ChaputGajewski2016; @AlcantaraPedrozainpress], systematically biased [@ClistEtAl2023], and duplicative of each other.
 <!-- TODO: example, X databases for Europe, Y for the rest of the world? -->
@@
 
-<!-- TODO: Include CalPal?? -->
+```{r data-fig-c14-global, include=FALSE}
+c14_global <- read_tsv(here("data", "c14_global.tsv"))
+c14_global_sum <- read_tsv(here("data", "c14_global_sum.tsv"))
+
+ppp_window <- st_transform(st_union(gisco_get_countries()), "+proj=moll")
+
+c14_global_density <- c14_global |>
+  drop_na(longitude, latitude) |>
+  nest(.by = source) |>
+  mutate(
+    sf = map(data, st_as_sf, coords = c("longitude", "latitude"), crs = 4326),
+    sf = map(sf, st_transform, crs = "+proj=moll"),
+    sf = map(sf, st_filter, ppp_window), # no dates in the sea, please
+    n_sf = map_int(sf, nrow),
+    ppp = map(sf, as.ppp, W = as.owin(ppp_window)),
+    density = map(ppp, density, sigma = 50000, adjust = 10, eps = 50000),
+    density = map(density, st_as_stars)
+  )
+```
+
+```{r fig-c14-global, message=FALSE, warning=FALSE}
+#| fig-cap: Geographic and temporal (sum calibration) of georeferenced dates in XRONOS and other global radiocarbon compilations
+#| fig-height: 6
+plot_c14_density <- function(density, title, coastline = coast) {
+  ggplot() +
+    geom_stars(
+      aes(fill = ecdf(v)(v)), # transform density to quantiles like plot.stars()
+      data = density
+    ) +
+    geom_sf(data = coastline, fill = NA) +
+    scale_fill_stepsn(
+      name = "Density",
+      n.breaks = 11,
+      colours = colour("turku", reverse = TRUE)(11),
+      labels = c("Low", rep("", 7), "High"),
+      na.value = NA,
+      guide = guide_none()
+    ) +
+    labs(title = title) +
+    coord_sf(crs = "+proj=moll") +
+    theme_minimal_grid() +
+    theme(
+      plot.title = element_text(hjust = 0.5),
+      plot.margin = unit(c(11, 5.5, 0, 5.5), "pt")
+    )
+}
+
+fig_density <- map2(c14_global_density$density, c14_global_density$source, 
+                    plot_c14_density)
+names(fig_density) <- c14_global_density$source
+
+# From https://stackoverflow.com/questions/11053899/how-to-get-a-reversed-log10-scale-in-ggplot2
+transform_reverse_log <- function(base = exp(1)) {
+  trans <- function(x) -log(x, base)
+  inv <- function(x) base^(-x)
+  scales::trans_new(paste0("reverselog-", format(base)), trans, inv, 
+                    scales::log_breaks(base = base), 
+                    domain = c(1e-100, Inf))
+}
+
+plot_c14_sum <- function(data, n) {
+  ggplot(data, aes(age, pdens)) +
+    geom_area(fill = "#3C3B39") + # 'XRONOS grey'
+    scale_x_continuous(name = NULL, breaks = c(50000, 5000, 500),
+                       labels = c("50000", "5000", "500 BP"),
+                       transform = transform_reverse_log()) +
+    scale_y_continuous(name = NULL, labels = NULL) +
+    labs(caption = paste0("N=", n)) +
+    theme_minimal_vgrid() + 
+    theme(
+      axis.line.y = element_blank(),
+      axis.ticks.y = element_blank(),
+      plot.margin = unit(c(0, 5.5, 11, 5.5), "pt")
+    )
+}
+
+fig_sum <- c14_global_sum |>
+  nest(.by = source) |>
+  pull(data) |>
+  map2(c14_global |> nest(.by = source) |> pull(data) |> map(nrow), plot_c14_sum)
+names(fig_sum) <- unique(c14_global_sum$source)
+
+# Lay out with patchwork
+fig_density[["c14bazAAR"]] + fig_density[["IntChron"]] +
+  fig_sum[["c14bazAAR"]] + fig_sum[["IntChron"]] +
+  fig_density[["p3k14c"]] + fig_density[["XRONOS"]] +
+  fig_sum[["p3k14c"]] + fig_sum[["XRONOS"]] +
+  plot_layout(ncol = 2, heights = rep(c(5, 1), 2))
+
+# Alt. plot by country
+# c14_global |>
+#   count(country, source) |>
+#   left_join(gisco_get_countries(), by = c("country" = "CNTR_ID")) |>
+#   mutate(n = replace_na(n, 0)) |>
+#   st_as_sf() |>
+#   ggplot() +
+#   facet_wrap(vars(source)) +
+#   geom_sf(data = gisco_get_countries(), fill = "white", colour = NA) +
+#   geom_sf(aes(fill = n), colour = NA) +
+#   geom_sf(data = coast, fill = NA) +
+#   scale_fill_turku(reverse = TRUE, labels = scales::comma) +
+#   labs(fill = "Dates") +
+#   coord_sf(crs = "+proj=moll") +
+#   theme_minimal_grid()
+```
 
@@
 Because c14bazAAR downloads data from its original source repositories, rather than mirroring it, it only includes resources that have been published in a fully open and machine-accessible format.
-Despite this limitation, it has global coverage and is still the largest collection of radiocarbon dates available (@FIG).
+Despite this limitation, it has global coverage and a large number of dates (@fig-c14-global), and was therefore our starting point for data collection for XRONOS.
 
@@
 
-As of writing, c14bazAAR includes <!-- TODO: N --> radiocarbon dates with unique laboratory identifiers, IntChron <!-- TODO: N --> (excluding those from non-archaeological contexts), and p3k14c <!-- TODO: N -->.
-The geographic distribution of dates from each is similar (@FIG), reflecting the large degree in overlap between the sources of c14bazAAR and p3k14c in particular.
-IntChron, which in practice is currently only used to publish dates associated with the Oxford Radiocarbon Lab, has dates from more diverse contexts, but is much smaller.
+As of writing, c14bazAAR has `r format(nrow(filter(c14_global, source == "c14bazAAR")), big.mark = ",")` radiocarbon dates with unique laboratory identifiers (excluding those sourced from p3k14c and XRONOS), IntChron has `r format(nrow(filter(c14_global, source == "IntChron")), big.mark = ",")` (excluding those from non-archaeological contexts), and p3k14c has `r format(nrow(filter(c14_global, source == "c14bazAAR")), big.mark = ",")`.
+The geographic distribution of dates from each is similar (@fig-c14-global), reflecting the large degree in overlap between the sources of each compilation.
+IntChron, which in practice is currently only used to publish dates associated with the Oxford Radiocarbon Lab, has dates from more diverse contexts, but less overall.
 
@@
 
-# Dear god why
+# Dear god why is this necessary
 xronos_dm_svg_file <- tempfile("xronos_dm_", fileext = ".svg")

### 2024-12-02 data/data_c14_global.R

@@ -1,0 +1,127 @@
+# data_c14_global.R
+# Summarise coverage of 'global' radiocarbon databases (c14bazAAR, IntChron,
+# p3k14c, XRONOS)
+library(c14)       # https://github.com/joeroe/c14
+library(c14bazAAR) # https://github.com/rOpenSci/c14bazAAR
+library(countrycode)
+library(dplyr, warn.conflicts = FALSE)
+library(future)
+library(here)
+library(purrr)
+library(readr)
+library(rintchron) # https://github.com/joeroe/rintchron
+library(tidyr)
+library(xronos)    # https://github.com/xronos-ch/xronos.R
+
+plan(multisession)
+
+# c14bazAAR (excl. p3k14c and XRONOS)
+c14bazAAR <- get_c14data("all") |>
+  filter(sourcedb != "p3k14c")
+
+# IntChron
+egyptdb <- intchron("egyptdb")
+nrcf <- intchron("nrcf", tabulate = FALSE)
+nrcf <- intchron_tabulate(nrcf[-c(778, 781, 783, 791, 792)]) # bad records
+oxa <- intchron("oxa")
+sadb <- intchron("sadb")
+intchron <- bind_rows(egyptdb, nrcf, oxa, sadb)
+
+# p3k14c
+p3k14c <- get_c14data("p3k14c")
+
+# XRONOS
+xronos <- chron_data(.everything = TRUE)
+
+# Merge basic information from each database
+c14_global <- bind_rows(
+  transmute(
+    c14bazAAR,  source = "c14bazAAR", 
+    lab_id = labnr, c14_age = c14age, c14_error = c14std, country, 
+    latitude = lat, longitude = lon
+  ),
+  transmute(
+    intchron, source = "IntChron",
+    lab_id = labcode, c14_age = r_date, c14_error = r_date_sigma, 
+    country = record_country, latitude = record_latitude, 
+    longitude = record_longitude
+  ),
+  transmute(
+    p3k14c, source = "p3k14c",
+    lab_id = labnr, c14_age = c14age, c14_error = c14std, country, 
+    latitude = lat, longitude = lon
+  ),
+  transmute(
+    xronos, source = "XRONOS",
+    lab_id = labnr, c14_age = bp, c14_error = std, country, 
+    latitude = as.numeric(lat), longitude = as.numeric(lng)
+  )
+)
+
[…]

### 2024-12-02 paper.qmd
Update TODO list

@@
 * [x] First draft
-* [ ] Complete references:
-  * [x] Introduction
-  * [ ] State of the art
-  * [ ] Concept
-  * [ ] Implementation
-  * [ ] Evaluation
-  * [ ] Appendix table
-* [ ] Figures and tables
-  * [x] FIGURE: graph of number of datasets through time
-  * [x] FIGURE: map of number of dates in datasets
-  * [ ] FIGURE: distribution of radiocarbon dates in c14bazAAR / IntChron / p3k14c / XRONOS
-  * [x] FIGURE: data model
-  * [x] FIGURE: GUI screenshots (data browser, c14 show view, version history, data curation)
-  * [ ] TABLE: list of 'issues'
-  * [ ] APPENDIX: table of datasets
+* [x] References
+* [x] Figures and tables
+* [ ] SM1 (data/c14_datasets.tsv)
+  * [ ] References
+  * [ ] Nice PDF version
 * [ ] Other TODOs
@@
 A complete and up-to-date list of acknowledgements for the project can be found at <https://xronos.ch/about/acknowledgements>.
+<!-- TODO: peer reviewers -->
 

### 2024-12-02 paper.qmd
Resolve several TODOs

@@
 
-```{r setup}
+```{r setup, cache=FALSE}
 #| include: false
@@
 library("dm")
+library("english")
 library("giscoR")
@@
 library("stars")
+library("stringr")
 library("tidyr")
@@
 Over the last two decades, archaeologists have answered this call by publishing an increasing number of compilations of dates from archaeological contexts as open data.
-These efforts have facilitated major reevaluations of previously-established chronologies [e.g.
-@HighamEtAl2014; @LoftusEtAl2019; @PratesEtAl2020; @KatsianisEtAl2020], important new insights into past processes [@RirisEtAl2024], and the development of novel ways of using chronological data [@Crema2022].
+These efforts have facilitated re-evaluations of chronologies themselves [e.g.
+@HighamEtAl2014; @LoftusEtAl2019; @PratesEtAl2020; @KatsianisEtAl2020] but also the development of novel ways of using chronological data [@MoodyEtAl2021; @Crema2022; @RirisEtAl2024].
 <!-- TODO: Would welcome more citations here -->
@@
 Though an explicit emphasis on 'open data' is a relatively recent phenomenon in archaeology [@Lake2012], the open publication of compiled radiocarbon dates has a substantial prehistory.
-<!-- TODO: check and update this about date lists... -->
 Arnold and Libby [-@ArnoldLibby1951] initiated the tradition of regularly publishing all the dates they had obtained.
@@
 'date lists' in printed periodicals and gazetteers were excluded.
-A full list of the datasets we identified is presented in appendix XXXX<!-- TODO -->.
+A full list of the datasets we identified can be found in the supplementary materials.
 
@@
 The shift from online databases towards more static but more preservable open data products is welcome, given how many databases from the first generation have subsequently ceased to be accessible.
-<!-- TODO: can this be quantified or visualised?
--MH -->
 Version-controlled repositories are particular well-suited to data compilation projects because they allow for continued updates whilst still providing snapshot 'releases' that are citeable and can be archived in long-term repositories.
@@
 
-In addition, the temporal and geographic coverage of these resources is uneven [@ChaputGajewski2016; @AlcantaraPedrozainpress], systematically biased [@ClistEtAl2023], and duplicative of each other.
-<!-- TODO: example, X databases for Europe, Y for the rest of the world? -->
-By our count, <!-- TODO: X --> of the databases are not 'open' according to the Open Knowledge Foundation's definition of data openness ["Open data and content can be freely used, modified, and shared by anyone for any purpose", @OpenKnowledgeFoundation], which both limits the access to and reuse potential of these datasets.
-<!-- TODO: X --> are not currently available in readily machine-readable formats (e.g. plain text or database files rather than PDFs or hypertext).
+The temporal and geographic coverage of these resources is uneven [@ChaputGajewski2016; @AlcantaraPedrozainpress], systematically biased [@ClistEtAl2023], and duplicative.
+For example, we identified `r english(sum(str_detect(c14_datasets$m49_region, "Western Europe"), na.rm = TRUE))` different databases covering Western Europe but none covering South Asia.
+The quality and accessibility of published compilations is also variable.
+`r str_to_sentence(english(sum(c14_datasets$open, na.rm = TRUE)))` of the `r english(nrow(c14_datasets))` resources we reviewed are not 'open' according to the Open Knowledge Foundation's definition of data openness ["Open data and content can be freely used, modified, and shared by anyone for any purpose", @OpenKnowledgeFoundation], which both limits the access to and reuse potential of these datasets.
+And even of these, many are not currently available in readily machine-readable formats (e.g. plain text or database files rather than PDFs or hypertext).
 
@@
 Like any archaeological date, a typological date is derived from a physical sample – the object or set of object from which a chronological estimate was derived.
-Typological dates on one class of object may well clash with other classes of object, or for that matter with scientific dates, but without this kind of metadata such inconsistencies are difficult to resolve.
-<!-- TODO: clarify - MH -->
+Typological dates on one class of object may well clash with other classes of object, or for that matter with scientific dates – does one trust the date on pottery, the date on architecture, or the radiocarbon date?
+Without additional metadata on e.g. who made the typological determination or what the radiocarbon date was obtained on, such inconsistencies are difficult to resolve 
 Similarly the absolute date range corresponding to a typological determination (e.g.
@@
 XRONOS is our answer to Kintigh's call [@Kintigh2006] for digital infrastructures that don't just provide access to chronological data but enables researchers to "archive, access, integrate, and mine disparate data sets".
-It parallels and draws inspiration from several similar initiatives within and outwith archaeology, such as <!-- TODO examples
-Examples:
-https://link.springer.com/article/10.1007/s10816-018-9399-6
-https://link.springer.com/article/10.1007/s10816-010-9098-4
-https://www.cambridge.org/core/journals/latin-american-antiquity/article/public-database-of-archaeological-resources-on-easter-island-rapa-nui-using-google-earth/F30677F6FC99B4C2BC762BCC98FD0966
-https://www.cambridge.org/core/journals/advances-in-archaeological-practice/article/data-integration-in-the-service-of-synthetic-research/E297C57441239AB35053F987AA758EB9
-https://www.cambridge.org/core/journals/advances-in-archaeological-practice/article/tdar/FCE1056A210A452E5FE65560763D4CC7
-https://www.cambridge.org/core/journals/advances-in-archaeological-practice/article/googlebased-freeware-solution-for-archaeological-field-survey-and-onsite-artifact-analysis/7914F673D1E1CAD96E5270E794CDFFD9
-https://link.springer.com/article/10.1007/s10816-015-9272-9
-https://link.springer.com/article/10.1007/s10816-015-9251-1
-https://link.springer.com/article/10.1007/s10816-015-9240-4
-https://anatomypubs.onlinelibrary.wiley.com/doi/10.1002/ar.23130
-https://academic.oup.com/dsh/article/29/3/349/2938128
-https://www.sciencedirect.com/science/article/pii/S0167739X13000678?via%3Dihub
-
--->.
+It complements several similar open data infrastructures within and outwith archaeology, such as SEAD for environmental archaeology [@Buckland2014], IMPACT for mummified human remains [@NelsonWade2015], Neotoma for palaeoecological data [@WilliamsEtAl2018], IsoArcH for stable isotope data [@PlompEtAl2022], and the 'Big Interdisciplinary Archaeological Database' (BIAD), an ambitious new initiative to combine many of these individual domains, including chronology [@ReiterEtAl2024].
 To improve upon existing global syntheses of radiocarbon dates (see @sec-global-compilations), we aimed to develop a living infrastructure that both continually collected data from diverse sources and presented a seamless single database to the user.
@@
   dm_add_fk("versions", "item_id", "typos") |>
-  # TODO: Self-references (primarily persuadable models), or too messy?
+  # Exclude self-references (primarily supersedable models) for readability, e.g.
   # dm_add_fk("contexts", "superseded_by", "contexts") |>
@@
 
-<!-- TODO: shorten -->
+<!-- TODO: shorten? -->
 
@@
 By passing a query as HTTP GET parameters to the index action of a resource, the list returned the user is modified to only include records that match that query.
-For example, <https://xronos.ch/sites?site[country_code]=CH> (the part of the URL after the `?` character encodes the SQL WHERE clause `country = 'CH'` as a GET parameter) lists sites in Switzerland <!-- TODO: check this actually works -->.
+For example, <https://xronos.ch/sites?site[country_code]=CH> (the part of the URL after the `?` character encodes the SQL WHERE clause `country = 'CH'` as a GET parameter) lists sites in Switzerland.
 More complex queries can be executed using nested parameters.
-For example, <https://xronos.ch/c14s?c14[sample][material][name]=charcoal> (encoding that the `c14` table should be joined to the `material` table via `sample`, followed by the WHERE clause `material.name = 'charcoal'`) lists radiocarbon dates obtained from charcoal samples <!-- TODO: check that this actually works...
--->.
+For example, <https://xronos.ch/c14s?sample[material][name]=charcoal> (encoding that the `c14` table should be joined to the `material` table via `sample`, followed by the WHERE clause `material.name = 'charcoal'`) lists radiocarbon dates obtained from charcoal samples.
 Uniquely, index actions can also respond with the result in a tabular data format (i.e. `.csv`).

### 2024-12-02 paper.qmd
Fix N for p3k14c

@@
 
-As of writing, c14bazAAR has `r format(nrow(filter(c14_global, source == "c14bazAAR")), big.mark = ",")` radiocarbon dates with unique laboratory identifiers (excluding those sourced from p3k14c and XRONOS), IntChron has `r format(nrow(filter(c14_global, source == "IntChron")), big.mark = ",")` (excluding those from non-archaeological contexts), and p3k14c has `r format(nrow(filter(c14_global, source == "c14bazAAR")), big.mark = ",")`.
+As of writing, c14bazAAR has `r format(nrow(filter(c14_global, source == "c14bazAAR")), big.mark = ",")` radiocarbon dates with unique laboratory identifiers (excluding those sourced from p3k14c and XRONOS), IntChron has `r format(nrow(filter(c14_global, source == "IntChron")), big.mark = ",")` (excluding those from non-archaeological contexts), and p3k14c has `r format(nrow(filter(c14_global, source == "p3k14c")), big.mark = ",")`.
 The geographic distribution of dates from each is similar (@fig-c14-global), reflecting the large degree in overlap between the sources of each compilation.

### 2024-12-04 paper.qmd
Clarify funding

@@
 
-Following a short pilot project in 2019, the first phase of development of XRONOS was completed in 2021–2024, supported by a project grant from the Swiss National Science Foundation.
+Following a short pilot project in 2019, the first phase of development of XRONOS was completed in 2021–2024.
 The web interface (<https://xronos.ch>) has been publicly accessible since July 2021.
@@
 
-This work was funded by the Swiss National Science Foundation ([SNSF Project #198152](https://data.snf.ch/grants/grant/198153)) and the University of Bern (Initiator Grant, Caroline Heitz, Project 'Time and Temporality').
+This work was funded by the Swiss National Science Foundation ([SNSF Project #198152](https://data.snf.ch/grants/grant/198153)) and the University of Bern (UniBE Initiator Grant 2019, Caroline Heitz, Project 'Time and Temporality').

### 2024-12-04 paper.qmd
Add references

@@
   XRONOS (<https://xronos.ch>) is an open data infrastructure for the backbone of the archaeological record – chronology.
-It aims provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world.
+It provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world.
 By collating a large number of existing regional and global compilations of dates, XRONOS offers the most comprehensive radiocarbon database yet published, with over 350,000 radiocarbon and 75,000 site records.
 It also provides a foundation for expanding the systematic collection of chronometric information beyond radiocarbon, with support for typological and dendrochronological dates and a generalisable data model that can be adapted to other methods of absolute dating.
-Automated and semi-automated quality control processes ensure that data from diverse sources is continuously integrated and standardised, making it significantly easier to filter data of interest and reducing the need for manual data cleaning by end users.
+Automated and semi-automated quality control processes ensure that data from diverse sources is continuously integrated and standardised, making it easier to find information of interest and reducing the need for manual data cleaning by end users.
 In this paper we describe the concept and implementation of XRONOS in relation to the state of the art in chronometric data-sharing, and evaluate its potential as a general-purpose open repository and curation platform for archaeological chronology.
@@
 <!--
-* [x] First draft
-* [x] References
-* [x] Figures and tables
 * [ ] SM1 (data/c14_datasets.tsv)
@@
 Chronology is the backbone of the archaeological record.
-As a necessary prerequisite to understanding the context of any past event or process [@Lucas2004], it is has unsurprisingly been at the forefront of methodological development in archaeology for as long as the discipline has existed: from putting finds and events in sequence [@Rowley-Conwy2007] to using scientific methods to place them on an absolute timescale [@Libby1955] and increasingly sophisticated statistical methods to refine them [@BronkRamsey2009].
+As a necessary prerequisite to understanding the context of any past event or process [@Lucas2004], it is has unsurprisingly been at the forefront of methodological development in archaeology for as long as the discipline has existed: from putting finds and events in sequence [@Thomsen1836; @Worsaae1843; @Petrie1899; @Ford1962; @Harris1979] to an increasingly wide array of scientific methods that place them on an absolute timescale [@Douglass1929; @DanielsEtAl1953; @Libby1955; @EverndenEtAl1965; @BadaHelfman1975] and an increasingly sophisticated set of statistical tools to build them into chronologies [@Suess1967; @BuckEtAl1991; @Mischka2004; @LevyEtAl2021; @Crema2024].
 If archaeology is to be an open science [@Lake2012], it is therefore critical that effective open access to chronological information be placed front and centre.
@@
 These efforts have facilitated re-evaluations of chronologies themselves [e.g.
-@HighamEtAl2014; @LoftusEtAl2019; @PratesEtAl2020; @KatsianisEtAl2020] but also the development of novel ways of using chronological data [@MoodyEtAl2021; @Crema2022; @RirisEtAl2024].
-<!-- TODO: Would welcome more citations here -->
+@HighamEtAl2014; @LoftusEtAl2019; @PratesEtAl2020; @KatsianisEtAl2020] but also the development of novel ways of using chronological data [e.g.
+@Grove2011; @Crema2022; @CremaEtAl2024; @RirisEtAl2024; @MaromWolkowski2024].
 The focus has been overwhelmingly on radiocarbon dating and most compilations focus on a single region and/or period.
@@
 
-Thus, even though radiocarbon data comes from a relatively limited number of sources [one of the 172 active labs, @RadiocarbonLabList] and has relatively standardised reporting conventions [@Millard2014; @Bayliss2015], in practice the only way to produce aggregated datasets in recent decades has been to manually search through relevant literature for dates reported by the submitter of the sample.
-This already laborious process is further hampered by a significant inconsistency in how much authors adhere to measurement reporting conventions, a lack of conventions on the reporting of *contextual* information, and weak or nonexistent disciplinary norms regarding the responsibility to publish results openly in a timely fashion.
+Thus, even though radiocarbon data comes from a relatively limited number of sources [some 172 active labs, @RadiocarbonLabList] and has relatively standardised reporting conventions [@Millard2014; @Bayliss2015], in practice the only way to produce aggregated datasets in recent decades has been to manually search through relevant literature for dates reported by the submitter of the sample.
+This already laborious process is further hampered by a significant inconsistency in how much authors adhere to measurement reporting conventions, a lack of conventions on the reporting of *contextual* information, weak or nonexistent disciplinary norms regarding the responsibility to publish results openly in a timely fashion, and a range of other issues affecting data reuse [@MoodyEtAl2021].
 
@@
 XRONOS is our answer to Kintigh's call [@Kintigh2006] for digital infrastructures that don't just provide access to chronological data but enables researchers to "archive, access, integrate, and mine disparate data sets".
-It complements several similar open data infrastructures within and outwith archaeology, such as SEAD for environmental archaeology [@Buckland2014], IMPACT for mummified human remains [@NelsonWade2015], Neotoma for palaeoecological data [@WilliamsEtAl2018], IsoArcH for stable isotope data [@PlompEtAl2022], and the 'Big Interdisciplinary Archaeological Database' (BIAD), an ambitious new initiative to combine many of these individual domains, including chronology [@ReiterEtAl2024].
+It complements several similar open data infrastructures within and outwith archaeology, such as the Strategic Environmental Archaeology Database [SEAD, @Buckland2014], IMPACT for mummified human remains [@NelsonWade2015], Neotoma for palaeoecological data [@WilliamsEtAl2018], IsoArcH for stable isotope data [@PlompEtAl2022], the International Soil Radiocarbon Database [ISRaD, @LawrenceEtAl2020], and the 'Big Interdisciplinary Archaeological Database' (BIAD), an ambitious new initiative to combine many of these individual domains, including chronology [@ReiterEtAl2024].
 To improve upon existing global syntheses of radiocarbon dates (see @sec-global-compilations), we aimed to develop a living infrastructure that both continually collected data from diverse sources and presented a seamless single database to the user.

### 2024-12-04 paper.qmd
Copyedit and resolve remaining pre-submission TODOs

@@
 
-<!--
-* [ ] SM1 (data/c14_datasets.tsv)
-  * [ ] References
-  * [ ] Nice PDF version
-* [ ] Other TODOs
-* [ ] Copyedit:
-  * [ ] Introduction
-  * [ ] State of the art
-  * [ ] Concept
-  * [ ] Implementation
-  * [ ] Evaluation
-  -->
-
 # Introduction
@@
 @HighamEtAl2014; @LoftusEtAl2019; @PratesEtAl2020; @KatsianisEtAl2020] but also the development of novel ways of using chronological data [e.g.
-@Grove2011; @Crema2022; @CremaEtAl2024; @RirisEtAl2024; @MaromWolkowski2024].
+@Grove2011; @SilvaSteele2014; @Crema2022; @CremaEtAl2024; @RirisEtAl2024; @MaromWolkowski2024].
 The focus has been overwhelmingly on radiocarbon dating and most compilations focus on a single region and/or period.
@@
 The publication of many overlapping, non-standardised and mostly static open data resources means that it is still difficult to obtain reliable and up-to-date chronological datasets, especially for applications that crosscut convential geographic and temporal domains of research.
-Initiatives towards synthesis this have improved this situation, but the goal of a global dataset that is both comprehensive and up-to-date remains elusive.
+Initiatives towards synthesis have improved this situation, but the goal of a global dataset that is both comprehensive and up-to-date remains elusive.
 
@@
 In the remainder of this paper, we describe the concept and implementation of XRONOS in relation to the state of the art in open chronometric data in archaeology, and evaluate our progress in achieving these goals as of writing.
-Since we envisage both XRONOS as a dataset and XRONOS as software to be continually developing resources, the description here should be read as a 'snapshot' rather than the final state of the project.
+Since we envisage both XRONOS as a dataset and XRONOS as software to be continually developing resources, the description here should be read as a 'snapshot' of the project as of writing rather than its final state.
 
@@
 
-Though an explicit emphasis on 'open data' is a relatively recent phenomenon in archaeology [@Lake2012], the open publication of compiled radiocarbon dates has a substantial prehistory.
-Arnold and Libby [-@ArnoldLibby1951] initiated the tradition of regularly publishing all the dates they had obtained.
-This practice was subsequently continued as radiocarbon laboratories periodically shared and compiled their own 'date lists', published mainly in the journals *Radiocarbon* and *Archaeometry*.
-However, as the number of labs and volume of radiocarbon dates being produced grew, this paper-based format became impractical and mostly disappeared [with exceptions, e.g.
-@NdeyeEtAl2022] without being replaced by another form of systematic data-sharing or dissemination [@BronkRamseyEtAl2019].
-Additionally, because the date lists were sourced from radiocarbon laboratories directly—not those who collected the sample—they typically included only very limited contextual information.
-On the eve of the AMS revolution there was an effort to create a computerised 'International Radiocarbon Database' [@Kra1988]—already by 1989 described as a "much needed, long overdue enterprise" [@Kra1989]—but it never came to fruition.
+Though an *explicit* emphasis on 'open data' is a relatively recent phenomenon in archaeology [@Lake2012], the open publication of compiled radiocarbon dates has a substantial prehistory.
+Arnold and Libby [-@ArnoldLibby1951] initiated the tradition of regularly publishing 'data lists', a practice was subsequently continued by radiocarbon laboratories as supplements to journals such as *Radiocarbon* and *Archaeometry*.
+However, as the number of labs and volume of radiocarbon dates being produced grew, this paper-based format became impractical and mostly disappeared [@BronkRamseyEtAl2019; c.f.
+e.g.
+@NdeyeEtAl2022], without being replaced by another form of systematic data-sharing or dissemination.
+Additionally, because date lists were sourced from radiocarbon laboratories directly—not from those who collected the sample—they typically included only very limited contextual information.
+On the eve of the AMS revolution there was an effort to create a computerised 'International Radiocarbon Database' [@Kra1988]—already by 1989 described as a "much needed, long overdue enterprise" [@Kra1989, p.
+1067]—but it never came to fruition.
 
-Thus, even though radiocarbon data comes from a relatively limited number of sources [some 172 active labs, @RadiocarbonLabList] and has relatively standardised reporting conventions [@Millard2014; @Bayliss2015], in practice the only way to produce aggregated datasets in recent decades has been to manually search through relevant literature for dates reported by the submitter of the sample.
-This already laborious process is further hampered by a significant inconsistency in how much authors adhere to measurement reporting conventions, a lack of conventions on the reporting of *contextual* information, weak or nonexistent disciplinary norms regarding the responsibility to publish results openly in a timely fashion, and a range of other issues affecting data reuse [@MoodyEtAl2021].
+Thus, even though radiocarbon data comes from a relatively limited number of sources [some 172 active labs, @RadiocarbonLabList] and has relatively standardised reporting conventions [@Millard2014; @Bayliss2015], in practice the only way to produce aggregated datasets in recent decades has been to manually search through relevant literature for dates reported secondarily by the submitter of the sample.
+This already laborious process is further hampered by a significant inconsistency in how much authors adhere to reporting conventions for measurements and sample metadata, a lack of conventions on the reporting of *contextual* information, weak or nonexistent disciplinary norms regarding the responsibility to publish results openly in a timely fashion, and a range of other issues affecting data reuse [@MoodyEtAl2021].
 
@@
 Our review of the literature identified `r n_c14_datasets` published since 1994.
-This is almost certainly an undercount, because our firsthand knowledge of regional literature was limited to Europe and West Asia and many resources only ever existed in 'grey' formats (e.g.
+This is almost certainly an undercount, because our firsthand knowledge of regional literature is limited to Europe and West Asia and many resources only ever existed in 'grey' formats (e.g.
 websites that were not indexed and no longer exist).
@@
 ```{r fig-c14-datasets-time}
-#| fig-cap: Cumulative number of radiocarbon compilations published since 1995 according to our survey (see Supplementary Material).
+#| fig-cap: Cumulative number of radiocarbon compilations published since 1995 according to our survey (see supplementary materials)
 c14_datasets |>
@@
   labs(x = NULL, y = "Datasets*",
-       caption = glue("* Excludes {n_c14_datasets_sans_publication_year} datasets with an unknown year of publication")) +
+       caption = glue("* Excludes {english(n_c14_datasets_sans_publication_year)} datasets with an unknown year of publication")) +
   theme_half_open()
@@
 
-The number of available compilations has steadily increased since around 1995 (@fig-c14-datasets-time).
-The first generation came around the turn of the century and consisted mostly of online databases with a web frontend.
-These included some databases operated by radiocarbon labs, for example the Oxford Radiocarbon Lab (ORAU) and the Belgian Royal Institute for Cultural Heritage (KIK-IRPA), and essentially represented a continuation of their date lists in a digital format.
-The majority, however, were compiled from the literature and individual researchers interested in a particular region.
+The number of available compilations has increased exponentially since around 1995 (@fig-c14-datasets-time).
+The first generation came around the turn of the century and consists mostly of online databases with a web frontend.
+These include some databases operated by radiocarbon labs, for example the Oxford Radiocarbon Lab (ORAU) and the Belgian Royal Institute for Cultural Heritage (KIK-IRPA), and essentially represent a continuation of their date lists in a digital format.
+The majority, however, were compiled from the literature by individual researchers interested in a particular region and/or period.
 Notable early examples include ANDES 14C in 1994 [Central Andes, @MichczynskiEtAl1995], CARD [Canada, @GajewskiEtAl2011] and RADON [Europe, @Raetzel-Fabian1999] in 1999, and CANEW in 2001 [Near East, @ReingruberThissen2005].
@@
 The geographic coverage of regional radiocarbon compilations is markedly uneven (@fig-c14-datasets-map).
-Europe and, especially, North America are very over-represented [as has already been observed by @ChaputGajewski2016 and others].
+Europe and, especially, North America are over-represented [@ChaputGajewski2016; @AlcantaraPedrozainpress].
 South America, West Asia, and East Asia are reasonably well-covered, but there practically no systematically compiled dates from East or West Africa, Central or South Asia, or Mainland Southeast Asia.
 This is probably explained in part by a lower volume of archaeological research and access to radiocarbon dating in these regions, but a lack of attention in compilation work must also be a factor.
-For example, radiocarbon dating has been an established part of Indian archaeology since at least 1961 [@KusumgarEtAl1963], but we have not able to locate a single systematic compilation of dates from the Indian subcontinent.
+For example, radiocarbon dating has been an established part of Indian archaeology since at least 1961 [@KusumgarEtAl1963], but we have not able to locate a single systematic compilation of dates from South Asia.^[We would be very happy to be corrected on this point.]
 
@@
 
-Datasets based on literature review also become out of date almost immediately upon publication, due the the constant stream of new dates.
+Datasets based on literature review also become out of date almost immediately upon publication, due the the constant production of new dates.
 Unfortunately this applies to many databases that are in theory continuously updated, as it is common to see them become unmaintained and or unexpectedly become unavailable.
-Of the `r n_c14_datasets` published datasets we identified, `r n_c14_datasets_maintained` were intended to be continuously updated, but only `r n_c14_datasets_recently_updated` had received updates in the last two years.
-The average 'lifespan' of a dataset from its publication to its last update is `r mean_c14_datasets_lifespan` years.
+Of the `r n_c14_datasets` published datasets we identified, `r n_c14_datasets_maintained` were intended to be continuously updated, but only `r n_c14_datasets_recently_updated` have received updates in the last two years.
+The average 'lifespan' of a dataset from its publication to its last update is around `r english(round(mean_c14_datasets_lifespan))` years.
 Most radiocarbon datasets we reviewed were compiled with a specific goal in mind (e.g.
@@
 
-The temporal and geographic coverage of these resources is uneven [@ChaputGajewski2016; @AlcantaraPedrozainpress], systematically biased [@ClistEtAl2023], and duplicative.
-For example, we identified `r english(sum(str_detect(c14_datasets$m49_region, "Western Europe"), na.rm = TRUE))` different databases covering Western Europe but none covering South Asia.
+Other outstanding problems with existing compilations include various systematic biases in data collection [@ClistEtAl2023] and a large degree of overlap and duplication between individual databases.
+For example, we identified `r english(sum(str_detect(c14_datasets$m49_region, "Western Europe"), na.rm = TRUE))` different resources covering Western Europe but none covering South Asia.
 The quality and accessibility of published compilations is also variable.
-`r str_to_sentence(english(sum(c14_datasets$open, na.rm = TRUE)))` of the `r english(nrow(c14_datasets))` resources we reviewed are not 'open' according to the Open Knowledge Foundation's definition of data openness ["Open data and content can be freely used, modified, and shared by anyone for any purpose", @OpenKnowledgeFoundation], which both limits the access to and reuse potential of these datasets.
+`r sum(c14_datasets$open, na.rm = TRUE)` of the `r nrow(c14_datasets)` resources we reviewed are not 'open' according to the Open Knowledge Foundation's definition of data openness ["Open data and content can be freely used, modified, and shared by anyone for any purpose", @OpenKnowledgeFoundation], which both limits the access to and reuse potential of these datasets.
 And even of these, many are not currently available in readily machine-readable formats (e.g. plain text or database files rather than PDFs or hypertext).
@@
 Although the core elements of a radiocarbon date—laboratory identifier, radiocarbon age, measurement error—are more or less standardised, there is no such consistency in contextual information on the sample or site.
-Such contextual information is important not just for the interpretation of dates, but for 'chronometric hygiene' [filtering out unreliable dates based on sample information, see e.g.
-@PettittEtAl2003] and for correcting for known systematic errors such as the marine reservoir effect [@AlvesEtAl2018].
+Such contextual information is important not just for the interpretation of dates, but for filtering out unreliable dates based on sample information ['chronometric hygiene' sensu @PettittEtAl2003] and for correcting for known systematic errors such as the marine reservoir effect [@AlvesEtAl2018].
 Most published datasets incorporate all or part of earlier compilations, meaning duplicate records are also very common, but deduplicating them is not a trivial problem due to format variations (see @sec-implementation-data).
-These issues are by no means impossible to overcome, but adds a significant amount of data-cleaning effort to a process that is otherwise very amenable to standardisation.
+These issues are by no means impossible to overcome, but adds a significant amount of data-cleaning effort to a process that would otherwise be very amenable to standardisation.
 
@@
 The profusion of radiocarbon compilations over the last decade has naturally prompted many to think globally.
-Three existing initiatives are particularly similar to XRONOS in their aims: c14bazAAR, IntChron, and p3k14c.
+Three existing initiatives in particular share similar aims to XRONOS (at least as far as radiocarbon is concerned): c14bazAAR, IntChron, and p3k14c.
 
@@
 
-Another indexical approach is the IntChron project [@BronkRamseyEtAl2019], which exposes data from multiple sources and exposes them with a common JSON-based web interface.
+Another indexical approach is taken by the IntChron project [@BronkRamseyEtAl2019], which exposes data from multiple sources and exposes them with a common JSON-based web interface.
 The IntChron specification is open, meaning that radiocarbon labs or compilation projects can implement it independently and thereby allow end users to access their data through a common interface (though to our knowledge it has so far only been adopted by databases associated with the Oxford Radiocarbon Lab).
@@
 
-As of writing, c14bazAAR has `r format(nrow(filter(c14_global, source == "c14bazAAR")), big.mark = ",")` radiocarbon dates with unique laboratory identifiers (excluding those sourced from p3k14c and XRONOS), IntChron has `r format(nrow(filter(c14_global, source == "IntChron")), big.mark = ",")` (excluding those from non-archaeological contexts), and p3k14c has `r format(nrow(filter(c14_global, source == "p3k14c")), big.mark = ",")`.
+As of December 2024, c14bazAAR had `r format(nrow(filter(c14_global, source == "c14bazAAR")), big.mark = ",")` radiocarbon dates with unique laboratory identifiers (excluding those sourced from p3k14c and XRONOS), IntChron had `r format(nrow(filter(c14_global, source == "IntChron")), big.mark = ",")` (excluding those from non-archaeological contexts), and p3k14c had `r format(nrow(filter(c14_global, source == "p3k14c")), big.mark = ",")`.
 The geographic distribution of dates from each is similar (@fig-c14-global), reflecting the large degree in overlap between the sources of each compilation.
-IntChron, which in practice is currently only used to publish dates associated with the Oxford Radiocarbon Lab, has dates from more diverse contexts, but less overall.
+IntChron, which in practice is currently only used to publish dates associated with the Oxford Radiocarbon Lab, has dates from more diverse contexts, but is an order of magnitude smaller.
 
@@
 Radiocarbon has been by far the most active area of open data compilation, but archaeological chronology incorporates a much more diverse range of sources of information [@Harding1999].
-In periods beyond the practical limit of radiocarbon dating (c. 50,000 BP), other types of radiometric (K–Ar, U–Pb, etc.), chemical or luminescence dating offer an alternative [@Aitken1999].
+In periods beyond the practical limit of radiocarbon dating (c. 55,000 BP), other types of radiometric (K–Ar, U–Pb, etc.), chemical or luminescence dating offer an alternative [@Aitken1999].
 Conversely, in historic periods, radiocarbon is often relatively underused compared to conventional typological dating (based on artefact characteristics), which in these periods can offer comparable or better temporal resolution, or direct dating based on epigraphy [@HermankovaEtAl2021], numismatics [@KemmersMyrberg2011] or historical sources.
@@
 Compared to radiocarbon, there are few examples of systematic, open compilations of any of these other types of data.
-
 This is most striking when it comes to other radiometric/scientific dating methods, as the data structures and publication modes are very similar to radiocarbon.
@@
 
-With regard to tree-ring data, some databases provide valuable resources for dendrochronological studies in general but are not primarily designed to collect tree-ring data from archaeological contexts.
+<!-- TODO: add references? -->
+
+With regard to tree-ring data, some databases provide valuable resources for dendrochronological studies in general but are not primarily intended for archaeological contexts.
 For instance, Dendro4Art specializes in dendrochronological data related to wooden art objects, such as sculptures and panel paintings.
@@
 The Digital Collaboratory for Cultural Dendrochronology (DCCD) presents itself as a potentially valuable international platform for dendrochronology, particularly through its integration with archaeological data services such as ARIADNE.
-However, it remains heavily biased toward datasets from the Netherlands, which account for more than two-thirds of its entries, while only 0.08% of its Quercus data pertain to Switzerland and just 2.5% represent prehistoric datasets.
+However, it remains heavily biased toward datasets from the Netherlands, which account for more than two-thirds of its entries, while only 0.08% of its *Quercus* data pertain to Switzerland and just 2.5% represent prehistoric datasets.
 Notably, these estimates date back to 2021, and an updated assessment is currently unattainable following the platform’s migration to DataverseNL, which now charges an annual fee of nearly €6,000 for access.
@@
 Typological dates—i.e.
-relative, expertise- or seriation-based dating based on artefact characteristics—is an interesting case because they are ubiquitous but rarely treated as a form of chronometric data in their own right.
+relative, expertise- or seriation-based dating based on artefact characteristics—are ubiquitous in archaeological studies but rarely treated as a form of chronometric data in their own right.
 For example, the majority of the radiocarbon datasets we reviewed (@sec-c14-compilation) included some form of typological chronological information in the form of a 'period' or 'culture' column.
@@
 PeriodO [@RabinowitzEtAl2016] is a linked open data infrastructure that includes a shared vocabulary of typological periods and corresponding calendar age estimates, and an important step towards addressing the latter problem.
-However, it remains to be systematically linked to actual compilations of typological dates.
+However, it remains to be systematically linked to compilations of typological dates [though there are some efforts in this direction e.g. @HannahEtAl2022].
 
-What is completely missing to date is a general-purpose infrastructure for combining all of these types of chronometric information on a global scale.
+---
+
+What is missing to date is a general-purpose infrastructure for combining all of these types of chronometric information on a global scale.
 This is the gap that XRONOS aims to fill, starting with three methods: radiocarbon, dendrochronology, and typological dating.
@@
 
-XRONOS is our answer to Kintigh's call [@Kintigh2006] for digital infrastructures that don't just provide access to chronological data but enables researchers to "archive, access, integrate, and mine disparate data sets".
-It complements several similar open data infrastructures within and outwith archaeology, such as the Strategic Environmental Archaeology Database [SEAD, @Buckland2014], IMPACT for mummified human remains [@NelsonWade2015], Neotoma for palaeoecological data [@WilliamsEtAl2018], IsoArcH for stable isotope data [@PlompEtAl2022], the International Soil Radiocarbon Database [ISRaD, @LawrenceEtAl2020], and the 'Big Interdisciplinary Archaeological Database' (BIAD), an ambitious new initiative to combine many of these individual domains, including chronology [@ReiterEtAl2024].
+XRONOS is our answer to Kintigh's call [-@Kintigh2006] for digital infrastructures that don't just provide access to chronological data but enables researchers to "archive, access, integrate, and mine disparate data sets".
+It complements several similar open data infrastructures within and outwith archaeology, such as the Global Biodiverisity Information Facility [GBIF, @CanhosEtAl2004], the Strategic Environmental Archaeology Database [SEAD, @Buckland2014], IMPACT for mummified human remains [@NelsonWade2015], Neotoma for palaeoecological data [@WilliamsEtAl2018], IsoArcH for stable isotope data [@PlompEtAl2022], the International Soil Radiocarbon Database [ISRaD, @LawrenceEtAl2020], and the 'Big Interdisciplinary Archaeological Database' (BIAD), an ambitious new initiative to combine many of these individual domains, including chronology [@ReiterEtAl2024].
 To improve upon existing global syntheses of radiocarbon dates (see @sec-global-compilations), we aimed to develop a living infrastructure that both continually collected data from diverse sources and presented a seamless single database to the user.
-The Global Biodiversity Information Facility [GBIF, <https://gbif.org>, @CanhosEtAl2004]—which provides a single, consistent interface to many sources of global biodiversity data—has served as an exemplar for us in this regard.
 
@@
 
-1. Combine all available sources of radiocarbon and other chronometric data in single database; 
-2. Develop robust tools for the continuous ingestion, collation and curation of this data;
-3. Disseminate the collated and curated data as linked open data within a FAIR framework.
+1. Combine all available sources of radiocarbon and other chronometric data in single database
+2. Develop robust tools for the continuous ingestion, collation and curation of this data
+3. Disseminate the collated and curated data as linked open data within a FAIR framework
 
-Meeting these goals required principally the development of a) a conceptual data model, including links to other open data resources, that is flexible enough for all forms of chronometric data; and b) an software implementation that supports the main functions of ingesting, curating, and disseminating this data.
+Meeting these goals required the development of a) a conceptual data model, including links to other open data resources, that is flexible enough for all forms of chronometric data; and b) a software implementation that supports the main functions of ingesting, curating, and disseminating this data.
 The individual components of this work are described in more depth below but, briefly, consist of a relational data model implemented in a PostgreSQL database; a Ruby application providing server-based tools for ingestion, curation and dissemination of data; and multiple graphical and programmatic interfaces to the resulting dataset.
@@
 
-<!-- TODO: shorten? -->
-
 At the base of the XRONOS data model (@fig-data-model) are sets of spatiotemporal coordinates or, as we call them, *chrons*.
@@
 the event occurred in the Neolithic) is conceptualised as a uniform probability distribution over the timespan between the known start and end dates of that period.
-A similar model is used in OxCal [@BronkRamsey2024, a direct inspiration for our approach] to integrate prior chronological information from diverse sources.
-In practical terms, this model means that the canonical representation the time component of any chron in XRONOS', regardless of source, is a probably distribution over the set of calendar years (arbitrarily measured in years Before Present) in which it could have plausibly occurred.
+A similar model is used in OxCal [@BronkRamsey2009, a direct inspiration for our approach] to integrate prior chronological information from diverse sources.
+In practical terms, this model means that the canonical representation the time component of any chron in XRONOS, regardless of source, is a probability distribution over the set of calendar years (arbitrarily measured in years Before Present) in which it could have plausibly occurred.
 Further statistics, e.g. a midpoint estimate or TPQ/TAQ range, can be derived from this distribution using well-known methods.
@@
 This allows it to encode arbitrary relational structures between contexts, whether they be hierarchical (e.g. phases and sub-phases) or graphical (e.g. stratigraphic).
-In this way, it can serve as a foundation for chronological modelling. <!-- TODO: reword, maybe a bit vague? -->
+In this way, it can serve as a foundation for chronological modelling.
 
@@
 Metadata is incorporated into XRONOS' data model at the level of the individual records (e.g.
-all records store their data of creation and last modification) and through two additional nodes: bibliographic references and versions.
+all records store their data of creation and last modification) and through two additional types of record: bibliographic *references* and *versions*.
 Bibliographic references store information on the source of a record in BibTeX format and can be linked through many-to-many associations to
@@
 The XRONOS data model presents several opportunities to link to other resources as linked open data (LOD).
-We use controlled vocabularies in the form of the GBIF Backbone Taxonomy (for taxonomic descriptions of samples) via its API, and in future plan to extend this to for example site types (using the Getty and/or DAI<!-- TODO: check & cite -->).
-Standardising field using these vocabularies to automatically then link the two entities.
-Beyond this, there is also a concordance between XRONOS entities and other specialised data infrastructures in archaeology, such as Perio.do [@RabinowitzEtAl2016, mapping to typological chrons] and gazetters like Pleiades [@Pleiades] or Vici [@Voorburg2012, mapping to site names].
+We use controlled vocabularies such as the GBIF Backbone Taxonomy [@GBIF2023, for taxonomic descriptions of samples] to both standardise fields and to link the two resources based on this shared concept.
+There are concordances between XRONOS entities and several other specialised data infrastructures in archaeology, such as PeriodO [@RabinowitzEtAl2016, mapping to typological chrons] and gazetters like Pleiades [@Pleiades] or Vici [@Voorburg2012, mapping to site names].
 Moving beyond archaeology, Wikidata (<https://wikidata.org>) already includes many of the concepts represented in the XRONOS model (e.g.
 an 'archaeological site', <https://www.wikidata.org/wiki/Q839954>).
-Linking to Wikidata is especially useful because it dissimenates—and thereby preserves—the data compiled in XRONOS beyond the narrow field of archaeology.
+Linking to Wikidata is especially useful because it dissimenates—and thereby preserves—the data compiled in XRONOS beyond the specialist/academic community.
 It also allows us to enrich the database with contextual information that otherwise be beyond our scope and resources, for example embedding multilingual descriptions of sites from Wikipedia articles on site records where the site has been linked to a Wikidata item.
@@
 ## Software architecture
-<!-- TODO: shorten -->
 
@@
 The XRONOS web application is written in Ruby and uses CRUD (Create, Read, Update, Delete) and MVC (Model, View, Controller) patterns as implemented in the Ruby on Rails framework.
-The choice of this somewhat traditional or 'boring' architecture—as opposed to the non-relational/semantic technologies that are more *à la mode* in the digital humanities [@Fan2018; @Hyvonen2020; @SchloenProsser2023]—was purposeful, motivated by a sense that it offers a better chance of longevity and maintanability.
-Our aim is for the software architecture to be as boring as possible so that the scientific contents can be as interesting as possible.
 
@@
 For example, <https://xronos.ch/c14s?sample[material][name]=charcoal> (encoding that the `c14` table should be joined to the `material` table via `sample`, followed by the WHERE clause `material.name = 'charcoal'`) lists radiocarbon dates obtained from charcoal samples.
-Uniquely, index actions can also respond with the result in a tabular data format (i.e. `.csv`).
+Index actions can also respond with the result in a tabular data format (i.e. `.csv`).
 
@@
 Our aim is not just to 'mirror' these sources as they are, but integrate them into a single curated and continuously updated database.
-<!-- TODO: too detailed?
--->For the purposes of ingestion, we classify data resources into three categories: static resources, such as supplementary data in published papers, which are imported once; versioned resources, updated on a periodic basis, which we import after each new version; and live resources, which are continuously updated and therefore continuously imported.
+For the purposes of ingestion, we classify data resources into three categories: static resources, such as supplementary data in published papers, which are imported once; versioned resources, updated on a periodic basis, which we import after each new version; and live resources, which are continuously updated and therefore continuously imported.
 Records of each type are imported into XRONOS in as close to their original state as possible, i.e. without any corrections or standardisation applied.
@@
 Controlled vocabularies are used in a number of places in the data model (@fig-data-model), and we use thesauri to automatically standardise these fields as much as possible.
-For example, the taxonomic description of samples is controlled using GBIF's backbone taxonomy [@GBIF2023], and we also use a thesaurus service provided by GBIF to automatically change variant or obsolete taxonomic names to the canonical version.
+For example, as mentioned above, the taxonomic description of samples is controlled using GBIF's backbone taxonomy [@GBIF2023], and we also use a thesaurus service provided by GBIF to automatically change variant or obsolete taxonomic names to the canonical version.
 If the system is not able to standardise a field using the available thesaurus, it is flagged for manual correction.
@@
 Pages representing REST resources directly are supplemented by a number of synthetic interfaces, for example the 'data browser' (<https://xronos.ch/data>, @fig-ui-data-browser), which facilitates more complex filtering, or the search interface (<https://xronos.ch/search>).
-The GUI also includes several resources which are not part of XRONOS' scientific data model, for example documentation pages, user profiles and news articles; these are currently not exposed in the API.
+The GUI also includes several resources which are not part of XRONOS' scientific data model, for example documentation pages, user profiles and news articles; these are not exposed in the API.
 
@@
 For this reason, there is no sharp division between a 'public' and 'private' areas – viewing/querying data and editing/curating data share the same architecture and interface patterns.
-<!-- TODO: say something about private data, even though it's not implemented yet? -->
 
@@
 Though users can make such requests manually and parse the data with one of several off-the-shelf tools, the primary intended uses of this interface is to provide access for 1) programmatic clients to XRONOS and 2) other web services.
-The XRONOS R package [@HinzRoe2024] is an example of a programmatic client; it uses the API to facilitate direct querying and retrieval of data from XRONOS in the R statistical programming language [@R2024], which is widely used for computational applications in archaeology [@BatistRoe2024].
+The XRONOS R package [@HinzRoe2024] is an example of a programmatic client; it uses the API to facilitate direct querying and retrieval of data from XRONOS in the R statistical programming language [@R2024].
 Similar libraries could be developed in other programming environments used for scientific computing, such as Python or Haskell.
@@
 
-<!-- Conclusion. TODO: do we need this? -->
 An overarching principle of this software architecture is that all interaction with XRONOS' data store, and as much of the data processing and 'business logic' of responding to REST requests as possible, is directed through the same server-side routines.
@@
 At the current stage of implementation, we argue XRONOS provides a framework for access to chronometric data that is more open, more reliable, and more comprehensive than the previously available global radiocarbon compilations.
-XRONOS blends aspects of the three existing approaches (c14bazAAR, IntChron, and p3k14c) to achieve the same aim of providing access to the global radiocarbon data through a common interface.
+XRONOS blends aspects of the three existing approaches (c14bazAAR, IntChron, and p3k14c; see Section @sec-global-compilations) to achieve the same aim of providing access to the global radiocarbon data through a common interface.
 Like c14bazAAR and IntChron, it is a 'metadatabase' that draws from existing data resources and maintains an explicit link to them.
@@
 
-Development of XRONOS will continue within the framework of 'ESTER', an ERC-funded research project on estimation of prehistoric population development from large, multiproxy datasets.
+From 2025, development of XRONOS will continue within the framework of 'ESTER', an ERC-funded research project on estimation of prehistoric population development from large, multiproxy datasets.
 Our immediate development goals include the incorporation of dendrochronology into the database, further refinement of our data curation pipelines, and the public release of the editing interfaces.
 
-Looking beyond the near term, we have endeavoured to create a sustainable infrastructure that can be maintained by a wider scholarly commons – though we must acknowledge that this is a difficult problem, and one that is more organisational than technical.
+Looking beyond the near term, we have endeavoured to create a sustainable infrastructure that can be maintained by a wider scholarly commons – though we must acknowledge that this is a difficult problem, and one that is as much organisational than technical.
 The source code for all the software components of the system are available online (at <https://github.com/xronos-ch>) under open licenses.

### 2024-12-05 paper.qmd
Add tbl-c14-datasets
Switching output format to one that supports multiple .bib files

@@
     orcid: 0000-0002-9904-6548
+    email: martin.hinz@unibe.ch
+    corresponding: true
     affiliation:
@@
     department: Institute of Archaeological Sciences
-    country: Switzerland
     url: https://www.iaw.unibe.ch/
@@
     name: Max Planck Institute for Geoanthropology
-    country: Germany
     url: https://www.shh.mpg.de/
@@
   - typological dating
-published: "Manuscript for submission to the *Journal of Computer Applications in Archaeology*"
+#published: "Manuscript for submission to the *Journal of Computer Applications in Archaeology*"
 number-sections: true
-bibliography: references.bib
+bibliography: 
+  - references.bib
+  - data/c14_datasets.bib
 format:
-  hikmah-pdf:
-    papersize: a4
-    mainfont: "Linux Libertine O" # had to adjust, otherwise was not rendering
-    sansfont: "URW Gothic"
-    biblatex-chicago: true
-    biblio-style: authordate
-    biblatexoptions:
-      - backend=biber
-      - autolang=hyphen
-      - isbn=false
-      - uniquename=false
+  elsevier-pdf:
+    keep-tex: true    
+    cite-method: citeproc
+    journal:
+      name: Journal of Computer Applications in Archaeology
+      formatting: doubleblind
+  # Pretty but broken:
+  # https://github.com/andrewheiss/hikmah-academic-quarto/issues/7
+  # hikmah-pdf:
+  #   papersize: a4
+  #   mainfont: "Linux Libertine O"
+  #   sansfont: "URW Gothic"
+  #   biblatex-chicago: true
+  #   biblio-style: authordate
+  #   biblatexoptions:
+  #     # - backend=biber
+  #     - autolang=hyphen
+  #     - isbn=false
+  #     - uniquename=false
 #  html:
@@
 
+```{r tbl-c14-datasets}
+#| tbl-cap: Summary of published compilations of radiocarbon dates. For full data, see supplementary materials.
+# References in tables broken by:
+# https://github.com/quarto-dev/quarto-cli/issues/9342
+c14_datasets |>
+  transmute(
+    database = glue("[{name}]({url})"),
+    publication_year, n_dates, citations
+  ) |>
+  arrange(publication_year) |>
+  gt() |>
+  cols_label(
+    publication_year = "published",
+    citations = "references",
+    n_dates = "dates"
+    ) |>
+  cols_label_with(everything(), str_to_sentence) |>
+  sub_missing(everything()) |>
+  fmt_markdown(c(database, citations)) |>
+  fmt_number(c(n_dates), decimals = 0) |>
+  cols_width(
+    database ~ pct(40),
+    publication_year ~ pct(15),
+    n_dates ~ pct(15),
+    citations ~ pct(30)
+  ) |>
+  cols_align(
+    "left", c(database)
+  ) |>
+  tab_options(
+    table.font.size = 13
+  )
+```
+
 Despite these inefficiencies, there have been a profusion of published radiocarbon compilations since the decline of the date list.
-Our review of the literature identified `r n_c14_datasets` published since 1994.
+Our review of the literature identified `r n_c14_datasets` published since 1994 (Table @tbl-c14-datasets and supplementary materials).
 This is almost certainly an undercount, because our firsthand knowledge of regional literature is limited to Europe and West Asia and many resources only ever existed in 'grey' formats (e.g.
@@
 'date lists' in printed periodicals and gazetteers were excluded.
-A full list of the datasets we identified can be found in the supplementary materials.
 
 ```{r fig-c14-datasets-time}
-#| fig-cap: Cumulative number of radiocarbon compilations published since 1995 according to our survey (see supplementary materials)
+#| fig-cap: Cumulative number of radiocarbon compilations published since 1995
 c14_datasets |>
@@
        caption = glue("* Excludes {english(n_c14_datasets_sans_publication_year)} datasets with an unknown year of publication")) +
-  theme_half_open()
+  theme_minimal_vgrid()
 ```
@@
 The source code for all the software components of the system are available online (at <https://github.com/xronos-ch>) under open licenses.
-The databases of the instance at <https://xronos.ch> are also archived with Zenodo (<!-- TODO: link -->).
+The databases of the instance at <https://xronos.ch> are also archived with Zenodo (<link omitted for blind review><!-- TODO: link -->).
 With these two resources we reduce XRONOS' 'bus factor'; if we are not able to continue operating XRONOS, somebody else can fully recreate it.

### 2024-12-05 paper.tex
Add tbl-c14-datasets
Switching output format to one that supports multiple .bib files

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  number,
  doubleblind]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{anyfontsize}
\usepackage{multirow}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Journal of Computer Applications in Archaeology}

\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={XRONOS: an open data infrastructure for archaeological chronology},
  pdfauthor={Joe Roe; Clemens Schmid; Setareh Ebrahimiabareghi; Caroline Heitz; Martin Hinz},
  pdfkeywords={open data, chronology, chronometry, radiocarbon
dating, dendrochronology, typological dating},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{XRONOS: an open data infrastructure for archaeological
chronology}
\author[1]{Joe Roe%
\corref{cor1}%
}
 \ead{joseph.a.roe@unibe.ch} 
\author[2]{Clemens Schmid%
%
}

\author[1]{Setareh Ebrahimiabareghi%
%
}

\author[1]{Caroline Heitz%
%
}

\author[1]{Martin Hinz%
\corref{cor1}%
}
 \ead{martin.hinz@unibe.ch}

\affiliation[1]{organization={University of Bern, Institute of
Archaeological Sciences},,postcodesep={}}
\affiliation[2]{organization={Max Planck Institute for
Geoanthropology},,postcodesep={}}

\cortext[cor1]{Corresponding author}

\begin{abstract}
XRONOS (\url{https://xronos.ch}) is an open data infrastructure for the
backbone of the archaeological record -- chronology. It provides open
access to published radiocarbon dates and other chronometric data from
any period, anywhere in the world. By collating a large number of
existing regional and global compilations of dates, XRONOS offers the
most comprehensive radiocarbon database yet published, with over 350,000
radiocarbon and 75,000 site records. It also provides a foundation for
expanding the systematic collection of chronometric information beyond
radiocarbon, with support for typological and dendrochronological dates
and a generalisable data model that can be adapted to other methods of
absolute dating. Automated and semi-automated quality control processes
ensure that data from diverse sources is continuously integrated and
standardised, making it easier to find information of interest and
reducing the need for manual data cleaning by end users. In this paper
we describe the concept and implementation of XRONOS in relation to the
state of the art in chronometric data-sharing, and evaluate its
potential as a general-purpose open repository and curation platform for
archaeological chronology.
\end{abstract}

\begin{keyword}
    open data \sep chronology \sep chronometry \sep radiocarbon
dating \sep dendrochronology \sep 
    typological dating
\end{keyword}
\end{frontmatter}

\section{Introduction}\label{introduction}

Chronology is the backbone of the archaeological record. As a necessary
prerequisite to understanding the context of any past event or process
(Lucas, 2004), it is has unsurprisingly been at the forefront of
methodological development in archaeology for as long as the discipline
has existed: from putting finds and events in sequence (Ford, 1962;
Harris, 1979; Petrie, 1899; Thomsen, 1836; Worsaae, 1843) to an
increasingly wide array of scientific methods that place them on an
absolute timescale (Bada and Helfman, 1975; Daniels et al., 1953;
Douglass, 1929; Evernden et al., 1965; Libby, 1955) and an increasingly
sophisticated set of statistical tools to build them into chronologies
(Buck et al., 1991; Crema, 2024; Levy et al., 2021; Mischka, 2004;
Suess, 1967). If archaeology is to be an open science (Lake, 2012), it
is therefore critical that effective open access to chronological
information be placed front and centre.

Over the last two decades, archaeologists have answered this call by
publishing an increasing number of compilations of dates from
archaeological contexts as open data. These efforts have facilitated
re-evaluations of chronologies themselves (e.g. Higham et al., 2014;
Katsianis et al., 2020; Loftus et al., 2019; Prates et al., 2020) but
also the development of novel ways of using chronological data (Crema et
al., 2024; Crema, 2022; e.g. Grove, 2011; Marom and Wolkowski, 2024;
Riris et al., 2024; Silva and Steele, 2014). The focus has been
overwhelmingly on radiocarbon dating and most compilations focus on a
single region and/or period. The profusion of open radiocarbon data in
particular has prompted several initiatives towards a global synthesis
(Bird et al., 2022; Bronk Ramsey et al., 2019; e.g. Schmid et al.,
2019).

At the same time, the broad range of other types of chronological
information used in archaeology---from other radiometric methods to
dendrochronology to typological dating and epigraphy---remains
relatively difficult to access as open data. Even when it comes to
radiocarbon data, the coverage of available compilations is patchy both
geographically and in time and of variable quality (see
Section~\ref{sec-c14-compilation}). The publication of many overlapping,
non-standardised and mostly static open data resources means that it is
still difficult to obtain reliable and up-to-date chronological
datasets, especially for applications that crosscut convential
geographic and temporal domains of research. Initiatives towards
synthesis have improved this situation, but the goal of a global dataset
that is both comprehensive and up-to-date remains elusive.

XRONOS is a new open data infrastructure that aims to provide access to
published radiocarbon dates and other chronometric data from any period,
anywhere in the world. It is our attempt to move the state of the art in
open archaeological chronology beyond the publication of static, one-off
resources (`uploading CSVs,' Batist, 2023, pp. 188--189), and towards a
living digital infrastructure (Kintigh, 2006) embedded in a transparent
and sustainable collabrative network. The core of XRONOS is a server
application that ingests chronological data from diverse sources,
facilitates semi-automated and manual curation of this data, and makes
it available via both a web-based graphical user interface (GUI) and
machine-readable application programming interface (API). The web
frontend can be accessed via \url{https://xronos.ch} and all components
of the software are developed as free and open source software with
source code available at \url{https://github.com/xronos-ch}.

In the remainder of this paper, we describe the concept and
implementation of XRONOS in relation to the state of the art in open
chronometric data in archaeology, and evaluate our progress in achieving
these goals as of writing. Since we envisage both XRONOS as a dataset
and XRONOS as software to be continually developing resources, the
description here should be read as a `snapshot' of the project as of
writing rather than its final state.

\section{State of the art}\label{state-of-the-art}

\subsection{Compilations of radiocarbon
dates}\label{sec-c14-compilation}

Though an \emph{explicit} emphasis on `open data' is a relatively recent
phenomenon in archaeology (Lake, 2012), the open publication of compiled
radiocarbon dates has a substantial prehistory. Arnold and Libby (Arnold
and Libby, 1951) initiated the tradition of regularly publishing `data
lists', a practice was subsequently continued by radiocarbon
laboratories as supplements to journals such as \emph{Radiocarbon} and
\emph{Archaeometry}. However, as the number of labs and volume of
radiocarbon dates being produced grew, this paper-based format became
impractical and mostly disappeared (Bronk Ramsey et al., 2019; c.f. e.g.
Ndeye et al., 2022), without being replaced by another form of
systematic data-sharing or dissemination. Additionally, because date
lists were sourced from radiocarbon laboratories directly---not from
those who collected the sample---they typically included only very
limited contextual information. On the eve of the AMS revolution there
was an effort to create a computerised `International Radiocarbon
Database' (Kra, 1988)---already by 1989 described as a ``much needed,
long overdue enterprise'' (Kra, 1989, p. 1067)---but it never came to
fruition.

Thus, even though radiocarbon data comes from a relatively limited
number of sources (some 172 active labs, Radiocarbon, 2024) and has
relatively standardised reporting conventions (Bayliss, 2015; Millard,
2014), in practice the only way to produce aggregated datasets in recent
decades has been to manually search through relevant literature for
dates reported secondarily by the submitter of the sample. This already
laborious process is further hampered by a significant inconsistency in
how much authors adhere to reporting conventions for measurements and
sample metadata, a lack of conventions on the reporting of
\emph{contextual} information, weak or nonexistent disciplinary norms
regarding the responsibility to publish results openly in a timely
fashion, and a range of other issues affecting data reuse (Moody et al.,
2021).

\begingroup
\fontsize{9.8pt}{11.7pt}\selectfont

\begin{longtable}{>{\raggedright\arraybackslash}p{\dimexpr 0.40\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 0.15\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 0.15\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedright\arraybackslash}p{\dimexpr 0.30\linewidth -2\tabcolsep-1.5\arrayrulewidth}}

\caption{\label{tbl-c14-datasets}Summary of published compilations of
radiocarbon dates. For full data, see supplementary materials.}

\tabularnewline

\toprule
Database & Published & Dates & References \\ 
\midrule\addlinespace[2.5pt]
\href{https://bda.huma-num.fr/}{Base de Données Archéologique} & 1994 & 7,000 & @Perrin2019 \\ 
\href{https://andesc14.pl}{ANDES 14C} & 1994 & 5,800 & @MichczynskiEtAl1995 \\ 
\href{http://www.tayproject.org/enghome.html}{Archaeological Settlements of Turkey} & 1998 & 1,600 & @TanindiErdogu2005 \\ 
\href{https://www.canadianarchaeology.ca/}{Canadian Archaeological Radiocarbon Database} & 1999 & 171,500 & @GajewskiEtAl2011; @KellyEtAl2022 \\ 
\href{https://radonb.ufg.uni-kiel.de/}{RADO.NB (incl. RADON and RADON-B)} & 1999 & 34,200 & @Raetzel-Fabian1999; @RADON; @RADONB; @RADO.NB \\ 
\href{https://www.waikato.ac.nz/nzcd}{New Zealand Radiocarbon Database} & 2000 & 2,000 & @McFadgenEtAl2000 \\ 
\href{http://web.archive.org/web/20080509082232/http://www.canew.org/index.html}{International Central Anatolian Neolithic e-Workshop databases} & 2001 & 1,000 & @ReingruberThissen2005; @ReingruberThissen2009 \\ 
\href{https://ees.kuleuven.be/geography/projects/14c-palaeolithic/}{Radiocarbon Palaeolithic Europe Database} & 2002 & 17,900 & @Vermeersch2024 \\ 
\href{https://zenodo.org/doi/10.5281/zenodo.7215741}{CalPal database} & 2002 & 49,800 & @Weninger2022 \\ 
\href{http://context-database.uni-koeln.de}{CONTEXT} & 2002 & 2,900 & @BohnerSchyle2004 \\ 
\href{http://pidba.utk.edu/dating.htm}{Paleoindian Database of the Americas} & 2003 & 1,300 & @AndersonEtAl2010 \\ 
\href{https://archaeologydataservice.ac.uk/archives/view/austarch_na_2014/}{AustArch (1, 2, and 3)} & 2008 & 5,000 & @WilliamsEtAl2008; @WilliamsSmith2012; @WilliamsSmith2013 \\ 
\href{https://sites.google.com/site/chapplearchaeology/irish-radiocarbon-dendrochronological-dates}{Irish Radiocarbon \& Dendrochronological Dates} & 2010 & 10,700 & @IRDD \\ 
\href{https://www.exoriente.org/associated_projects/ppnd.php}{Platform for Neolithic Radiocarbon Dates (PPND)} & 2010 & 800 & @Benz2010 \\ 
\href{http://www.paleoanthro.org/media/journal/content/PA20110001.pdf}{PACEA Geo-Referenced Radiocarbon Database} & 2011 & 6,000 & @dErricoEtAl2011 \\ 
\href{https://doi.org/10.1016/j.quaint.2012.08.2052}{Peru archaeological radiocarbon database} & 2013 & 300 & @RademakerEtAl2013 \\ 
\href{https://telearchaeology.org///EUBAR/}{EUBAR C14 database} & 2014 & 1,700 & @CapuzzoEtAl2014 \\ 
\href{https://doi.org/10.1016/j.quascirev.2014.05.015}{Wang 2014} & 2014 & 4,700 & @WangEtAl2014 \\ 
\href{http://www.14sea.org/}{14SEA} & 2015 & 3,000 & @ReingruberThissen2017 \\ 
\href{https://discovery.ucl.ac.uk/id/eprint/1469811/}{EUROEVOL} & 2015 & 14,100 & @ManningEtAl2016 \\ 
\href{https://doi.org/10.1016/j.quascirev.2015.06.022}{Flohr et al. 2015} & 2015 & 3,000 & @FlohrEtAl2016 \\ 
\href{https://doi.org/10.1016/j.quaint.2014.09.076}{South Central Andes Radiocarbon} & 2015 & 1,700 & @GayoEtAl2015 \\ 
\href{https://github.com/dirkseidensticker/aDRAC}{Archive des datations radiocarbones d'Afrique centrale} & 2016 & 1,900 & @SeidenstickerSchmid2021 \\ 
\href{https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/NJLNRJ}{KITE East Africa} & 2016 & 800 & @CourtneyMustaphi2016 \\ 
\href{https://doi.org/10.14324/000.ds.1570274}{Palmisano et al. 2017} & 2017 & 1,900 & @PalmisanoEtAl2017 \\ 
\href{http://vmtropicar-proto.ird.fr/archeologie/}{Plateforme des datations archéologiques intertropicales} & 2017 & 1,300 & @deSaulieuEtAl2017 \\ 
\href{http://www.idearqueologia.org/c14/}{Archivo de Dataciones Radiocarbónicas de la Prehistoria Recient (IDEARQ)} & 2017 & — & @UriarteGonzalezEtAl2017 \\ 
\href{https://www.rekihaku.ac.jp/up-cgi/login.pl?p=param/esrd/db_param}{Database of Radiocarbon Dates Published in Japanese Archaeological Reports} & 2018 & 44,000 & @Kudo2018; KudoEtAl2023 \\ 
\href{https://c14.arch.ox.ac.uk/sadb/db}{Southern African Radiocarbon Database} & 2019 & 2,700 & @LoftusEtAl2019 \\ 
\href{https://doi.org/10.1016/j.quascirev.2019.105878}{Douglass et al. 2019} & 2019 & 200 & @DouglassEtAl2019 \\ 
\href{https://theia.arch.cam.ac.uk/MedAfriCarbon/}{MedAfriCarbon} & 2020 & 1,600 & @LucariniEtAl2020 \\ 
\href{https://github.com/clipo/rapanui-radiocarbon}{A repository of radiocarbon data for Rapa Nui} & 2020 & 800 & @Lipo2020 \\ 
\href{https://github.com/jgregoriods/rxpand}{rxpand} & 2020 & 2,800 & @GregoriodeSouza2020 \\ 
\href{https://core.tdar.org/project/455305/mesoamerican-radiocarbon-database-mesorad}{Mesoamerican Radiocarbon Database (MesoRAD)} & 2020 & 1,800 & @HoggarthEtAl2021 \\ 
\href{https://zenodo.org/records/4541470}{AgriChange} & 2020 & 3,700 & @Martinez-GrauEtAl2021 \\ 
\href{https://github.com/ercrema/jomonPhasesAndPopulation}{Crema \& Kobayashi 2020} & 2020 & 2,100 & @CremaEtAl2016 \\ 
\href{https://rdr.ucl.ac.uk/articles/dataset/Dataset_for_An_Aegean_history_and_archaeology_written_through_radiocarbon_dates/12489137/1}{Katsianis et al. 2020} & 2020 & 3,200 & @KatsianisEtAl2020 \\ 
\href{https://www.arqueodata.com/}{ArqueoData} & 2021 & 800 & @Alcantara2021 \\ 
\href{https://github.com/philriris/caribbean-14C}{caribbean-14C} & 2021 & 2,100 & @Riris2021 \\ 
\href{https://github.com/apalmisano82/NERD}{Near East Radiocarbon Dates (NERD)} & 2021 & 11,000 & @PalmisanoEtAl2022a \\ 
\href{https://doi.org/10.13131/archelogicadata-yb11-yb66}{NeoNet} & 2021 & 2,500 & @HuetEtAl2022; @HuetEtAl2024 \\ 
\href{https://github.com/ercrema/NeolithicKoreaDemography}{Kim et al. 2021} & 2021 & 900 & @KimEtAl2021 \\ 
\href{https://doi.org/10.1371/journal.pone.0251407}{Cochrane et al. 2021} & 2021 & 100 & @CochraneEtAl2021 \\ 
\href{https://www.waikato.ac.nz/research/research-services-facilities/radiocarbon-dating/research//nz-radiocarbon-database}{Aotearoa New Zealand Radiocarbon Databas} & 2022 & 4,100 & @PetcheyEtAl2022 \\ 
\href{https://github.com/apalmisano82/AIDA}{Archive of Italian Radiocarbon Dates (AIDA)} & 2022 & 4,000 & @PalmisanoEtAl2022 \\ 
\href{https://www.p3k14c.org/}{p3k14c} & 2022 & 179,700 & @BirdEtAl2022 \\ 
\href{https://research.jcu.edu.au/data/published/7c74f590a2ba11edb22c156e754c4bda/}{Torres Strait Radiocarbon Database} & 2022 & 300 & @LinnenluckeEtAl2023 \\ 
\href{http://holoceno.iaas.ull.es/14Canarias_ULL/}{14Canarias} & 2023 & 700 & @Pardo-GordoEtAl2023 \\ 
\href{https://doi.org/10.1016/j.jasrep.2023.103944}{Hoebe et al. 2023} & 2023 & 6,500 & @HoebeEtAl2023 \\ 
\href{https://core.tdar.org/collection/71234}{Bolivian Radiocarbon Database} & 2023 & 3,000 & @Capriles2023 \\ 
\href{https://doi.org/10.5281/zenodo.8334722}{An Annotated Compilation of Chronometric Dates for the Middle-Upper Palaeolithic Transition (45-30 ka BP) in Northern Iberia (Spain)} & 2023 & 200 & @Diaz-RodriguezEtAl2023 \\ 
\href{https://doi.org/10.1371/journal.pone.0291956}{Großman et al. 2023} & 2023 & 3,400 & @GrossmannEtAl2023 \\ 
\href{https://doi.org/10.1080/0067270X.2023.2215649}{Datations absolues, Inventaires archéologiques et Bibliographies en Afrique Centrale} & 2023 & 1,800 & @ClistEtAl2023 \\ 
\href{https://doi.org/10.1016/j.quaint.2024.01.012}{Updated Peru archaeological radiocarbon database} & 2024 & 500 & @Rademaker2024 \\ 
\href{https://banadora.mom.fr/}{Banque Nationale de Données Radiocarbone pour l'Europe et le Proche Orient} & — & — & @BANADORA \\ 
\href{https://c14.arch.ox.ac.uk/database}{Oxford Radicoarbon Accelerator Unit database} & — & 8,500 & @GillespieEtAl1984; @BronkRamseyEtAl2009 \\ 
\href{https://c14.arch.ox.ac.uk/egyptdb/db.php}{Egyptian Radiocarbon Database} & — & 1,600 & @RamseyEtAl2010; @DeeEtAl2012; @DeeEtAl2013 \\ 
\href{https://c14.arch.ox.ac.uk/database}{NERC Radiocarbon Facility} & — & 6,100 & @GarnettEtAl2023 \\ 
\href{http://www.adias-uae.com/radiocarbon.html}{Abu Dhabi Islands Archaeological Survey Radiocarbon Database} & — & 100 & @Al-Abyadh-Balghelam-Dalma-JebelEtAl \\ 
\href{http://c14.kikirpa.be/}{Royal Institute for Cultural Heritage web based Radiocarbon database} & — & 7,300 & @VanStrydonckDeRoock \\ 
\href{https://telearchaeology.org///c14/}{La base de dades radiocarbòniques de Catalunya} & — & — & @BarceloAlvarezEtAl2013 \\ 
\bottomrule

\end{longtable}

\endgroup

Despite these inefficiencies, there have been a profusion of published
radiocarbon compilations since the decline of the date list. Our review
of the literature identified 61 published since 1994 (Table
Table~\ref{tbl-c14-datasets} and supplementary materials). This is
almost certainly an undercount, because our firsthand knowledge of
regional literature is limited to Europe and West Asia and many
resources only ever existed in `grey' formats (e.g.~websites that were
not indexed and no longer exist). We also restricted ourselves to
structured datasets disseminated primarily in a digital format; `date
lists' in printed periodicals and gazetteers were excluded.

\begin{figure}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-c14-datasets-time-1.pdf}}

}

\caption{\label{fig-c14-datasets-time}Cumulative number of radiocarbon
compilations published since 1995}

\end{figure}%

The number of available compilations has increased exponentially since
around 1995 (Figure~\ref{fig-c14-datasets-time}). The first generation
came around the turn of the century and consists mostly of online
databases with a web frontend. These include some databases operated by
radiocarbon labs, for example the Oxford Radiocarbon Lab (ORAU) and the
Belgian Royal Institute for Cultural Heritage (KIK-IRPA), and
essentially represent a continuation of their date lists in a digital
format. The majority, however, were compiled from the literature by
individual researchers interested in a particular region and/or period.
Notable early examples include ANDES 14C in 1994 (Central Andes,
Michczyński et al., 1995), CARD (Canada, Gajewski et al., 2011) and
RADON (Europe, Raetzel-Fabian, 1999) in 1999, and CANEW in 2001 (Near
East, Reingruber and Thissen, 2005). From 2010, coinciding with broader
shifts in scientific publishing (Tenopir et al., 2011), it became more
common to publish standalone `open data' products in the form of journal
supplements, archives in repositories and/or data papers; the
\emph{\href{https://openarchaeologydata.metajnl.com}{Journal of Open
Archaeology Data}}, launched in 2012, has been a prominent venue for
this latter category. Most recently there has been a trend towards
providing version-controlled plain text data via platforms such as
\href{https://github.com}{GitHub}, reflecting the broader adoption of
these tools amongst computational archaeologists over the last decade
(Batist and Roe, 2024). The shift from online databases towards more
static but more preservable open data products is welcome, given how
many databases from the first generation have subsequently ceased to be
accessible. Version-controlled repositories are particular well-suited
to data compilation projects because they allow for continued updates
whilst still providing snapshot `releases' that are citeable and can be
archived in long-term repositories.

\begin{figure}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-c14-datasets-map-1.pdf}}

}

\caption{\label{fig-c14-datasets-map}Geographic coverage of published
regional radiocarbon compilations according to our survey (see
Supplementary Material).}

\end{figure}%

Although this body of work has greatly improved the accessibility of
radiocarbon dates and supported significant methodological advances
(Crema et al., 2024; Crema, 2022), some limitations are apparent.

The geographic coverage of regional radiocarbon compilations is markedly
uneven (Figure~\ref{fig-c14-datasets-map}). Europe and, especially,
North America are over-represented (Alcántara and Pedroza, in press;
Chaput and Gajewski, 2016). South America, West Asia, and East Asia are
reasonably well-covered, but there practically no systematically
compiled dates from East or West Africa, Central or South Asia, or
Mainland Southeast Asia. This is probably explained in part by a lower
volume of archaeological research and access to radiocarbon dating in
these regions, but a lack of attention in compilation work must also be
a factor. For example, radiocarbon dating has been an established part
of Indian archaeology since at least 1961 (Kusumgar et al., 1963), but
we have not able to locate a single systematic compilation of dates from
South Asia.\footnote{We would be very happy to be corrected on this
  point.}

Datasets based on literature review also become out of date almost
immediately upon publication, due the the constant production of new
dates. Unfortunately this applies to many databases that are in theory
continuously updated, as it is common to see them become unmaintained
and or unexpectedly become unavailable. Of the 61 published datasets we
identified, 33 were intended to be continuously updated, but only 13
have received updates in the last two years. The average `lifespan' of a
dataset from its publication to its last update is around 4 years. Most
radiocarbon datasets we reviewed were compiled with a specific goal in
mind (e.g.~a particular analysis) and, even where there is the intention
to keep them updated afterwards, the exigencies of scientific production
combined with the labour-intensive nature of the process make that
difficult to achieve in practice.

Laboratory databases solve the problem of currency, but tend to have
more arbitary coverage, since the inclusion of data is determined by who
submits dates to that lab, not any form of principled curation. There
are also comparatively few of them -- most active labs no longer
directly publish dates that they produce (if they ever did).

Other outstanding problems with existing compilations include various
systematic biases in data collection (Clist et al., 2023) and a large
degree of overlap and duplication between individual databases. For
example, we identified 9 different resources covering Western Europe but
none covering South Asia. The quality and accessibility of published
compilations is also variable. 50 of the 61 resources we reviewed are
not `open' according to the Open Knowledge Foundation's definition of
data openness (``Open data and content can be freely used, modified, and
shared by anyone for any purpose,'' Open Knowledge Foundation, n.d.),
which both limits the access to and reuse potential of these datasets.
And even of these, many are not currently available in readily
machine-readable formats (e.g.~plain text or database files rather than
PDFs or hypertext).

The fragmentation of the radiocarbon record into regional datasets also
hinders analysis at larger scales. Although the core elements of a
radiocarbon date---laboratory identifier, radiocarbon age, measurement
error---are more or less standardised, there is no such consistency in
contextual information on the sample or site. Such contextual
information is important not just for the interpretation of dates, but
for filtering out unreliable dates based on sample information
(`chronometric hygiene' sensu Pettitt et al., 2003) and for correcting
for known systematic errors such as the marine reservoir effect (Alves
et al., 2018). Most published datasets incorporate all or part of
earlier compilations, meaning duplicate records are also very common,
but deduplicating them is not a trivial problem due to format variations
(see Section~\ref{sec-implementation-data}). These issues are by no
means impossible to overcome, but adds a significant amount of
data-cleaning effort to a process that would otherwise be very amenable
to standardisation.

\subsection{Global radiocarbon
compilations}\label{sec-global-compilations}

\begin{figure}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-c14-global-1.pdf}}

}

\caption{\label{fig-c14-global}Geographic and temporal (sum calibration)
of georeferenced dates in XRONOS and other global radiocarbon
compilations}

\end{figure}%

The profusion of radiocarbon compilations over the last decade has
naturally prompted many to think globally. Three existing initiatives in
particular share similar aims to XRONOS (at least as far as radiocarbon
is concerned): c14bazAAR, IntChron, and p3k14c.

The first available synthetic radiocarbon database was c14bazAAR (Schmid
et al., 2019), an R package that provides an index of openly published
radiocarbon databases and a common interface for retrieving them and
performing basic data cleaning. Because c14bazAAR downloads data from
its original source repositories, rather than mirroring it, it only
includes resources that have been published in a fully open and
machine-accessible format. Despite this limitation, it has global
coverage and a large number of dates (Figure~\ref{fig-c14-global}), and
was therefore our starting point for data collection for XRONOS.

Another indexical approach is taken by the IntChron project (Bronk
Ramsey et al., 2019), which exposes data from multiple sources and
exposes them with a common JSON-based web interface. The IntChron
specification is open, meaning that radiocarbon labs or compilation
projects can implement it independently and thereby allow end users to
access their data through a common interface (though to our knowledge it
has so far only been adopted by databases associated with the Oxford
Radiocarbon Lab). The JSON format also lends itself to the
implementation of wrapper libraries, for example the rIntChron package
gives direct access to IntChron-indexed databases in R (Roe, 2024).

p3k14c (Bird et al., 2022) instead compiles multiple source databases
into a single flat file dataset, with a similar level of coverage to
c14bazAAR. The major advantage of this approach is that the data is made
internally consistent and has been manually cleaned to an extent, which
makes it particularly well-suited to global analyses. The downside is
that without the continuous link to the source databases present in the
c14bazAAR and IntChron, it can only be kept up to date manually with
periodic re-releases. An accompanying package (Bird et al., 2024)
provides direct access to the p3k14c dataset in R.

As of December 2024, c14bazAAR had 118,071 radiocarbon dates with unique
laboratory identifiers (excluding those sourced from p3k14c and XRONOS),
IntChron had 12,388 (excluding those from non-archaeological contexts),
and p3k14c had 176,016. The geographic distribution of dates from each
is similar (Figure~\ref{fig-c14-global}), reflecting the large degree in
overlap between the sources of each compilation. IntChron, which in
practice is currently only used to publish dates associated with the
Oxford Radiocarbon Lab, has dates from more diverse contexts, but is an
order of magnitude smaller.

\subsection{Beyond radiocarbon}\label{beyond-radiocarbon}

Radiocarbon has been by far the most active area of open data
compilation, but archaeological chronology incorporates a much more
diverse range of sources of information (Harding, 1999). In periods
beyond the practical limit of radiocarbon dating (c.~55,000 BP), other
types of radiometric (K--Ar, U--Pb, etc.), chemical or luminescence
dating offer an alternative (Aitken, 1999). Conversely, in historic
periods, radiocarbon is often relatively underused compared to
conventional typological dating (based on artefact characteristics),
which in these periods can offer comparable or better temporal
resolution, or direct dating based on epigraphy (Heřmánková et al.,
2021), numismatics (Kemmers and Myrberg, 2011) or historical sources. In
places where it is widely available, dendrochronology (Baillie, 2014)
also produces significantly better resolved chronologies and therefore
tends to be the main source of chronometric data. Other more
application-specific chronological methods include shoreline dating
(Brøgger, 1905; Roalkvam, 2023), lichenometry (Benedict, 2009) and rock
weathering dating (Bednarik, 2020; Whitley, 2012).

Compared to radiocarbon, there are few examples of systematic, open
compilations of any of these other types of data. This is most striking
when it comes to other radiometric/scientific dating methods, as the
data structures and publication modes are very similar to radiocarbon.
The `Radiocarbon Palaeolithic Europe Database' (Vermeersch, 2020),
despite the name, includes a significant number of thermo- and optically
stimulated luminescence, electon spin resonance, uranium--thorium and
amino acid dates. Similarly, the AustArch database (Williams et al.,
2014) includes luminescence dates alongside radiocarbon, but is limited
to Australia and was last updated in 2013. Apart from these and a few
other exceptions where other scientific dates are collected alongside
radiocarbon, we are not aware of any open compilations of them.

\subsubsection{Dendrochronology}\label{dendrochronology}

With regard to tree-ring data, some databases provide valuable resources
for dendrochronological studies in general but are not primarily
intended for archaeological contexts. For instance, Dendro4Art
specializes in dendrochronological data related to wooden art objects,
such as sculptures and panel paintings. While this focus serves art
historians and conservationists well, its utility for studying
prehistoric datasets is minimal. Similarly, the Dendrochronological
Picture Database, maintained by the Swiss Federal Institute for Snow and
Avalanche Research (SLF), offers a visual archive of approximately 1,400
images documenting dendrochronological phenomena. Although valuable as
an educational resource, it does not provide raw data necessary for
chronological or archaeological analyses. Additionally, the OLDLIST and
Eastern OLDLIST databases focus on documenting the maximum ages of trees
worldwide. Their emphasis on biological longevity, while significant for
ecological research, limits their applicability to archaeological or
prehistoric investigations.

Among databases that do provide dendrochronological data, the degree to
which they support prehistoric research varies substantially. The NOAA
International Tree-Ring Data Bank (ITRDB) serves as a global repository
of tree-ring measurements. However, its focus remains predominantly on
North America, with only 34 datasets representing European prehistoric
contexts. This restricts its relevance for studying the European past.
Similarly, the ADS database, maintained by the UK-based Vernacular
Architecture Group, compiles dendrochronological data from the UK but is
limited to medieval and later periods, making it unsuitable for
prehistoric studies.

DendroDB, hosted by the Swiss Federal Institute for Forest, Snow and
Landscape Research (WSL), emphasizes ecological and climate studies over
archaeological wood material. While it claims a broad scope, the
database remains non-functional, rendering it ineffective for research
needs. The CFS-TRenD database, managed by the Canadian Forest Service,
compiles over 4,600 datasets from Canadian forests, primarily focusing
on boreal ecosystems. Despite its extensive coverage for North America,
its geographical specificity and lack of open access restrict its
utility for European prehistoric contexts. Similarly, the QUB
Dendrochronology Database, managed by Queen's University Belfast, offers
valuable datasets for Ireland and the UK but lacks significant
representation of prehistoric material, limiting its application in
broader archaeological investigations. The Building Archaeology Research
Database (BARD) contains over 24,000 records, including
dendrochronological data from more than 2,700 buildings. However, its
focus on medieval and post-medieval timber-frame construction further
narrows its utility for studies involving prehistoric wood samples.

The Digital Collaboratory for Cultural Dendrochronology (DCCD) presents
itself as a potentially valuable international platform for
dendrochronology, particularly through its integration with
archaeological data services such as ARIADNE. However, it remains
heavily biased toward datasets from the Netherlands, which account for
more than two-thirds of its entries, while only 0.08\% of its
\emph{Quercus} data pertain to Switzerland and just 2.5\% represent
prehistoric datasets. Notably, these estimates date back to 2021, and an
updated assessment is currently unattainable following the platform's
migration to DataverseNL, which now charges an annual fee of nearly
€6,000 for access. Furthermore, database activity has declined
significantly, from 3,846 new project records between 2010 and 2014 to
only 83 by the end of 2019. Although 519 additional records have been
reported since 2021, it remains unclear whether this figure includes
revisions to pre-existing entries, potentially inflating the count. The
database's narrow focus and restrictive access model significantly limit
its broader utility for prehistoric research.

Finally, the Strategic Environmental Archaeology Database (SEAD), hosted
by Umeå University, integrates multiple environmental proxies, including
dendrochronological data. However, the dendrochronological component is
largely confined to Swedish data, with limited relevance to prehistoric
contexts. While SEAD aims for broader applications, its dendro component
has limited utility for studies outside Sweden.

The utility of dendrochronological databases for prehistoric research
varies widely. Global resources such as NOAA's ITRDB and DCCD offer
substantial datasets but face significant limitations in practical
geographical and temporal scope. Similarly, platforms like DendroDB and
BARD primarily cater to historical studies, leaving critical gaps in
prehistoric coverage. Specialized resources like OLDLIST, Dendro4Art,
and the Dendrochronological Picture Database provide valuable
contributions but lack the direct relevance necessary for archaeological
tree-ring analysis. Consequently, researchers focused on prehistoric
dendrochronology must navigate a fragmented landscape of databases, each
offering distinct strengths and limitations. Addressing these gaps
remains crucial for advancing the field.

\subsubsection{Typological dating}\label{typological-dating}

Typological dates---i.e.~relative, expertise- or seriation-based dating
based on artefact characteristics---are ubiquitous in archaeological
studies but rarely treated as a form of chronometric data in their own
right. For example, the majority of the radiocarbon datasets we reviewed
(Section~\ref{sec-c14-compilation}) included some form of typological
chronological information in the form of a `period' or `culture' column.
This is also typically present in many other forms of systematic
compilation work in archaeology, for example site gazetteers. Aggregated
typological information from such sources are often used in aoristic
analysis and related methods (Crema, 2024; Mischka, 2004). What is
lacking in this presentation of typological dating is metadata on how
the determination was made and how exactly it is to be understood. Like
any archaeological date, a typological date is derived from a physical
sample -- the object or set of object from which a chronological
estimate was derived. Typological dates on one class of object may well
clash with other classes of object, or for that matter with scientific
dates -- does one trust the date on pottery, the date on architecture,
or the radiocarbon date? Without additional metadata on e.g.~who made
the typological determination or what the radiocarbon date was obtained
on, such inconsistencies are difficult to resolve Similarly the absolute
date range corresponding to a typological determination (e.g.~``Late
Neolithic'') can be interpreted in multiple ways depending on the region
and intentions of the expert making the determination. PeriodO
(Rabinowitz et al., 2016) is a linked open data infrastructure that
includes a shared vocabulary of typological periods and corresponding
calendar age estimates, and an important step towards addressing the
latter problem. However, it remains to be systematically linked to
compilations of typological dates (though there are some efforts in this
direction e.g. Hannah et al., 2022).

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

What is missing to date is a general-purpose infrastructure for
combining all of these types of chronometric information on a global
scale. This is the gap that XRONOS aims to fill, starting with three
methods: radiocarbon, dendrochronology, and typological dating. These
were chosen because they are widely used and relatively advanced in
terms of open data, but an important aim of the project is to develop a
generalisable data model that can easily scale to any and all types of
archaeological chronology (see Section~\ref{sec-data-model}).

\section{Concept}\label{concept}

XRONOS inherits its basic structure from RADON (Hinz et al., 2012;
Kneisel et al., 2014; Raetzel-Fabian, 1999; Rinne et al., 2024), with a
database-backed web application and a data model that separates
radiocarbon dates, contextual information, and sites. Our overall aims
in developing XRONOS is to bring this model, which RADON has operated on
for more than twenty years, up to date, to generalise it to other types
of chronometric information, and to transform it from an online database
to a data infrastructure that supports the continuous ingestion,
curation, and open dissemination of archaeological chronologies from
diverse sources.

\subsection{Design goals}\label{design-goals}

XRONOS is our answer to Kintigh's call (Kintigh, 2006) for digital
infrastructures that don't just provide access to chronological data but
enables researchers to ``archive, access, integrate, and mine disparate
data sets''. It complements several similar open data infrastructures
within and outwith archaeology, such as the Global Biodiverisity
Information Facility (GBIF, Canhos et al., 2004), the Strategic
Environmental Archaeology Database (SEAD, Buckland, 2014), IMPACT for
mummified human remains (Nelson and Wade, 2015), Neotoma for
palaeoecological data (Williams et al., 2018), IsoArcH for stable
isotope data (Plomp et al., 2022), the International Soil Radiocarbon
Database (ISRaD, Lawrence et al., 2020), and the `Big Interdisciplinary
Archaeological Database' (BIAD), an ambitious new initiative to combine
many of these individual domains, including chronology (Reiter et al.,
2024). To improve upon existing global syntheses of radiocarbon dates
(see Section~\ref{sec-global-compilations}), we aimed to develop a
living infrastructure that both continually collected data from diverse
sources and presented a seamless single database to the user.

Our principal goals for the software were therefore to:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Combine all available sources of radiocarbon and other chronometric
  data in single database
\item
  Develop robust tools for the continuous ingestion, collation and
  curation of this data
\item
  Disseminate the collated and curated data as linked open data within a
  FAIR framework
\end{enumerate}

Meeting these goals required the development of a) a conceptual data
model, including links to other open data resources, that is flexible
enough for all forms of chronometric data; and b) a software
implementation that supports the main functions of ingesting, curating,
and disseminating this data. The individual components of this work are
described in more depth below but, briefly, consist of a relational data
model implemented in a PostgreSQL database; a Ruby application providing
server-based tools for ingestion, curation and dissemination of data;
and multiple graphical and programmatic interfaces to the resulting
dataset.

\subsection{Data model}\label{sec-data-model}

\begin{figure}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-data-model-1.pdf}}

}

\caption{\label{fig-data-model}Simplified entity relationship diagram
showing the XRONOS data model}

\end{figure}%

At the base of the XRONOS data model (Figure~\ref{fig-data-model}) are
sets of spatiotemporal coordinates or, as we call them, \emph{chrons}.
In an archaeological context, we conceptualise a chron as an assertion
linking human activity with a particular point in space and time. Our
data model currently encompasses three types of chron: radiocarbon
dates, typological dates (e.g.~`Early Neolithic') and
dendrochronological dates. However we anticipate that the concept will
accommodate other types of absolute and relative dating techniques, as
the scope of the database expands.

Chrons are conceptually useful because they emphasise that different
types of archaeological `dates', drawn from different sources, have
essentially the same information content: the location of an event in
space and time. We thereby avoid privileging certain sources of
chronological data over (as might be the case if, for example, we
treated `period' as a fixed attribute of a site) and can accommodate
contradictory (e.g.~differences of opinion on typological
classification). This is important given that XRONOS aspires to be an
authoratative `backbone' with a global scope, so we cannot realistically
impose a single chronological scheme or resolve conflicting information
provided by specialists. They are useful practically because they expose
a common interface for attributes that all types of chronological
information share, such as a \emph{terminus post quem} (TPQ),
\emph{terminus ante quem} (TAQ), and midpoint estimate. This allows
applications that use XRONOS' data model (including XRONOS itself) to
collate chronological data from multiple sources, without necessarily
having to be aware of the pecularities of each type of dating.

In order to unify chronological information in the form of a chron, we
need a common chronological `coordinate system'. The natural choice is a
\emph{calendar probability distribution}, which expresses the
probability that an event occurred as a function of time on a calendric
scale. Most archaeologists are familiar with working with this kind of
representation in the form of calibrated radiocarbon dates, but it can
be extended and generalised to essentially any kind of chronological
information. For example, in aoristic analysis (Mischka, 2004), a
periodic time estimate (e.g.~the event occurred in the Neolithic) is
conceptualised as a uniform probability distribution over the timespan
between the known start and end dates of that period. A similar model is
used in OxCal (Bronk Ramsey, 2009, a direct inspiration for our
approach) to integrate prior chronological information from diverse
sources. In practical terms, this model means that the canonical
representation the time component of any chron in XRONOS, regardless of
source, is a probability distribution over the set of calendar years
(arbitrarily measured in years Before Present) in which it could have
plausibly occurred. Further statistics, e.g.~a midpoint estimate or
TPQ/TAQ range, can be derived from this distribution using well-known
methods. In this way, we can support many different types of date and
much of the implementation of XRONOS can be agnostic to the source of
chronological information.

Chrons are located in space through association to a \emph{sample} --
the physical object from which a chronological determination was made.
The location of samples is represented with geographical coordinates and
an associated coordinate reference system (CRS), though since in
practice the precise location of single samples is rarely available,
this property is usually inherited from the site. We also record
relevant metadata on the nature of the sample. For radiocarbon dates,
for example, we follow established conventions (Millard, 2014) in
recording the type (e.g.~charcoal, charred seed) and, where applicable,
taxonomic designation (e.g.~\emph{Quercus}, \emph{Triticum dicoccum}) of
the organic material used for dating. For typological dates, an ideal
scenario would be for the sample to represent the particular object from
which an inference was made (e.g.~`Natufian' might be inferred from
`lunate-type microlith'). In practice, the best we can glean from most
published datasets is the type of material used (e.g.~`pottery',
`lithics'). The same sample can be associated with multiple chrons,
including different types of chron. This is useful, for example, for
representing replicate radiocarbon dates on the same sample, or
radiocarbon dates and dendrochronological made on the same section of
wood for wiggle-matching.

Further contextual information is associated with \emph{contexts} and
\emph{sites}. The site is the primary geographic container for
chronological information. As already mentioned, we typically record the
spatial location of chrons using this entity, though it is possible to
modify this by providing specific coordinates at the sample level. Sites
also have attributes describing their conventional name or names in
different languages and are associated with a flexible `site type'
typology that combines information on their form and function.

A context represents the specific find-context of a sample, e.g.~an
architectural feature, stratigraphic unit, or phase. Since the units and
conventions for recording such information vary greatly between
different regions and archaeological traditions---and XRONOS is designed
with global data in mind---we leave the question of what a context
precisely represents open, and only record an unstandardised, free text
label for it. Crucially, however, contexts can have a self-referential
association to other contexts belonging to the same site. This allows it
to encode arbitrary relational structures between contexts, whether they
be hierarchical (e.g.~phases and sub-phases) or graphical
(e.g.~stratigraphic). In this way, it can serve as a foundation for
chronological modelling.

The series of relations
\texttt{{[}chron{]}\ \textgreater{}\ sample\ \textgreater{}\ context\ \textgreater{}\ site}
links the chronological and contextual sides of the XRONOS data model.
Each step is a many-to-one association, meaning for example that it is
possible to attach multiple chrons to the same sample (e.g.~replicated
radiocarbon dates on the same material), multiple \emph{types} of chrons
to the same sample (e.g.~radiocarbon dates on tree-rings for
wiggle-matching). Since this kind of information is rarely
systematically recorded in our source databases, there are currently few
actual records that make use of this feature of the data model. However,
we hope it will provide a foundation for more nuanced chronological
modelling in the future.

Metadata is incorporated into XRONOS' data model at the level of the
individual records (e.g.~all records store their data of creation and
last modification) and through two additional types of record:
bibliographic \emph{references} and \emph{versions}. Bibliographic
references store information on the source of a record in BibTeX format
and can be linked through many-to-many associations to sites or chrons.
Versions are a special type of record that are associated with all other
records (including bibliographic references) and store the previous
versions of those records as a series of changesets. In this way all
changes to data are recorded and can be reconstructed (or reversed)
precisely. This `paper trail' also stores contextual metadata, e.g.~who
made the change and why. It also means that records that are deleted can
be reviewed or restored from their stored version history, which is
never discarded. Together these two systems provide a transparent record
where the data in XRONOS comes from and how it has been altered, which
we view as essential in a scientific data infrastructure.

\subsection{Linked data}\label{linked-data}

The XRONOS data model presents several opportunities to link to other
resources as linked open data (LOD). We use controlled vocabularies such
as the GBIF Backbone Taxonomy (GBIF Secretariat, 2023, for taxonomic
descriptions of samples) to both standardise fields and to link the two
resources based on this shared concept. There are concordances between
XRONOS entities and several other specialised data infrastructures in
archaeology, such as PeriodO (Rabinowitz et al., 2016, mapping to
typological chrons) and gazetters like Pleiades (Bagnall et al., 2006)
or Vici (Voorburg, 2012, mapping to site names). Moving beyond
archaeology, Wikidata (\url{https://wikidata.org}) already includes many
of the concepts represented in the XRONOS model (e.g.~an `archaeological
site', \url{https://www.wikidata.org/wiki/Q839954}). Linking to Wikidata
is especially useful because it dissimenates---and thereby
preserves---the data compiled in XRONOS beyond the specialist/academic
community. It also allows us to enrich the database with contextual
information that otherwise be beyond our scope and resources, for
example embedding multilingual descriptions of sites from Wikipedia
articles on site records where the site has been linked to a Wikidata
item.

Conversely, we encourage others to use XRONOS as a linked open data
resource by providing stable URLs and a machine-readable interface for
every record. Such usages could look like, for example, using a XRONOS
URL as a canonical representation of a single radiocarbon date
(e.g.~\url{https://xronos.ch/c14s/SMU-257} -\textgreater{}
\url{https://xronos.ch/c14s/23410}). We also plan to implement
additional view formats to facilitate this, such as IntChron-format
JSON, to allow it to be indexed through the IntChron service, or RDF,
which is widely used by many linked open data resources in the digital
humanities.

\section{Implementation}\label{implementation}

Following a short pilot project in 2019, the first phase of development
of XRONOS was completed in 2021--2024. The web interface
(\url{https://xronos.ch}) has been publicly accessible since July 2021.
Though we envisage XRONOS as a continuously-developing open source
project and `living database', the following offers a snapshot of
progress at the end of our first grant-funded implementation phase. We
do not aim to be comprehensive, but rather to describe some key elements
of XRONOS' current implementation that illustrate how our concept has
been realised in practice.

\subsection{Software architecture}\label{software-architecture}

The XRONOS data model is implemented as a relational database using the
free and open source database management system PostgreSQL. However,
apart from backups and other routine maintenance procedures, all
interaction with the database is via a web application, which thus forms
the core of XRONOS' architecture. The XRONOS web application is written
in Ruby and uses CRUD (Create, Read, Update, Delete) and MVC (Model,
View, Controller) patterns as implemented in the Ruby on Rails
framework.

The XRONOS web application exposes two distinct user interfaces: a
graphical user interface accessed through a web browser; and an
application programming interface (API). Both interfaces follow a REST
(Representational State Transfer) pattern (Verborgh et al., 2015), where
each resource (e.g.~a single radiocarbon date, a single bibliographic
reference, or a single user) is statelessly mapped to a single address.
Users can then interact with resources at these addresses using a
preditable and uniform interface based on HTTP verbs. For example, the
radiocarbon date RTD-8904 is represented by the address
\url{https://xronos.ch/c14s/156205}. Users can view information on this
resource by sending a GET request to that address, regardless of which
interface they are using, and authorised users can modify it using POST,
PATCH, or DELETE requests. The bibliographic reference associated with
this date (Richter et al., 2017) is similarly represented at the address
\url{https://xronos.ch/references/17778}, and can be accessed at that
address using the same interface as the radiocarbon date. These uniform
REST interfaces are another example of a boring architectural choice
that make it easier for us to enrich the scientific contents of XRONOS,
by adding new types of modular resource that represent new scientific
entities.

This basic REST pattern is augmented by seven `actions' (following the
standard pattern in Rails application) that express different ways of
interacting with a resource: index, show, destroy, new, create, edit,
and update. The `show' action represents interaction with a single
resource, as described above. The `index' action, which lists resources
of a given type (e.g.~\url{https://xronos.ch/c14s} for radiocarbon
dates), is worth special mention because it is through this that the
filtering logic at the core of XRONOS' two interfaces is implemented. By
passing a query as HTTP GET parameters to the index action of a
resource, the list returned the user is modified to only include records
that match that query. For example,
\url{https://xronos.ch/sites?site\%5Bcountry_code\%5D=CH} (the part of
the URL after the \texttt{?} character encodes the SQL WHERE clause
\texttt{country\ =\ \textquotesingle{}CH\textquotesingle{}} as a GET
parameter) lists sites in Switzerland. More complex queries can be
executed using nested parameters. For example,
\url{https://xronos.ch/c14s?sample\%5Bmaterial\%5D\%5Bname\%5D=charcoal}
(encoding that the \texttt{c14} table should be joined to the
\texttt{material} table via \texttt{sample}, followed by the WHERE
clause
\texttt{material.name\ =\ \textquotesingle{}charcoal\textquotesingle{}})
lists radiocarbon dates obtained from charcoal samples. Index actions
can also respond with the result in a tabular data format
(i.e.~\texttt{.csv}).

\subsection{Data ingestion and curation}\label{sec-implementation-data}

The chronological data in XRONOS comes from a variety of sources,
including published structured datasets in repositories and journal
supplements, other online databases, literature review, and direct input
from collaborators. Our aim is not just to `mirror' these sources as
they are, but integrate them into a single curated and continuously
updated database. For the purposes of ingestion, we classify data
resources into three categories: static resources, such as supplementary
data in published papers, which are imported once; versioned resources,
updated on a periodic basis, which we import after each new version; and
live resources, which are continuously updated and therefore
continuously imported. Records of each type are imported into XRONOS in
as close to their original state as possible, i.e.~without any
corrections or standardisation applied. This ensures that any subsequent
changes (even immediate, automatic ones) are entered into the record's
version history, so that the source of any deviations or potential
errors can always be reconstructed. The version history also records the
direct source of the data for attribution purposes. In addition, a
bibliographic reference to the original resource is attached to ensure
that the source is clearly attributed even if the record is merged with
another one.

\begingroup
\fontsize{9.8pt}{11.7pt}\selectfont

\begin{longtable}{>{\raggedright\arraybackslash}p{\dimexpr 0.33\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 0.33\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedright\arraybackslash}p{\dimexpr 0.33\linewidth -2\tabcolsep-1.5\arrayrulewidth}}

\caption{\label{tbl-issues}Automatically-recognised data quality issues
currently implemented in XRONOS}

\tabularnewline

\toprule
Issue & N & Description \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{3}{>{\raggedright\arraybackslash}m{1.32\linewidth}}{Sites} \\[2.5pt] 
\midrule\addlinespace[2.5pt]
MISSING\_COORDINATES & 4452 & Missing geographic coordinates \\ 
INVALID\_COORDINATES & 0 & Geographic coordinates fall outside the earth's ellipsoid \\ 
MISSING\_COUNTRY\_CODE & 1221 & Missing data on what country the site is in \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{3}{>{\raggedright\arraybackslash}m{1.32\linewidth}}{Samples} \\[2.5pt] 
\midrule\addlinespace[2.5pt]
MISSING\_MATERIAL & 138177 & Missing data on the sample material \\ 
MISSING\_TAXON & 138182 & Missing data on the sample taxon \\ 
MISSING\_CRS & 0 & Sample has coordinates but the coordinate reference system is not given \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{3}{>{\raggedright\arraybackslash}m{1.32\linewidth}}{Taxons} \\[2.5pt] 
\midrule\addlinespace[2.5pt]
UNKNOWN\_TAXON & 9260 & Sample taxon has not been matched to the GBIF Backbone Taxonomy \\ 
LONG\_TAXON & 509 & Description of the sample taxon is implausibly long \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{3}{>{\raggedright\arraybackslash}m{1.32\linewidth}}{Radiocarbon dates} \\[2.5pt] 
\midrule\addlinespace[2.5pt]
MISSING\_C14\_AGE & 447 & Missing radiocarbon age \\ 
VERY\_OLD\_C14 & 76 & Radiocarbon age older than the effective range of the method (50 ka) \\ 
MISSING\_C14\_ERROR & 585 & Missing measurement error \\ 
MISSING\_D14C & 238875 & Missing δ13C measurement \\ 
MISSING\_D14C\_ERROR & 238875 & Missing δ13C measurement error \\ 
MISSING\_C14\_METHOD & 242233 & Missing data on radiocarbon dating method (conventional, AMS, etc.) \\ 
MISSING\_C14\_LAB\_ID & 1233 & Missing laboratory identifier \\ 
INVALID\_LAB\_ID & 16053 & Laboratory identifier does not match the standard format (e.g. 'Abc-1234') \\ 
MISSING\_C14\_LAB & 350190 & Missing data on the radiocarbon laboratory \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{3}{>{\raggedright\arraybackslash}m{1.32\linewidth}}{Bibliographic references} \\[2.5pt] 
\midrule\addlinespace[2.5pt]
MIXED\_REFERENCE & 21933 & Bibliographic reference appears to combine multiple publications \\ 
MISSING\_BIBTEX & 42109 & Bibliographic reference without structured data in BibTeX format \\ 
\bottomrule

\end{longtable}

\endgroup

Once ingested, we apply a number of automated and semi-automated quality
control processes to integrate new data into the existing database.
Controlled vocabularies are used in a number of places in the data model
(Figure~\ref{fig-data-model}), and we use thesauri to automatically
standardise these fields as much as possible. For example, as mentioned
above, the taxonomic description of samples is controlled using GBIF's
backbone taxonomy (GBIF Secretariat, 2023), and we also use a thesaurus
service provided by GBIF to automatically change variant or obsolete
taxonomic names to the canonical version. If the system is not able to
standardise a field using the available thesaurus, it is flagged for
manual correction. A wide variety of other potential data quality issues
(e.g.~missing data on what country a site is in) are also flagged for
human review by this system Table~\ref{tbl-issues}, which can often be
semi-automated (e.g.~suggesting close matches in the thesaurus or the
country indicated by the record's coordinates).

A final critical component of XRONOS' data curation system is duplicate
handling. We import data from many overlapping resources (many of which
incorporate each other either in whole or in part), so duplicate records
are common (as recently discussed by Reiter et al., 2024). The end
result of standardising and correcting a record is also often to create
a duplicate: e.g.~the same sample imported from one source as `oak' but
another as `\emph{Quercus} sp.' will become a duplicate pair as
`\emph{Quercus}', and thus be recognised as a single sample. Such exact
duplicates can be merged automatically, with the oldest record becoming
the authoratitive version, but detecting fuzzier duplicated information
(e.g.~differences in the spelling of site names) has proved a more
difficult problem. As of writing there are therefore still many
duplicate records in XRONOS that need to be manually resolved, but we
hope to automate much more of this work in the future.

\subsection{User interfaces}\label{user-interfaces}

\begin{figure}

\begin{minipage}{0.50\linewidth}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{figures/xronos-ui-data-browser.png}}

}

\subcaption{\label{fig-ui-data-browser}Data browser}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{figures/xronos-ui-show.png}}

}

\subcaption{\label{fig-ui-show}Site record}

\end{minipage}%
\newline
\begin{minipage}{0.50\linewidth}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{figures/xronos-ui-papertrail.png}}

}

\subcaption{\label{fig-ui-papertrail}Record change log}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{figures/xronos-ui-curation.png}}

}

\subcaption{\label{fig-ui-curation}Curation interface}

\end{minipage}%

\caption{\label{fig-ui}Views in the XRONOS web graphical user interface}

\end{figure}%

The graphical user interface (GUI) to XRONOS, accessed through a web
browser (e.g.~at \url{https://xronos.ch}), uses REST resources and
actions as the building blocks for various interfaces through which
users can browse, search, retrieve, and analyse chronometric data
(Figure~\ref{fig-ui}). Each action on each resource is represented by a
page, though not all of these are publicly accessible. The most import
of these are the `index' views, which list and summaries all instances
of a resource and support filtering and sorting; and `show' views (e.g.
Figure~\ref{fig-ui-show} for a site), which give a more comprehensive
information on an individual record along with visualisations, links to
related records, external linked open data resources, and a log of
changes made to the data since it was imported into XRONOS
(Figure~\ref{fig-ui-papertrail}). Pages representing REST resources
directly are supplemented by a number of synthetic interfaces, for
example the `data browser' (\url{https://xronos.ch/data},
Figure~\ref{fig-ui-data-browser}), which facilitates more complex
filtering, or the search interface (\url{https://xronos.ch/search}). The
GUI also includes several resources which are not part of XRONOS'
scientific data model, for example documentation pages, user profiles
and news articles; these are not exposed in the API.

Access to various `backstage' interfaces for creating, editing, and
deleting data, and monitoring data quality is managed using a user
permissions system. Currently only authorised users affiliated with the
XRONOS project can access these, but in the future we intend to support
open registration and expose editing interfaces to all authenticated
users. For this reason, there is no sharp division between a `public'
and `private' areas -- viewing/querying data and editing/curating data
share the same architecture and interface patterns.

The XRONOS API uses the same addresses as the web-based GUI (with the
exception of some of the synthetic interfaces mentioned above) but
responds with machine-readable data in JSON format, rather than a HTML
page. This response can be triggered by appending \texttt{.json} to the
address or by including a HTTP \texttt{content-format} header in the
request. Though users can make such requests manually and parse the data
with one of several off-the-shelf tools, the primary intended uses of
this interface is to provide access for 1) programmatic clients to
XRONOS and 2) other web services. The XRONOS R package (Hinz and Roe,
2024) is an example of a programmatic client; it uses the API to
facilitate direct querying and retrieval of data from XRONOS in the R
statistical programming language (R Core Team, 2024). Similar libraries
could be developed in other programming environments used for scientific
computing, such as Python or Haskell. The API also provides the
foundation for other web services to access XRONOS directly, to embed
chronological information in other contexts or otherwise make use of its
data resources.

An overarching principle of this software architecture is that all
interaction with XRONOS' data store, and as much of the data processing
and `business logic' of responding to REST requests as possible, is
directed through the same server-side routines. First and foremost, this
allows us to provide multiple interfaces (i.e.~the GUI and API, perhaps
more in the future) without duplicating these elements of our codebase.
It also improves accessibility for users accessing XRONOS through
devices with limited processing capability or through text-only
browsers. More broadly, avoiding reliance on client-side processing,
e.g.~with Javascript or Web Assembly (WASM)---which would be the other
option---allows us keep our client interfaces simple (in most cases
plain, semantic HTML pages and self-contained stylesheets) and
therefore, we hope, sustainable in the face of constantly-evolving
client-side technologies and standards. It does have the weakness that,
in practical terms, XRONOS is difficult to run and relies on the
continued existence of a maintained external server. We have however
tried to mitigate this by providing clearly-documented source code and
regular data dumps so that, if our instance of XRONOS disappears, or if
one simply does not want to use it, it is possible for others to host a
XRONOS server of their own. As the sustainability of scientific software
and data infrastructures is a pressing problem, in the future it may be
desirable to support further decentralisation through, for example, a
federated server-server model.

\section{Evaluation}\label{evaluation}

In this paper we have outlined the conceptual and technical
infrastructure developed during the initial, SNF-funded phases of
development on XRONOS in 2019 and 2021--2024. These include a
generalised data model for radiocarbon and typological dates, extendible
to other chronometric information, and associated contextual
information; an R- and Ruby-based pipeline for continuous ingestion of
data from a variety of sources; continuous, semi-automated data cleaning
protocols; a Ruby-on-Rails application providing a web-based frontend to
the data and a REST API for programmatic access; and an R package for
interfacing with the API. These systems are in production and publicly
accessible at \url{https://xronos.ch}.

At the current stage of implementation, we argue XRONOS provides a
framework for access to chronometric data that is more open, more
reliable, and more comprehensive than the previously available global
radiocarbon compilations. XRONOS blends aspects of the three existing
approaches (c14bazAAR, IntChron, and p3k14c; see Section
Section~\ref{sec-global-compilations}) to achieve the same aim of
providing access to the global radiocarbon data through a common
interface. Like c14bazAAR and IntChron, it is a `metadatabase' that
draws from existing data resources and maintains an explicit link to
them. But like p3k14c, it integrates these into a single database and
applies data curation processes to harmonise them and improve the
quality of the information. It has a wider scope than c14bazAAR or
IntChron, as it mirrors rather than directly retrieves the source data
(allowing us to use resources that aren't openly published), and does
not rely on the authors of these sources to implement a common
specification. It also goes beyond the functionality of p3k14c by
providing systems for the continuous ingestion and curation of new data
as it is published. With that said, we see the approaches followed by as
complementary rather than competing. A c14bazAAR parser for XRONOS is in
development (\url{https://github.com/ropensci/c14bazAAR/pull/150}), and
we also aim to provide an IntChron interface to XRONOS' data in the near
future. New data and corrections from p3k14c are incorporated into
XRONOS as they are released.

From 2025, development of XRONOS will continue within the framework of
`ESTER', an ERC-funded research project on estimation of prehistoric
population development from large, multiproxy datasets. Our immediate
development goals include the incorporation of dendrochronology into the
database, further refinement of our data curation pipelines, and the
public release of the editing interfaces.

Looking beyond the near term, we have endeavoured to create a
sustainable infrastructure that can be maintained by a wider scholarly
commons -- though we must acknowledge that this is a difficult problem,
and one that is as much organisational than technical. The source code
for all the software components of the system are available online (at
\url{https://github.com/xronos-ch}) under open licenses. The databases
of the instance at \url{https://xronos.ch} are also archived with Zenodo
(). With these two resources we reduce XRONOS' `bus factor'; if we are
not able to continue operating XRONOS, somebody else can fully recreate
it. Equally importantly, we enable the `right to fork' should others
wish to take the software and/or database in another direction. But
aside from these extreme scenarios, the long-term sustainability of
XRONOS is contingent on the existence of a community of scholars that
use and contribute back to it.

\section{Acknowledgements}\label{acknowledgements}

The XRONOS project was carried under the direction of Albert Hafner.
Individual contributors to the XRONOS database to date include Chiara
Huwiler, Rivana Moser, Tomasz Chmielewski, and Stephanie Döppler. A
complete and up-to-date list of acknowledgements for the project can be
found at \url{https://xronos.ch/about/acknowledgements}.

\section{Funding statement}\label{funding-statement}

This work was funded by the Swiss National Science Foundation
(\href{https://data.snf.ch/grants/grant/198153}{SNSF Project \#198152})
and the University of Bern (UniBE Initiator Grant 2019, Caroline Heitz,
Project `Time and Temporality').

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-Aitken1999}
Aitken, M.J., 1999. Science-{Based Dating} in {Archaeology}. Routledge,
London. \url{https://doi.org/10.4324/9781315836645}

\bibitem[\citeproctext]{ref-AlcantaraPedrozainpress}
Alcántara, A., Pedroza, I., in press. Latin {American Radiocarbon} on
the {Web Databases} and {Datasets}: {A Mexican} and {Brazilian
Perspectives}. Radiocarbon.

\bibitem[\citeproctext]{ref-AlvesEtAl2018}
Alves, E.Q., Macario, K., Ascough, P., Bronk Ramsey, C., 2018. The
{Worldwide Marine Radiocarbon Reservoir Effect}: {Definitions},
{Mechanisms}, and {Prospects}. Reviews of Geophysics 56, 278--305.
\url{https://doi.org/10.1002/2017RG000588}

\bibitem[\citeproctext]{ref-ArnoldLibby1951}
Arnold, J.R., Libby, W.F., 1951. Radiocarbon {Dates}. Science 113,
111--120. \url{https://doi.org/10.1126/science.113.2927.111}

\bibitem[\citeproctext]{ref-BadaHelfman1975}
Bada, J.L., Helfman, P.M., 1975. Amino acid racemization dating of
fossil bones. World Archaeology 7, 160--173.
\url{https://doi.org/10.1080/00438243.1975.9979630}

\bibitem[\citeproctext]{ref-Pleiades}
Bagnall, R., Talbert, R.J.A., Tom, E., et al., 2006.
\href{https://pleiades.stoa.org/}{Pleiades: {A Community-Built
Gazetteer} and {Graph} of {Ancient Places}}.

\bibitem[\citeproctext]{ref-Baillie2014}
Baillie, M.G.L., 2014. Tree-ring {Dating} and {Archaeology}. Routledge,
London. \url{https://doi.org/10.4324/9781315748689}

\bibitem[\citeproctext]{ref-Batist2023}
Batist, Z., 2023. Archaeological data work as continuous and
collaborative practice. \url{https://doi.org/10.5281/zenodo.8373390}

\bibitem[\citeproctext]{ref-BatistRoe2024}
Batist, Z., Roe, J., 2024. Open {Archaeology}, {Open Source}?
{Collaborative} practices in an emerging community of archaeological
software engineers. Internet Archaeol.
\url{https://doi.org/10.11141/ia.67.13}

\bibitem[\citeproctext]{ref-Bayliss2015}
Bayliss, A., 2015. Quality in {Bayesian} chronological models in
archaeology. World Archaeol. 47, 677--700.
\url{https://doi.org/10.1080/00438243.2015.1067640}

\bibitem[\citeproctext]{ref-Bednarik2020}
Bednarik, R.G., 2020. The use of weathering indices in rock art science
and archaeology. Rock Art Research: The Journal of the Australian Rock
Art Research Association (AURA) 29, 59--84.
\url{https://doi.org/10.3316/ielapa.275310392544915}

\bibitem[\citeproctext]{ref-Benedict2009}
Benedict, J.B., 2009. A {Review} of {Lichenometric Dating} and {Its
Applications} to {Archaeology}. American Antiquity 74, 143--172.
\url{https://doi.org/10.1017/S0002731600047545}

\bibitem[\citeproctext]{ref-BirdEtAl2024}
Bird, D., Bocinsky, K., Vander Linden, M., 2024.
\href{https://github.com/people3k/p3k14c}{p3k14c: P3k14C: A synthetic
global database of archaeological radiocarbon dates}.

\bibitem[\citeproctext]{ref-BirdEtAl2022}
Bird, D., Miranda, L., Vander Linden, M., Robinson, E., Bocinsky, R.K.,
Nicholson, C., Capriles, J.M., Finley, J.B., Gayo, E.M., Gil, A.,
d'Alpoim Guedes, J., Hoggarth, J.A., Kay, A., Loftus, E., Lombardo, U.,
Mackie, M., Palmisano, A., Solheim, S., Kelly, R.L., Freeman, J., 2022.
P3k14c, a synthetic global database of archaeological radiocarbon dates.
Scientific Data 9, 27. \url{https://doi.org/10.1038/s41597-022-01118-7}

\bibitem[\citeproctext]{ref-Brogger1905}
Brøgger, W.C., 1905. Strandliniens beliggenhed under stenalderen i det
sydøstlige Norge. I kommission hos H. Aschehoug \& Company.

\bibitem[\citeproctext]{ref-BronkRamsey2009}
Bronk Ramsey, C., 2009. Bayesian {Analysis} of {Radiocarbon Dates}.
Radiocarbon 51, 337--360.
\url{https://doi.org/10.1017/S0033822200033865}

\bibitem[\citeproctext]{ref-BronkRamseyEtAl2019}
Bronk Ramsey, C., Blaauw, M., Kearney, R., Staff, R.A.S., 2019. The
importance of open access to chronological information: The {IntChron}
initiative. Radiocarbon 61, 1--11.
\url{https://doi.org/10.1017/RDC.2019.21}

\bibitem[\citeproctext]{ref-BuckEtAl1991}
Buck, C.E., Kenworthy, J.B., Litton, C.D., Smith, A.F.M., 1991.
Combining archaeological and radiocarbon information: A {Bayesian}
approach to calibration. Antiquity 65, 808--821.
\url{https://doi.org/10.1017/S0003598X00080534}

\bibitem[\citeproctext]{ref-Buckland2014}
Buckland, P.I., 2014. {SEAD} - {The Strategic Environmental Archaeology
Database Inter-linking Multiproxy Environmental Data} with
{Archaeological Investigations} and {Ecology}, in: Verhagen, P., Earl,
G. (Eds.), Archaeology in the {Digital Era}: {Papers} from the 40th
{Annual Conference} of {Computer Applications} and {Quantitative
Methods} in {Archaeology} ({CAA}), {Southampton}, 26-29 {March} 2012.
Amsterdam University Press, pp. 320--331.
\url{https://doi.org/10.1515/9789048519590-035}

\bibitem[\citeproctext]{ref-CanhosEtAl2004}
Canhos, V.P., Souza, S. de, Giovanni, R.D., Canhos, D.A.L., 2004. Global
{Biodiversity Informatics}: Setting the scene for a "new world" of
ecological forecasting. Biodiversity Informatics 1.
\url{https://doi.org/10.17161/bi.v1i0.3}

\bibitem[\citeproctext]{ref-ChaputGajewski2016}
Chaput, M.A., Gajewski, K., 2016. Radiocarbon dates as estimates of
ancient human population size. Anthropocene 15, 3--12.
\url{https://doi.org/10.1016/j.ancene.2015.10.002}

\bibitem[\citeproctext]{ref-ClistEtAl2023}
Clist, B., Denbow, J., Lanfranchi, R., 2023. Using the radiocarbon dates
of {Central Africa} for studying long-term demographic trends of the
last 50,000 years: Potential and pitfalls. Azania: Archaeological
Research in Africa 58, 235--293.
\url{https://doi.org/10.1080/0067270X.2023.2215649}

\bibitem[\citeproctext]{ref-Crema2024}
Crema, E.R., 2024. A {Bayesian} alternative for {Aoristic} analyses in
archaeology. \url{https://doi.org/10.31219/osf.io/98qkx}

\bibitem[\citeproctext]{ref-Crema2022}
Crema, E.R., 2022. Statistical {Inference} of {Prehistoric Demography}
from {Frequency Distributions} of {Radiocarbon Dates}: {A Review} and a
{Guide} for the {Perplexed}. J Archaeol Method Theory 29, 1387--1418.
\url{https://doi.org/10.1007/s10816-022-09559-5}

\bibitem[\citeproctext]{ref-CremaEtAl2024}
Crema, E.R., Bloxam, A., Stevens, C.J., Vander Linden, M., 2024.
Modelling diffusion of innovation curves using radiocarbon data. Journal
of Archaeological Science 165, 105962.
\url{https://doi.org/10.1016/j.jas.2024.105962}

\bibitem[\citeproctext]{ref-DanielsEtAl1953}
Daniels, F., Boyd, C.A., Saunders, D.F., 1953. Thermoluminescence as a
{Research Tool}. Science 117, 343--349.
\url{https://doi.org/10.1126/science.117.3040.343}

\bibitem[\citeproctext]{ref-Douglass1929}
Douglass, A.E., 1929. The secret of the {Southwest} solved by talkative
tree rings. National Geographic 56, 737--770.

\bibitem[\citeproctext]{ref-EverndenEtAl1965}
Evernden, J.F., Curtis, G.H., Bishop, W., Brace, C.L., Clark, J.D.,
Damon, P.E., Hay, R.L., Hopkins, D.M., Howell, F.C., Knopf, A., Kretzoi,
M., Leakey, L.S.B., Maude, H.E., Richards, J.R., Savage, D.E., Wright,
H.E., 1965. The {Potassium-Argon Dating} of {Late Cenozoic Rocks} in
{East Africa} and {Italy} {[}and {Comments} and {Reply}{]}. Current
Anthropology 6, 342--385. \url{https://doi.org/10.1086/200619}

\bibitem[\citeproctext]{ref-Ford1962}
Ford, J.A., 1962. A quantitative method for deriving cultural
chronology, Manuales técnicos. Pan American Union, Washington.

\bibitem[\citeproctext]{ref-GajewskiEtAl2011}
Gajewski, K., Munoz, S., Peros, M., Viau, A., Morlan, R., Betts, M.,
2011. The {Canadian Archaeological Radiocarbon Database} ({Card}):
{Archaeological 14C Dates} in {North America} and {Their
Paleoenvironmental Context}. Radiocarbon 53, 371--394.
\url{https://doi.org/10.1017/S0033822200056630}

\bibitem[\citeproctext]{ref-GBIF2023}
GBIF Secretariat, 2023. {GBIF Backbone Taxonomy}.
\url{https://doi.org/10.15468/39OMEI}

\bibitem[\citeproctext]{ref-Grove2011}
Grove, M., 2011. A {Spatio-Temporal Kernel Method} for {Mapping Changes}
in {Prehistoric Land-Use Patterns}. Archaeometry 53, 1012--1030.
\url{https://doi.org/10.1111/j.1475-4754.2010.00578.x}

\bibitem[\citeproctext]{ref-HannahEtAl2022}
Hannah, E., Carlin, N., Griffiths, S., Gaydarska, B., Swan, R., Corns,
A., 2022. \href{https://www.jstor.org/stable/27198744}{Project
{Radiocarbon}}. Archaeology Ireland 36, 14--17.

\bibitem[\citeproctext]{ref-Harding1999}
Harding, A., 1999. Establishing {Archaeological Chronologies}, in:
Companion {Encyclopedia} of {Archaeology}. Routledge.

\bibitem[\citeproctext]{ref-Harris1979}
Harris, E.C., 1979. The laws of archaeological stratigraphy. World
Archaeology 11, 111--117.
\url{https://doi.org/10.1080/00438243.1979.9979753}

\bibitem[\citeproctext]{ref-HermankovaEtAl2021}
Heřmánková, P., Kaše, V., Sobotková, A., 2021. Inscriptions as data:
Digital epigraphy in macro-historical perspective. Journal of Digital
History 1, 99--141. \url{https://doi.org/10.1515/jdh-2021-1004}

\bibitem[\citeproctext]{ref-HighamEtAl2014}
Higham, T., Douka, K., Wood, R., Ramsey, C.B., Brock, F., Basell, L.,
Camps, M., Arrizabalaga, A., Baena, J., Barroso-Ruíz, C., Bergman, C.,
Boitard, C., Boscato, P., Caparrós, M., Conard, N.J., Draily, C.,
Froment, A., Galván, B., Gambassini, P., Garcia-Moreno, A., Grimaldi,
S., Haesaerts, P., Holt, B., Iriarte-Chiapusso, M.-J., Jelinek, A.,
Jordá Pardo, J.F., Maíllo-Fernández, J.-M., Marom, A., Maroto, J.,
Menéndez, M., Metz, L., Morin, E., Moroni, A., Negrino, F.,
Panagopoulou, E., Peresani, M., Pirson, S., de la Rasilla, M.,
Riel-Salvatore, J., Ronchitelli, A., Santamaria, D., Semal, P., Slimak,
L., Soler, J., Soler, N., Villaluenga, A., Pinhasi, R., Jacobi, R.,
2014. The timing and spatiotemporal patterning of {Neanderthal}
disappearance. Nature 512, 306--309.
\url{https://doi.org/10.1038/nature13621}

\bibitem[\citeproctext]{ref-RADON}
Hinz, M., Furholt, M., Müller, J., Raetzel-Fabian, D., Rinne, C.,
Sjögren, K.-G., Wotzka, H.-P., 2012. {RADON} - {Radiocarbon} dates
online 2012. {Central European} database of {14C} dates for the
{Neolithic} and {Early Bronze Age}. Jungsteinsite 14, 1--4.

\bibitem[\citeproctext]{ref-HinzRoe2024}
Hinz, M., Roe, J., 2024.
\href{https://github.com/xronos-ch/xronos.R}{Xronos: Client for the
'XRONOS' chronological database}.

\bibitem[\citeproctext]{ref-KatsianisEtAl2020}
Katsianis, M., Bevan, A., Styliaras, G., Maniatis, Y., 2020. An {Aegean
History} and {Archaeology Written} through {Radiocarbon Dates}. Journal
of Open Archaeology Data 8. \url{https://doi.org/10.5334/joad.65}

\bibitem[\citeproctext]{ref-KemmersMyrberg2011}
Kemmers, F., Myrberg, N., 2011. Rethinking numismatics. {The}
archaeology of coins. Archaeological Dialogues 18, 87--108.
\url{https://doi.org/10.1017/S1380203811000146}

\bibitem[\citeproctext]{ref-Kintigh2006}
Kintigh, K., 2006. The promise and challenge of archaeological data
integration. Am. Antiq. 71, 567--578.
\url{https://doi.org/10.1017/S0002731600039810}

\bibitem[\citeproctext]{ref-RADONB}
Kneisel, J., Hinz, M., Rinne, C., 2014. {RADON-B} -- {Radiocarbon} dates
online (version 2014). {Database} for {European 14C} dates for the
{Bronze} and {Early Iron Age}.

\bibitem[\citeproctext]{ref-Kra1989}
Kra, R., 1989. The {International Radiocarbon Data Base}: {A Progress
Report}. Radiocarbon 31, 1067--1075.
\url{https://doi.org/10.1017/S003382220001273X}

\bibitem[\citeproctext]{ref-Kra1988}
Kra, R., 1988. Updating the past: {The Establishment} of the
{International Radiocarbon Data Base}. American Antiquity 53, 118--125.
\url{https://doi.org/10.2307/281158}

\bibitem[\citeproctext]{ref-KusumgarEtAl1963}
Kusumgar, S., Lal, D., Sarna, R.P., 1963. Tata {Institute Radiocarbon
Date List I}. Radiocarbon 5, 273--282.
\url{https://doi.org/10.1017/S0033822200036894}

\bibitem[\citeproctext]{ref-Lake2012}
Lake, M., 2012. Open archaeology. World Archaeology 44, 471--478.
\url{https://doi.org/10.1080/00438243.2012.748521}

\bibitem[\citeproctext]{ref-LawrenceEtAl2020}
Lawrence, C.R., Beem-Miller, J., Hoyt, A.M., Monroe, G., Sierra, C.A.,
Stoner, S., Heckman, K., Blankinship, J.C., Crow, S.E., McNicol, G.,
Trumbore, S., Levine, P.A., Vindušková, O., Todd-Brown, K., Rasmussen,
C., Hicks Pries, C.E., Schädel, C., McFarlane, K., Doetterl, S., Hatté,
C., He, Y., Treat, C., Harden, J.W., Torn, M.S., Estop-Aragonés, C.,
Asefaw Berhe, A., Keiluweit, M., Della Rosa Kuhnen, Á., Marin-Spiotta,
E., Plante, A.F., Thompson, A., Shi, Z., Schimel, J.P., Vaughn, L.J.S.,
von Fromm, S.F., Wagai, R., 2020. An open-source database for the
synthesis of soil radiocarbon data: {International Soil Radiocarbon
Database} ({ISRaD}) version 1.0. Earth System Science Data 12, 61--76.
\url{https://doi.org/10.5194/essd-12-61-2020}

\bibitem[\citeproctext]{ref-LevyEtAl2021}
Levy, E., Geeraerts, G., Pluquet, F., Piasetzky, E., Fantalkin, A.,
2021. Chronological networks in archaeology: {A} formalised scheme.
Journal of Archaeological Science 127, 105225.
\url{https://doi.org/10.1016/j.jas.2020.105225}

\bibitem[\citeproctext]{ref-Libby1955}
Libby, W.F., 1955. Radiocarbon {Dating}, Second. ed. University of
Chicago Press.

\bibitem[\citeproctext]{ref-LoftusEtAl2019}
Loftus, E., Mitchell, P.J., Ramsey, C.B., 2019. An archaeological
radiocarbon database for southern {Africa}. Antiquity 93, 870--885.
\url{https://doi.org/10.15184/aqy.2019.75}

\bibitem[\citeproctext]{ref-Lucas2004}
Lucas, G., 2004. Beyond chronology, in: The {Archaeology} of {Time}.
Routledge.

\bibitem[\citeproctext]{ref-MaromWolkowski2024}
Marom, N., Wolkowski, U., 2024. A note on predator-prey dynamics in
radiocarbon datasets. Peer Community Journal 4.
\url{https://doi.org/10.24072/pcjournal.395}

\bibitem[\citeproctext]{ref-MichczynskiEtAl1995}
Michczyński, A., Krzanowski, A., Pazdur, M.F., Ziołkowski, M.S., 1995. A
{Computer-Based Database} for {Radiocarbon Dates} of {Central Andean
Archaeology}. Radiocarbon 37, 337--343.
\url{https://doi.org/10.1017/S0033822200030812}

\bibitem[\citeproctext]{ref-Millard2014}
Millard, A.R., 2014. Conventions for reporting radiocarbon
determinations. Radiocarbon 56, 555--559.
\url{https://doi.org/10.2458/56.17455}

\bibitem[\citeproctext]{ref-Mischka2004}
Mischka, D., 2004. Aoristische Analyse in der Archäologie.
Archäologische Informationen 27, 233--243.
\url{https://doi.org/10.11588/ai.2004.2.12685}

\bibitem[\citeproctext]{ref-MoodyEtAl2021}
Moody, B., Dye, T., May, K., Wright, H., Buck, C., 2021. Digital
chronological data reuse in archaeology: {Three} case studies with
varying purposes and perspectives. Journal of Archaeological Science:
Reports 40, 103188. \url{https://doi.org/10.1016/j.jasrep.2021.103188}

\bibitem[\citeproctext]{ref-NdeyeEtAl2022}
Ndeye, M., Athie, A.H., Kébé, D., Sène, M., 2022. {IFAN CH}.{DIOP
RADIOCARBON LABORATORY MEASUREMENTS II}. Radiocarbon 64, 397--414.
\url{https://doi.org/10.1017/RDC.2022.8}

\bibitem[\citeproctext]{ref-NelsonWade2015}
Nelson, A.J., Wade, A.D., 2015. Impact: {Development} of a {Radiological
Mummy Database}. The Anatomical Record 298, 941--948.
\url{https://doi.org/10.1002/ar.23130}

\bibitem[\citeproctext]{ref-OpenKnowledgeFoundation}
Open Knowledge Foundation, n.d. The {Open Definition}: {Defining Open}
in {Open Data}, {Open Content} and {Open Knowledge} {[}WWW Document{]}.
URL \url{https://opendefinition.org/} (accessed 11.25.2024).

\bibitem[\citeproctext]{ref-Petrie1899}
Petrie, W.M.F., 1899. Sequences in {Prehistoric Remains}. The Journal of
the Anthropological Institute of Great Britain and Ireland 29, 295--301.
\url{https://doi.org/10.2307/2843012}

\bibitem[\citeproctext]{ref-PettittEtAl2003}
Pettitt, P.B., Davies, W., Gamble, C.S., Richards, M.B., 2003.
Palaeolithic radiocarbon chronology: Quantifying our confidence beyond
two half-lives. J. Archaeol. Sci. 30, 1685--1693.
\url{https://doi.org/10.1016/S0305-4403(03)00070-0}

\bibitem[\citeproctext]{ref-PlompEtAl2022}
Plomp, E., Stantis, C., James, H.F., Cheung, C., Snoeck, C., Kootker,
L., Kharobi, A., Borges, C., Moreiras Reynaga, D.K., Pospieszny, Ł.,
Fulminante, F., Stevens, R., Alaica, A.K., Becker, A., de Rochefort, X.,
Salesse, K., 2022. The {IsoArcH} initiative: {Working} towards an open
and collaborative isotope data culture in bioarchaeology. Data in Brief
45, 108595. \url{https://doi.org/10.1016/j.dib.2022.108595}

\bibitem[\citeproctext]{ref-PratesEtAl2020}
Prates, L., Politis, G.G., Perez, S.I., 2020. Rapid radiation of humans
in {South America} after the last glacial maximum: {A} radiocarbon-based
study. PLOS ONE 15, e0236023.
\url{https://doi.org/10.1371/journal.pone.0236023}

\bibitem[\citeproctext]{ref-R2024}
R Core Team, 2024. \href{https://www.R-project.org/}{R: A language and
environment for statistical computing}. R Foundation for Statistical
Computing, Vienna, Austria.

\bibitem[\citeproctext]{ref-RabinowitzEtAl2016}
Rabinowitz, A., Shaw, R., Buchanan, S., Golden, P., Kansa, E., 2016.
Making {Sense} of the {Ways We Make Sense} of the {Past}: {The Periodo
Project}. Bulletin of the Institute of Classical Studies 59, 42--55.
\url{https://doi.org/10.1111/j.2041-5370.2016.12037.x}

\bibitem[\citeproctext]{ref-RadiocarbonLabList}
Radiocarbon, 2024. Radiocarbon {Lab Information} {[}WWW Document{]}. URL
\url{https://radiocarbon.webhost.uits.arizona.edu/laboratories}
(accessed 11.25.2024).

\bibitem[\citeproctext]{ref-Raetzel-Fabian1999}
Raetzel-Fabian, D., 1999. Editorial. Jungsteinsite.

\bibitem[\citeproctext]{ref-ReingruberThissen2005}
Reingruber, A., Thissen, L., 2005. {14C} database for the {Aegean}
catchment ({Eastern Greece}, southern {Balkans} and western {Turkey})
10,000--5500 cal {BC}, in: Lichter, C. (Ed.), How {Did Farming Reach
Europe}?, Byzas. Ege Yayinlar, Istanbul, pp. 295--327.

\bibitem[\citeproctext]{ref-ReiterEtAl2024}
Reiter, S.S., Staniuk, R., Kolář, J., Bulatović, J., Rose, H.A.,
Ryabogina, N.E., Speciale, C., Schjerven, N., Paulsson, B.S., Lee,
V.Y.K., Canteri, E., Revill, A., Dahlberg, F., Sabatini, S., Frei, K.M.,
Racimo, F., Ivanova-Bieg, M., Traylor, W., Kate, E.J., Derenne, E.,
Frank, L., Woodbridge, J., Fyfe, R., Shennan, S., Kristiansen, K.,
Thomas, M.G., Timpson, A., 2024. The BIAD standards: Recommendations for
archaeological data publication and insights from the big
interdisciplinary archaeological database. Open Archaeology 10,
20240015. \url{https://doi.org/doi:10.1515/opar-2024-0015}

\bibitem[\citeproctext]{ref-RichterEtAl2017}
Richter, T., Arranz-Otaegui, A., Yeomans, L., Boaretto, E., 2017. High
{Resolution AMS Dates} from {Shubayqa} 1, northeast {Jordan Reveal
Complex Origins} of {Late Epipalaeolithic Natufian} in the {Levant}.
Sci. Rep. 7, 17025. \url{https://doi.org/10.1038/s41598-017-17096-5}

\bibitem[\citeproctext]{ref-RADONB2024}
Rinne, C., Kneisel, J., Hinz, M., Furholt, M., Krischke, N., Müller, J.,
Raetzel-Fabian, D., Rodens, M., Sjögren, K.-G., Vandkilde, H., Wotzka,
H.-P., 2024. Rado.{NB}.

\bibitem[\citeproctext]{ref-RirisEtAl2024}
Riris, P., Silva, F., Crema, E., Palmisano, A., Robinson, E., Siegel,
P.E., French, J.C., Jørgensen, E.K., Maezumi, S.Y., Solheim, S., Bates,
J., Davies, B., Oh, Y., Ren, X., 2024. Frequent disturbances enhanced
the resilience of past human populations. Nature 1--6.
\url{https://doi.org/10.1038/s41586-024-07354-8}

\bibitem[\citeproctext]{ref-Roalkvam2023}
Roalkvam, I., 2023. Shoredate: {An R} package for shoreline dating
coastal {Stone Age} sites. Journal of Open Source Software 8, 5337.
\url{https://doi.org/10.21105/joss.05337}

\bibitem[\citeproctext]{ref-Roe2024}
Roe, J., 2024. \href{https://github.com/joeroe/rintchron}{Rintchron:
Interface to IntChron}.

\bibitem[\citeproctext]{ref-SchmidEtAl2019}
Schmid, C., Seidensticker, D., Hinz, M., 2019. {c14bazAAR}: An r package
for downloading and preparing C14 dates from different source databases.
{Journal of Open Source Software} 4, 1914.
\url{https://doi.org/10.21105/joss.01914}

\bibitem[\citeproctext]{ref-SilvaSteele2014}
Silva, F., Steele, J., 2014. New methods for reconstructing geographical
effects on dispersal rates and routes from large-scale radiocarbon
databases. Journal of Archaeological Science 52, 609--620.
\url{https://doi.org/10.1016/j.jas.2014.04.021}

\bibitem[\citeproctext]{ref-Suess1967}
Suess, H.E., 1967.
\href{https://www.osti.gov/etdeweb/biblio/20895312}{Bristlecone pine
calibration of the radiocarbon time scale from 4100 {B}.{C}. To 1500
{B}.{C}}.

\bibitem[\citeproctext]{ref-TenopirEtAl2011}
Tenopir, C., Allard, S., Douglass, K., Aydinoglu, A.U., Wu, L., Read,
E., Manoff, M., Frame, M., 2011. Data {Sharing} by {Scientists}:
{Practices} and {Perceptions}. PLOS ONE 6, e21101.
\url{https://doi.org/10.1371/journal.pone.0021101}

\bibitem[\citeproctext]{ref-Thomsen1836}
Thomsen, C.J., 1836. Ledetrad til {Nordiske Oldkindighed}. Copenhagen.

\bibitem[\citeproctext]{ref-VerborghEtAl2015}
Verborgh, R., van Hooland, S., Cope, A.S., Chan, S., Mannens, E., Van de
Walle, R., 2015. The fallacy of the multi-{API} culture: {Conceptual}
and practical benefits of {Representational State Transfer} ({REST}).
Journal of Documentation 71, 233--252.
\url{https://doi.org/10.1108/JD-07-2013-0098}

\bibitem[\citeproctext]{ref-Vermeersch2020}
Vermeersch, P.M., 2020. Radiocarbon {Palaeolithic Europe} database: {A}
regularly updated dataset of the radiometric data regarding the
{Palaeolithic} of {Europe}, {Siberia} included. Data Brief 31, 105793.
\url{https://doi.org/10.1016/j.dib.2020.105793}

\bibitem[\citeproctext]{ref-Voorburg2012}
Voorburg, R., 2012. Archaeological {Atlas} of {Antiquity} - {Vici}.org
{[}WWW Document{]}. URL \url{https://vici.org/} (accessed 11.25.2024).

\bibitem[\citeproctext]{ref-Whitley2012}
Whitley, D.S., 2012. In {Suspect Terrain}: {Dating Rock Engravings}, in:
A {Companion} to {Rock Art}. John Wiley \& Sons, Ltd, pp. 605--624.
\url{https://doi.org/10.1002/9781118253892.ch34}

\bibitem[\citeproctext]{ref-WilliamsEtAl2014}
Williams, A., Ulm, S., Smith, M., Reid, J., 2014. {AustArch}: {A
Database} of {14C} and {Non-14C Ages} from {Archaeological Sites} in
{Australia} - {Composition}, {Compilation} and {Review} ({Data Paper}).
Internet Archaeology. \url{https://doi.org/10.11141/ia.36.6}

\bibitem[\citeproctext]{ref-WilliamsEtAl2018}
Williams, J.W., Grimm, E.C., Blois, J.L., Charles, D.F., Davis, E.B.,
Goring, S.J., Graham, R.W., Smith, A.J., Anderson, M., Arroyo-Cabrales,
J., Ashworth, A.C., Betancourt, J.L., Bills, B.W., Booth, R.K.,
Buckland, P.I., Curry, B.B., Giesecke, T., Jackson, S.T., Latorre, C.,
Nichols, J., Purdum, T., Roth, R.E., Stryker, M., Takahara, H., 2018.
The {Neotoma Paleoecology Database}, a multiproxy, international,
community-curated data resource. Quaternary Research 89, 156--177.
\url{https://doi.org/10.1017/qua.2017.105}

\bibitem[\citeproctext]{ref-Worsaae1843}
Worsaae, J.J.A., 1843. Danmarks Oldtid oplyst ved Oldsager og Gravhøie.
Selskabet for Trykkefrihedens rette Brug, Copenhagen.

\end{CSLReferences}

\end{document}

### 2024-12-05 paper.qmd
Add data availability statement and license

@@
     orcid: 0000-0002-1011-1244
-    email: joseph.a.roe@unibe.ch
+    email: joeroe@hey.com
     corresponding: true
@@
       name: Journal of Computer Applications in Archaeology
-      formatting: doubleblind
+      formatting: review
   # Pretty but broken:
@@
 The source code for all the software components of the system are available online (at <https://github.com/xronos-ch>) under open licenses.
-The databases of the instance at <https://xronos.ch> are also archived with Zenodo (<link omitted for blind review><!-- TODO: link -->).
+The databases of the instance at <https://xronos.ch> are also archived with Zenodo (*link omitted for blind review*<!-- TODO: link -->).
 With these two resources we reduce XRONOS' 'bus factor'; if we are not able to continue operating XRONOS, somebody else can fully recreate it.
@@
 This work was funded by the Swiss National Science Foundation ([SNSF Project #198152](https://data.snf.ch/grants/grant/198153)) and the University of Bern (UniBE Initiator Grant 2019, Caroline Heitz, Project 'Time and Temporality').
+
+# Data accessibility statement
+
+The data and R code used to produce our analysis of the state of the art in radiocarbonc compilation, including all the figures presented here, can be accessed via Zenodo at (<!-- TODO: add link -->*link omitted for blind review, included in submission materials*)
+
+The database and software described in this paper is open source and can be accessed at <https://github.com/xronos-ch>.

### 2024-12-05 paper.tex
Add data availability statement and license

@@
   number,
-  doubleblind]{elsarticle}
+  review]{elsarticle}
 
@@
 }
- \ead{joseph.a.roe@unibe.ch} 
+ \ead{joeroe@hey.com} 
 \author[2]{Clemens Schmid%
@@
 of the instance at \url{https://xronos.ch} are also archived with Zenodo
-(). With these two resources we reduce XRONOS' `bus factor'; if we are
-not able to continue operating XRONOS, somebody else can fully recreate
-it. Equally importantly, we enable the `right to fork' should others
-wish to take the software and/or database in another direction. But
-aside from these extreme scenarios, the long-term sustainability of
-XRONOS is contingent on the existence of a community of scholars that
-use and contribute back to it.
+(\emph{link omitted for blind review}). With these two resources we
+reduce XRONOS' `bus factor'; if we are not able to continue operating
+XRONOS, somebody else can fully recreate it. Equally importantly, we
+enable the `right to fork' should others wish to take the software
+and/or database in another direction. But aside from these extreme
+scenarios, the long-term sustainability of XRONOS is contingent on the
+existence of a community of scholars that use and contribute back to it.
 
@@
 Project `Time and Temporality').
+
+\section{Data accessibility
+statement}\label{data-accessibility-statement}
+
+The data and R code used to produce our analysis of the state of the art
+in radiocarbonc compilation, including all the figures presented here,
+can be accessed via Zenodo at (\emph{link omitted for blind review,
+included in submission materials})
+
+The database and software described in this paper is open source and can
+be accessed at \url{https://github.com/xronos-ch}.
 

### 2024-12-05 paper.qmd
Fix table width

@@
   cols_label(issue_code = "Issue", n = "N", issue_description = "Description") |>
-  cols_width(everything() ~ pct(33)) |> # TODO: fix...
+  cols_width(
+    issue_code ~ pct(30),
+    n ~ pct(20),
+    issue_description ~ pct(50)
+  ) |>
   tab_options(

### 2024-12-05 paper.qmd
Add missing references

@@
   - data/c14_datasets.bib
+nocite: | # workaround for lack of table references
+  @Al-Abyadh-Balghelam-Dalma-JebelEtAl, @Alcantara2021, @AndersonEtAl2010, @BANADORA, @BarceloAlvarezEtAl2013, @Benz2010, @BirdEtAl2022, @BohnerSchyle2004, @BronkRamseyEtAl2009, @Capriles2023, @CapuzzoEtAl2014, @ClistEtAl2023, @CochraneEtAl2021, @CourtneyMustaphi2016, @CremaEtAl2016, @DeeEtAl2012, @DeeEtAl2013, @dErricoEtAl2011, @deSaulieuEtAl2017, @Diaz-RodriguezEtAl2023, @DouglassEtAl2019, @FlohrEtAl2016, @GajewskiEtAl2011, @GarnettEtAl2023, @GayoEtAl2015, @GillespieEtAl1984, @GregoriodeSouza2020, @GrossmannEtAl2023, @HoebeEtAl2023, @HoggarthEtAl2021, @HuetEtAl2022, @HuetEtAl2024, @IRDD, @KatsianisEtAl2020, @KellyEtAl2022, @KimEtAl2021, @Kudo2018, @KudoEtAl2023, @LinnenluckeEtAl2023, @Lipo2020, @LoftusEtAl2019, @LucariniEtAl2020, @ManningEtAl2016, @Martinez-GrauEtAl2021, @McFadgenEtAl2000, @MichczynskiEtAl1995, @PalmisanoEtAl2017, @PalmisanoEtAl2022, @PalmisanoEtAl2022a, @Pardo-GordoEtAl2023, @Perrin2019, @PetcheyEtAl2022, @Rademaker2024, @RademakerEtAl2013, @RADON, @RADONB, @RADONB2024, @Raetzel-Fabian1999, @RamseyEtAl2010, @ReingruberThissen2005, @ReingruberThissen2009, @ReingruberThissen2017, @Riris2021, @SeidenstickerSchmid2021, @TanindiErdogu2005, @UriarteGonzalezEtAl2017, @VanStrydonckDeRoock, @Vermeersch2024, @WangEtAl2014, @Weninger2022, @WilliamsEtAl2008, @WilliamsEtAl2014, @WilliamsSmith2012, @WilliamsSmith2013
 format:
@@
 The database and software described in this paper is open source and can be accessed at <https://github.com/xronos-ch>.
+
+# References

### 2024-12-05 paper.tex
Fix table width
Add missing references

@@
 
-\begingroup
-\fontsize{9.8pt}{11.7pt}\selectfont
-
-\begin{longtable}{>{\raggedright\arraybackslash}p{\dimexpr 0.33\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 0.33\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedright\arraybackslash}p{\dimexpr 0.33\linewidth -2\tabcolsep-1.5\arrayrulewidth}}
+\begin{table}
 
@@
 
-\tabularnewline
-
+\centering{
+
+\fontsize{9.8pt}{11.7pt}\selectfont
+\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 0.30\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 0.20\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedright\arraybackslash}p{\dimexpr 0.50\linewidth -2\tabcolsep-1.5\arrayrulewidth}}
 \toprule
@@
 \midrule\addlinespace[2.5pt]
-\multicolumn{3}{>{\raggedright\arraybackslash}m{1.32\linewidth}}{Sites} \\[2.5pt] 
+\multicolumn{3}{l}{Sites} \\[2.5pt] 
 \midrule\addlinespace[2.5pt]
@@
 \midrule\addlinespace[2.5pt]
-\multicolumn{3}{>{\raggedright\arraybackslash}m{1.32\linewidth}}{Samples} \\[2.5pt] 
+\multicolumn{3}{l}{Samples} \\[2.5pt] 
 \midrule\addlinespace[2.5pt]
@@
 \midrule\addlinespace[2.5pt]
-\multicolumn{3}{>{\raggedright\arraybackslash}m{1.32\linewidth}}{Taxons} \\[2.5pt] 
+\multicolumn{3}{l}{Taxons} \\[2.5pt] 
 \midrule\addlinespace[2.5pt]
@@
 \midrule\addlinespace[2.5pt]
-\multicolumn{3}{>{\raggedright\arraybackslash}m{1.32\linewidth}}{Radiocarbon dates} \\[2.5pt] 
+\multicolumn{3}{l}{Radiocarbon dates} \\[2.5pt] 
 \midrule\addlinespace[2.5pt]
@@
 \midrule\addlinespace[2.5pt]
-\multicolumn{3}{>{\raggedright\arraybackslash}m{1.32\linewidth}}{Bibliographic references} \\[2.5pt] 
+\multicolumn{3}{l}{Bibliographic references} \\[2.5pt] 
 \midrule\addlinespace[2.5pt]
@@
 \bottomrule
-
-\end{longtable}
-
-\endgroup
+\end{tabular*}
+
+}
+
+\end{table}%
 
@@
 
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
+
 \phantomsection\label{refs}
@@
 
+\bibitem[\citeproctext]{ref-Al-Abyadh-Balghelam-Dalma-JebelEtAl}
+Al-Abyadh-Balghelam-Dalma-Jebel, A., Dhanna-Marawah-Marawah, M.R., Yas,
+R.G.-R.-S.B., n.d. {ADIAS Radiocarbon Archive}.
+
+\bibitem[\citeproctext]{ref-Alcantara2021}
+Alcántara, A., 2021. Entre tepalcates. {Hacia} una comprensi{ó}n
+historiogr{á}fica de la dataci{ó}n arqueom{é}trica en la arqueolog{í}a
+mexicana (1950-2020) (PhD thesis).
+\url{https://doi.org/10.13140/RG.2.2.24447.74400/2}
+
 \bibitem[\citeproctext]{ref-AlcantaraPedrozainpress}
@@
 
+\bibitem[\citeproctext]{ref-AndersonEtAl2010}
+Anderson, D.G., Miller, D.S., Yerka, S.J., Gillam, J.C., Johanson, E.N.,
+Anderson, D.T., Goodyear, A.C., Smallwood, A.M., 2010.
+\href{https://www.jstor.org/stable/40914542}{{PIDBA} ({Paleoindian
+Database} of the {Americas}) 2010: {Current Status} and {Findings}}.
+Archaeology of Eastern North America 38, 63--89.
+
 \bibitem[\citeproctext]{ref-ArnoldLibby1951}
@@
 London. \url{https://doi.org/10.4324/9781315748689}
+
+\bibitem[\citeproctext]{ref-BANADORA}
+Banque {Nationale} de {Donn{é}es} {Radiocarbone} ({BANADORA}), n.d.
+
+\bibitem[\citeproctext]{ref-BarceloAlvarezEtAl2013}
+Barceló Álvarez, J.A., Bogdanović, I., Capuzzo, G., 2013. New
+developments of tecnological and archaeological data bases: {The}
+example of radiocarbon dating, in: Arqueolog{í}a: {Actas} Del {Primer
+Congreso Internacional} de {Buenas Pr{á}cticas} En {Patrimonio Mundial},
+2013, {ISBN} 978-84-941030-9-4, p{á}gs. 733-749. JAS Arqueolog{í}a, pp.
+733--749.
 
@@
 \url{https://doi.org/10.1017/S0002731600047545}
+
+\bibitem[\citeproctext]{ref-Benz2010}
+Benz, M., 2010. {PPND} -- {The Platform} for {Neolithic Radiocarbon
+Dates}.
 
@@
 
+\bibitem[\citeproctext]{ref-BohnerSchyle2004}
+Böhner, U., Schyle, D., 2004. Radiocarbon {CONTEXT} database.
+
 \bibitem[\citeproctext]{ref-Brogger1905}
@@
 \url{https://doi.org/10.1017/RDC.2019.21}
+
+\bibitem[\citeproctext]{ref-BronkRamseyEtAl2009}
+Bronk Ramsey, C., Higham, T.F.G., Brock, F., Baker, D., Ditchfield, P.,
+2009. Radiocarbon {Dates} from the {Oxford Ams System}: {Archaeometry
+Da}TEList 33. Archaeometry 51, 323--349.
+\url{https://doi.org/10.1111/j.1475-4754.2008.00457.x}
 
@@
 
+\bibitem[\citeproctext]{ref-Capriles2023}
+Capriles, J.M., 2023. The {Bolivian Radiocarbon Database}: {A
+Countrywide Compilation} of {Radiocarbon Dates}. Journal of Open
+Archaeology Data 11. \url{https://doi.org/10.5334/joad.104}
+
+\bibitem[\citeproctext]{ref-CapuzzoEtAl2014}
+Capuzzo, G., Boaretto, E., Barceló, J.A., 2014. {EUBAR}: {A Database} of
+{14C Measurements} for the {European Bronze Age}. {A Bayesian Analysis}
+of {14C-Dated Archaeological Contexts} from {Northern Italy} and
+{Southern France}. Radiocarbon 56, 851--869.
+\url{https://doi.org/10.2458/56.17453}
+
+\bibitem[\citeproctext]{ref-IRDD}
+Chapple, R.M., 2019. Catalogue of {Radiocarbon Determinations} \&
+{Dendrochronology Dates} ({August} 2019 release).
+
 \bibitem[\citeproctext]{ref-ChaputGajewski2016}
@@
 
+\bibitem[\citeproctext]{ref-CochraneEtAl2021}
+Cochrane, E.E., Rieth, T.M., Filimoehala, D., 2021. The first
+quantitative assessment of radiocarbon chronologies for initial pottery
+in {Island Southeast Asia} supports multi-directional {Neolithic}
+dispersal. PLOS ONE 16, e0251407.
+\url{https://doi.org/10.1371/journal.pone.0251407}
+
+\bibitem[\citeproctext]{ref-CourtneyMustaphi2016}
+Courtney Mustaphi, C., 2016. Radiocarbon dates from eastern {Africa} in
+the {CARD2}.0 format. \url{https://doi.org/10.7910/DVN/NJLNRJ}
+
 \bibitem[\citeproctext]{ref-Crema2024}
@@
 
+\bibitem[\citeproctext]{ref-CremaEtAl2016}
+Crema, E.R., Habu, J., Kobayashi, K., Madella, M., 2016. Summed
+{Probability Distribution} of {14C Dates Suggests Regional Divergences}
+in the {Population Dynamics} of the {Jomon Period} in {Eastern Japan}.
+PLoS One 11, e0154809.
+\url{https://doi.org/10.1371/journal.pone.0154809}
+
+\bibitem[\citeproctext]{ref-dErricoEtAl2011}
+d'Errico, F., Banks, W.E., Vanhaeren, M., Laroulandie, V., Langlais, M.,
+2011. {PACEA Geo-Referenced Radiocarbon Database}. PaleoAnthropology
+2011, 1--12.
+
 \bibitem[\citeproctext]{ref-DanielsEtAl1953}
@@
 
+\bibitem[\citeproctext]{ref-deSaulieuEtAl2017}
+de Saulieu, G., Oslisly, R., Nlend, P., Garcin, Y., Chevillotte, H.,
+Kuete, N., 2017. Plateforme des datations arch{é}ologiques d'{Afrique
+Centrale}. Version 2, 03.
+
+\bibitem[\citeproctext]{ref-DeeEtAl2013}
+Dee, M., Wengrow, D., Shortland, A., Stevenson, A., Brock, F., Girdland
+Flink, L., Bronk Ramsey, C., 2013. An absolute chronology for early
+{Egypt} using radiocarbon dating and {Bayesian} statistical modelling.
+Proceedings of the Royal Society A: Mathematical, Physical and
+Engineering Sciences 469, 20130395.
+\url{https://doi.org/10.1098/rspa.2013.0395}
+
+\bibitem[\citeproctext]{ref-DeeEtAl2012}
+Dee, M.W., Rowland, J.M., Higham, T.F.G., Shortland, A.J., Brock, F.,
+Harris, S.A., Ramsey, C.B., 2012. Synchronising radiocarbon dating and
+the {Egyptian} historical chronology by improved sample selection.
+Antiquity 86, 868--883. \url{https://doi.org/10.1017/S0003598X00047979}
+
+\bibitem[\citeproctext]{ref-Diaz-RodriguezEtAl2023}
+Díaz-Rodríguez, M., Nielsen, T.K., Maier, A., Riede, F., 2023. An
+{Annotated Compilation} of {Chronometric Dates} for the {Middle-Upper
+Palaeolithic Transition} (45-30 ka {BP}) in {Northern Iberia} ({Spain}).
+Journal of Open Archaeology Data 11.
+\url{https://doi.org/10.5334/joad.113}
+
 \bibitem[\citeproctext]{ref-Douglass1929}
@@
 tree rings. National Geographic 56, 737--770.
+
+\bibitem[\citeproctext]{ref-DouglassEtAl2019}
+Douglass, K., Hixon, S., Wright, H.T., Godfrey, L.R., Crowley, B.E.,
+Manjakahery, B., Rasolondrainy, T., Crossland, Z., Radimilahy, C., 2019.
+A critical review of radiocarbon dates clarifies the human settlement of
+{Madagascar}. Quat. Sci. Rev. 221, 105878.
+\url{https://doi.org/10.1016/j.quascirev.2019.105878}
 
@@
 
+\bibitem[\citeproctext]{ref-FlohrEtAl2016}
+Flohr, P., Fleitmann, D., Matthews, R., Matthews, W., Black, S., 2016.
+Evidence of resilience to past climate change in {Southwest Asia}:
+{Early} farming communities and the 9.2 and 8.2 ka events. Quat. Sci.
+Rev. 136, 23--39. \url{https://doi.org/10.1016/j.quascirev.2015.06.022}
+
 \bibitem[\citeproctext]{ref-Ford1962}
@@
 
+\bibitem[\citeproctext]{ref-GarnettEtAl2023}
+Garnett, M.H., Bryant, C.L., Freeman, S.P.H.T., Gulliver, P., Ascough,
+P.L., 2023. A {DATABASE OF NERC RADIOCARBON MEASUREMENTS DETERMINED BY
+ACCELERATOR MASS SPECTROMETRY}. Radiocarbon 1--4.
+\url{https://doi.org/10.1017/RDC.2023.18}
+
+\bibitem[\citeproctext]{ref-GayoEtAl2015}
+Gayo, E.M., Latorre, C., Santoro, C.M., 2015. Timing of occupation and
+regional settlement patterns revealed by time-series analyses of an
+archaeological radiocarbon database for the {South-Central Andes}
+(16{\(^\circ\)}--25{\(^\circ\)}{S}). Quaternary International,
+Palaeodemography in {Southern South America} 356, 4--14.
+\url{https://doi.org/10.1016/j.quaint.2014.09.076}
+
 \bibitem[\citeproctext]{ref-GBIF2023}
@@
 \url{https://doi.org/10.15468/39OMEI}
+
+\bibitem[\citeproctext]{ref-GillespieEtAl1984}
+Gillespie, R., Gowlett, J.a.J., Hall, E.T., Hedges, R.E.M., 1984.
+Radiocarbon {Measurement} by {Accelerator Mass Spectrometry}: {An Early
+Selection} of {Dates}. Archaeometry 26, 15--20.
+\url{https://doi.org/10.1111/j.1475-4754.1984.tb00313.x}
+
+\bibitem[\citeproctext]{ref-GregoriodeSouza2020}
+Gregorio de Souza, J., 2020. Rxpand: {Radiocarbon} dates for the spread
+of farming and ceramics in tropical {South America}.
+
+\bibitem[\citeproctext]{ref-GrossmannEtAl2023}
+Großmann, R., Weinelt, M., Müller, J., 2023. Demographic dynamics
+between 5500 and 3500 {calBP} (3550--1550 {BCE}) in selected study
+regions of {Central Europe} and the role of regional climate influences.
+PLOS ONE 18, e0291956.
+\url{https://doi.org/10.1371/journal.pone.0291956}
 
@@
 
+\bibitem[\citeproctext]{ref-HoebeEtAl2023}
+Hoebe, P.W., Peeters, J.H.M., Arnoldussen, S., 2023. Parsing prehistoric
+patterns: {Prospects} and limitations of a big radiocarbon dataset for
+understanding the impact of climate on {Late Palaeolithic} and
+{Mesolithic} populations in northwest {Europe} (16--7.5~ka {calBP}).
+Journal of Archaeological Science: Reports 49, 103944.
+\url{https://doi.org/10.1016/j.jasrep.2023.103944}
+
+\bibitem[\citeproctext]{ref-HoggarthEtAl2021}
+Hoggarth, J.A., Ebert, C.E., Castelazo-Calva, V.E., 2021. {MesoRAD}: {A
+New Radiocarbon Data Set} for {Archaeological Research} in
+{Mesoamerica}. Journal of Open Archaeology Data 9, 10.
+\url{https://doi.org/10.5334/joad.83}
+
+\bibitem[\citeproctext]{ref-HuetEtAl2024}
+Huet, T., Basílio, A.C., Carvalho, A.F., Cubas, M., Gibaja, J.F.,
+López-Romero, E., Oms, F.X., Mazzucco, N., 2024. {NeoNet Atlantic}.
+{Radiocarbon Dates} for the {Late Mesolithic}/{Early Neolithic
+Transition} in the {Southern European Atlantic Coast} 12, 2.
+\url{https://doi.org/10.5334/joad.120}
+
+\bibitem[\citeproctext]{ref-HuetEtAl2022}
+Huet, T., Cubas, M., Gibaja, J.F., Oms, F.X., Mazzucco, N., 2022.
+{NeoNet Dataset}. {Radiocarbon Dates} for the {Late Mesolithic}/{Early
+Neolithic Transition} in the {North Central-Western Mediterranean
+Basin}. Journal of Open Archaeology Data 10, 3.
+\url{https://doi.org/10/gpk67w}
+
 \bibitem[\citeproctext]{ref-KatsianisEtAl2020}
@@
 
+\bibitem[\citeproctext]{ref-KellyEtAl2022}
+Kelly, R.L., Mackie, M.E., Robinson, E., Meyer, J., Berry, M.,
+Boulanger, M., Codding, B.F., Freeman, J., Garland, C.J., Gingerich, J.,
+Hard, R., Haug, J., Martindale, A., Meeks, S., Miller, M., Miller, S.,
+Perttula, T., Railey, J.A., Reid, K., Scharlotta, I., Spangler, J.,
+Thomas, D.H., Thompson, V., White, A., 2022. A {New Radiocarbon
+Database} for the {Lower} 48 {States}. American Antiquity 87, 581--590.
+\url{https://doi.org/10.1017/aaq.2021.157}
+
 \bibitem[\citeproctext]{ref-KemmersMyrberg2011}
@@
 
+\bibitem[\citeproctext]{ref-KimEtAl2021}
+Kim, H., Lee, G.-A., Crema, E.R., 2021. Bayesian analyses question the
+role of climate in {Chulmun} demography. Scientific Reports 11, 23797.
+\url{https://doi.org/10.1038/s41598-021-03180-4}
+
 \bibitem[\citeproctext]{ref-Kintigh2006}
@@
 \url{https://doi.org/10.2307/281158}
+
+\bibitem[\citeproctext]{ref-Kudo2018}
+Kudo, Y., 2018. Approach for {Creating Database} of the {Radiocarbon
+Dates Published} on the {Archaeological Research Reports} in {Japan}.
+Bulletin of the National Museum of Japanese History 212, 251.
+
+\bibitem[\citeproctext]{ref-KudoEtAl2023}
+Kudo, Y., Sakamoto, M., Hakozaki, M., Stevens, C.J., Crema, E.R., 2023.
+An {Archaeological Radiocarbon Database} of {Japan}. Journal of Open
+Archaeology Data 11. \url{https://doi.org/10.5334/joad.115}
 
@@
 
+\bibitem[\citeproctext]{ref-LinnenluckeEtAl2023}
+Linnenlucke, L., Bird, M.I., Petchey, F., Alliston, G., McNiven, I.J.,
+David, B., Ulm, S., 2023. Assessing the {Reliability} of a {Small-Scale
+Legacy Radiocarbon Dataset Using Chronometric Transparency Approaches}:
+{Torres Strait Radiocarbon Database}. Journal of Open Archaeology Data
+11. \url{https://doi.org/10.5334/joad.95}
+
+\bibitem[\citeproctext]{ref-Lipo2020}
+Lipo, C., 2020. A repository of radiocarbon data for {Rapa Nui}.
+
 \bibitem[\citeproctext]{ref-LoftusEtAl2019}
@@
 
+\bibitem[\citeproctext]{ref-LucariniEtAl2020}
+Lucarini, G., Wilkinson, T., Crema, E.R., Palombini, A., Bevan, A.,
+Broodbank, C., 2020. The {MedAfriCarbon Radiocarbon Database} and {Web
+Application}. {Archaeological Dynamics} in {Mediterranean Africa}, ca.
+9600--700 {BC}. Journal of Open Archaeology Data 8.
+\url{https://doi.org/10.5334/joad.60}
+
 \bibitem[\citeproctext]{ref-Lucas2004}
@@
 Routledge.
+
+\bibitem[\citeproctext]{ref-ManningEtAl2016}
+Manning, K., Colledge, S., Crema, E., Shennan, S., Timpson, A., 2016.
+The {Cultural Evolution} of {Neolithic Europe}. {EUROEVOL Dataset} 1:
+{Sites}, {Phases} and {Radiocarbon Data}. Journal of Open Archaeology
+Data 5, 275. \url{https://doi.org/10.5334/joad.40}
 
@@
 \url{https://doi.org/10.24072/pcjournal.395}
+
+\bibitem[\citeproctext]{ref-Martinez-GrauEtAl2021}
+Martínez-Grau, H., Morell-Rovira, B., Antolín, F., 2021. Radiocarbon
+{Dates Associated} to {Neolithic Contexts} ({Ca}. 5900 -- 2000 {Cal BC})
+from the {Northwestern Mediterranean Arch} to the {High Rhine Area}.
+Journal of Open Archaeology Data 9.
+\url{https://doi.org/10.5334/joad.72}
+
+\bibitem[\citeproctext]{ref-McFadgenEtAl2000}
+McFadgen, B., Higham, T., Sparks, R.J., 2000. New {Zealand} radiocarbon
+database for archaeology. Archaeology in New Zealand 43, 297--300.
 
@@
 URL \url{https://opendefinition.org/} (accessed 11.25.2024).
+
+\bibitem[\citeproctext]{ref-PalmisanoEtAl2022}
+Palmisano, A., Bevan, A., Kabelindde, A., Roberts, N., Shennan, S.,
+2022a. {AIDA}: {Archive} of {Italian} radiocarbon {DAtes}.
+
+\bibitem[\citeproctext]{ref-PalmisanoEtAl2022a}
+Palmisano, A., Bevan, A., Lawrence, D., Shennan, S., 2022b. The {NERD
+Dataset}: {Near East Radiocarbon Dates} between 15,000 and 1,500 cal.
+Yr. {BP} 10, 2. \url{https://doi.org/10.5334/joad.90}
+
+\bibitem[\citeproctext]{ref-PalmisanoEtAl2017}
+Palmisano, A., Bevan, A., Shennan, S., 2017. Data and code for
+demographic trends in the paper "{Human} responses and non-responses to
+climatic variations during the {Last Glacial-Interglacial} transition in
+the eastern {Mediterranean}".
+\url{https://doi.org/10.14324/000.ds.1570274}
+
+\bibitem[\citeproctext]{ref-Pardo-GordoEtAl2023}
+Pardo-Gordó, S., Vidal-Matutano, P., González-Marrero, M. del C.,
+Chávez-Álvarez, M.E., 2023. The {14Canarias} web application. {An}
+interactive radiocarbon database for the {Canary Islands}. Journal of
+Open Archaeology Data 11. \url{https://doi.org/10.5334/joad.105}
+
+\bibitem[\citeproctext]{ref-Perrin2019}
+Perrin, T., 2019. \href{https://www.jstor.org/stable/26910790}{{BDA}:
+Une {Base} de {Donn{é}es Arch{é}ologique} collaborative en ligne}.
+Bulletin de la Soci{é}t{é} pr{é}historique fran{ç}aise 116, 159--162.
+
+\bibitem[\citeproctext]{ref-PetcheyEtAl2022}
+Petchey, F., Bickler, S., Hughes, L., Bunbury, M., 2022. The
+{Aotearoa}/{New Zealand Radiocarbon Database Upgrade}. Archaeology in
+New Zealand 65, 32--40.
 
@@
 
+\bibitem[\citeproctext]{ref-Rademaker2024}
+Rademaker, K., 2024. Updated {Peru} archaeological radiocarbon database,
+20,000--7000 {14C BP}. Quaternary International.
+\url{https://doi.org/10.1016/j.quaint.2024.01.012}
+
+\bibitem[\citeproctext]{ref-RademakerEtAl2013}
+Rademaker, K., Bromley, G.R.M., Sandweiss, D.H., 2013. Peru
+archaeological radiocarbon database, 13,000--7000 {14C B}.{P}.
+Quaternary International, A {Late Pleistocene}/early {Holocene}
+archaeological {14C} database for {Central} and {South America}:
+Palaeoenvironmental contexts and demographic interpretations 301,
+34--45. \url{https://doi.org/10.1016/j.quaint.2012.08.2052}
+
 \bibitem[\citeproctext]{ref-RadiocarbonLabList}
@@
 Raetzel-Fabian, D., 1999. Editorial. Jungsteinsite.
+
+\bibitem[\citeproctext]{ref-RamseyEtAl2010}
+Ramsey, C.B., Dee, M.W., Rowland, J.M., Higham, T.F.G., Harris, S.A.,
+Brock, F., Quiles, A., Wild, E.M., Marcus, E.S., Shortland, A.J., 2010.
+Radiocarbon-{Based Chronology} for {Dynastic Egypt}. Science 328,
+1554--1557. \url{https://doi.org/10.1126/science.1189395}
+
+\bibitem[\citeproctext]{ref-ReingruberThissen2017}
+Reingruber, A., Thissen, L., 2017. The {14SEA Project}: {A 14C} database
+for {Southeast Europe} and {Anatolia} (10,000--3000 {calBC}).
+
+\bibitem[\citeproctext]{ref-ReingruberThissen2009}
+Reingruber, A., Thissen, L., 2009. Depending on {14C Data}:
+{Chronological Frameworks} in the {Neolithic} and {Chalcolithic} of
+{Southeastern Europe}. Radiocarbon 51, 751--770.
+\url{https://doi.org/10.1017/S0033822200056071}
 
@@
 
+\bibitem[\citeproctext]{ref-Riris2021}
+Riris, P., 2021. Carribean-{14C}: {A} compilation of 2147 anthropogenic
+radiocarbon ({14C}) dates for the {Caribbean} region from 504 sites
+across 57 islands.
+
 \bibitem[\citeproctext]{ref-RirisEtAl2024}
@@
 
+\bibitem[\citeproctext]{ref-SeidenstickerSchmid2021}
+Seidensticker, D., Schmid, C., 2021. Dirkseidensticker/{aDRAC}: V2.0
+{Pikunda}. \url{https://doi.org/10.5281/zenodo.4530717}
+
 \bibitem[\citeproctext]{ref-SilvaSteele2014}
@@
 
+\bibitem[\citeproctext]{ref-TanindiErdogu2005}
+Tanindi, O., Erdoğu, B., 2005. Recording the past: Towards an inventory
+of the archaeological sites of {Turkey}. Conservation and Management of
+Archaeological Sites 7, 125--130.
+\url{https://doi.org/10.1179/135050305793137468}
+
 \bibitem[\citeproctext]{ref-TenopirEtAl2011}
@@
 Thomsen, C.J., 1836. Ledetrad til {Nordiske Oldkindighed}. Copenhagen.
+
+\bibitem[\citeproctext]{ref-UriarteGonzalezEtAl2017}
+Uriarte González, A., Fernández Freire, C., Fraguas Bravo, A., Castañeda
+Clemente, N., Capdevila Montes, E., Salas Tovar, E., Gilman, A., Bosque
+González, I. del, Vicent García, J.M., 2017. {IDEArq-C14: Una
+Infraestructura de Datos Espaciales para la cronolog{í}a
+radiocarb{ó}nica de la Prehistoria Reciente ib{é}rica}.
+\url{https://doi.org/10.13039/501100003329}
+
+\bibitem[\citeproctext]{ref-VanStrydonckDeRoock}
+Van Strydonck, M., De Roock, E., n.d. Radiocarbon and stable isotope
+measurements. Royal Institute for Cultural Heritage.
 
@@
 
+\bibitem[\citeproctext]{ref-Vermeersch2024}
+Vermeersch, P.M., 2024. Radiocarbon {Palaeolithic Europe Database}.
+
 \bibitem[\citeproctext]{ref-Vermeersch2020}
@@
 
+\bibitem[\citeproctext]{ref-WangEtAl2014}
+Wang, C., Lu, H., Zhang, J., Gu, Z., He, K., 2014. Prehistoric
+demographic fluctuations in {China} inferred from radiocarbon data and
+their linkage with climate change over the past 50,000 years. Quaternary
+Science Reviews 98, 45--59.
+\url{https://doi.org/10.1016/j.quascirev.2014.05.015}
+
+\bibitem[\citeproctext]{ref-Weninger2022}
+Weninger, B., 2022. {CalPal}.
+\url{https://doi.org/10.5281/ZENODO.7215741}
+
 \bibitem[\citeproctext]{ref-Whitley2012}
@@
 \url{https://doi.org/10.1002/9781118253892.ch34}
+
+\bibitem[\citeproctext]{ref-WilliamsSmith2013}
+Williams, A.N., Smith, M., 2013.
+\href{https://www.jstor.org/stable/23621983}{{AustArch3}: {A} database
+of 14 {C} and luminescence ages from archaeological sites in southern
+{Australia}}. Australian Archaeology 102--102.
+
+\bibitem[\citeproctext]{ref-WilliamsSmith2012}
+Williams, A.N., Smith, M.A., 2012.
+\href{https://www.jstor.org/stable/23621556}{{AustArch2}: {A} database
+of 14 {C} and luminescence ages from archaeological sites in the {Top
+End}}. Australian Archaeology 146--146.
+
+\bibitem[\citeproctext]{ref-WilliamsEtAl2008}
+Williams, A.N., Smith, M.A., Turney, C.S.M., Cupper, M.L., 2008.
+\href{https://www.jstor.org/stable/20839288}{{AustArch1}: {A Database}
+of {\(^{14}\)}{C} and {Luminescence Ages} from {Archaeological Sites} in
+the {Australian Arid Zone}}. Australian Archaeology 99--99.
 

### 2024-12-05 paper.tex
Use title case
Per JCAA guidelines

@@
 \hypersetup{
-  pdftitle={XRONOS: an open data infrastructure for archaeological chronology},
+  pdftitle={XRONOS: An Open Data Infrastructure for Archaeological Chronology},
   pdfauthor={Joe Roe; Clemens Schmid; Setareh Ebrahimiabareghi; Caroline Heitz; Martin Hinz},
@@
 \begin{frontmatter}
-\title{XRONOS: an open data infrastructure for archaeological
-chronology}
+\title{XRONOS: An Open Data Infrastructure for Archaeological
+Chronology}
 \author[1]{Joe Roe%
@@
 
-\section{State of the art}\label{state-of-the-art}
+\section{State of the Art}\label{state-of-the-art}
 
@@
 
-\section{Funding statement}\label{funding-statement}
+\section{Funding Statement}\label{funding-statement}
 
@@
 
-\section{Data accessibility
-statement}\label{data-accessibility-statement}
+\section{Data Accessibility
+Statement}\label{data-accessibility-statement}
 

### 2024-12-05 paper.tex
Save figures in figures/

@@
 
-\begingroup
-\fontsize{9.8pt}{11.7pt}\selectfont
-
-\begin{longtable}{>{\raggedright\arraybackslash}p{\dimexpr 0.40\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 0.15\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 0.15\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedright\arraybackslash}p{\dimexpr 0.30\linewidth -2\tabcolsep-1.5\arrayrulewidth}}
+\begin{table}
 
@@
 
-\tabularnewline
-
+\centering{
+
+\fontsize{9.8pt}{11.7pt}\selectfont
+\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 0.40\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 0.15\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 0.15\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedright\arraybackslash}p{\dimexpr 0.30\linewidth -2\tabcolsep-1.5\arrayrulewidth}}
 \toprule
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-c14-datasets-time-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/fig-c14-datasets-time-1.pdf}}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-c14-datasets-map-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/fig-c14-datasets-map-1.pdf}}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-c14-global-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/fig-c14-global-1.pdf}}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-data-model-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{figures/fig-data-model-1.pdf}}
 

### 2024-12-05 paper.qmd
Use title case
Per JCAA guidelines
Save figures in figures/
Generate double blind version
Add Zenodo DOI

@@
 ---
-title: "XRONOS: an open data infrastructure for archaeological chronology"
+title: "XRONOS: An Open Data Infrastructure for Archaeological Chronology"
 short-title: XRONOS
@@
       name: Journal of Computer Applications in Archaeology
-      formatting: review
+      formatting: doubleblind
   # Pretty but broken:
@@
   cache: true
+knitr:
+  opts_chunk:
+    fig.path: figures/
 ---
@@
 
-# State of the art
+# State of the Art
 
@@
 
-# Funding statement
+# Funding Statement
 
@@
 
-# Data accessibility statement
+# Data Accessibility Statement
 
-The data and R code used to produce our analysis of the state of the art in radiocarbonc compilation, including all the figures presented here, can be accessed via Zenodo at (<!-- TODO: add link -->*link omitted for blind review, included in submission materials*)
+The data and R code used to produce our analysis of the state of the art in radiocarbon compilation, including all the figures presented here, can be accessed via Zenodo at <https://doi.org/10.5281/zenodo.14282598>.
 

### 2024-12-05 paper.tex
Generate double blind version
Add Zenodo DOI

@@
   number,
-  review]{elsarticle}
+  doubleblind]{elsarticle}
 
@@
 The data and R code used to produce our analysis of the state of the art
-in radiocarbonc compilation, including all the figures presented here,
-can be accessed via Zenodo at (\emph{link omitted for blind review,
-included in submission materials})
+in radiocarbon compilation, including all the figures presented here,
+can be accessed via Zenodo at
+\url{https://doi.org/10.5281/zenodo.14282598}.
 

### 2025-04-04 paper.qmd
Rerender as preprint
Incorporate @nevrome's suggestions
Did what I could with fig-c14-datasets-map... the problem is that adding
extra elements (i.e. scales) causes patchwork to squeeze the plot area
even more.

@@
   - id: mpg
-    name: Max Planck Institute for Geoanthropology
+    name: Max Planck Institute for Evolutionary Anthropology
     url: https://www.shh.mpg.de/
@@
       name: Journal of Computer Applications in Archaeology
-      formatting: doubleblind
+      formatting: preprint
   # Pretty but broken:
@@
 In the remainder of this paper, we describe the concept and implementation of XRONOS in relation to the state of the art in open chronometric data in archaeology, and evaluate our progress in achieving these goals as of writing.
-Since we envisage both XRONOS as a dataset and XRONOS as software to be continually developing resources, the description here should be read as a 'snapshot' of the project as of writing rather than its final state.
+Since we envisage both XRONOS as a dataset and XRONOS as software to be continually developing resources, the description here should be read as a 'snapshot' of the project as of version 1.2.0.
 
@@
        caption = glue("* Excludes {english(n_c14_datasets_sans_publication_year)} datasets with an unknown year of publication")) +
-  theme_minimal_vgrid()
+  theme_minimal_vgrid(font_size = 11)
 ```
 
-The number of available compilations has increased exponentially since around 1995 (@fig-c14-datasets-time).
+The number of available compilations has increased signifcantly since around 1995 (@fig-c14-datasets-time).
 The first generation came around the turn of the century and consists mostly of online databases with a web frontend.
@@
 ```{r fig-c14-datasets-map, message=FALSE}
-#| fig-cap: Geographic coverage of published regional radiocarbon compilations according to our survey (see Supplementary Material).
+#| fig-cap: Geographic coverage of published regional radiocarbon compilations according to our survey (binned by UN M49 macroregion).
 c14_datasets |>
@@
   coord_sf(crs = "+proj=moll") +
-  theme_minimal_grid()
+  theme_minimal_grid(font_size = 11)
 ```
@@
 ```{r fig-c14-global, message=FALSE, warning=FALSE}
-#| fig-cap: Geographic and temporal (sum calibration) of georeferenced dates in XRONOS and other global radiocarbon compilations
-#| fig-height: 6
+#| fig-cap: Geographic (kernel density) and temporal (sum calibration) of georeferenced dates in XRONOS and other global radiocarbon compilations
+#| fig-height: 5
 plot_c14_density <- function(density, title, coastline = coast) {
@@
     coord_sf(crs = "+proj=moll") +
-    theme_minimal_grid() +
+    theme_minimal_grid(font_size = 11) +
     theme(
@@
     labs(caption = paste0("N=", n)) +
-    theme_minimal_vgrid() + 
+    theme_minimal_vgrid(font_size = 11) + 
     theme(
@@
 #   coord_sf(crs = "+proj=moll") +
-#   theme_minimal_grid()
+#   theme_minimal_grid(font_size = 11)
 ```
@@
 In periods beyond the practical limit of radiocarbon dating (c. 55,000 BP), other types of radiometric (K–Ar, U–Pb, etc.), chemical or luminescence dating offer an alternative [@Aitken1999].
-Conversely, in historic periods, radiocarbon is often relatively underused compared to conventional typological dating (based on artefact characteristics), which in these periods can offer comparable or better temporal resolution, or direct dating based on epigraphy [@HermankovaEtAl2021], numismatics [@KemmersMyrberg2011] or historical sources.
+Conversely, in historic periods, or on calibration 'plateaus' [@GuldersonEtAl2005], radiocarbon is often less important than conventional typological dating (based on artefact characteristics), which in these periods can offer comparable or better temporal resolution, or direct dating based on epigraphy [@HermankovaEtAl2021], numismatics [@KemmersMyrberg2011] or historical sources.
 In places where it is widely available, dendrochronology [@Baillie2014] also produces significantly better resolved chronologies and therefore tends to be the main source of chronometric data.
@@
 DendroDB, hosted by the Swiss Federal Institute for Forest, Snow and Landscape Research (WSL), emphasizes ecological and climate studies over archaeological wood material.
-While it claims a broad scope, the database remains non-functional, rendering it ineffective for research needs.
 The CFS-TRenD database, managed by the Canadian Forest Service, compiles over 4,600 datasets from Canadian forests, primarily focusing on boreal ecosystems.
@@
 Typological dates on one class of object may well clash with other classes of object, or for that matter with scientific dates – does one trust the date on pottery, the date on architecture, or the radiocarbon date?
-Without additional metadata on e.g. who made the typological determination or what the radiocarbon date was obtained on, such inconsistencies are difficult to resolve 
+Without additional metadata on e.g. who made the typological determination or what the radiocarbon date was obtained on, such inconsistencies are difficult to resolve.
 Similarly the absolute date range corresponding to a typological determination (e.g.
@@
 XRONOS is our answer to Kintigh's call [-@Kintigh2006] for digital infrastructures that don't just provide access to chronological data but enables researchers to "archive, access, integrate, and mine disparate data sets".
-It complements several similar open data infrastructures within and outwith archaeology, such as the Global Biodiverisity Information Facility [GBIF, @CanhosEtAl2004], the Strategic Environmental Archaeology Database [SEAD, @Buckland2014], IMPACT for mummified human remains [@NelsonWade2015], Neotoma for palaeoecological data [@WilliamsEtAl2018], IsoArcH for stable isotope data [@PlompEtAl2022], the International Soil Radiocarbon Database [ISRaD, @LawrenceEtAl2020], and the 'Big Interdisciplinary Archaeological Database' (BIAD), an ambitious new initiative to combine many of these individual domains, including chronology [@ReiterEtAl2024].
+It complements several similar open data infrastructures within and outwith archaeology, such as the Global Biodiverisity Information Facility [GBIF, @CanhosEtAl2004], the Strategic Environmental Archaeology Database [SEAD, @Buckland2014], IMPACT for mummified human remains [@NelsonWade2015], Neotoma for palaeoecological data [@WilliamsEtAl2018], IsoArcH for stable isotope data [@PlompEtAl2022], Poseidon for archaeogenetic data [@SchmidEtAl2024], the International Soil Radiocarbon Database [ISRaD, @LawrenceEtAl2020], and the 'Big Interdisciplinary Archaeological Database' (BIAD), an ambitious new initiative to combine many of these individual domains, including chronology [@ReiterEtAl2024].
 To improve upon existing global syntheses of radiocarbon dates (see @sec-global-compilations), we aimed to develop a living infrastructure that both continually collected data from diverse sources and presented a seamless single database to the user.
@@
 Chrons are conceptually useful because they emphasise that different types of archaeological 'dates', drawn from different sources, have essentially the same information content: the location of an event in space and time.
-We thereby avoid privileging certain sources of chronological data over (as might be the case if, for example, we treated 'period' as a fixed attribute of a site) and can accommodate contradictory (e.g.
+We thereby avoid privileging certain sources of chronological data over others (as might be the case if, for example, we treated 'period' as a fixed attribute of a site) and can accommodate contradictory (e.g.
 differences of opinion on typological classification).
-This is important given that XRONOS aspires to be an authoratative 'backbone' with a global scope, so we cannot realistically impose a single chronological scheme or resolve conflicting information provided by specialists.
-They are useful practically because they expose a common interface for attributes that all types of chronological information share, such as a *terminus post quem* (TPQ), *terminus ante quem* (TAQ), and midpoint estimate.
-This allows applications that use XRONOS' data model (including XRONOS itself) to collate chronological data from multiple sources, without necessarily having to be aware of the pecularities of each type of dating.
+This is important given that XRONOS aspires to be an authoritative 'backbone' with a global scope, so we cannot realistically impose a single chronological scheme or resolve conflicting information provided by specialists.
+Chrons are useful practically because they expose a common interface for attributes that all types of chronological information share, such as a *terminus post quem* (TPQ), *terminus ante quem* (TAQ), and midpoint estimate.
+This allows applications to query chronological data from multiple sources, without necessarily having to be aware of the pecularities of each type of dating.
 
@@
 
-The series of relations `[chron] > sample > context > site` links the chronological and contextual sides of the XRONOS data model.
+The series of relations `(chron) > sample > context > site` links the chronological and contextual sides of the XRONOS data model.
 Each step is a many-to-one association, meaning for example that it is possible to attach multiple chrons to the same sample (e.g.
@@
 The bibliographic reference associated with this date [@RichterEtAl2017] is similarly represented at the address <https://xronos.ch/references/17778>, and can be accessed at that address using the same interface as the radiocarbon date.
-These uniform REST interfaces are another example of a boring architectural choice that make it easier for us to enrich the scientific contents of XRONOS, by  adding new types of modular resource that represent new scientific entities.
 
@@
 <https://xronos.ch/c14s> for radiocarbon dates), is worth special mention because it is through this that the filtering logic at the core of XRONOS' two interfaces is implemented.
-By passing a query as HTTP GET parameters to the index action of a resource, the list returned the user is modified to only include records that match that query.
+By passing a query as HTTP GET parameters to the index action of a resource, the list returned to the user is modified to only include records that match that query.
 For example, <https://xronos.ch/sites?site[country_code]=CH> (the part of the URL after the `?` character encodes the SQL WHERE clause `country = 'CH'` as a GET parameter) lists sites in Switzerland.
@@
 A wide variety of other potential data quality issues (e.g.
-missing data on what country a site is in) are also flagged for human review by this system [@tbl-issues], which can often be semi-automated (e.g.
+missing data on what country a site is in) are also flagged for human review by this system (@tbl-issues), which can often be semi-automated (e.g.
 suggesting close matches in the thesaurus or the country indicated by the record's coordinates).
@@
 We import data from many overlapping resources (many of which incorporate each other either in whole or in part), so duplicate records are common [as recently discussed by @ReiterEtAl2024].
-The end result of standardising and correcting a record is also often to create a duplicate: e.g.
-the same sample imported from one source as 'oak' but another as '*Quercus* sp.' will become a duplicate pair as '*Quercus*', and thus be recognised as a single sample.
+The end result of standardising and correcting a record is also often to reveal a duplicate: e.g.
+the same sample imported from one source as 'oak' but another as '*Quercus* sp.' will become a recognisable duplicate record as '*Quercus*', and thus can be automatically merged to a single sample.
 Such exact duplicates can be merged automatically, with the oldest record becoming the authoratitive version, but detecting fuzzier duplicated information (e.g.
@@
 and an R package for interfacing with the API. 
-These systems are in production and publicly accessible at <https://xronos.ch>.
+These systems are deployed and publicly accessible at <https://xronos.ch>.
 
@@
 With that said, we see the approaches followed by as complementary rather than competing.
-A c14bazAAR parser for XRONOS is in development (<https://github.com/ropensci/c14bazAAR/pull/150>), and we also aim to provide an IntChron interface to XRONOS' data in the near future.
+A parser for XRONOS has been contributed back to c14bazAAR and we also aim to provide an IntChron interface to XRONOS' data in the near future.
 New data and corrections from p3k14c are incorporated into XRONOS as they are released.

### 2025-04-04 paper.tex
Rerender as preprint
Incorporate @nevrome's suggestions
Did what I could with fig-c14-datasets-map... the problem is that adding
extra elements (i.e. scales) causes patchwork to squeeze the plot area
even more.

@@
   number,
-  doubleblind]{elsarticle}
+  preprint]{elsarticle}
 
@@
 Archaeological Sciences},,postcodesep={}}
-\affiliation[2]{organization={Max Planck Institute for
-Geoanthropology},,postcodesep={}}
+\affiliation[2]{organization={Max Planck Institute for Evolutionary
+Anthropology},,postcodesep={}}
 
@@
 description here should be read as a `snapshot' of the project as of
-writing rather than its final state.
+version 1.2.0.
 
@@
 
-The number of available compilations has increased exponentially since
+The number of available compilations has increased signifcantly since
 around 1995 (Figure~\ref{fig-c14-datasets-time}). The first generation
@@
 \caption{\label{fig-c14-datasets-map}Geographic coverage of published
-regional radiocarbon compilations according to our survey (see
-Supplementary Material).}
+regional radiocarbon compilations according to our survey (binned by UN
+M49 macroregion).}
 
@@
 
-\caption{\label{fig-c14-global}Geographic and temporal (sum calibration)
-of georeferenced dates in XRONOS and other global radiocarbon
-compilations}
+\caption{\label{fig-c14-global}Geographic (kernel density) and temporal
+(sum calibration) of georeferenced dates in XRONOS and other global
+radiocarbon compilations}
 
@@
 dating offer an alternative (Aitken, 1999). Conversely, in historic
-periods, radiocarbon is often relatively underused compared to
-conventional typological dating (based on artefact characteristics),
-which in these periods can offer comparable or better temporal
-resolution, or direct dating based on epigraphy (Heřmánková et al.,
-2021), numismatics (Kemmers and Myrberg, 2011) or historical sources. In
-places where it is widely available, dendrochronology (Baillie, 2014)
-also produces significantly better resolved chronologies and therefore
-tends to be the main source of chronometric data. Other more
-application-specific chronological methods include shoreline dating
-(Brøgger, 1905; Roalkvam, 2023), lichenometry (Benedict, 2009) and rock
-weathering dating (Bednarik, 2020; Whitley, 2012).
+periods, or on calibration `plateaus' (Guilderson et al., 2005),
+radiocarbon is often less important than conventional typological dating
+(based on artefact characteristics), which in these periods can offer
+comparable or better temporal resolution, or direct dating based on
+epigraphy (Heřmánková et al., 2021), numismatics (Kemmers and Myrberg,
+2011) or historical sources. In places where it is widely available,
+dendrochronology (Baillie, 2014) also produces significantly better
+resolved chronologies and therefore tends to be the main source of
+chronometric data. Other more application-specific chronological methods
+include shoreline dating (Brøgger, 1905; Roalkvam, 2023), lichenometry
+(Benedict, 2009) and rock weathering dating (Bednarik, 2020; Whitley,
+2012).
 
@@
 Landscape Research (WSL), emphasizes ecological and climate studies over
-archaeological wood material. While it claims a broad scope, the
-database remains non-functional, rendering it ineffective for research
-needs. The CFS-TRenD database, managed by the Canadian Forest Service,
-compiles over 4,600 datasets from Canadian forests, primarily focusing
-on boreal ecosystems. Despite its extensive coverage for North America,
-its geographical specificity and lack of open access restrict its
-utility for European prehistoric contexts. Similarly, the QUB
-Dendrochronology Database, managed by Queen's University Belfast, offers
-valuable datasets for Ireland and the UK but lacks significant
-representation of prehistoric material, limiting its application in
-broader archaeological investigations. The Building Archaeology Research
-Database (BARD) contains over 24,000 records, including
-dendrochronological data from more than 2,700 buildings. However, its
-focus on medieval and post-medieval timber-frame construction further
-narrows its utility for studies involving prehistoric wood samples.
+archaeological wood material. The CFS-TRenD database, managed by the
+Canadian Forest Service, compiles over 4,600 datasets from Canadian
+forests, primarily focusing on boreal ecosystems. Despite its extensive
+coverage for North America, its geographical specificity and lack of
+open access restrict its utility for European prehistoric contexts.
+Similarly, the QUB Dendrochronology Database, managed by Queen's
+University Belfast, offers valuable datasets for Ireland and the UK but
+lacks significant representation of prehistoric material, limiting its
+application in broader archaeological investigations. The Building
+Archaeology Research Database (BARD) contains over 24,000 records,
+including dendrochronological data from more than 2,700 buildings.
+However, its focus on medieval and post-medieval timber-frame
+construction further narrows its utility for studies involving
+prehistoric wood samples.
 
@@
 the typological determination or what the radiocarbon date was obtained
-on, such inconsistencies are difficult to resolve Similarly the absolute
-date range corresponding to a typological determination (e.g.~``Late
-Neolithic'') can be interpreted in multiple ways depending on the region
-and intentions of the expert making the determination. PeriodO
-(Rabinowitz et al., 2016) is a linked open data infrastructure that
-includes a shared vocabulary of typological periods and corresponding
-calendar age estimates, and an important step towards addressing the
-latter problem. However, it remains to be systematically linked to
-compilations of typological dates (though there are some efforts in this
-direction e.g. Hannah et al., 2022).
+on, such inconsistencies are difficult to resolve. Similarly the
+absolute date range corresponding to a typological determination
+(e.g.~``Late Neolithic'') can be interpreted in multiple ways depending
+on the region and intentions of the expert making the determination.
+PeriodO (Rabinowitz et al., 2016) is a linked open data infrastructure
+that includes a shared vocabulary of typological periods and
+corresponding calendar age estimates, and an important step towards
+addressing the latter problem. However, it remains to be systematically
+linked to compilations of typological dates (though there are some
+efforts in this direction e.g. Hannah et al., 2022).
 
@@
 palaeoecological data (Williams et al., 2018), IsoArcH for stable
-isotope data (Plomp et al., 2022), the International Soil Radiocarbon
-Database (ISRaD, Lawrence et al., 2020), and the `Big Interdisciplinary
+isotope data (Plomp et al., 2022), Poseidon for archaeogenetic data
+(Schmid et al., 2024), the International Soil Radiocarbon Database
+(ISRaD, Lawrence et al., 2020), and the `Big Interdisciplinary
 Archaeological Database' (BIAD), an ambitious new initiative to combine
@@
 space and time. We thereby avoid privileging certain sources of
-chronological data over (as might be the case if, for example, we
+chronological data over others (as might be the case if, for example, we
 treated `period' as a fixed attribute of a site) and can accommodate
@@
 classification). This is important given that XRONOS aspires to be an
-authoratative `backbone' with a global scope, so we cannot realistically
+authoritative `backbone' with a global scope, so we cannot realistically
 impose a single chronological scheme or resolve conflicting information
-provided by specialists. They are useful practically because they expose
-a common interface for attributes that all types of chronological
+provided by specialists. Chrons are useful practically because they
+expose a common interface for attributes that all types of chronological
 information share, such as a \emph{terminus post quem} (TPQ),
 \emph{terminus ante quem} (TAQ), and midpoint estimate. This allows
-applications that use XRONOS' data model (including XRONOS itself) to
-collate chronological data from multiple sources, without necessarily
-having to be aware of the pecularities of each type of dating.
+applications to query chronological data from multiple sources, without
+necessarily having to be aware of the pecularities of each type of
+dating.
 
@@
 The series of relations
-\texttt{{[}chron{]}\ \textgreater{}\ sample\ \textgreater{}\ context\ \textgreater{}\ site}
+\texttt{(chron)\ \textgreater{}\ sample\ \textgreater{}\ context\ \textgreater{}\ site}
 links the chronological and contextual sides of the XRONOS data model.
@@
 \url{https://xronos.ch/references/17778}, and can be accessed at that
-address using the same interface as the radiocarbon date. These uniform
-REST interfaces are another example of a boring architectural choice
-that make it easier for us to enrich the scientific contents of XRONOS,
-by adding new types of modular resource that represent new scientific
-entities.
+address using the same interface as the radiocarbon date.
 
@@
 passing a query as HTTP GET parameters to the index action of a
-resource, the list returned the user is modified to only include records
-that match that query. For example,
+resource, the list returned to the user is modified to only include
+records that match that query. For example,
 \url{https://xronos.ch/sites?site\%5Bcountry_code\%5D=CH} (the part of
@@
 (e.g.~missing data on what country a site is in) are also flagged for
-human review by this system Table~\ref{tbl-issues}, which can often be
+human review by this system (Table~\ref{tbl-issues}), which can often be
 semi-automated (e.g.~suggesting close matches in the thesaurus or the
@@
 are common (as recently discussed by Reiter et al., 2024). The end
-result of standardising and correcting a record is also often to create
+result of standardising and correcting a record is also often to reveal
 a duplicate: e.g.~the same sample imported from one source as `oak' but
-another as `\emph{Quercus} sp.' will become a duplicate pair as
-`\emph{Quercus}', and thus be recognised as a single sample. Such exact
-duplicates can be merged automatically, with the oldest record becoming
-the authoratitive version, but detecting fuzzier duplicated information
-(e.g.~differences in the spelling of site names) has proved a more
-difficult problem. As of writing there are therefore still many
-duplicate records in XRONOS that need to be manually resolved, but we
-hope to automate much more of this work in the future.
+another as `\emph{Quercus} sp.' will become a recognisable duplicate
+record as `\emph{Quercus}', and thus can be automatically merged to a
+single sample. Such exact duplicates can be merged automatically, with
+the oldest record becoming the authoratitive version, but detecting
+fuzzier duplicated information (e.g.~differences in the spelling of site
+names) has proved a more difficult problem. As of writing there are
+therefore still many duplicate records in XRONOS that need to be
+manually resolved, but we hope to automate much more of this work in the
+future.
 
@@
 the data and a REST API for programmatic access; and an R package for
-interfacing with the API. These systems are in production and publicly
+interfacing with the API. These systems are deployed and publicly
 accessible at \url{https://xronos.ch}.
@@
 as it is published. With that said, we see the approaches followed by as
-complementary rather than competing. A c14bazAAR parser for XRONOS is in
-development (\url{https://github.com/ropensci/c14bazAAR/pull/150}), and
-we also aim to provide an IntChron interface to XRONOS' data in the near
-future. New data and corrections from p3k14c are incorporated into
-XRONOS as they are released.
+complementary rather than competing. A parser for XRONOS has been
+contributed back to c14bazAAR and we also aim to provide an IntChron
+interface to XRONOS' data in the near future. New data and corrections
+from p3k14c are incorporated into XRONOS as they are released.
 
@@
 
+\bibitem[\citeproctext]{ref-GuldersonEtAl2005}
+Guilderson, T.P., Reimer, P.J., Brown, T.A., 2005. The {Boon} and {Bane}
+of {Radiocarbon Dating}. Science 307, 362--364.
+\url{https://doi.org/10.1126/science.1104164}
+
 \bibitem[\citeproctext]{ref-HannahEtAl2022}
@@
 
+\bibitem[\citeproctext]{ref-SchmidEtAl2024}
+Schmid, C., Ghalichi, A., Lamnidis, T.C., Mudiyanselage, D.B.A., Haak,
+W., Schiffels, S., 2024. Poseidon -- {A} framework for archaeogenetic
+human genotype data management. eLife 13.
+\url{https://doi.org/10.7554/eLife.98317.1}
+
 \bibitem[\citeproctext]{ref-SchmidEtAl2019}

### 2025-04-04 paper.qmd
Update in response to reviews

@@
 Of the `r n_c14_datasets` published datasets we identified, `r n_c14_datasets_maintained` were intended to be continuously updated, but only `r n_c14_datasets_recently_updated` have received updates in the last two years.
-The average 'lifespan' of a dataset from its publication to its last update is around `r english(round(mean_c14_datasets_lifespan))` years.
+The mean 'lifespan' of a dataset from its publication to its last update is around `r english(round(mean_c14_datasets_lifespan))` years.
 Most radiocarbon datasets we reviewed were compiled with a specific goal in mind (e.g.
@@
 ```{r fig-c14-global, message=FALSE, warning=FALSE}
-#| fig-cap: Geographic (kernel density) and temporal (sum calibration) of georeferenced dates in XRONOS and other global radiocarbon compilations
+#| fig-cap: Geographic (quantile kernel density) and temporal (sum calibration) of georeferenced dates in XRONOS and other global radiocarbon compilations
 #| fig-height: 5
@@
 In places where it is widely available, dendrochronology [@Baillie2014] also produces significantly better resolved chronologies and therefore tends to be the main source of chronometric data.
-Other more application-specific chronological methods include shoreline dating [@Brogger1905; @Roalkvam2023], lichenometry [@Benedict2009] and rock weathering dating [@Whitley2012; @Bednarik2020].
+Other more application-specific chronological methods include shoreline dating [@Brogger1905; @Roalkvam2023], lichenometry [@Benedict2009] and rock weathering dating [@Whitley2012; @Bednarik2020; @AndreaeEtAl2020].
 
@@
 Aggregated typological information from such sources are often used in aoristic analysis and related methods [@Mischka2004; @Crema2024].
-What is lacking in this presentation of typological dating is metadata on how the determination was made and how exactly it is to be understood.
+But this presentation of typological dating lacks metadata on how the determination was made and how exactly it is to be understood.
 Like any archaeological date, a typological date is derived from a physical sample – the object or set of object from which a chronological estimate was derived.
@@
 an 'archaeological site', <https://www.wikidata.org/wiki/Q839954>).
-Linking to Wikidata is especially useful because it dissimenates—and thereby preserves—the data compiled in XRONOS beyond the specialist/academic community.
+Linking to Wikidata is especially useful because it disseminates—and thereby preserves—the data compiled in XRONOS beyond the specialist/academic community.
 It also allows us to enrich the database with contextual information that otherwise be beyond our scope and resources, for example embedding multilingual descriptions of sites from Wikipedia articles on site records where the site has been linked to a Wikidata item.

### 2025-04-04 paper.tex
Update in response to reviews

@@
 identified, 33 were intended to be continuously updated, but only 13
-have received updates in the last two years. The average `lifespan' of a
+have received updates in the last two years. The mean `lifespan' of a
 dataset from its publication to its last update is around 4 years. Most
@@
 
-\caption{\label{fig-c14-global}Geographic (kernel density) and temporal
-(sum calibration) of georeferenced dates in XRONOS and other global
-radiocarbon compilations}
+\caption{\label{fig-c14-global}Geographic (quantile kernel density) and
+temporal (sum calibration) of georeferenced dates in XRONOS and other
+global radiocarbon compilations}
 
@@
 include shoreline dating (Brøgger, 1905; Roalkvam, 2023), lichenometry
-(Benedict, 2009) and rock weathering dating (Bednarik, 2020; Whitley,
-2012).
+(Benedict, 2009) and rock weathering dating (Andreae et al., 2020;
+Bednarik, 2020; Whitley, 2012).
 
@@
 typological information from such sources are often used in aoristic
-analysis and related methods (Crema, 2024; Mischka, 2004). What is
-lacking in this presentation of typological dating is metadata on how
-the determination was made and how exactly it is to be understood. Like
-any archaeological date, a typological date is derived from a physical
+analysis and related methods (Crema, 2024; Mischka, 2004). But this
+presentation of typological dating lacks metadata on how the
+determination was made and how exactly it is to be understood. Like any
+archaeological date, a typological date is derived from a physical
 sample -- the object or set of object from which a chronological
@@
 site', \url{https://www.wikidata.org/wiki/Q839954}). Linking to Wikidata
-is especially useful because it dissimenates---and thereby
+is especially useful because it disseminates---and thereby
 preserves---the data compiled in XRONOS beyond the specialist/academic
@@
 
+\bibitem[\citeproctext]{ref-AndreaeEtAl2020}
+Andreae, M.O., Al-Amri, A., Andreae, C.M., Guagnin, M., Jochum, K.P.,
+Stoll, B., Weis, U., 2020. Archaeometric studies on the petroglyphs and
+rock varnish at {Kilwa} and {Sakaka}, northern {Saudi Arabia}. Arabian
+Archaeology and Epigraphy 31, 219--244.
+\url{https://doi.org/10.1111/aae.12167}
+
 \bibitem[\citeproctext]{ref-ArnoldLibby1951}

### 2025-06-09 paper.qmd
Add CH'S SNSF grant to acknowledgements

@@
 The source code for all the software components of the system are available online (at <https://github.com/xronos-ch>) under open licenses.
-The databases of the instance at <https://xronos.ch> are also archived with Zenodo (*link omitted for blind review*<!-- TODO: link -->).
+The databases of the instance at <https://xronos.ch> are also archived with Zenodo at <https://doi.org/10.5281/zenodo.14282598>.
 With these two resources we reduce XRONOS' 'bus factor'; if we are not able to continue operating XRONOS, somebody else can fully recreate it.
@@
 A complete and up-to-date list of acknowledgements for the project can be found at <https://xronos.ch/about/acknowledgements>.
-<!-- TODO: peer reviewers -->
+
+We are also grateful to the two anonymous reviewers of our manuscript, whose comments have improved the finished paper.
 
@@
 
-This work was funded by the Swiss National Science Foundation ([SNSF Project #198152](https://data.snf.ch/grants/grant/198153)) and the University of Bern (UniBE Initiator Grant 2019, Caroline Heitz, Project 'Time and Temporality').
+This work was funded by the Swiss National Science Foundation ([SNSF Project #198152](https://data.snf.ch/grants/grant/198153); [SNSF Project #194326](https://data.snf.ch/grants/grant/194326)) and the University of Bern (UniBE Initiator Grant 2019).
 

### 2025-06-09 paper.qmd
Split table 1 to separate file

@@
 In the remainder of this paper, we describe the concept and implementation of XRONOS in relation to the state of the art in open chronometric data in archaeology, and evaluate our progress in achieving these goals as of writing.
-Since we envisage both XRONOS as a dataset and XRONOS as software to be continually developing resources, the description here should be read as a 'snapshot' of the project as of version 1.2.0.
+Since we envisage both XRONOS as a dataset and XRONOS as software to be continually developing resources, the description here should be read as a 'snapshot' of the project as of [version 1.1.0](https://github.com/xronos-ch/xronos.rails/releases/tag/v1.1.0).
 
@@
 # https://github.com/quarto-dev/quarto-cli/issues/9342
-c14_datasets |>
-  transmute(
-    database = glue("[{name}]({url})"),
-    publication_year, n_dates, citations
-  ) |>
-  arrange(publication_year) |>
-  gt() |>
-  cols_label(
-    publication_year = "published",
-    citations = "references",
-    n_dates = "dates"
-    ) |>
-  cols_label_with(everything(), str_to_sentence) |>
-  sub_missing(everything()) |>
-  fmt_markdown(c(database, citations)) |>
-  fmt_number(c(n_dates), decimals = 0) |>
-  cols_width(
-    database ~ pct(40),
-    publication_year ~ pct(15),
-    n_dates ~ pct(15),
-    citations ~ pct(30)
-  ) |>
-  cols_align(
-    "left", c(database)
-  ) |>
-  tab_options(
-    table.font.size = 13
-  )
+#c14_datasets |>
+#  transmute(
+#    database = glue("[{name}]({url})"),
+#    publication_year, n_dates, citations
+#  ) |>
+#  arrange(publication_year) |>
+#  gt() |>
+#  cols_label(
+#    publication_year = "published",
+#    citations = "references",
+#    n_dates = "dates"
+#    ) |>
+#  cols_label_with(everything(), str_to_sentence) |>
+#  sub_missing(everything()) |>
+#  fmt_markdown(c(database, citations)) |>
+#  fmt_number(c(n_dates), decimals = 0) |>
+#  cols_width(
+#    database ~ pct(40),
+#    publication_year ~ pct(15),
+#    n_dates ~ pct(15),
+#    citations ~ pct(30)
+#  ) |>
+#  cols_align(
+#    "left", c(database)
+#  ) |>
+#  tab_options(
+#    table.font.size = 13
+#  )
+"omitted from manuscript – see attached file `table1.html`"
 ```
@@
 ### Dendrochronology
-
-<!-- TODO: add references? -->
 

### 2025-06-09 paper.tex
Split table 1 to separate file

@@
 description here should be read as a `snapshot' of the project as of
-version 1.2.0.
+\href{https://github.com/xronos-ch/xronos.rails/releases/tag/v1.1.0}{version
+1.1.0}.
 
@@
 
-\fontsize{9.8pt}{11.7pt}\selectfont
-\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 0.40\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 0.15\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 0.15\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedright\arraybackslash}p{\dimexpr 0.30\linewidth -2\tabcolsep-1.5\arrayrulewidth}}
-\toprule
-Database & Published & Dates & References \\ 
-\midrule\addlinespace[2.5pt]
-\href{https://bda.huma-num.fr/}{Base de Données Archéologique} & 1994 & 7,000 & @Perrin2019 \\ 
-\href{https://andesc14.pl}{ANDES 14C} & 1994 & 5,800 & @MichczynskiEtAl1995 \\ 
-\href{http://www.tayproject.org/enghome.html}{Archaeological Settlements of Turkey} & 1998 & 1,600 & @TanindiErdogu2005 \\ 
-\href{https://www.canadianarchaeology.ca/}{Canadian Archaeological Radiocarbon Database} & 1999 & 171,500 & @GajewskiEtAl2011; @KellyEtAl2022 \\ 
-\href{https://radonb.ufg.uni-kiel.de/}{RADO.NB (incl. RADON and RADON-B)} & 1999 & 34,200 & @Raetzel-Fabian1999; @RADON; @RADONB; @RADO.NB \\ 
-\href{https://www.waikato.ac.nz/nzcd}{New Zealand Radiocarbon Database} & 2000 & 2,000 & @McFadgenEtAl2000 \\ 
-\href{http://web.archive.org/web/20080509082232/http://www.canew.org/index.html}{International Central Anatolian Neolithic e-Workshop databases} & 2001 & 1,000 & @ReingruberThissen2005; @ReingruberThissen2009 \\
-\href{https://ees.kuleuven.be/geography/projects/14c-palaeolithic/}{Radiocarbon Palaeolithic Europe Database} & 2002 & 17,900 & @Vermeersch2024 \\ 
-\href{https://zenodo.org/doi/10.5281/zenodo.7215741}{CalPal database} & 2002 & 49,800 & @Weninger2022 \\ 
-\href{http://context-database.uni-koeln.de}{CONTEXT} & 2002 & 2,900 & @BohnerSchyle2004 \\ 
-\href{http://pidba.utk.edu/dating.htm}{Paleoindian Database of the Americas} & 2003 & 1,300 & @AndersonEtAl2010 \\ 
-\href{https://archaeologydataservice.ac.uk/archives/view/austarch_na_2014/}{AustArch (1, 2, and 3)} & 2008 & 5,000 & @WilliamsEtAl2008; @WilliamsSmith2012; @WilliamsSmith2013 \\ 
-\href{https://sites.google.com/site/chapplearchaeology/irish-radiocarbon-dendrochronological-dates}{Irish Radiocarbon \& Dendrochronological Dates} & 2010 & 10,700 & @IRDD \\ 
-\href{https://www.exoriente.org/associated_projects/ppnd.php}{Platform for Neolithic Radiocarbon Dates (PPND)} & 2010 & 800 & @Benz2010 \\ 
-\href{http://www.paleoanthro.org/media/journal/content/PA20110001.pdf}{PACEA Geo-Referenced Radiocarbon Database} & 2011 & 6,000 & @dErricoEtAl2011 \\ 
-\href{https://doi.org/10.1016/j.quaint.2012.08.2052}{Peru archaeological radiocarbon database} & 2013 & 300 & @RademakerEtAl2013 \\ 
-\href{https://telearchaeology.org///EUBAR/}{EUBAR C14 database} & 2014 & 1,700 & @CapuzzoEtAl2014 \\ 
-\href{https://doi.org/10.1016/j.quascirev.2014.05.015}{Wang 2014} & 2014 & 4,700 & @WangEtAl2014 \\ 
-\href{http://www.14sea.org/}{14SEA} & 2015 & 3,000 & @ReingruberThissen2017 \\ 
-\href{https://discovery.ucl.ac.uk/id/eprint/1469811/}{EUROEVOL} & 2015 & 14,100 & @ManningEtAl2016 \\ 
-\href{https://doi.org/10.1016/j.quascirev.2015.06.022}{Flohr et al. 2015} & 2015 & 3,000 & @FlohrEtAl2016 \\ 
-\href{https://doi.org/10.1016/j.quaint.2014.09.076}{South Central Andes Radiocarbon} & 2015 & 1,700 & @GayoEtAl2015 \\ 
-\href{https://github.com/dirkseidensticker/aDRAC}{Archive des datations radiocarbones d'Afrique centrale} & 2016 & 1,900 & @SeidenstickerSchmid2021 \\ 
-\href{https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/NJLNRJ}{KITE East Africa} & 2016 & 800 & @CourtneyMustaphi2016 \\ 
-\href{https://doi.org/10.14324/000.ds.1570274}{Palmisano et al. 2017} & 2017 & 1,900 & @PalmisanoEtAl2017 \\ 
-\href{http://vmtropicar-proto.ird.fr/archeologie/}{Plateforme des datations archéologiques intertropicales} & 2017 & 1,300 & @deSaulieuEtAl2017 \\ 
-\href{http://www.idearqueologia.org/c14/}{Archivo de Dataciones Radiocarbónicas de la Prehistoria Recient (IDEARQ)} & 2017 & — & @UriarteGonzalezEtAl2017 \\ 
-\href{https://www.rekihaku.ac.jp/up-cgi/login.pl?p=param/esrd/db_param}{Database of Radiocarbon Dates Published in Japanese Archaeological Reports} & 2018 & 44,000 & @Kudo2018; KudoEtAl2023 \\ 
-\href{https://c14.arch.ox.ac.uk/sadb/db}{Southern African Radiocarbon Database} & 2019 & 2,700 & @LoftusEtAl2019 \\ 
-\href{https://doi.org/10.1016/j.quascirev.2019.105878}{Douglass et al. 2019} & 2019 & 200 & @DouglassEtAl2019 \\ 
-\href{https://theia.arch.cam.ac.uk/MedAfriCarbon/}{MedAfriCarbon} & 2020 & 1,600 & @LucariniEtAl2020 \\ 
-\href{https://github.com/clipo/rapanui-radiocarbon}{A repository of radiocarbon data for Rapa Nui} & 2020 & 800 & @Lipo2020 \\ 
-\href{https://github.com/jgregoriods/rxpand}{rxpand} & 2020 & 2,800 & @GregoriodeSouza2020 \\ 
-\href{https://core.tdar.org/project/455305/mesoamerican-radiocarbon-database-mesorad}{Mesoamerican Radiocarbon Database (MesoRAD)} & 2020 & 1,800 & @HoggarthEtAl2021 \\ 
-\href{https://zenodo.org/records/4541470}{AgriChange} & 2020 & 3,700 & @Martinez-GrauEtAl2021 \\ 
-\href{https://github.com/ercrema/jomonPhasesAndPopulation}{Crema \& Kobayashi 2020} & 2020 & 2,100 & @CremaEtAl2016 \\ 
-\href{https://rdr.ucl.ac.uk/articles/dataset/Dataset_for_An_Aegean_history_and_archaeology_written_through_radiocarbon_dates/12489137/1}{Katsianis et al. 2020} & 2020 & 3,200 & @KatsianisEtAl2020 \\ 
-\href{https://www.arqueodata.com/}{ArqueoData} & 2021 & 800 & @Alcantara2021 \\ 
-\href{https://github.com/philriris/caribbean-14C}{caribbean-14C} & 2021 & 2,100 & @Riris2021 \\ 
-\href{https://github.com/apalmisano82/NERD}{Near East Radiocarbon Dates (NERD)} & 2021 & 11,000 & @PalmisanoEtAl2022a \\ 
-\href{https://doi.org/10.13131/archelogicadata-yb11-yb66}{NeoNet} & 2021 & 2,500 & @HuetEtAl2022; @HuetEtAl2024 \\ 
-\href{https://github.com/ercrema/NeolithicKoreaDemography}{Kim et al. 2021} & 2021 & 900 & @KimEtAl2021 \\ 
-\href{https://doi.org/10.1371/journal.pone.0251407}{Cochrane et al. 2021} & 2021 & 100 & @CochraneEtAl2021 \\ 
-\href{https://www.waikato.ac.nz/research/research-services-facilities/radiocarbon-dating/research//nz-radiocarbon-database}{Aotearoa New Zealand Radiocarbon Databas} & 2022 & 4,100 & @PetcheyEtAl2022 \\
-\href{https://github.com/apalmisano82/AIDA}{Archive of Italian Radiocarbon Dates (AIDA)} & 2022 & 4,000 & @PalmisanoEtAl2022 \\ 
-\href{https://www.p3k14c.org/}{p3k14c} & 2022 & 179,700 & @BirdEtAl2022 \\ 
-\href{https://research.jcu.edu.au/data/published/7c74f590a2ba11edb22c156e754c4bda/}{Torres Strait Radiocarbon Database} & 2022 & 300 & @LinnenluckeEtAl2023 \\ 
-\href{http://holoceno.iaas.ull.es/14Canarias_ULL/}{14Canarias} & 2023 & 700 & @Pardo-GordoEtAl2023 \\ 
-\href{https://doi.org/10.1016/j.jasrep.2023.103944}{Hoebe et al. 2023} & 2023 & 6,500 & @HoebeEtAl2023 \\ 
-\href{https://core.tdar.org/collection/71234}{Bolivian Radiocarbon Database} & 2023 & 3,000 & @Capriles2023 \\ 
-\href{https://doi.org/10.5281/zenodo.8334722}{An Annotated Compilation of Chronometric Dates for the Middle-Upper Palaeolithic Transition (45-30 ka BP) in Northern Iberia (Spain)} & 2023 & 200 & @Diaz-RodriguezEtAl2023 \\
-\href{https://doi.org/10.1371/journal.pone.0291956}{Großman et al. 2023} & 2023 & 3,400 & @GrossmannEtAl2023 \\ 
-\href{https://doi.org/10.1080/0067270X.2023.2215649}{Datations absolues, Inventaires archéologiques et Bibliographies en Afrique Centrale} & 2023 & 1,800 & @ClistEtAl2023 \\ 
-\href{https://doi.org/10.1016/j.quaint.2024.01.012}{Updated Peru archaeological radiocarbon database} & 2024 & 500 & @Rademaker2024 \\ 
-\href{https://banadora.mom.fr/}{Banque Nationale de Données Radiocarbone pour l'Europe et le Proche Orient} & — & — & @BANADORA \\ 
-\href{https://c14.arch.ox.ac.uk/database}{Oxford Radicoarbon Accelerator Unit database} & — & 8,500 & @GillespieEtAl1984; @BronkRamseyEtAl2009 \\ 
-\href{https://c14.arch.ox.ac.uk/egyptdb/db.php}{Egyptian Radiocarbon Database} & — & 1,600 & @RamseyEtAl2010; @DeeEtAl2012; @DeeEtAl2013 \\ 
-\href{https://c14.arch.ox.ac.uk/database}{NERC Radiocarbon Facility} & — & 6,100 & @GarnettEtAl2023 \\ 
-\href{http://www.adias-uae.com/radiocarbon.html}{Abu Dhabi Islands Archaeological Survey Radiocarbon Database} & — & 100 & @Al-Abyadh-Balghelam-Dalma-JebelEtAl \\ 
-\href{http://c14.kikirpa.be/}{Royal Institute for Cultural Heritage web based Radiocarbon database} & — & 7,300 & @VanStrydonckDeRoock \\ 
-\href{https://telearchaeology.org///c14/}{La base de dades radiocarbòniques de Catalunya} & — & — & @BarceloAlvarezEtAl2013 \\ 
-\bottomrule
-\end{tabular*}
+\begin{verbatim}
+[1] "omitted from manuscript – see attached file `table1.html`"
+\end{verbatim}
 
@@
 of the instance at \url{https://xronos.ch} are also archived with Zenodo
-(\emph{link omitted for blind review}). With these two resources we
-reduce XRONOS' `bus factor'; if we are not able to continue operating
-XRONOS, somebody else can fully recreate it. Equally importantly, we
-enable the `right to fork' should others wish to take the software
-and/or database in another direction. But aside from these extreme
-scenarios, the long-term sustainability of XRONOS is contingent on the
-existence of a community of scholars that use and contribute back to it.
+at \url{https://doi.org/10.5281/zenodo.14282598}. With these two
+resources we reduce XRONOS' `bus factor'; if we are not able to continue
+operating XRONOS, somebody else can fully recreate it. Equally
+importantly, we enable the `right to fork' should others wish to take
+the software and/or database in another direction. But aside from these
+extreme scenarios, the long-term sustainability of XRONOS is contingent
+on the existence of a community of scholars that use and contribute back
+to it.
 
@@
 
+We are also grateful to the two anonymous reviewers of our manuscript,
+whose comments have improved the finished paper.
+
 \section{Funding Statement}\label{funding-statement}
@@
 This work was funded by the Swiss National Science Foundation
-(\href{https://data.snf.ch/grants/grant/198153}{SNSF Project \#198152})
-and the University of Bern (UniBE Initiator Grant 2019, Caroline Heitz,
-Project `Time and Temporality').
+(\href{https://data.snf.ch/grants/grant/198153}{SNSF Project \#198152};
+\href{https://data.snf.ch/grants/grant/194326}{SNSF Project \#194326})
+and the University of Bern (UniBE Initiator Grant 2019).
 

### 2025-06-09 table1.qmd
Split table 1 to separate file

---
title: "Table 1"
subtitle: "From Roe et al., 'XRONOS: An Open Data Infrastructure for Archaeological Chronology'"
bibliography: 
  - references.bib
  - data/c14_datasets.bib
execute:
  echo: false
---

This table was omitted from the main manuscript due to formatting restrictions.

```{r deps}
library("dplyr", warn.conflicts = FALSE)
library("glue")
library("gt")
library("readr")
library("stringr")
```

```{r data-c14-datasets}
c14_datasets <- read_tsv("data/c14_datasets.tsv", show_col_types = FALSE)

n_c14_datasets <- nrow(c14_datasets)
n_c14_datasets_sans_publication_year <- nrow(filter(c14_datasets, is.na(publication_year)))
```

```{r tbl-c14-datasets}
#| tbl-cap: Summary of published compilations of radiocarbon dates. For full data, see supplementary materials.
# References in tables broken by:
# https://github.com/quarto-dev/quarto-cli/issues/9342
c14_datasets |>
  transmute(
    database = glue("[{name}]({url})"),
    publication_year, n_dates, citations
  ) |>
  arrange(publication_year) |>
  gt() |>
  cols_label(
    publication_year = "published",
    citations = "references",
    n_dates = "dates"
    ) |>
  cols_label_with(everything(), str_to_sentence) |>
  sub_missing(everything()) |>
  fmt_markdown(c(database, citations)) |>
  fmt_number(c(n_dates), decimals = 0) |>
  cols_width(
    database ~ pct(40),
    publication_year ~ pct(15),
    n_dates ~ pct(15),
    citations ~ pct(30)
  ) |>
  cols_align(
    "left", c(database)
  ) |>
  tab_options(
    table.font.size = 13
  )
```
